### 2012-02-02 MIMOSA.tex
Initial checkin of paper

% XeLaTeX can use any Mac OS X font. See the setromanfont command below.
% Input to XeLaTeX is full Unicode, so Unicode characters can be typed directly into the source.

% The next lines tell TeXShop to typeset with xelatex, and to open and save the source with Unicode encoding.

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[11pt]{article}
\usepackage{geometry,authblk,multicol,amsmath}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}

% Will Robertson's fontspec.sty can be used to simplify font choices.
% To experiment, open /Applications/Font Book to examine the fonts provided on Mac OS X,
% and change "Hoefler Text" to any of these choices.

\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}
\setromanfont[Mapping=tex-text]{Times New Roman}
\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Gill Sans}
\setmonofont[Scale=MatchLowercase]{Andale Mono}

\title{Mixture Models for Single Cell Assays}
\author[1]{Greg Finak}
\author[1]{\ldots Others \ldots}
\author[1]{Raphael Gottardo}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
Flow cytometry, specifically the intracellular cytokine staining (ICS) assay is an important tool to assess immune response to vaccines in clinical trials, providing both phenotypic and functional information about the immune system. The goal of the ICS assay is to determine whether an individual's immune system responds to a vaccine, as well as the nature of the response. ICS assays test multiple antigens and multiple cytokines across hundreds of individuals, with the usual goal of identifying individuals in whom some rare subpopulation of cells is greater in a stimulated conditions than in a control. Maximizing sensitivity and specificity to detect responders to vaccination is one of the most important considerations in designing an ICS assay, particularly due to the rarity of the cell subpopulations being measured. The measured cell subpopulations are usually rare, such as antigen--specific cytokine producing T--cells, and are often measured as a fraction of their parent population. One assay typically measures multiple cytokines simultaneously. Consequently, ICS data is typically analyzed as a series of 2x2 contingency tables using Fisher's exact test. This is problematic for a number of reasons. Fisher's exact test can be overly conservative for small counts. Additionally, the data in the 2x2 table for the ICS assay does not truly have fixed margins since the stimulated and unstimulated cells come from independent experiments. In this paper we present a coherent empirical-Bayes mixture model framework that models cytokine–specific T-cell response across all individuals in a study simultaneously, while modelling the stimulated and unstimulated cell counts independently. We show that our model increases the sensitivity and specificity for positivity calls compared to the classical approach, using simulations and real data from vaccine trials.

\end{abstract}
\section{Introduction}
Intracellular cytokine staining assays are used in the context of vaccine trials to provide a sensitive and quantitative assessment of antigen--specific T cell response to vaccine, including phenotyping and measurement of multiple effector functions~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.

%An automated system for analyzing data from ICS assays has been developed by LabKey Software in collaboration with the HVTN, and is being used for current HIV vaccine trials~\cite{Shulman:2008jv}. A limitation of this system is that it relies on manual gating using external tools, and thus all the limitations of manual gating apply. The process is time consuming, error prone and subjective. Laboratories using the system overcome this drawback by performing highly controlled experiments that are suitable for use with template gates and requiring only occasional manual adjustment by the analyst~\cite{Horton:2007tsa,Shulman:2008jv}. Despite these mitigating factors, gating rare cytokine positive cell populations in ICS assays remains challenging, and biologically interesting markers with antibodies that have high staining variability cannot be analyzed in an automated fashion since they require considerable effort to make manual adjustments.
%
%In order to make further advancements in computational approaches to flow cytometry data analysis, computational researchers must be able to access the current ``gold--standard'' manual gating analysis to make quantitative comparisons. Accessibility of manual gating and data preprocessing pipelines for flow cytometry studies has been sporadic at best. LabKey allows access to population statistics extracted from flowJo workspaces, but doesn't allow access to the manual gates directly~\cite{Shulman:2008jv}. The BioConductor package \textit{flowFlowJo} allows import of a subset of older flowJo workspaces ($\leq$ v7.0) into BioConductor for use with \textit{flowCore} and other packages\cite{Hahne:2009vv, Gosink:2009iq}. However, this package does not support newer flowJo workspaces (>7.0) generated by the Macintosh version of the software, which are used by many laboratories, and limitations of the data structures representing an FCM workflow in BioConductor do not readily allow for sample to sample variability in the gate definitions for corresponding populations. Furthermore \textit{flowFlowJo} and the BioConductor flow tools are impractical for the analysis of realistically sized clinical flow cytometry data sets, which often exceed the memory capacity of the workstations running the software.

%To address problems gating cell subpopulations when there is substantial sample--to--sample variability algorithms have been developed for FCM data normalization, including \textit{fdaNorm} and \textit{gaussNorm} in the \textit{flowStats} package in BioConductor~\cite{Hahne:2010uh}. These algorithms are designed to normalize FCM data on a per--channel basis across multiple samples. Briefly, these normalization algorithms perform density estimation, feature extraction (peak identification), peak--matching (registration), and finally alignment. However, their utility has remained largely unproven in real--world applications, partly due to the inherent limitations already mentioned of working with large FCM data sets in BioConductor, partly due to difficulties with the peak--matching step when the data densities are multi--modal (i.e. three or more modes) and variable across samples, and partly due to difficulties in accessing real--world manually gated data for direct comparison. 
%
%These usability issues have hampered the adoption of the otherwise powerful and cutting--edge FCM data analysis toolset available in BioConductor. In this paper, we present a series of packages and extensions to the BioConductor FCM toolset that address these problems.

%The \textit{flowWorkspace} package allows the import of manual gates, compensation matrices and data transformations from flowJo workspaces generated by Macintosh versions of flowJo, making a majority of the FCM analyses archived in laboratories available for analysis in BioConductor. \textit{flowWorkspace} defines several new data structures for working with FCM data in BioConductor, including the \textit{GatingHierarchy} representing the hierarchical set of gates applied to an FCM file, and the \textit{GatingSet} representing a collection of \textit{GatingHierarchies} on multiple FCM files. Importantly, while the \textit{GatingSet} requires all samples to have the same subpopulations defined, it allows the precise gate definitions to vary across samples, more accurately reflecting the manner in which gates may be adjusted for individual samples within \textit{flowJo} by the data analyst. 
%
%The \textit{flowWorkspace} package is fully integrated with the new \textit{ncdfFlow} package that provides disk--based access to FCM data via the netCDF C library. The package defines a new \textit{ncdfFlowSet} object that inherits from flowCore's \textit{flowSet}. We have further extended the \textit{flowCore} package to ensure existing methods and algorithms are compatible with \textit{ncdfFlow}. Together, \textit{flowWorkspace} and \textit{ncdfFlow} allow  real--world FCM datasets many Gigabytes in size to be analyzed using the BioConductor toolset on regular workstations.
%
%Finally, we have made several improvements to the \textit{fdaNorm} FCM data normalization algorithm in \textit{flowStats}, making it more robust, and amenable to integration in an FCM data analysis pipeline. First, we have adapted the algorithm to take advantage of the \textit{ncdfFlow} package allowing it to work on large data sets. We have also added the ability to chunk the data into several parts, making the algorithm more memory efficient, and importantly, we have added the ability to normalize FCM data to a target distribution defined by a specific FCM file. Support for \textit{flowWorkspace} has been added to the normalization algorithm, allowing the user to easily include a normalization step for flowJo defined manual gates. Finally, and perhaps most importantly, we have developed a new sequential normalization strategy in the context of manual gating. Rather than normalizing FCM data once, at the beginning of gating, our approach performs a normalization step prior to each hierarchical gating step. This has several benefits: primarily, population matching is simplified for each normalization step since overlapping populations appearing higher in the gating hierarchy are excluded from the density. Such a systematic sequential normalization strategy improves upon normalization after dump gates only. 
%
%We combine these tools to develop an integrated data analysis pipeline for ICS assays and use it to analyze two real--world, full size, vaccine trial, ICS data sets from the HVTN, comparing the results against gold standard manual gating.

\section{Materials and Methods}
\subsection{Vaccine Trial ICS Dataset Description}
HVTN054 is a phase 1 (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}. The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}. Vaccine was given at two increasing doses, as well as a placebo. T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}. The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}. The sample size consisted of 20 vaccine and four placebo recipients. Statistical analysis of the original positivity calls is described in the original publication~\cite{Peiperl:2010ej}.

%HVTN080 is a phase 1 clinical trial to evaluate the safety and
%immunogenicity of PENNVAX\texttrademark-B (gag, pol and env) vaccine, with or without IL-12 DNA plasmid, delivered via electroporation in healthy, HIV-1 uninfected adult participants.

\subsection{Data Import, Preprocessing, and Gating}
The gated ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4). Data were preprocessed using the flowJo--defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T--cells for each sample and stimulation~\cite{Hahne:2009vv}.

\subsection{Statistical Analysis of Responder and Non--responder calls}
Below, we summarize the methods for statistical analysis of responder and non--responder calls in the published trial, as well as the methods compared in this paper.
\subsubsection{Statistical Analysis in the Published Trial}
The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}. In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine--positive T-cells in the antigen--stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine--positive T-cells for the negative control (unstimulated) sample from the same individual. In the original trial, significance was measured via one--sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual. A discrete Bonferroni adjustment for multiple comparisons was applied, and stimulations with an adjusted p--value below $\alpha = 0.00001$ were called positive.

\subsection{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold. Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods. Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines \todo[inline]{IS THIS CORRECT, OR IS IT WITHIN INDIVIDUALS ACROSS STIMULATIONS?}. In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.

\subsection{Two Competing Beta--Binomial Models}
Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach. We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples). For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively. Note that usually, $N_u \ne N_s$. The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants. For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table: 
% latex table generated in R 2.14.0 by xtable 1.5-6 package
% Wed Oct 12 09:41:50 2011
\begin{table}[ht]
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for cytokine positive and cytokine negative events between stimulated and unstimulated conditions}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}

The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as:
\begin{align}
  \text{if }&p_{si}=p_{ui}\equiv p_{0i};& n_{si} \sim \mathrm{Bin}(N_{si},p_u);\text{ }& n_{ui} \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null}\\
 \text{if }&p_{si}>p_{ui};& n_{si} \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& n_{ui} \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:alternate}
 \end{align}
Where $p_{si}$ and $p_{ui}$ are the unobserved proportions. Equation~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control. Equation~\eqref{eq:alternate} represents the  alternate hypothesis where the cytokine response is stronger in the stimulation than in the control. We place a common Beta prior on the $p_{si}$ and $p_{ui}$ across individuals, as shown:

 \begin{align}
 p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s)\label{eq:stimprior}\\
 p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:nullprior}
 \end{align}
If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models. Given this formulation, the posterior probability of the data given that it is generated by model \eqref{eq:null}, is:
 \begin{align}
  	\mathrm{Pr}(y_i|\alpha_u,\beta_u)
	&=\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}\\
	\intertext{with marginal log--likelihood:}
	\begin{split}
	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N_{si}}{n_{si}}}+\log{\binom{N_{ui}}{n_{ui}}}+\right.\\
	\left.\log{\left(\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLL}
	\end{split}
 \end{align} 
Thus, in the case of no response to stimulation, the counts for the stimulated and unstimulated samples are modelled as draws from the same "unstimulated" Beta--binomial distribution. Note that the unobserved parameters, $p_{si}, p_{ui}$ have been integrated out to give the marginal log--likelihood.

If the data is generated by model~\eqref{eq:alternate}, the posterior probability of the data is given by:
\begin{align}
	\begin{split}
\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) =\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
	 \frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
\label{eq:model2post}
\end{split}
	\intertext{with marginal log--likelihood:}
\begin{split}
\mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N_{ui}}{n_{ui}}+ \log\binom{N_{si}}{n_{si}}+ \log\left(\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\right)+\right. \\ \left. \log\left(\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)\right) + \right. \\ \left. \log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right) \right. \right. \\  \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)}\left(\mathrm{I_{1-p_{ui}}}(N_{si}-n_{si}+\beta_s,n_{si}+\alpha_s)\right)d p_{ui}\right] \\ -  \left. \log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right) \right. \right. \\  \left. \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)} \left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}\right] \right\}\label{eq:model2MLL}
\end{split}
\end{align}

The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions. $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, which is just the CDF of Beta distribution with parameters $\alpha_s,\beta_s$, leaving a 1--dimensional integration for the ratio of normalizing constants.

\subsection{The Mixture of Beta--Binomials}
Although we have specified the two models for the data, we do not know which observation was generated by which model. Clearly, not all individuals are expected to exhibit an immune response to a stimulation. Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}. We capture this uncertainty with a mixture framework of the two competing beta--binomial models. The likelihood for the mixture is given by:
\begin{align}
\begin{split}
%L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi_1 Pr(y_i|\alpha_u,\beta_u) +\pi_2 Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) \right] ,\\ \sum\limits_{k=1}\limits^2\pi_k=1\\
L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi_1 f_1(y_i|\theta_1) +\pi_2 f_2(y_i|\theta_2) \right] ,\\ \sum\limits_{k=1}\limits^2\pi_k=1
\end{split}
\end{align}
Where $\theta_1=\{\alpha_u,\beta_u\}, \theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, $\pi_1$ is the fraction of observations exhibiting no response to stimulation, $\pi_2$ the fraction of observations exhibiting a response to stimulation, and $f_1 = Pr(y_i|\alpha_u,\beta_u), f_2 = Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)$ from \eqref{eq:model1post} and \eqref{eq:model2post}, above.

The unobserved component memberships are treated as missing data and modelled as random variables  $\mathbf{z}_i = \left\{z_{i1},(1-z_{i1})\right\}$

$$
z_{ik} = \left\{ \begin{array}{rl}
1 &\mbox{ if observation $i$ is from the $k$'th model (component)} \\
0&\mbox{ otherwise}
\end{array} \right.
$$

Each $\mathbf{z}_{i}$ follows an independent multinomial distribution with one trial and  parameters $\boldsymbol{\pi}=\left\{\pi_1,1-\pi_1\right\}$. Given the $z_i$'s, the complete data log--likelihood is:

\begin{align}
\begin{split}
\mathcal{L}_c(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y},\mathbf{z})=\sum\limits_{i=1}\limits^P\sum\limits_{k=1}\limits^2{z_{ik}\left[ \log\pi_k+\log f_k(y_i|\theta_k)\right]}\label{eq:CDLL}
\end{split}
\end{align}

In this form, we use the expectation--maximization (EM) algorithm~\cite{Dempster:1977ul} to fit the model.

\subsubsection*{E--step}
Given the model parameters $\boldsymbol{\Psi}=\left\{\alpha_u,\beta_u,\alpha_s,\beta_s,\pi_k\right\}$, and the data $\mathbf{y}$, we estimate the unobserved component memberships, $\mathbf{Z_i}$  by computing the conditional expectation of the $\mathbf{Z_i}$'s,  $\mathbb{E}\boldsymbol{_\Psi}(\mathbf{Z_i}|\mathbf{y}_i)$:
\begin{align}
\tilde z_{ik} &= \frac{\pi_k f_k(\mathbf{y}_i|\theta_k)}{\sum\limits_{k=1}\limits^{2}\pi_kf_k(\mathbf{y}_i|\theta_k)}
\end{align}

\subsubsection*{M--step}
Finally, given the $\tilde{z}_{ik}$, we update the estimates of the model parameters to maximize the conditional expectation of the complete--data log--likelihood. The mixing proportions are given by:
\begin{align}
\hat\pi_k = \frac{ \sum_i^P \tilde z_{ik}}{n}
\end{align}

There is no closed form for the model hyper--parameters, $\alpha_u,\beta_u,\alpha_s,\beta_s$, and they are estimated via numerical optimization using R's \textit{optim} function. For this purpose they are re--parameterized as $\mu_u=\frac{\alpha_u}{\alpha_u+\beta_u}$ and $S=\alpha_u+\beta_u$ (likewise for the $\alpha_s,\beta_s$), corresponding to the mean and sample size of the prior distributions.

\subsubsection*{Initialization}
We initialize the $z_{ik}$'s using Fisher's exact test to assign each observation to either the $p_{si}=p_{ui}$ or $p_{si} > p_{ui}$ components. We then use the $\hat{z}_i$'s to initialize the hyper--parameters to their method--of--moments estimates:
\begin{align}
\hat{\alpha} = \hat{\mu}\left(\frac{\hat{\mu}(1-\hat{\mu})}{\hat{\sigma}^2} -1\right)\\
\hat{\beta} =  (1-\hat{\mu}\left(\frac{\hat{\mu}(1-\hat{\mu})}{\hat{\sigma}^2} -1\right)
\end{align}

Where $\hat{\mu}$ and $\hat{\sigma}^2$ are the sample mean and sample variance estimates, given the $z_{ik}$'s.

\subsection*{Generalization to Multiple Cytokines and Polyfunctionality with the Multinomial Dirichlet}
The model can be generalized to handle multiple cytokines in a single stimulation, in order to assess polyfunctional cytokine responses of T--cells. We use the Multinomial--Dirichlet family of distributions to model counts of events in two \textit{different} 2x2 contingency tables. The observed data can be represented in the following way:

\begin{table}[ht]
\centering
\parbox{0.8\linewidth}{
\caption{Contingency tables for counts of cells expressing two cytokines between stimulated and unstimulated conditions. $n_{\{s,u\}j}$ denotes observed counts for stimulated or unstimulated table cell $j$, and individual $i$  }\label{tab:multdir}
\begin{minipage}[b]{0.5\linewidth}
\centering
\begin{tabular}{rrr}
\multicolumn{3}{c}{Stimulated}\\
  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine A}\\
 & Negative & Positive \\ 
 \multicolumn{1}{l}{Cytokine B}&&\\
  \hline
Negative &   $n_{si1}$ &   $n_{si2}$ \\ 
Positive &   $n_{si3}$ &   $n_{si4}$ \\ 
   \hline
\end{tabular}
\end{minipage}
\begin{minipage}[b]{0.5\linewidth}
\centering
\begin{tabular}{rrr}
\multicolumn{3}{c}{Unstimulated}\\
  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine A}\\
 & Negative & Positive \\ 
 \multicolumn{1}{l}{Cytokine B}&&\\
  \hline
Negative &   $n_{ui1}$ &   $n_{ui2}$ \\ 
Positive &   $n_{ui3}$ &   $n_{ui4}$ \\ 
   \hline
\end{tabular}
\end{minipage}
}
\end{table}

Where the vector of observed counts for individual $i$ in the stimulated or unstimulated sample is denoted: $\bar{n}_{\{s,u\}i} = \{n_{\{s,u\}ij}\} ; j\in\{1\ldots 4\}$, and $j$ indexes the cells of the appropriate contingency table shown in~Table~\ref{tab:multdir}. The counts are modelled as draws from different multinomial distributions:
\begin{align}
\text{if } & \bar{p}_{si}=\bar{p}_{ui};& \bar{n}_{ui} \sim \mathcal{M}(\bar{p}_{ui},N_{ui});\bar{n}_{si} \sim \mathcal{M}(\bar{p}_{ui},N_{si})\\
\text{if } & \bar{p}_{si} \ne \bar{p}_{ui};& \bar{n}_{ui} \sim \mathcal{M}(\bar{p}_{ui},N_{ui});\bar{n}_{si} \sim \mathcal{M}(\bar{p}_{si},N_{si})
\end{align}
with Dirichlet priors on the proportions:
\begin{align}
\bar{p}_{si} \sim \mathrm{Dir}(\bar{\alpha}_s) ; \bar{p}_{ui} \sim \mathrm{Dir}(\bar{\alpha}_u)
\end{align}

For the null component, where $\bar{p}_{s}=\bar{p}_{u}$ the marginal likelihood is given by: 
\begin{align}
\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u) &= \prod_{i=0}^P\frac{ \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\\
\end{align}
Where $\mathrm{B_j}$ is the $\mathrm{j}$--dimensional Beta function: $\frac{\prod_{j=1}^J\Gamma(\alpha_j)}{\Gamma(\sum\alpha_j)}$.

The marginal likelihood for a component where $p_{sj} \ne p_{uj}$ for all $j$, is given by:
\begin{align}
\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u,\bar{\alpha}_s) &= \prod_{i=0}^P\frac{  \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}) \mathrm{B_j}(\bar{\alpha}_{s}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_s)\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\label{eq:postmult}
\end{align}

Without loss of generality, if only some $p_j$ are different between stimulated and unstimulated samples, the appropriate components of $\alpha_j$  can be substituted in the calculation of the likelihood eq~\eqref{eq:postmult}.

\subsubsection*{Mixture Model Complexity}
We may wish to detect any of $2^3 = 8$ different possible scenarios where the proportion of events in corresponding cells of the contingency tables are either equal or unequal between stimulated and unstimulated conditions. Such a model would have 8 components and 55 parameters. However, if we recognize that the models can be nested, i.e. that parameters can be shared across components with similar outcomes, then the number of parameters can be reduced to 19, and further to 15 if we only consider components where any one cell of the tables differs between stimulated and unstimulated conditions. This is outlined in Table~\ref{tab:nesting}.
\begin{table}
\centering
\parbox{0.8\linewidth}{\caption{Nesting of models and parameter counts. Each row is a model component. The three columns correspond to cells two, three, and four of the contingency tables shown in Table~\ref{tab:multdir}. An open circle at a position indicates that the component models $p_{sj}=p_{uj}$, and a filled circle indicates that the component models $p_{sj}\ne p_{uj}$. The number of additional parameters that need to be estimated by including each additional component in the mixture model is in the fourth column (number of parameters for proportions + number of parameters for component weights).}\label{tab:nesting}
\centering
\begin{tabular}{ccccc}
\multicolumn{3}{c}{Cell of Table}\\
\hline
cell 2&cell 3&cell 4&\# of parameters\\
\hline
$\circ$&$\circ$&$\circ$&6+1\\
$\circ$&$\circ$&$\bullet$&2+1\\
$\circ$&$\bullet$&$\circ$&2+1\\
$\bullet$&$\circ$&$\circ$&2+1\\
$\circ$&$\bullet$&$\bullet$&0+1\\
$\bullet$&$\bullet$&$\circ$&0+1\\
$\bullet$&$\circ$&$\bullet$&0+1\\
$\bullet$&$\bullet$&$\bullet$&0\\
\end{tabular}
}
\end{table}

\section*{Results}
\subsection*{Simulations}
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{Figures/simulations} 
   \caption{Performance of the Beta-binomial mixture vs Fisher's exact test in simulated data. Data were simulated from a model with hyper--parameters estimated from day 28, Gag1 stimulated, CD4+, IL2 expressing T--cells in the HVTN054 data set. We simulated 500 observations, with a response rate of 40\%, and increasing numbers of cells, ten times each. The performance, measured by the AUC, of the beta--binomial mixture compared to Fisher's exact test is shown in the first panel, as a function of increasing number of events. The estimated and true prior distributions are shown in the second panel for one simulated data set, with N=150,000 events. The ROC curve for Fisher's exact test and the Beta--binomial model for the same simulated data set are shown in the third panel. The observed vs expected false discovery rate for Fisher's exact test and the Beta--binomial model are shown for the same data set in the fourth panel.}
   \label{fig:simulations}
\end{figure}

\subsection*{HVTN054 ICS Data}
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6in]{Figures/HVTN054LikelihoodAndVolcanoplots} 
   \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28. A significant difference between control and stimulated samples is called at the 1\% FDR threshold (red) for the Beta--binomial model, and at 1\% for FDR adjusted p--values from Fisher's exact test (green). The likelihood surface shows the observed counts from stimulated and unstimulated samples. The volcano plot shows the difference between the proportion of cytokine positive cells in the stimulated and unstimulated samples, for the maximum likelihood estimates of the proportions (triangles) and for the MAP estimates (crosses)}
   \label{fig:simulations}
\end{figure}

\section*{Discussion}

\section*{Conclusions}

%\renewcommand{\thesection}{S.\arabic{section}}
%\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
%\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}
%\section{Supplementary Information}
%\subsection{Derivation of the Beta--Binomial Model for $p_s=p_u$ and $p_s>p_u$}
%\label{sec:derivation}
%
%We derive the posterior predictive distribution and marginal log--likelihood for the null model ($p_s=p_u$) and the alternative model ($p_s>p_u$) below: patient indices, $i$ on the $\left<N_s,n_s,N_u,n_u\right>$ are omitted for clarity. 
%For model \eqref{eq:null}, ($p_s=p_u, \alpha_s=\alpha_u, \beta_s=\beta_u$), the posterior predictive distribution of the data given the model is:
% \begin{align}
% 	\mathrm{Pr}(y_i|\alpha_u,\beta_u) &= \int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i,p_u|\alpha_u,\beta_u)dp_u\\
%	&=\int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u\\
%	\begin{split}
%		=\int\limits_{p_u=0}\limits^{1}\binom{N_s+n_s}{n_s}p_u^{n_s}(1-p_u)^{N_s}\binom{N_u+n_u}{n_u}p_u^{n_u}(1-p_u)^{N_u}\cdot \\ 
%		\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1}dp_u
%	 \end{split}\\
%	 \begin{split}
% 	=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\cdot\\ \int\limits_{p_u=0}\limits^{1}p_u^{n_s+n_u+\alpha_u-1}(1-p_u)^{N_s+N_u+\beta_u-1}dp_u
%	\end{split}\\
%	\intertext{The integrand is the kernel of a beta distribution with parameters $\left(n_u+n_s+\alpha_u,N_u+N_s+\beta_u\right)$, giving a closed form expression for the  posterior predictive distribution of model \eqref{eq:null}:}
%	\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
%	\intertext{with marginal log--likelihood:}
%	\begin{split}
%	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N^i_{s}+n^i_{s}}{n^i_{s}}}+\log{\binom{N^i_{u}+n^i_{u}}{n^i_{u}}}+\right.\\
%	\left.\log{\left(\mathrm{B}(n^i_s+n^i_u+\alpha_u,N^i_s+N^i_u+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLLsupp}
%	\end{split}
% \end{align} 
% 
%For model \eqref{eq:alternate}, ($p_s>p_u$), the posterior predictive distribution of the data given the model is:
%\begin{align}
% 	\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) &= \int\limits_{p_u=0}\limits^1\int\limits_{p_s>p_u}\limits^{1} \mathrm{Pr}(y_i,p_u,p_s|\alpha_u,\beta_u,\alpha_s,\beta_s) dp_u  dp_s\label{eq:postpredmodel2}\\
%	\intertext{Assuming independence of stimulated and unstimulated observations in \eqref{eq:postpredmodel2}:} 
%	&=\int\limits_{p_u=0}\limits^{1}\int\limits_{p_s>p_u}\limits^1 \mathrm{Pr}(n_s|p_s)\mathrm{Pr}(n_u|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)\mathrm{Pr}(p_s|\alpha_s,\beta_s)dp_u dp_s\\
%		&=\int\limits_{p_u=0}\limits^{1}\mathrm{Pr}(n_u|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u \int\limits_{p_s>p_u}\limits^1 \mathrm{Pr}(n_s|p_s)\mathrm{Pr}(p_s|\alpha_s,\beta_s)dp_s\\
%		\begin{split}
%	=\int\limits_{p_u=0}\limits^{1}\left[ \binom{N_u+n_u}{n_u}p_u^{n_u}(1-p_u)^{N_u} \frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1}d p_u\right]\cdot \\ \int\limits_{p_s>p_u}\limits^{1}\left[ \binom{N_s+n_s}{n_s}p_s^{n_s}(1-p_s)^{N_s} \frac{1}{\mathrm{B}(\alpha_s,\beta_s)}p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}d p_s\right]
%	\end{split}\\
%	\begin{split}
%	 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
%	 \int\limits_{p_u=0}\limits^{1}\left[p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1} d p_u\right]\int\limits_{p_s>p_u}\limits^{1}\left[p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s+\beta_s-1}d p_s\right]
%	\end{split}
%	\end{align}
%	\begin{align}
%	\begin{split}
%	 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
%	 \int\limits_{p_u=0}\limits^{1}\left[p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1}d p_u \right]\left(1-\int\limits^{p_u}\limits_{p_s=0}\left[p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s+\beta_s-1}d p_s\right]\right)
%	\end{split}
%	\intertext{The second integral can be expressed as the CDF of the beta distribution, also known as the regularized incomplete beta function, $\mathrm{I}_x(\alpha,\beta)$}
%	\begin{split}
%	 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_s+\alpha_s,N_s+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
%	 \int\limits_{p_u=0}\limits^{1}\left[p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1} \right]\left(1-\mathrm{I_{p_s}}(n_s+\alpha_s,N_s+\beta_s)\right)d p_u
%	\end{split}\\
%	\intertext{giving the posterior predictive distribution of model \eqref{eq:alternate}}
%	\begin{split}
%		 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{\mathrm{B}(n_u+\alpha_u,N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_s+\alpha_s,N_s+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
%	 \int\limits_{p_u=0}\limits^{1}\left[\frac{1}{\mathrm{B}(n_u+\alpha_u,N_u+\beta_u)}p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1} \right]\left(1-\mathrm{I_{p_u}}(n_s+\alpha_s,N_s+\beta_s)\right)d p_u\label{eq:model2postsupp}
%	\end{split}
%\end{align}
%The integral in equation~\eqref{eq:model2postsupp} is computed numerically. The marginal log--likelihood of model~\eqref{eq:alternate} is then:
%\begin{equation}
%		 \begin{split}
%		 \mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}+(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N^i_u+n^i_u}{n^i_u}+ \log\binom{N^i_s+n^i_s}{n^i_s}+ \log\left(\mathrm{B}(n^i_u+\alpha_u,N^i_u+\beta_u)\right)+ \log\left(\mathrm{B}(n^i_s+\alpha_s,N^i_s+\beta_s)\right)+\right. \\ \left. \log\left(\hspace{1em}\int\limits_{p_u=0}\limits^{1}\left[\frac{1}{\mathrm{B}(n^i_u+\alpha_u,N^i_u+\beta_u)}p_u^{n^i_u+\alpha_u-1}(1-p_u)^{N^i_u+\beta_u-1} \right]\left(1-\mathrm{I_{p_u}}(n^i_s+\alpha_s,N^i_s+\beta_s)\right)d p_u\right)\right\}\label{eq:model2MLLsupp}
%\end{split}
%\end{equation}
%\subsection{The Beta-Binomial Mixture Model}
%For a given stimulation, not all individuals are expected to exhibit an immune response to that stimulation. Therefore, for a set of ICS data, any single observation, ($n^i_s,n^i_u$) could either be derived from model~\eqref{eq:null}, or from model~\eqref{eq:alternate}. We model this situation explicitly using a mixture--model framework.
\clearpage
\bibliographystyle{unsrtnat}
\bibliography{MIMOSA}

\end{document}

### 2012-02-04 MIMOSA.tex
modified the abstract and intro to be more generic, referring more to single cell assays in general.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \begin{abstract}
-Flow cytometry, specifically the intracellular cytokine staining (ICS) assay is an important tool to assess immune response to vaccines in clinical trials, providing both phenotypic and functional information about the immune system.
-The goal of the ICS assay is to determine whether an individual's immune system responds to a vaccine, as well as the nature of the response.
-ICS assays test multiple antigens and multiple cytokines across hundreds of individuals, with the usual goal of identifying individuals in whom some rare subpopulation of cells is greater in a stimulated conditions than in a control.
-Maximizing sensitivity and specificity to detect responders to vaccination is one of the most important considerations in designing an ICS assay, particularly due to the rarity of the cell subpopulations being measured.
-The measured cell subpopulations are usually rare, such as antigen--specific cytokine producing T--cells, and are often measured as a fraction of their parent population.
-One assay typically measures multiple cytokines simultaneously.
-Consequently, ICS data is typically analyzed as a series of 2x2 contingency tables using Fisher's exact test.
-This is problematic for a number of reasons.
-Fisher's exact test can be overly conservative for small counts.
-Additionally, the data in the 2x2 table for the ICS assay does not truly have fixed margins since the stimulated and unstimulated cells come from independent experiments.
-In this paper we present a coherent empirical-Bayes mixture model framework that models cytokine–specific T-cell response across all individuals in a study simultaneously, while modelling the stimulated and unstimulated cell counts independently.
-We show that our model increases the sensitivity and specificity for positivity calls compared to the classical approach, using simulations and real data from vaccine trials.
-
+Immunological endpoints in vaccine trials are measured through a variety of different assays that, often, provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of specific genes in single cells from specific cell sub--populations.
+Although these measurements are continuous, they are generally discretized for analysis, with each cell called either positive or negative for a measured variable, depending on some predetermined threshold.
+A motivating example is the intracellular cytokine staining (ICS) flow cytometry assay.
+This assay is used to assess an individual's immune response to a vaccine by measuring the abundance of antigen--specific T--cell subpopulations producing one or more specific cytokines.
+In this way, the assay provides phenotypic and functional information about the immune system.
+ICS assays are used to test multiple antigens and multiple cytokines across hundreds of individuals in a vaccine study.
+The threshold for calling an individual cell as positive or negative for a given cytokine is predetermined by the gating step of the analysis.
+The thresholded data is discretized and treated as count data of the number of cytokine positive and negative cells in each stimulated and control sample.
+The goal of the analysis is to identify the ``vaccine responders''.
+These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline (the unstimulated control).
+The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern.
+The typical approach to analysis using Fisher's exact test is problematic for two reasons.
+It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test.
+Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.
+In this paper we present a cohort framework based on an empirical-Bayes mixture of Beta--binomial or Dirichlet--Multinomial distributions, for analyzing count data derived from such thresholded single--cell assays.
+In general, any single--cell assay in which continuous data can be thresholded to classify individual cells into either ``positive'' or ``negative'' categories with respect to one or more variables and compared across different conditions, is suitable for analysis using our method and the extensions presented herein.
+Using the ICS assay as a motivating example, our method models cytokine–specific T-cell response across all individuals simultaneously while modelling the stimulated and unstimulated cell counts independently.
+Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test.
 \end{abstract}
 \section{Introduction}
-Intracellular cytokine staining assays are used in the context of vaccine trials to provide a sensitive and quantitative assessment of antigen--specific T cell response to vaccine, including phenotyping and measurement of multiple effector functions~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Single--cell assays are an important tool in vaccine trials to measure different characteristics of specific immune cell subpopulations.
+These assays typically measure multiple variables simultaneously for each individual cell in a homogeneous mixture such as whole blood.
+These variables can be used to classify individual cells in the mixture into more homogeneous phenotypes and functional categories.
+These, in turn, provide a functional read--out of the immune system.
+A motivating example is the intracellular cytokine staining assay from flow cytometry, which, in the context of vaccine studies, is used to provide a sensitive and quantitative assessment of antigen--specific T--cell response to a vaccine.
+It provides both phenotypic information about cell subpopulations (CD4 and CD8 T--cells) in whole blood, and functional information (cytokine expression) about how those subpopulations express different cytokines in response to an antigen stimulus~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.
-
-%An automated system for analyzing data from ICS assays has been developed by LabKey Software in collaboration with the HVTN, and is being used for current HIV vaccine trials~\cite{Shulman:2008jv}.
-A limitation of this system is that it relies on manual gating using external tools, and thus all the limitations of manual gating apply.
-The process is time consuming, error prone and subjective.
-Laboratories using the system overcome this drawback by performing highly controlled experiments that are suitable for use with template gates and requiring only occasional manual adjustment by the analyst~\cite{Horton:2007tsa,Shulman:2008jv}.
-Despite these mitigating factors, gating rare cytokine positive cell populations in ICS assays remains challenging, and biologically interesting markers with antibodies that have high staining variability cannot be analyzed in an automated fashion since they require considerable effort to make manual adjustments.
-%
-%In order to make further advancements in computational approaches to flow cytometry data analysis, computational researchers must be able to access the current ``gold--standard'' manual gating analysis to make quantitative comparisons.
-Accessibility of manual gating and data preprocessing pipelines for flow cytometry studies has been sporadic at best.
-LabKey allows access to population statistics extracted from flowJo workspaces, but doesn't allow access to the manual gates directly~\cite{Shulman:2008jv}.
-The BioConductor package \textit{flowFlowJo} allows import of a subset of older flowJo workspaces ($\leq$ v7.0) into BioConductor for use with \textit{flowCore} and other packages\cite{Hahne:2009vv, Gosink:2009iq}.
-However, this package does not support newer flowJo workspaces (>7.0) generated by the Macintosh version of the software, which are used by many laboratories, and limitations of the data structures representing an FCM workflow in BioConductor do not readily allow for sample to sample variability in the gate definitions for corresponding populations.
-Furthermore \textit{flowFlowJo} and the BioConductor flow tools are impractical for the analysis of realistically sized clinical flow cytometry data sets, which often exceed the memory capacity of the workstations running the software.
-
-%To address problems gating cell subpopulations when there is substantial sample--to--sample variability algorithms have been developed for FCM data normalization, including \textit{fdaNorm} and \textit{gaussNorm} in the \textit{flowStats} package in BioConductor~\cite{Hahne:2010uh}.
-These algorithms are designed to normalize FCM data on a per--channel basis across multiple samples.
-Briefly, these normalization algorithms perform density estimation, feature extraction (peak identification), peak--matching (registration), and finally alignment.
-However, their utility has remained largely unproven in real--world applications, partly due to the inherent limitations already mentioned of working with large FCM data sets in BioConductor, partly due to difficulties with the peak--matching step when the data densities are multi--modal (i.e.
-three or more modes) and variable across samples, and partly due to difficulties in accessing real--world manually gated data for direct comparison.
-%
-%These usability issues have hampered the adoption of the otherwise powerful and cutting--edge FCM data analysis toolset available in BioConductor.
-In this paper, we present a series of packages and extensions to the BioConductor FCM toolset that address these problems.
-
-%The \textit{flowWorkspace} package allows the import of manual gates, compensation matrices and data transformations from flowJo workspaces generated by Macintosh versions of flowJo, making a majority of the FCM analyses archived in laboratories available for analysis in BioConductor.
-\textit{flowWorkspace} defines several new data structures for working with FCM data in BioConductor, including the \textit{GatingHierarchy} representing the hierarchical set of gates applied to an FCM file, and the \textit{GatingSet} representing a collection of \textit{GatingHierarchies} on multiple FCM files.
-Importantly, while the \textit{GatingSet} requires all samples to have the same subpopulations defined, it allows the precise gate definitions to vary across samples, more accurately reflecting the manner in which gates may be adjusted for individual samples within \textit{flowJo} by the data analyst.
-%
-%The \textit{flowWorkspace} package is fully integrated with the new \textit{ncdfFlow} package that provides disk--based access to FCM data via the netCDF C library.
-The package defines a new \textit{ncdfFlowSet} object that inherits from flowCore's \textit{flowSet}.
-We have further extended the \textit{flowCore} package to ensure existing methods and algorithms are compatible with \textit{ncdfFlow}.
-Together, \textit{flowWorkspace} and \textit{ncdfFlow} allow  real--world FCM datasets many Gigabytes in size to be analyzed using the BioConductor toolset on regular workstations.
-%
-%Finally, we have made several improvements to the \textit{fdaNorm} FCM data normalization algorithm in \textit{flowStats}, making it more robust, and amenable to integration in an FCM data analysis pipeline.
-First, we have adapted the algorithm to take advantage of the \textit{ncdfFlow} package allowing it to work on large data sets.
-We have also added the ability to chunk the data into several parts, making the algorithm more memory efficient, and importantly, we have added the ability to normalize FCM data to a target distribution defined by a specific FCM file.
-Support for \textit{flowWorkspace} has been added to the normalization algorithm, allowing the user to easily include a normalization step for flowJo defined manual gates.
-Finally, and perhaps most importantly, we have developed a new sequential normalization strategy in the context of manual gating.
-Rather than normalizing FCM data once, at the beginning of gating, our approach performs a normalization step prior to each hierarchical gating step.
-This has several benefits: primarily, population matching is simplified for each normalization step since overlapping populations appearing higher in the gating hierarchy are excluded from the density.
-Such a systematic sequential normalization strategy improves upon normalization after dump gates only.
-%
-%We combine these tools to develop an integrated data analysis pipeline for ICS assays and use it to analyze two real--world, full size, vaccine trial, ICS data sets from the HVTN, comparing the results against gold standard manual gating.
 

### 2012-02-06 MIMOSA.tex
Added figure for unconstrained model fit to constrained model data

@@
    \includegraphics[width=4in]{Figures/simulations} 
-   \caption{Performance of the Beta-binomial mixture vs Fisher's exact test in simulated data.
-Data were simulated from a model with hyper--parameters estimated from day 28, Gag1 stimulated, CD4+, IL2 expressing T--cells in the HVTN054 data set.
+   \caption{Performance of the constrained Beta-binomial mixture vs Fisher's exact test in simulated data from a constrained model.
+Data were simulated from a constrained model with hyper--parameters estimated from day 28, Gag1 stimulated, CD4+, IL2 expressing T--cells in the HVTN054 data set.
 We simulated 500 observations, with a response rate of 40\%, and increasing numbers of cells, ten times each.
-The performance, measured by the AUC, of the beta--binomial mixture compared to Fisher's exact test is shown in the first panel, as a function of increasing number of events.
+The performance, measured by the AUC, of the constrained beta--binomial mixture compared to Fisher's exact test is shown in the first panel, as a function of increasing number of events.
 The estimated and true prior distributions are shown in the second panel for one simulated data set, with N=150,000 events.
-The ROC curve for Fisher's exact test and the Beta--binomial model for the same simulated data set are shown in the third panel.
+The ROC curve for Fisher's exact test and the constrained Beta--binomial model for the same simulated data set are shown in the third panel.
+The observed vs expected false discovery rate for Fisher's exact test and the constrained Beta--binomial model are shown for the same data set in the fourth panel.}
+   \label{fig:simulations}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=4in]{Figures/simulations_negt} 
+   \caption{Performance of the unconstrained Beta-binomial mixture (unconstrained model) vs Fisher's exact test in simulated data from a constrained model.
+Data were simulated from a constrained model with hyper--parameters estimated from day 28, Gag1 stimulated, CD4+, IL2 expressing T--cells in the HVTN054 data set.
+We simulated 500 observations, with a response rate of 40\%, and increasing numbers of cells, ten times each.
+The performance, measured by the AUC, of the unconstrained beta--binomial mixture compared to Fisher's exact test is shown in the first panel, as a function of increasing number of events.
+The estimated and true prior distributions are shown in the second panel for one simulated data set, with N=150,000 events.
+The ROC curve for Fisher's exact test and the unconstrained Beta--binomial model for the same simulated data set are shown in the third panel.
 The observed vs expected false discovery rate for Fisher's exact test and the Beta--binomial model are shown for the same data set in the fourth panel.}

### 2012-02-09 MIMOSA.tex
revised simulation figures
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
+\subsection{Simulations}
+We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
+Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
+The constrained model was fit to this data and the sensitivity and specificity of the model to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated ROC curve analysis and compared against Fisher's exact test.
+(Figure~\ref{fig:simulations}).
+The nominal vs observed false discovery rate was also examined for the model and for Fisher's exact test, as well as the accuracy of the hyper--parameters estimated by the model.
+In order to assess whether the unconstrained model could be used in place of the constrained model in order to benefit from faster computations using an exact closed form expression for the posterior, an identical analysis was performed after fitting the unconstrained model to data simulated from the constrained model (Figure~\ref{fig:simulations_negt}.
+
 \section*{Results}
 \subsection*{Simulations}
+We ran simulation studies to assess the performance of the constrained mode as described in the Methods.
+We found that the constrained model out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}).
+Furthermore, the model gave reasonable estimates of the true hyper--parameters, such that the estimated and true prior distributions closely resembled each other (Figure~\ref{fig:simulations}).
+
+Since the constrained model relies on Monte--Carlo integration, which can be computationally costly, to estimate the normalizing constant in the likelihood calculation, we examined whether the unconstrained model could be used to accurately fit data generated from the constrained model.
+We found that the unconstrained model performed as well as the constrained model when fitting data generated from the constrained model (Figure~\ref{fig:simulations_negt}).
+
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
 The observed vs expected false discovery rate for Fisher's exact test and the Beta--binomial model are shown for the same data set in the fourth panel.}
-   \label{fig:simulations}
+   \label{fig:simulations_negt}
 \end{figure}
@@
 The volcano plot shows the difference between the proportion of cytokine positive cells in the stimulated and unstimulated samples, for the maximum likelihood estimates of the proportions (triangles) and for the MAP estimates (crosses)}
-   \label{fig:simulations}
+   \label{fig:icsdata}
 \end{figure}

### 2012-02-16 MIMOSA.tex
Update MIMOSA paper with more ICS data
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

--- 2012-02-04 MIMOSA.tex
@@
 \begin{abstract}
-Immunological endpoints in vaccine trials are measured through a variety of different assays that, often, provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of specific genes in single cells from specific cell sub--populations.
-Although these measurements are continuous, they are generally discretized for analysis, with each cell called either positive or negative for a measured variable, depending on some predetermined threshold.
-A motivating example is the intracellular cytokine staining (ICS) flow cytometry assay.
-This assay is used to assess an individual's immune response to a vaccine by measuring the abundance of antigen--specific T--cell subpopulations producing one or more specific cytokines.
-In this way, the assay provides phenotypic and functional information about the immune system.
-ICS assays are used to test multiple antigens and multiple cytokines across hundreds of individuals in a vaccine study.
-The threshold for calling an individual cell as positive or negative for a given cytokine is predetermined by the gating step of the analysis.
-The thresholded data is discretized and treated as count data of the number of cytokine positive and negative cells in each stimulated and control sample.
-The goal of the analysis is to identify the ``vaccine responders''.
-These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline (the unstimulated control).
+Immunological endpoints in vaccine trials are measured through a number of different assays that often provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific cell sub--populations.
+While these measurements are continuous, they are generally discretized during data analysis.
+For example, in flow cytometry, individual cells are often classified as either positive or negative for a marker based on a predetermined threshold.
+One such assay is the intracellular cytokine staining assay that is used to assess an individual's immune response to a vaccine by measuring the abundance antigen--specific T--cell subpopulations producing different cytokines.
+Cells are classified as cytokine positive or negative based on some threshold, and the resulting discretized count data is used to test for an increase in cytokine producing cells between a treatment and control sample.
+ICS assays are generally used to identify ``vaccine responders''.
+These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline.
 The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern.
@@
 \end{abstract}
+
 \section{Introduction}
-Single--cell assays are an important tool in vaccine trials to measure different characteristics of specific immune cell subpopulations.
-These assays typically measure multiple variables simultaneously for each individual cell in a homogeneous mixture such as whole blood.
-These variables can be used to classify individual cells in the mixture into more homogeneous phenotypes and functional categories.
-These, in turn, provide a functional read--out of the immune system.
-A motivating example is the intracellular cytokine staining assay from flow cytometry, which, in the context of vaccine studies, is used to provide a sensitive and quantitative assessment of antigen--specific T--cell response to a vaccine.
-It provides both phenotypic information about cell subpopulations (CD4 and CD8 T--cells) in whole blood, and functional information (cytokine expression) about how those subpopulations express different cytokines in response to an antigen stimulus~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
+These assays typically measure multiple variables simultaneously on individual cells in a homogeneous mixture such as whole blood.
+These variables are used to classify individual cells in the mixture into more homogeneous sub populations based on phenotypic or functional differences.
+Such single--cell assays provide a snapshot of immune function at a given time and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{}.
+
+A motivating example from vaccine trial research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine.
+Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells.
+Not all T--cells can recognize all antigens.
+Those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines that further promote the immune response.
+After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future.
+None the less, these antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells.
+The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that was present in the original vaccine.
+Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected.
+Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
+These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences.
 Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.
 
+The ICS assay in flow cytometry is just one example of the applications of single--cell technologies to immunology.
+Single--cell gene expression technologies such as Fluidigm are enabling researchers to measure the expression of 96 genes in 96 single cells from a homogeneous population sorted by flow cytometry\cite{Jang:2011uj}.
+This technology has been used to identify signatures of immune cell sub--populations that correlate with different types of vaccine~\cite{Flatz:2011jb}.
+
 \section{Materials and Methods}
+
 \subsection{Vaccine Trial ICS Dataset Description}
@@
 Statistical analysis of the original positivity calls is described in the original publication~\cite{Peiperl:2010ej}.
-
-%HVTN080 is a phase 1 clinical trial to evaluate the safety and
-%immunogenicity of PENNVAX\texttrademark-B (gag, pol and env) vaccine, with or without IL-12 DNA plasmid, delivered via electroporation in healthy, HIV-1 uninfected adult participants.
 
@@
 The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions.
-$I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, which is just the CDF of Beta distribution with parameters $\alpha_s,\beta_s$, leaving a 1--dimensional integration for the ratio of normalizing constants.
+We call this the \emph{constrained} model.
+
+The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, is just the CDF of Beta distribution with parameters $\alpha_s,\beta_s$, leaving a 1--dimensional integration for the ratio of normalizing constants.
+
+Without constraints, ($p_s\ne p_u$), the ratio of integrals for the normalizing constant in \eqref{eq:model2post} is dropped. We call this the \emph{unconstrained} model.
 
@@
 We use the Multinomial--Dirichlet family of distributions to model counts of events in two \textit{different} 2x2 contingency tables.
+Here we consider only the \emph{unconstrained} case ($p_s\ne p_u$).
 The observed data can be represented in the following way:
@@
 
+\subsection{Simulations}
+We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
+Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
+The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test.
+(Figure~\ref{fig:simulations}).
+The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions.
+In order to assess whether the unconstrained model could be used to effectively fit data from a constrained model and benefit from faster computations using exact closed form expressions rather than Monte-Carlo integration, we fit the same data simulated from the constrained model using the unconstrained model (Figure~\ref{fig:simulations}.
+
+To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
+The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN054 data set (Figure~\ref{fig:simulations_trunc}).
+
 \section*{Results}
 \subsection*{Simulations}
+We ran simulation studies to assess the performance of the constrained and unconstrained models, as described in the Methods.
+We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}).
+Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).
+
+Since the constrained model relies on Monte--Carlo integration, which can be computationally costly, to estimate the normalizing constant in the likelihood calculation, we examined whether the unconstrained model could be used to accurately fit data generated from the constrained model.
+We found that the unconstrained model performed as well as the constrained model when fitting data generated from the constrained model (Figure~\ref{fig:simulations}).
+
+In order to assess the sensitivity of the model to deviations from model assumptions, we fit the unconstrained model to simulated data where the proportions were drawn from truncated normal distributions (Figure~\ref{fig:simulations_trunc}).
+Even in these circumstances the unconstrained MIMOSA model outperformed Fisher's exact test and in fact performed about as well as the constrained model fit to constrained data.
+
+\subsection*{HVTN054 ICS data}
+
+We tested our method on ICS data from HVTN054, a safety and immunogenicity vaccine trial for a replication defective Adenovirus vector HIV vaccine.
+The data set consisted of 48 individuals who received two doses of vaccine or placebo, and ICS time points for these individuals were available at day 0 and day 28 after vaccination.
+Responders for each cytokine and stimulation combination were classified using Fisher's exact test and the constrained MIMOSA model at the 1\% FDR level, then the response rate within the control and treatment groups were calculated for both methods.
+Figure~\ref{fig:positivityrates} shows the response rates for a subset of the ICS data (Gag2 and Pol2 stimulations, for the IL2 and IFNg cytokines, at days 0 and 28, in both CD4 and CD8 T--cell subpopulations).
+On day 0, we found that the response rates from the MIMOSA model were zero, and equal to response rates computed from Fisher's exact test across all treatment and control groups.
+The same was true for the response rate in the control groups at day 28.
+In the treatment groups, at day 28, the response rates for the MIMOSA model were  generally higher or equal to those for Fisher's exact test, demonstrating the increased sensitivity and specificity of our method.
+
+The MIMOSA framework generates predictions for individual samples.
+To better understand the differences between the predictions from the MIMOSA model and Fisher's exact test, we examined the model fit from Gag2 stimulated CD4+ T--cells at day 28 producing IFNg (Figure~\ref{fig:icsdata}).
+We see that two individuals are predicted as responders by the MIMOSA model which are not detected by Fisher's exact test.
+Both individuals have a larger proportion of IFNg producing antigen specific T--cells in antigen stimulated than unstimulated samples, as measure by the MAP or ML estimates of the proportions (Figure~\ref{fig:icsdata}).
+
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
    \centering
-   \includegraphics[width=4in]{Figures/simulations} 
-   \caption{Performance of the Beta-binomial mixture vs Fisher's exact test in simulated data.
-Data were simulated from a model with hyper--parameters estimated from day 28, Gag1 stimulated, CD4+, IL2 expressing T--cells in the HVTN054 data set.
-We simulated 500 observations, with a response rate of 40\%, and increasing numbers of cells, ten times each.
-The performance, measured by the AUC, of the beta--binomial mixture compared to Fisher's exact test is shown in the first panel, as a function of increasing number of events.
+   \includegraphics[width=4in]{Figures/simulations_all} 
+   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data.
+Data were simulated from a constrained model with hyper--parameters estimated from a real data of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial.
+For total cell counts from 10,000 to 150,000, we simulated ten data sets each of 500 observations with a response rate of 40\%.
+The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta--binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells.
 The estimated and true prior distributions are shown in the second panel for one simulated data set, with N=150,000 events.
-The ROC curve for Fisher's exact test and the Beta--binomial model for the same simulated data set are shown in the third panel.
-The observed vs expected false discovery rate for Fisher's exact test and the Beta--binomial model are shown for the same data set in the fourth panel.}
+The ROC curve for Fisher's exact test and the constrained and unconstrained Beta--binomial model for the same simulation are shown in the third panel.
+The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta--binomial model are shown in the fourth panel.}
    \label{fig:simulations}
@@
 
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
+   \caption{Sensitivity to departures from model assumptions.
+We generated data from variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions.
+The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity).
+The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel.
+The estimated and true prior distributions are shown in the second panel.
+The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
+   \label{fig:simulations_trunc}
+\end{figure}
+
 \subsection*{HVTN054 ICS Data}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=6in]{Figures/PositivityRates} 
+   \caption{Comparison of MIMOSA and Fisher's exact test for calling responders in ICS data.
+Response rates for IL2 and IFNg cytokine positivity in Gag2 and Pol2 stimulated samples at days 0 and 28 in the CD4 and CD8 T--cell subpopulations of the HVTN054 ICS data.
+In the treatment groups, response rates for the MIMOSA model are greater than or equal to those for Fisher's exact test at day 28 but are equal (and zero) at day 0 as expected.
+In the control groups, response rates are equal (and zero) between Fisher's exact test and the MIMOSA model at day 0 and at day 28, as expected.}
+   \label{fig:positivityrates}
+\end{figure}
+
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
 The likelihood surface shows the observed counts from stimulated and unstimulated samples.
-The volcano plot shows the difference between the proportion of cytokine positive cells in the stimulated and unstimulated samples, for the maximum likelihood estimates of the proportions (triangles) and for the MAP estimates (crosses)}
-   \label{fig:simulations}
+The volcano plot shows the difference between the proportion of cytokine positive cells in the stimulated and unstimulated samples, for the maximum likelihood estimates of the proportions (triangles) and for the MAP estimates (crosses).
+The effect of shrinking the MAP estimates towards zero can be seen in the volcano plot.}
+   \label{fig:icsdata}
 \end{figure}

### 2012-03-05 MIMOSA.tex
minor updates
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
-\subsection{Simulations}
+\subsection*{Simulation Studies}
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
@@
 \section*{Results}
-\subsection*{Simulations}
+\subsection*{Simulations Show MIMOSA Outperforms Fisher's Exact Test Event When Model Assumptions are Violated}
 We ran simulation studies to assess the performance of the constrained and unconstrained models, as described in the Methods.
@@
 
-\subsection*{HVTN054 ICS data}
+\subsection*{MIMOSA Outperforms Fisher's Exact Test in Real--World Vaccine Trial Data from HVTN054}
 
@@
    \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data.
-Data were simulated from a constrained model with hyper--parameters estimated from a real data of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial.
+Data were simulated from a constrained model with hyper--parameters estimated from a real data set of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial.
 For total cell counts from 10,000 to 150,000, we simulated ten data sets each of 500 observations with a response rate of 40\%.
 The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta--binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells.
-The estimated and true prior distributions are shown in the second panel for one simulated data set, with N=150,000 events.
+The beta distributions for the estimated and true hyper parameters are shown in the second panel for one simulated data set, with N=150,000 events.
 The ROC curve for Fisher's exact test and the constrained and unconstrained Beta--binomial model for the same simulation are shown in the third panel.
@@
 The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel.
-The estimated and true prior distributions are shown in the second panel.
+The beta distributions for the estimated and true hyper parameters  are shown in the second panel.
 The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
@@
 \end{figure}
-
-\subsection*{HVTN054 ICS Data}
 

### 2012-03-09 MIMOSA.tex
Updated multinomial dirichlet description
Added gradient and hessian for responder and non-responder components.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \begin{align}
-\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u) &= \prod_{i=0}^P\frac{ \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\\
+\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u) &= \prod_{i=0}^P\frac{ \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\label{NRmd}
 \end{align}
@@
 Without loss of generality, if only some $p_j$ are different between stimulated and unstimulated samples, the appropriate components of $\alpha_j$  can be substituted in the calculation of the likelihood eq~\eqref{eq:postmult}.
+
+The marginal likelihood of the non--responding component, given in eq~\eqref{NRmd} can be rewritten as:
+\begin{align}
+\begin{split}
+l(\bar{\alpha}_u|D)=\sum_i \left[ \log \Gamma (\sum_j \alpha_{uj}) - \log \Gamma (\sum_j \alpha_{uj}+n_{uij}+n_{sij})+ \sum_j (\log \Gamma(\alpha_{uj}+n_{uij}+n_{sij}) \right.\\ \left.
+-\log\Gamma(\alpha_{uj}))+ \log\Gamma(N_{si}+1) +\log\Gamma(N_{ui}+1)-\right.\\ \left.\left[\sum_j(\log\Gamma(n_{sij}+1)+\log\Gamma(n_{uij}+1))\right]\right]\label{mdloglike}
+\end{split}
+\end{align}
+
+The gradient is given by:
+\begin{align}
+\frac{\partial l}{ \partial \alpha_{uj}} = \sum_i \Psi(\sum_j \alpha_{uj}) - \Psi (\sum_j \alpha_{uj}+n_{uij}+n_{sij}) + \Psi(\alpha_{uj}+n_{uij}+n_{sij}) - \Psi(\alpha_{uj})\label{mdgrad}
+\end{align}
+
+and the Hessian is:
+\begin{align}
+j&\ne k; &\frac{\partial^2 l}{\partial \alpha_{uj}\alpha_{uk}} & = \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})\\
+j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{iuj})
+\end{align}
+
+The marginal likelihood of the responding component in eq~\ref{eq:postmult} can be rewritten as:
+\begin{align}
+\begin{split}
+\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
+-\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
+-\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
+\end{split}
+\end{align}
+with gradient:
+\begin{align}
+\frac{\partial l}{\partial\alpha_{\phi j}} = \sum_i \Psi(\alpha_{\phi j}+n_{\phi ij})-\Psi(\alpha_{\phi j})+\Psi(\sum_j \alpha_{\phi j}) - \Psi(\sum_j \alpha_{\phi j}+n_{\phi ij}): (\phi \in {s,u})
+\end{align}
+and Hessian:
+\begin{align}
+\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
+\frac{\partial^2l}{\partial \alpha_{\phi j}^2}&=\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
+\end{align}
 

### 2012-03-12 MIMOSA.tex
gradient and hessian for responder and non-responder marginal likelihoods
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 j&\ne k; &\frac{\partial^2 l}{\partial \alpha_{uj}\alpha_{uk}} & = \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})\\
-j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{iuj})
+j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{uj})
 \end{align}
@@
 \begin{align}
-\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
-\frac{\partial^2l}{\partial \alpha_{\phi j}^2}&=\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
+\frac{\partial^2l}{\partial \alpha_{\phi j}^2}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}+n_{\phi ij}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
+\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}&\overset{k\ne j}{=}\sum_i\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
 \end{align}

### 2012-03-16 MIMOSA.tex
Updating write-up of multivariate MIMOSA
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \usepackage{amssymb}
+\usepackage{booktabs}
 
@@
 \author[1]{Greg Finak}
-\author[1]{\ldots Others \ldots}
+\author[2]{SC De Rosa}
 \author[1]{Raphael Gottardo}
@@
 \affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
+\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, Wa}
 \date{\today}
@@
 Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
-Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines \todo[inline]{IS THIS CORRECT, OR IS IT WITHIN INDIVIDUALS ACROSS STIMULATIONS?}.
+Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
 In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
@@
 
+\subsection*{Detecting coexpression via the Multinomial-Dirichlet model}
+We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
+As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model with no difference between stimulations, and one component for the alternative model with a difference between stimulations.
+We fit the model to the IL4 and TNFa cytokines for Env1 stimulated CD4 T--cells from the day 28 time point.
+The classification of responder and non--responder individuals was compared between the multivariate MIMOSA model, univariate MIMOSA models applied to each cytokine combination, and Fisher's exact test applied to each cytokine combination.
+
+The multivariate MIMOSA model detected two additional responders than Fisher's exact test at the 1\% FDR level (Table~\ref{tab:MDFisher, top left.}.
+The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).
+In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4--fold and 2.5--fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples.
+These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7--fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
+
+\begin{table}[ht]
+\begin{center}
+\begin{minipage}{0.9\linewidth}
+\caption{Confusion matrices comparing responders and non--responders identified by the MIMOSA models and Fisher's exact test.
+Top left: Fisher's exact test vs the multivariate MIMOSA model for any combination of cytokines.Top right: Fisher's exact test vs MIMOSA for the univariate combination of IL4 and not TNFa.
+Bottom left: Fisher's exact test vs univariate MIMOSA for TNFa and not IL4.
+Bottom right: Fisher's exact test vs univariate MIMOSA for IL4 and TNFa.
+False discovery rate thresholds were set at 1\% for each cytokine combination (except in the multivariate model, where it was 1\% for each observation).
+}\label{tab:MDFisher}
+\end{minipage}
+\begin{minipage}{0.49\linewidth}
+\center
+\begin{tabular}{rrr}
+  \hline
+  \multicolumn{3}{c}{Any Combination}\\
+  \hline
+&\multicolumn{2}{c}{MIMOSA}\\
+\cmidrule(r){2-3}
+ & NR & R \\ 
+\cmidrule(r){1-1}
+Fisher's Exact Test&&\\
+  \hline
+NR &  32 &   2 \\ 
+  R &   0 &   4 \\ 
+   \hline
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.49\linewidth}
+\center
+\begin{tabular}{rrr}
+\hline
+\multicolumn{3}{c}{IL4 and not TNFa}\\
+  \hline
+&\multicolumn{2}{c}{MIMOSA}\\
+\cmidrule(r){2-3}
+ & NR & R \\ 
+\cmidrule(r){1-1}
+Fisher's Exact Test&&\\
+  \hline
+NR &  38 &   0 \\ 
+  R &   0 &   0 \\ 
+   \hline
+\end{tabular}
+\end{minipage}
+
+\begin{minipage}{0.49\linewidth}
+\center
+\begin{tabular}{rrr}
+  \hline
+  \multicolumn{3}{c}{TNFa and not IL4}\\
+  \hline
+&\multicolumn{2}{c}{MIMOSA}\\
+\cmidrule(r){2-3}
+ & NR & R \\ 
+\cmidrule(r){1-1}
+Fisher's Exact Test&&\\
+  \hline
+NR &  36 &   2 \\ 
+  R &   0 &   0 \\ 
+   \hline
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.49\linewidth}
+\center
+\begin{tabular}{rrr}
+  \hline
+  \multicolumn{3}{c}{IL4 and TNFa}\\
+  \hline
+&\multicolumn{2}{c}{MIMOSA}\\
+\cmidrule(r){2-3}
+ & NR & R \\ 
+\cmidrule(r){1-1}
+Fisher's Exact Test&&\\
+  \hline
+NR &  33 &   1 \\ 
+  R &   0 &   4 \\ 
+   \hline
+\end{tabular}
+\end{minipage}
+\end{center}
+\end{table}
+
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page

### 2012-03-21 MIMOSA.tex
More revisions to the paper text.
Updated figures for multinomial dirichlet.
Rewriting intro.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

--- 2012-03-12 MIMOSA.tex
@@
 
-\title{Mixture Models for Single Cell Assays}
+\title{Combinatorial Mixture Models for Single Cell Assays}
 \author[1]{Greg Finak}
-\author[1]{\ldots Others \ldots}
+\author[2]{SC De Rosa}
 \author[1]{Raphael Gottardo}
@@
 Immunological endpoints in vaccine trials are measured through a number of different assays that often provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific cell sub--populations.
-While these measurements are continuous, they are generally discretized during data analysis.
+While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells.
 For example, in flow cytometry, individual cells are often classified as either positive or negative for a marker based on a predetermined threshold.
 One such assay is the intracellular cytokine staining assay that is used to assess an individual's immune response to a vaccine by measuring the abundance antigen--specific T--cell subpopulations producing different cytokines.
-Cells are classified as cytokine positive or negative based on some threshold, and the resulting discretized count data is used to test for an increase in cytokine producing cells between a treatment and control sample.
+Cells are classified as cytokine positive or negative and the resulting cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells.
 ICS assays are generally used to identify ``vaccine responders''.
@@
 The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern.
-The typical approach to analysis using Fisher's exact test is problematic for two reasons.
+The standard approach to analyze such data, Fisher's exact test, is problematic for two reasons.
 It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test.
 Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.
-In this paper we present a cohort framework based on an empirical-Bayes mixture of Beta--binomial or Dirichlet--Multinomial distributions, for analyzing count data derived from such thresholded single--cell assays.
-In general, any single--cell assay in which continuous data can be thresholded to classify individual cells into either ``positive'' or ``negative'' categories with respect to one or more variables and compared across different conditions, is suitable for analysis using our method and the extensions presented herein.
-Using the ICS assay as a motivating example, our method models cytokine–specific T-cell response across all individuals simultaneously while modelling the stimulated and unstimulated cell counts independently.
+In this paper we present a coherent framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from single--cell assays.
+Using the ICS assay as a motivating example, our method models cytokine–specific T-cell response across all individuals simultaneously, sharing information across observations using empirical Bayes methods.
+The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently.
 Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test.
@@
 These assays typically measure multiple variables simultaneously on individual cells in a homogeneous mixture such as whole blood.
-These variables are used to classify individual cells in the mixture into more homogeneous sub populations based on phenotypic or functional differences.
-Such single--cell assays provide a snapshot of immune function at a given time and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{}.
+These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences.
+Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
@@
 This technology has been used to identify signatures of immune cell sub--populations that correlate with different types of vaccine~\cite{Flatz:2011jb}.
+To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements.
+However, the proportion of single cells expressing individual genes also carries information and should be evaluated.
 
-\section{Materials and Methods}
+Here we present a combinatorial mixture model framework for the analysis of single--cell assay data in which multivariate measurement are made individual cells.
+The model is used to contrast two conditions (i.e.
+treatment and control) and identify observations where a significant differences exists between the treatment and control samples with respect to the proportions of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.
 
-\subsection{Vaccine Trial ICS Dataset Description}
+We develop our framework in the context of ICS data.
+While there is no standard approach to analyzing ICS assay data, Fisher's exact test, as well as more ad--hoc methods, are commonly applied on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa}.
+The drawbacks of Fisher's exact test applied to 2 x 2 contingency table data when the assumptions of the test are not met, have been reviewed in the literature~\cite{Proschan:2009ks}.
+Briefly, for single--cell assay data the assumptions of fixed margins are not met, since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect.
+More importantly, Fisher's exact test does not share information across observations, resulting in low power when counts are small.
+The framework developed in this paper addresses these issues explicitly.
+
+\section*{Materials and Methods}
+
+\subsection*{Vaccine Trial ICS Dataset Description}
 HVTN054 is a phase 1 (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
@@
 
-\subsection{Data Import, Preprocessing, and Gating}
+\subsection*{Data Import, Preprocessing, and Gating}
 The gated ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
@@
 
-\subsection{Statistical Analysis of Responder and Non--responder calls}
+\subsection*{Statistical Analysis of Responder and Non--responder calls}
 Below, we summarize the methods for statistical analysis of responder and non--responder calls in the published trial, as well as the methods compared in this paper.
-\subsubsection{Statistical Analysis in the Published Trial}
+\subsubsection*{Statistical Analysis in the Published Trial}
 The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
@@
 
-\subsection{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
+\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
 Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold.
 Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
-Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines \todo[inline]{IS THIS CORRECT, OR IS IT WITHIN INDIVIDUALS ACROSS STIMULATIONS?}.
+Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
 In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
 
-\subsection{Two Competing Beta--Binomial Models}
+\subsection*{Two Competing Beta--Binomial Models}
 Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach.
@@
 
+\subsection*{Testing Multifunctional Response via the Multinomial-Dirichlet model}
+We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
+As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
+We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
+
+The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations.
+We compared multivariate MIMOSA against Fisher's exact test.
+Fisher's exact test was used to test whether the number of cells expressing a specific combination of cytokines (i.e TNFa+/IL4-, TNFa+/IL4+, and so forth) was significantly higher in the stimulated samples than in the unstimulated samples.
+In this comparison, Fisher's exact test detected only one of the 8 responders detected by the MIMOSA model in the Env3 stimulated, day 28 data, and detected two false positives in the day zero data.
+Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.
+
+%The classification of responder and non--responder individuals was compared between the multivariate MIMOSA model, univariate MIMOSA models applied to each cytokine combination, and Fisher's exact test applied to each cytokine combination.
+
+%The multivariate MIMOSA model detected two additional responders than Fisher's exact test at the 1\% FDR level (Table~\ref{tab:MDFisher}, top left).
+The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).
+In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4--fold and 2.5--fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples.
+These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7--fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
+%
+%\begin{table}[ht]
+%\begin{center}
+%\begin{minipage}{0.9\linewidth}
+%\caption{Confusion matrices comparing responders and non--responders identified by the MIMOSA models and Fisher's exact test.
+Top left: Fisher's exact test vs the multivariate MIMOSA model for any combination of cytokines.Top right: Fisher's exact test vs MIMOSA for the univariate combination of IL4 and not TNFa.
+Bottom left: Fisher's exact test vs univariate MIMOSA for TNFa and not IL4.
+Bottom right: Fisher's exact test vs univariate MIMOSA for IL4 and TNFa.
+False discovery rate thresholds were set at 1\% for each cytokine combination (except in the multivariate model, where it was 1\% for each observation).
+}\label{tab:MDFisher}
+%\end{minipage}
+%\begin{minipage}{0.49\linewidth}
+%\center
+%\begin{tabular}{rrr}
+%  \hline
+%  \multicolumn{3}{c}{Any Combination}\\
+%  \hline
+%&\multicolumn{2}{c}{MIMOSA}\\
+%\cmidrule(r){2-3}
+% & NR & R \\ 
+%\cmidrule(r){1-1}
+%Fisher's Exact Test&&\\
+%  \hline
+%NR &  32 &   2 \\ 
+%  R &   0 &   4 \\ 
+%   \hline
+%\end{tabular}
+%\end{minipage}
+%\begin{minipage}{0.49\linewidth}
+%\center
+%\begin{tabular}{rrr}
+%\hline
+%\multicolumn{3}{c}{IL4 and not TNFa}\\
+%  \hline
+%&\multicolumn{2}{c}{MIMOSA}\\
+%\cmidrule(r){2-3}
+% & NR & R \\ 
+%\cmidrule(r){1-1}
+%Fisher's Exact Test&&\\
+%  \hline
+%NR &  38 &   0 \\ 
+%  R &   0 &   0 \\ 
+%   \hline
+%\end{tabular}
+%\end{minipage}
+%
+%\begin{minipage}{0.49\linewidth}
+%\center
+%\begin{tabular}{rrr}
+%  \hline
+%  \multicolumn{3}{c}{TNFa and not IL4}\\
+%  \hline
+%&\multicolumn{2}{c}{MIMOSA}\\
+%\cmidrule(r){2-3}
+% & NR & R \\ 
+%\cmidrule(r){1-1}
+%Fisher's Exact Test&&\\
+%  \hline
+%NR &  36 &   2 \\ 
+%  R &   0 &   0 \\ 
+%   \hline
+%\end{tabular}
+%\end{minipage}
+%\begin{minipage}{0.49\linewidth}
+%\center
+%\begin{tabular}{rrr}
+%  \hline
+%  \multicolumn{3}{c}{IL4 and TNFa}\\
+%  \hline
+%&\multicolumn{2}{c}{MIMOSA}\\
+%\cmidrule(r){2-3}
+% & NR & R \\ 
+%\cmidrule(r){1-1}
+%Fisher's Exact Test&&\\
+%  \hline
+%NR &  33 &   1 \\ 
+%  R &   0 &   4 \\ 
+%   \hline
+%\end{tabular}
+%\end{minipage}
+%\end{center}
+%\end{table}
+%
+%
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
    \label{fig:icsdata}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=6in]{Figures/Polyfunctionality_ggplot2} 
+   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--2 peptide stimulated CD4+ T--cells.}
+   \label{fig:polyfunctionality}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=6in]{Figures/Polyfunctionality_Env1} 
+   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--1 peptide stimulated CD4+ T--cells.}
+   \label{fig:polyfunctionalityenv1}
 \end{figure}

### 2012-03-30 MIMOSA.tex
Begin restructuring
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \usepackage{booktabs}
+\usepackage{url}
 
@@
 
-Here we present a combinatorial mixture model framework for the analysis of single--cell assay data in which multivariate measurement are made individual cells.
-The model is used to contrast two conditions (i.e.
-treatment and control) and identify observations where a significant differences exists between the treatment and control samples with respect to the proportions of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.
-
 We develop our framework in the context of ICS data.
 While there is no standard approach to analyzing ICS assay data, Fisher's exact test, as well as more ad--hoc methods, are commonly applied on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa}.
-The drawbacks of Fisher's exact test applied to 2 x 2 contingency table data when the assumptions of the test are not met, have been reviewed in the literature~\cite{Proschan:2009ks}.
-Briefly, for single--cell assay data the assumptions of fixed margins are not met, since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect.
+Problems with Fisher's exact when the test assumptions are not met have been reviewed in the literature~\cite{Proschan:2009ks}.
+In the context of single--cell assay data the assumption of fixed margins is not satisfied since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect.
 More importantly, Fisher's exact test does not share information across observations, resulting in low power when counts are small.
+
 The framework developed in this paper addresses these issues explicitly.
+We present a combinatorial mixture model framework for the analysis of single--cell assay data in which multivariate measurement are made individual cells.
+The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.
 
@@
 
+\subsubsection*{Model simplifications for K>2 cytokines}
+In the case of multiple cytokines, there is a combinatorial explosion in the number of parameters of the resulting model.
+We propose several alternative solutions to reducing the number of parameters in general, when the dimensionality of the Multinomial Dirichlet increases.
+If we model $N$ cytokines simultaneously, this results in $2^N$ possible combinations.
+We let $K$ be the number of possible cytokine combinations being modelled, with hyperparameter vector $\bar{\alpha}_0=\{\alpha_{00},\ldots,\alpha_{0K-1}\}$, and $\bar{\alpha}_s=\{\alpha_{s0},\ldots,\alpha_{sK-1}\}$ for the unstimulated and stimulated distributions, respectively.
+We formulate several strategies for reducing the number of parameters required to fit MIMOSA to high dimensional data by sharing hyperparameters across some of the $K$ combinations.
+
+\subsubsection*{Fully shared model hyperparameters}
+In the most simplistic case, we set up a two mixture component model (responders and non--responders), with a goal of identifying observations with no response (equal to the null), or some response in any of the possible combinations of measured cytokines.
+For ICS data, the first component of the hyperparameter vector represents cells not expressing any cytokine combination, and is usually much larger than the remaining components.
+As a first approximation, we can set the $K-1$ remaining parameters to be all equal $\{\alpha_{01}=\{\alpha_{01},\ldots,\alpha_{0K-1}\}$ and thus all variation across cells expressing different cytokine combinations in the unstimulated samples are captured by a single parameter.
+The same simplification is made for modeling the distribution of stimulated samples.
+As stated above, such a model can only be fit to two mixture components (the null component and the responder component), but requires estimation of only four parameters.
+In practice, such a model would likely have a vary high false positive and false negative rate.
+
+\subsubsection*{Second order hyperparameters as functions of first--order hyperparameters}
+Moving towards a more realistic case, we can define first--order hyperparameters as those elements of the $\alpha$ vectors corresponding to cells expressing individual cytokines (i.e.
+cytokine A only, or cytokine B only).
+Second--order and higher parameters are defined as cells expressing pairs, or triplets of cytokines, and so forth  (i.e.
+A and B, B and C, C and D).
+Higher order parameters can be expressed as functions of the first--order parameters (i.e.
+cells expressing A and B can be parameterized as $\alpha_{0ab} = \frac{\alpha_{0a}+\alpha_{0b}}{2}$ or $\alpha_{0ab}=\alpha_{0a}\alpha_{0b}$).
+This parameterization can be extended to models with more than two mixture components, allowing us to identify different types of responders.
+For example there may be observations where there are differences in the proportion of cells expressing combination A and B, but also observations where there are differences in the proportion of cells expressing A only, or B only.
+Distinguishing between these possibilities can increase the power of the model.
+
 \subsection*{Simulation Studies}
@@
 \section*{Results}
-\subsection*{Simulations Show MIMOSA Outperforms Fisher's Exact Test Event When Model Assumptions are Violated}
+\subsection*{Simulations Show MIMOSA Outperforms Fisher's Exact Test and is Robust to Departures from Model Assumptions}
 We ran simulation studies to assess the performance of the constrained and unconstrained models, as described in the Methods.
@@
 \section*{Discussion}
+The variety of single--cell assays being adopted by the immunology community is increasing.
+Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data.
+Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.
+Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}.
+Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
+
+The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model cell counts in non--responding and responding individuals.
+Information is shared across non--responders and across responders through an empirical--Bayes prior, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}).
+The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample, our simulations show that we can relax this assumption to the unconstrained case where $p_s\ne p_u$ without compromising the observed false positive or false negative rates (Figures~\ref{fig:simulations} and \ref{fig:simulations_trunc}), while saving on computation time.
+
+The analysis of real--world ICS data from a vaccine trial demonstrated that the MIMOSA model outperformed Fisher's exact test in detecting individuals who were responders and non--responders to vaccine across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:positivityrates}).
+Importantly, the MIMOSA model was fit within each antigen stimulation $\times$ cytokine $\times$ time point combination but the model was blinded to placebo and vaccine treatment.
+Despite this, MIMOSA demonstrated a lower false positive rate at day zero (increased specificity), where no response is expected, than Fisher's exact test (Figure~\ref{fig:positivityrates} Gag2, IL2, CD4+ T--cells, day 0, treatment group T1).
+Similarly, none of the placebo treated samples were identified as responders by the MIMOSA model (Figure~\ref{fig:positivityrates} treatment group C).
+MIMOSA also exhibited an increased sensitivity in real--world data compared to Fisher's exact test, as demonstrated by the higher response rates for 7 of the 8 stimulation/cytokine/cell type combinations at day 28 to which the model was fit, and where a response is expected to be observed.
+Although it is not possible to validate the responders identified at day 28 for all cases, the conclusion of increased sensitivity is consistent with both the performance of the model in simulations, and with the increased specificity observed on day 0 and in the placebo treated samples.
+
+We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
+For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test.
+These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions.
+This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model.
+This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.
+
+Polyfunctionality (identifying cells coexpressing multiple proteins, cytokines, or genes, is important in single cells assays since it allows the identification of functionally distinct subsets of cells in heterogeneous cell populations~\cite{Milush:2009bz}.
+In vaccine trials, these polyfunctional cell subsets may be correlated with vaccine response or efficacy.
+Existing methods for identifying polyfunctional cytokine profiles from ICS data either test individual combinations of cytokines separately, perform global tests at the expense of decreased power when only specific differences are of interest, sometimes  empirical in nature, and generally don't share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
+As others have pointed out, in order to have the most power to detect a true difference, the statistical test selected to detect differences in various combinations of cytokines should take into consideration the cytokine combinations that are of interest~\cite{Nason:2006dx}.
+A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different.
+This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.
+
+We show a proof of principle example with two components to identify responder and non--responder observations for the four--dimensional (two cytokine) case.
+Sharing of parameters across nested models in the two--cytokine case allows us to reduce the number of parameters.
+The model is easily extended.
+First, to multiple components to identify not just responders and non--responders, but responders with respect to each combination of cytokines.
+Second, extensions to multiple cytokines are straightforward, only requiring increasing the dimensionality of the Multinomial--Dirichlet distribution.
+With the increase in dimensionality there is a combinatorial explosion in the number of combinations of cytokines, and the resulting number of parameters in the model.
+We present several possible approaches to reducing the number of parameters in the model by sharing the hyperparamters across different combinations of cytokines.
+A number of these strategies can be generalized to models with multiple components that allow us to identify observations with differences in specific cytokine combinations between treatment and control samples.
 
 \section*{Conclusions}
-
+We have developed a combinatorial mixture model framework for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays.
+The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
 %\renewcommand{\thesection}{S.\arabic{section}}

### 2012-04-04 MIMOSA.tex
Restructuring..
ongoing.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \maketitle
+
 \begin{abstract}
-Immunological endpoints in vaccine trials are measured through a number of different assays that often provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific cell sub--populations.
-While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells.
-For example, in flow cytometry, individual cells are often classified as either positive or negative for a marker based on a predetermined threshold.
-One such assay is the intracellular cytokine staining assay that is used to assess an individual's immune response to a vaccine by measuring the abundance antigen--specific T--cell subpopulations producing different cytokines.
-Cells are classified as cytokine positive or negative and the resulting cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells.
-ICS assays are generally used to identify ``vaccine responders''.
+Immunological endpoints in vaccine trials are measured through a variety of assays that  provide single--cell measurements of multiple genes and proteins in specific immunological cell populations.
+%intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific immunological cell populations.
+Using single--cell data, we consider the problem of identifying subjects where these cell populations exhibit differential responses under different experimental conditions.  
+%While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells. 
+For example, in the intracellular cytokine staining assay from flow cytometry, individual cells are classified as either positive or negative for a marker based on a predetermined threshold.
+%One such assay is the intracellular cytokine staining assay that
+The assay is used to assess an individual's immune response to a vaccine by measuring the number of antigen--specific cells  producing different cytokines in different T--cell subpopulations in response to different antigen stimulations.
+Individuals whose T--cells exhibit increased production of a cytokine in response to stimulation are termed ``positive'' for that cytokine, and multiple such ``positivity calls'' are used to identify vaccine responders.
+%Cells are classified as cytokine positive or negative and the resulting The cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells.
+The ICS assay is generally used to identify ``vaccine responders''.
 These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline.
@@
 Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.
-In this paper we present a coherent framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from single--cell assays.
-Using the ICS assay as a motivating example, our method models cytokine–specific T-cell response across all individuals simultaneously, sharing information across observations using empirical Bayes methods.
+Here we present a framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from such single--cell assays.
+Our method models cellular responses in a marker--specific manner, treating the responding and non--responding observations as separate components in the model.
+Cell counts from the different experimental conditions are modelled independently, while sharing information across responding and non--responding observations through empirical Bayes priors in order to increase the sensitivity and specificity of positivity calls.
+We compare our method against Fisher's exact test and show how it can be extended to model multivariate (multiple markers) cellular responses.
+In simulations and in HIV vaccine trial data we find that our method has higher sensitivity and specificity than Fisher's exact test for positivity calls.
+% simultaneously, sharing information across observations using empirical Bayes methods.
 The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently.
@@
 Single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
-These assays typically measure multiple variables simultaneously on individual cells in a homogeneous mixture such as whole blood.
+These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood.
 These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences.
@@
 
-A motivating example from vaccine trial research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine.
+A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine.
 Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells.
-Not all T--cells can recognize all antigens.
-Those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines that further promote the immune response.
-After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future.
-None the less, these antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells.
+While not all T--cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
+After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future\cite{McKinstry:2010ei}.
+These antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells.
 The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that was present in the original vaccine.
@@
 These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences.
-Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.
+Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw, Bendall:2011wf}.
 
@@
 
-We develop our framework in the context of ICS data.
-While there is no standard approach to analyzing ICS assay data, Fisher's exact test, as well as more ad--hoc methods, are commonly applied on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa}.
-Problems with Fisher's exact when the test assumptions are not met have been reviewed in the literature~\cite{Proschan:2009ks}.
+Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}.
+These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
+
+%on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data Problems with Fisher's exact when the test assumptions are not met have been reviewed in the literature~\cite{Proschan:2009ks}.
 In the context of single--cell assay data the assumption of fixed margins is not satisfied since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect.
@@
 The framework developed in this paper addresses these issues explicitly.
-We present a combinatorial mixture model framework for the analysis of single--cell assay data in which multivariate measurement are made individual cells.
+We present a combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells.
 The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.
+Importantly, our approach shares information across observations by means of prior distributions on the responding, stimulated samples and the non--responding samples thereby increasing sensitivity and specificity to detect responding cell populations.
+The cell counts from the stimulated and unstimulated experiments are modelled independently, and different combinations of markers are represented as different mixture components in the framework.
+This approach allows us to omit certain combinations of markers that are not observable (i.e., two cytokines may never be co--expressed), or to explicitly represent combinations that are of interest (i.e., we can explicitly represent and test for a difference in a pair of cytokines in an experiment measuring many cytokines).
+This is a flexible approach that avoids the some of the drawbacks of global tests when only a few of the many measured markers show differences~\cite{Nason:2006dx}.
 
-\section*{Materials and Methods}
+\section{Data structure and notation} 
+We analyzed two data sets from two different immunological single--cell assays typically used in vaccine trials, one flow cytometry data set, and one single--cell gene expression data set.
 
-\subsection*{Vaccine Trial ICS Dataset Description}
-HVTN054 is a phase 1 (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
-The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.
-Vaccine was given at two increasing doses, as well as a placebo.
-T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}.
-The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}.
-The sample size consisted of 20 vaccine and four placebo recipients.
-Statistical analysis of the original positivity calls is described in the original publication~\cite{Peiperl:2010ej}.
+\textit{Flow cytometry}: This ICS data is from a phase--I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post--vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
+The statistical analysis of responder and non--responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
+The goal of this data set was to assess response rates of CD4 and CD8 T--cell populations to different antigens.
 
-\subsection*{Data Import, Preprocessing, and Gating}
-The gated ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
-Data were preprocessed using the flowJo--defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T--cells for each sample and stimulation~\cite{Hahne:2009vv}.
+\textit{Fluidigm single--cell gene expression}: This is a single--cell gene expression data set of flow--sorted CD8 T--cells from sixteen individuals.
+The T--cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single--cell level and compared to paired, unstimulated controls.
 
-\subsection*{Statistical Analysis of Responder and Non--responder calls}
-Below, we summarize the methods for statistical analysis of responder and non--responder calls in the published trial, as well as the methods compared in this paper.
-\subsubsection*{Statistical Analysis in the Published Trial}
-The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
-In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine--positive T-cells in the antigen--stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine--positive T-cells for the negative control (unstimulated) sample from the same individual.
-In the original trial, significance was measured via one--sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual.
-A discrete Bonferroni adjustment for multiple comparisons was applied, and stimulations with an adjusted p--value below $\alpha = 0.00001$ were called positive.
+In the remainder of the paper, we use the following notation to describe our model.
+We observe cell counts from $P$ subjects in two conditions (stimulated and unstimulated).
+Each cell can either be positive or negative for a marker.
+We let $i$ index the subjects and $k$ index the cytokines.
+The subscripts $s$,$u$ refer to the cell counts from stimulated and unstimulated conditions, respectively.
+The total number of cells is denoted by $N$, and given in our model, while $n$, the number of positive cells is part of the observe data.
+Thus for a given cytokine, $N_{si}$ denotes the total number of cells collected for subject $i$ in the stimulated condition and $n_{si}$ denotes how many of those cells were cytokine positive.
+Similarly, $N_{ui}$, $n_{ui}$ represent the total, and cytokine positive cell counts in the unstimulated condition for the same subject.
+For each cytokine and each stimulation, given the total cell counts, the observed data is a matrix of size $2 \times P$, where the $i$'th row is a subject, $\mathbf{y}_i = <n_{si},n_{ui}>$.
 
-\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
-Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold.
-Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
-Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
-In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
+\section{Differential expression with one marker}
+The data for a single marker can be represented in contingency table form:
 
-\subsection*{Two Competing Beta--Binomial Models}
-Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach.
-We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples).
-For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively.
-Note that usually, $N_u \ne N_s$.
-The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants.
-For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table:
-% latex table generated in R 2.14.0 by xtable 1.5-6 package
-% Wed Oct 12 09:41:50 2011
 \begin{table}[ht]
@@
 \end{table}
+We consider a marker to be differentially expressed between conditions if the number of cells in the stimulated samples is significantly different from the number of cells in the control sample.
+The model is described below.
 
-The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as:
+\subsection{Beta--Binomial Model for One Marker}
+The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as  binomial random variables.
 \begin{align}
-  \text{if }&p_{si}=p_{ui}\equiv p_{0i};& n_{si} \sim \mathrm{Bin}(N_{si},p_u);\text{ }& n_{ui} \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null}\\
- \text{if }&p_{si}>p_{ui};& n_{si} \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& n_{ui} \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:alternate}
+(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non--responder}\\
+(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
  \end{align}
-Where $p_{si}$ and $p_{ui}$ are the unobserved proportions.
-Equation~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control.
-Equation~\eqref{eq:alternate} represents the  alternate hypothesis where the cytokine response is stronger in the stimulation than in the control.
-We place a common Beta prior on the $p_{si}$ and $p_{ui}$ across individuals, as shown:
+
+Where $p_{si}$, $p_{ui}$ are the unobserved proportions.
+Model~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control, and thus no difference between the proportion of cytokine positive cells in stimulated and control samples.
+Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in cytokine response between the stimulation and the control, and thus a difference in the proportion of cells between stimulated and control samples.
+In the case of no response, the counts for the stimulated and unstimulated samples are independent draws from the same  binomial distribution.
+
+\subsection{Prior}
+Our model shares information across responders and non--responders, respectively. This is done by placing Beta priors on the $p_{si}$, $p_{ui}$, across subjects:
 
  \begin{align}
- p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s)\label{eq:stimprior}\\
- p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:nullprior}
+ p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s);&
+ p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior}
  \end{align}
-If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models.
+
+Additional sharing of information is done through the $\alpha_u, \beta_u$ hyper--parameters, which are common for the null and alternative models. 
+%If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models.
 Given this formulation, the posterior probability of the data given that it is generated by model \eqref{eq:null}, is:
+\subsection{Posetrior}
+\subsubsection{Null Model}
+Given the above formulation, the posterior for the null model is Beta--binomial, with marginal likelihood (integrating out $p_u$):
+
  \begin{align}
-  	\mathrm{Pr}(y_i|\alpha_u,\beta_u)
-	&=\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}\\
-	\intertext{with marginal log--likelihood:}
+  	\mathrm{L}(\alpha_u,\beta_u|\mathbf{y})
+	&=\prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
+%	\intertext{Integrating out the proportions gives the marginal log--likelihood.}
+%	\begin{split}
+%	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N_{si}}{n_{si}}}+\log{\binom{N_{ui}}{n_{ui}}}+\right.\\
+%	\left.\log{\left(\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLL}
+%	\end{split}
+ \end{align} 
+ \subsubsection{Alternative Model - Unconstrained}
+ If the data is generated by model~\eqref{eq:alternate} then we have  $p_s\ne p_u$, and the marginal likelihood of the alternative is:
+\begin{align}
 	\begin{split}
-	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N_{si}}{n_{si}}}+\log{\binom{N_{ui}}{n_{ui}}}+\right.\\
-	\left.\log{\left(\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLL}
+\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) =\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\label{model2:unconstrained}
 \end{split}
 \end{align}
-Thus, in the case of no response to stimulation, the counts for the stimulated and unstimulated samples are modelled as draws from the same "unstimulated" Beta--binomial distribution.
-Note that the unobserved parameters, $p_{si}, p_{ui}$ have been integrated out to give the marginal log--likelihood.
 
-If the data is generated by model~\eqref{eq:alternate}, the posterior probability of the data is given by:
+\subsubsection{Alternative Model - Constrained}
+For ICS data, however, we are generally interested in the one--sided case where $p_s>p_u$, which is consistent with a positive response to stimulation (i.e., more cells are producing the marker after stimulation).
+In this constrained case, integrating out $p_s$, $p_u$ leads to an additional factor, $C$, in the marginal likelihood, which is just the ratio of normalizing constants for the prior and the likelihood with constraints:
 \begin{align}
 \begin{split}
-\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) =\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
-	 \frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
+	 C=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
 \label{eq:model2post}
 \end{split}
-	\intertext{with marginal log--likelihood:}
-\begin{split}
-\mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N_{ui}}{n_{ui}}+ \log\binom{N_{si}}{n_{si}}+ \log\left(\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\right)+\right.
+%	\intertext{with marginal log--likelihood:}
+%\begin{split}
+%\mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N_{ui}}{n_{ui}}+ \log\binom{N_{si}}{n_{si}}+ \log\left(\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\right)+\right.
 \\ \left.
@@
 \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)} \left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}\right] \right\}\label{eq:model2MLL}
-\end{split}
+%\end{split}
 \end{align}
 
-The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions.
+%The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions.
 We call this the \emph{constrained} model.
 
-The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, is just the CDF of Beta distribution with parameters $\alpha_s,\beta_s$, leaving a 1--dimensional integration for the ratio of normalizing constants.
+The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional numerical integration for calculating the ratio of normalizing constants.
 
-Without constraints, ($p_s\ne p_u$), the ratio of integrals for the normalizing constant in \eqref{eq:model2post} is dropped. We call this the \emph{unconstrained} model.
+%Without constraints, ($p_s\ne p_u$), the ratio of integrals for the normalizing constant in \eqref{eq:model2post} is dropped. We call this the \emph{unconstrained} model.
 
-\subsection{The Mixture of Beta--Binomials}
-Although we have specified the two models for the data, we do not know which observation was generated by which model.
-Clearly, not all individuals are expected to exhibit an immune response to a stimulation.
-Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}.
-We capture this uncertainty with a mixture framework of the two competing beta--binomial models.
-The likelihood for the mixture is given by:
+\subsection{The Mixture Model Framework}
+
+The two models (null and alternative), are represented as two components in our mixture model framework.
+The observed data is thus a mixture of observations drawn from one or the other of these components (i.e., each subject is either a responder, or a non--responder).
+Setting  $\theta_1=\{\alpha_u,\beta_u\},\theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, the likelihood for the mixture model is:
+
+%\begin{align}
+%f(y_i|\alpha_{u},\beta{u},\alpha_{s},\beta_{s}) = (1-\pi)f_1(y_i|\alpha_u,\beta_u) + \pi f_2(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)
+%\end{align}
+
  \begin{align}
@@
 %L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi_1 Pr(y_i|\alpha_u,\beta_u) +\pi_2 Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) \right] ,\\ \sum\limits_{k=1}\limits^2\pi_k=1\\
-L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi_1 f_1(y_i|\theta_1) +\pi_2 f_2(y_i|\theta_2) \right] ,\\ \sum\limits_{k=1}\limits^2\pi_k=1
+L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi f_1(y_i|\alpha_u,\beta_u) +(1-\pi)f_2(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) \right] 
 \end{split}
 \end{align}
-Where $\theta_1=\{\alpha_u,\beta_u\}, \theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, $\pi_1$ is the fraction of observations exhibiting no response to stimulation, $\pi_2$ the fraction of observations exhibiting a response to stimulation, and $f_1 = Pr(y_i|\alpha_u,\beta_u), f_2 = Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)$ from \eqref{eq:model1post} and \eqref{eq:model2post}, above.
+%,\\ \sum\limits_{k=1}\limits^2\pi_k=1
+%Where $\theta_1=\{\alpha_u,\beta_u\}, \theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, $\pi_1$ is the fraction of observations exhibiting no response to stimulation, $\pi_2$ the fraction of observations exhibiting a response to stimulation, and $f_1 = Pr(y_i|\alpha_u,\beta_u), f_2 = Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)$ from \eqref{eq:model1post} and \eqref{eq:model2post}, above.
+The terms $f_1$,$f_2$ are the marginal likelihoods of the null and alternative models, respectively, $\pi$ is the proportion of responders.
 
-The unobserved component memberships are treated as missing data and modelled as random variables  $\mathbf{z}_i = \left\{z_{i1},(1-z_{i1})\right\}$
+The component membership of each subject is unobserved and  treated as missing data. They are modelled as random variables  $\mathbf{z}_i = \left\{z_{i1},(1-z_{i1})\right\}$, where:
 
@@
 Each $\mathbf{z}_{i}$ follows an independent multinomial distribution with one trial and  parameters $\boldsymbol{\pi}=\left\{\pi_1,1-\pi_1\right\}$.
-Given the $z_i$'s, the complete data log--likelihood is:
+They are interpreted as the posterior probabilities of response.
+Given the $z_i$'s the complete data likelihood is written as:
 
@@
 \begin{split}
-\mathcal{L}_c(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y},\mathbf{z})=\sum\limits_{i=1}\limits^P\sum\limits_{k=1}\limits^2{z_{ik}\left[ \log\pi_k+\log f_k(y_i|\theta_k)\right]}\label{eq:CDLL}
+\mathbf{L_c}(\theta_k,\pi_k|\mathbf{y},\mathbf{z}) = \prod\limits_{i=1}\limits^P\prod\limits_{k=1}\limits^2\left[\pi_k f_k(y_i|\theta_k)\right]^{z_{ik}}\label{eq:cll}
+%\mathcal{L}_c(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi|\mathbf{y},\mathbf{z})=\sum\limits_{i=1}\limits^P\sum\limits_{k=1}\limits^2{z_{ik}\left[ \log\pi_k+\log f_k(y_i|\theta_k)\right]}\label{eq:CDLL}
 \end{split}
@@
 
-In this form, we use the expectation--maximization (EM) algorithm~\cite{Dempster:1977ul} to fit the model.
+\subsection{Parameter estimation}
+In the from of~\eqref{eq:cll}, we use the expectation--maximization (EM) algorithm~\cite{Dempster:1977ul} to fit the model.
 
-\subsubsection*{E--step}
+\subsubsection{E--step}
 Given the model parameters $\boldsymbol{\Psi}=\left\{\alpha_u,\beta_u,\alpha_s,\beta_s,\pi_k\right\}$, and the data $\mathbf{y}$, we estimate the unobserved component memberships, $\mathbf{Z_i}$  by computing the conditional expectation of the $\mathbf{Z_i}$'s,  $\mathbb{E}\boldsymbol{_\Psi}(\mathbf{Z_i}|\mathbf{y}_i)$:
@@
 
-\subsubsection*{M--step}
-Finally, given the $\tilde{z}_{ik}$, we update the estimates of the model parameters to maximize the conditional expectation of the complete--data log--likelihood. The mixing proportions are given by:
+\subsubsection{M--step}
+Given the $\tilde{z}_{ik}$, we update the estimates of the model parameters to maximize the conditional expectation of the complete--data log--likelihood. The mixing proportions are given by:
 \begin{align}
@@
 
-There is no closed form for the model hyper--parameters, $\alpha_u,\beta_u,\alpha_s,\beta_s$, and they are estimated via numerical optimization using R's \textit{optim} function.
-For this purpose they are re--parameterized as $\mu_u=\frac{\alpha_u}{\alpha_u+\beta_u}$ and $S=\alpha_u+\beta_u$ (likewise for the $\alpha_s,\beta_s$), corresponding to the mean and sample size of the prior distributions.
+There is no closed form for the model hyper--parameter estimates, $\alpha_u,\beta_u,\alpha_s,\beta_s$.
+They are estimated numerically using R's \textit{optim} function.
+To simplify the optimization, they are re--parameterized as $\mu_u=\frac{\alpha_u}{\alpha_u+\beta_u}$ and $S=\alpha_u+\beta_u$ (likewise for the $\alpha_s,\beta_s$), corresponding to the mean and sample size of the prior distributions.
 
-\subsubsection*{Initialization}
-We initialize the $z_{ik}$'s using Fisher's exact test to assign each observation to either the $p_{si}=p_{ui}$ or $p_{si} > p_{ui}$ components.
+\subsubsection{Initialization}
+We initialize the $z_{ik}$'s using Fisher's exact test to assign each observation to either the null or alternative model components.
 We then use the $\hat{z}_i$'s to initialize the hyper--parameters to their method--of--moments estimates:
@@
 Where $\hat{\mu}$ and $\hat{\sigma}^2$ are the sample mean and sample variance estimates, given the $z_{ik}$'s.
+
+\section{Results}
+\subsection{ICS}
+\subsection{Fluidigm}
+\section{Differential expression across combinatorial marker combinations}
+\subsection{Model}
+\subsection{Prior}
+\subsection{Parameter estimation}
+\subsection{ICS example with two markers}
+\section{Discussion}
+\section{Appendix}
+
+%\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
+%Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold.
+Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
+Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
+In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
+%
+%Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach.
+We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples).
+For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively.
+Note that usually, $N_u \ne N_s$.
+The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants.
+For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table:
+% latex table generated in R 2.14.0 by xtable 1.5-6 package
+% Wed Oct 12 09:41:50 2011
+
+%\subsection{The Mixture of Beta--Binomials}
+%Although we have specified the two models for the data, we do not know which observation was generated by which model.
+Clearly, not all individuals are expected to exhibit an immune response to a stimulation.
+Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}.
+We capture this uncertainty with a mixture framework of the two competing beta--binomial models.
+The likelihood for the mixture is given by:
 
@@
 The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
-%\renewcommand{\thesection}{S.\arabic{section}}
-%\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
+\renewcommand{\thesection}{S.\arabic{section}}
+\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
+\setcounter{section}{0}
+\setcounter{subsection}{0}
 %\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}
-%\section{Supplementary Information}
+\section*{Supplementary Information}
+\section{Vaccine Trial ICS Dataset Description}
+\label{supp:data}
+HVTN054 is a phase 1 (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
+The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.
+Vaccine was given at two increasing doses, as well as a placebo.
+T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}.
+The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}.
+The sample size consisted of 20 vaccine and four placebo recipients.
+The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}.
+
+\section{Data Import, Preprocessing, and Gating}
+The ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
+Data were preprocessed using the flowJo--defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T--cells for each sample and stimulation~\cite{Hahne:2009vv}.
+
+\section{Statistical Analysis in the Published Trial}
+\label{supp:statpublished}
+The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
+In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine--positive T-cells in the antigen--stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine--positive T-cells for the negative control (unstimulated) sample from the same individual.
+In the original trial, significance was measured via one--sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual.
+In order to control the false positive rate for responder calls, made based on positivity for multiple cytokines, a discrete Bonferroni adjustment for multiple comparisons (across cytokines) was applied, and stimulations with an adjusted p--value below $\alpha = 0.00001$ were called positive~\cite{Horton:2007tsa}.
+
 %\subsection{Derivation of the Beta--Binomial Model for $p_s=p_u$ and $p_s>p_u$}

### 2012-04-04 MIMOSA.tex
restructuring
continued

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \usepackage{booktabs}
+\usepackage{multirow}
 \usepackage{url}
@@
 \section{Results}
+The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation.
+The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells.
+
 \subsection{ICS}
+\subsubsection*{MIMOSA Outperforms Fisher's Exact Test in Real--World Vaccine Trial Data from HVTN054}
+
+We tested our method on ICS data from HVTN054, a safety and immunogenicity vaccine trial for a replication defective Adenovirus vector HIV vaccine.
+The data set consisted of 48 individuals who received two doses of vaccine or placebo, and ICS time points for these individuals were available at day 0 and day 28 after vaccination.
+Responders for each cytokine and stimulation combination were classified using Fisher's exact test and the constrained MIMOSA model at the 1\% FDR level, then the response rates for \textit{any cytokine} being positive within the control and treatment groups were calculated for both methods.
+Table~\ref{tab:positivityrates} tabulates the response rates for a subset of the ICS data (Gag1, Gag2, Env1, Env2, and Pol2 stimulations, for any cytokine positive, at days 0 and 28, in both CD4 and CD8 T--cell subpopulations).
+
+% latex table generated in R 2.16.0 by xtable 1.7-0 package
+% Wed Apr  4 09:53:11 2012
+\begin{table}[ht]
+\begin{center}
+\caption{Response rates for Pol2, Gag1, Gag2, Env1, and Env2 stimulated CD4+ and CD8+ T--cells on Day 0 and Day 28.
+C, T1, T2 are the Control, and increasing dose treatment groups, respectively.
+No response is expected on Day 0 in any group, while a response is expected on Day 28 in the treatment groups.}\label{tab:positivityrates}
+\begin{tabular}{rrrrrrrrr}
+  \hline
+ &&\multicolumn{3}{c}{Day 0}&\multicolumn{3}{c}{Day 28}\\
+  \cmidrule{3-8}
+&Model & C  & T1 & T2  &   C  & T1  & T2  \\ 
+  \cmidrule{2-8}
+\multirow{2}{*}{CD8+ Pol2}&Beta--binomial & 0.00 & \textbf{0.00} & 0.00 &   0.00 & \textbf{0.89} & 0.57 \\ 
+& Fisher & 0.00 & 0.05 & 0.00 &   0.00 & 0.83 & 0.57 \\ 
+   \hline
+\multirow{2}{*}{CD4+ Pol2}&Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &   0.00 & \textbf{0.89} & \textbf{0.71} \\ 
+&Fisher & 0.00 & 0.05 & 0.00  & 0.00 & 0.72 & 0.50 \\ 
+ \hline
+\multirow{2}{*}{CD8+ Gag2}& Beta-binomial & 0.00 & 0.00 & 0.00  & 0.00 & 0.39 & \textbf{0.21} \\ 
+& Fisher & 0.00 & 0.00 & 0.00  & 0.00 & 0.39 & 0.14 \\ 
+   \hline
+\multirow{2}{*}{CD4+ Gag2}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &  0.00 & \textbf{0.44} & \textbf{0.50} \\ 
+ & Fisher & 0.00 & 0.05 & 0.00 &  0.00 & 0.33 & 0.43 \\ 
+   \hline
+\multirow{2}{*}{ CD8+ Env1}& Beta-binomial  & 0.00 & 0.00 & 0.00 & 0.29 & \textbf{0.71} & 0.86 \\ 
+   & Fisher & 0.00 & 0.00 & 0.00 &  \textbf{0.00} & 0.65 & 0.86 \\ 
+   \hline
+\multirow{2}{*}{  CD4+ Env1}& Beta-binomial   & 0.00 & 0.00 & 0.00 &  0.14 & \textbf{1.00} & \textbf{0.86} \\ 
+ & Fisher & 0.00 & 0.00 & 0.00 &  0.14 & 0.94 & 0.79 \\ 
+   \hline
+\multirow{2}{*}{CD8+ Env2}  & Beta-binomial & 0.00 & 0.05 & 0.08  & 0.29 & 0.83 & \textbf{0.57} \\ 
+  & Fisher & 0.00 & \textbf{0.00} & 0.08 & \textbf{0.14} & 0.83 & 0.43 \\ 
+   \hline
+\multirow{2}{*}{CD4+ Env2}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &  0.00 & \textbf{0.78} & 0.64 \\ 
+  & Fisher & 0.00 & 0.05 & 0.00 &  0.00 & 0.72 & 0.64 \\ 
+   \hline
+\multirow{2}{*}{CD8+ Gag1}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00  & 0.00 & \textbf{0.82} & 0.79 \\ 
+ & Fisher & 0.00 & 0.05 & 0.00  & 0.00 & 0.76 & 0.79 \\ 
+   \hline
+\multirow{2}{*}{CD4+ Gag1}& Beta-binomial & 0.00 & 0.00 & 0.00  & 0.00 & 0.71 & 0.79 \\ 
+ & Fisher & 0.00 & 0.00 & 0.00  & 0.00 & 0.71 & 0.79 \\ 
+   \hline
+
+\end{tabular}
+\end{center}
+\end{table}
+
+On day zero, where no response is expected, response rates from the beta--binomial model were either smaller than (5 cases), or no greater (23 cases) than those from Fisher's exact test, with the exception of CD8+ Env2 stimulated T--cells, where Fisher's exact test (0\% response rate) outperformed the beta--binomial model (5\% response rate).
+On day 28, where a response is expected in the treatment groups, the beta--binomial model had a higher response rate in 12 cases and was no different from Fisher's exact test in the remaining eight cases.
+In the control group, the beta--binomial model was equal to Fisher's exact test in eight cases, and performed slightly worse than Fisher's exact test in two cases (Env2 and Env1 stimulated CD8+ cells).
+%On day zero, we found that the response rates from the MIMOSA model were zero, and equal to response rates computed from Fisher's exact test across all treatment and control groups.
+The same was true for the response rate in the control groups at day 28.
+In the treatment groups, at day 28, the response rates for the MIMOSA model were  generally higher or equal to those for Fisher's exact test, demonstrating the increased sensitivity and specificity of our method.
+
+%The MIMOSA framework generates predictions for individual samples.
+To better understand the differences between the predictions from the MIMOSA model and Fisher's exact test, we examined the model fit from Gag2 stimulated CD4+ T--cells at day 28 producing IFNg (Figure~\ref{fig:icsdata}).
+We see that two individuals are predicted as responders by the MIMOSA model which are not detected by Fisher's exact test.
+Both individuals have a larger proportion of IFNg producing antigen specific T--cells in antigen stimulated than unstimulated samples, as measure by the MAP or ML estimates of the proportions (Figure~\ref{fig:icsdata}).
+
 \subsection{Fluidigm}
+
+\subsection{Simulation Studies}
+We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
+Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
+The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test.
+(Figure~\ref{fig:simulations}).
+The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions.
+
+We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}).
+Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).
+
+Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly,  we examined whether the unconstrained model could be used to accurately fit data generated from the constrained model.
+We found that the unconstrained model performed as well as the constrained model when fitting data generated from the constrained model (Figure~\ref{fig:simulations}).
+
+To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
+The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN054 data set (Figure~\ref{fig:simulations_trunc}).
+Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained model fit to constrained data.
+
 \section{Differential expression across combinatorial marker combinations}
 \subsection{Model}
-\subsection{Prior}
-\subsection{Parameter estimation}
-\subsection{ICS example with two markers}
-\section{Discussion}
-\section{Appendix}
-
-%\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
-%Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold.
-Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
-Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
-In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
-%
-%Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach.
-We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples).
-For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively.
-Note that usually, $N_u \ne N_s$.
-The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants.
-For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table:
-% latex table generated in R 2.14.0 by xtable 1.5-6 package
-% Wed Oct 12 09:41:50 2011
-
-%\subsection{The Mixture of Beta--Binomials}
-%Although we have specified the two models for the data, we do not know which observation was generated by which model.
-Clearly, not all individuals are expected to exhibit an immune response to a stimulation.
-Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}.
-We capture this uncertainty with a mixture framework of the two competing beta--binomial models.
-The likelihood for the mixture is given by:
-
-\subsection*{Generalization to Multiple Cytokines and Polyfunctionality with the Multinomial Dirichlet}
+%\subsection{Generalization to Multiple Cytokines and Polyfunctionality with the Multinomial Dirichlet}
 The model can be generalized to handle multiple cytokines in a single stimulation, in order to assess polyfunctional cytokine responses of T--cells.
-We use the Multinomial--Dirichlet family of distributions to model counts of events in two \textit{different} 2x2 contingency tables.
+Without loss of generality, for two markers we use the Multinomial--Dirichlet family of distributions to model counts of events in two \textit{different} 2x2 contingency tables.
 Here we consider only the \emph{unconstrained} case ($p_s\ne p_u$).
-The observed data can be represented in the following way:
-
-\begin{table}[ht]
+The observed data can be represented as shown in Table~\ref{tab:multdir}.
+\begin{table}[h]
 \centering
@@
 
-Where the vector of observed counts for individual $i$ in the stimulated or unstimulated sample is denoted: $\bar{n}_{\{s,u\}i} = \{n_{\{s,u\}ij}\} ; j\in\{1\ldots 4\}$, and $j$ indexes the cells of the appropriate contingency table shown in~Table~\ref{tab:multdir}.
+The vector of observed counts for individual $i$ in the stimulated or unstimulated sample is denoted: $\bar{n}_{\{s,u\}i} = \{n_{\{s,u\}ij}\} ; j\in\{1\ldots 4\}$, and $j$ indexes the cells of the appropriate contingency table shown in~Table~\ref{tab:multdir}.
 The counts are modelled as draws from different multinomial distributions:
+
 \begin{align}
-\text{if } & \bar{p}_{si}=\bar{p}_{ui};& \bar{n}_{ui} \sim \mathcal{M}(\bar{p}_{ui},N_{ui});\bar{n}_{si} \sim \mathcal{M}(\bar{p}_{ui},N_{si})\\
-\text{if } & \bar{p}_{si} \ne \bar{p}_{ui};& \bar{n}_{ui} \sim \mathcal{M}(\bar{p}_{ui},N_{ui});\bar{n}_{si} \sim \mathcal{M}(\bar{p}_{si},N_{si})
+ &(\bar{n}_{ui}|p_{ui},) \sim \mathcal{M}(\bar{p}_{ui},N_{ui}); (\bar{n}_{si}|p_{si}) \sim \mathcal{M}(\bar{p}_{ui},N_{si}); \text{if $i$ is a non--responder}\\
+ &(\bar{n}_{ui}|p_{ui}) \sim \mathcal{M}(\bar{p}_{ui},N_{ui}); (\bar{n}_{si}|p_{si}) \sim \mathcal{M}(\bar{p}_{si},N_{si});\text{if $i$ is a responder}
 \end{align}
-with Dirichlet priors on the proportions:
+where the $N_{\{s,u\}i}=\sum\limits_{j=1}\limits^{J} n_{\{s,u\}ij}$.
+\subsection{Prior}
+As with the one--dimensional case, we share information across subjects. We put Dirichlet priors on the proportions.
 \begin{align}
@@
 
-For the null component, where $\bar{p}_{s}=\bar{p}_{u}$ the marginal likelihood is given by: 
+For the null component, the marginal likelihood of the resulting multinomial--dirichlet is given by: 
 \begin{align}
@@
 
-The marginal likelihood for a component where $p_{sj} \ne p_{uj}$ for all $j$, is given by:
+Cells can express different combinations of markers.
+Since we now consider multiple markers simultaneously, the different marker combinations can be represented as separate components in the model.
+We examine the different possibilities below.
+
+\subsection{The simplest model: all marker combinations are different}
+The simplest case is where all possible marker combinations are expressed at different levels between the stimulated and unstimulated samples.
+In this case, we require only one additional component representing the alternative model, where $p_{sj} \ne p_{uj}$ for all $j$.
+An observation is assigned either to the null component (if there is no difference in any of the cell proportions), or to the alternative component, implying a difference in all the cell proportions.
+The marginal likelihood is given by:
 \begin{align}
@@
 
-Without loss of generality, if only some $p_j$ are different between stimulated and unstimulated samples, the appropriate components of $\alpha_j$  can be substituted in the calculation of the likelihood eq~\eqref{eq:postmult}.
+However, it is possible to have a situation where only some combinations of markers exhibit differences between stimulation and control.
+These can be represented explicitly by additional components in the model.
+Without loss of generality, if only some $p_j$ are different between stimulated and unstimulated samples, the appropriate mixture components can be represented by simply substituting the relevant $\alpha_j$ in the calculation of the likelihood eq~\eqref{eq:postmult}.
 
+\subsection{Parameter estimation}
+The multivariate model is fit using the EM algorithm. 
 The marginal likelihood of the non--responding component, given in eq~\eqref{NRmd} can be rewritten as:
@@
 
-\subsection*{Simulation Studies}
-We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
-Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
-The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test.
-(Figure~\ref{fig:simulations}).
-The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions.
-In order to assess whether the unconstrained model could be used to effectively fit data from a constrained model and benefit from faster computations using exact closed form expressions rather than Monte-Carlo integration, we fit the same data simulated from the constrained model using the unconstrained model (Figure~\ref{fig:simulations}.
+\subsection{ICS example with two markers}
+\section{Discussion}
+\section{Appendix}
 
-To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN054 data set (Figure~\ref{fig:simulations_trunc}).
+%\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
+%Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold.
+Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
+Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
+In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
+%
+%Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach.
+We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples).
+For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively.
+Note that usually, $N_u \ne N_s$.
+The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants.
+For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table:
+% latex table generated in R 2.14.0 by xtable 1.5-6 package
+% Wed Oct 12 09:41:50 2011
+
+%\subsection{The Mixture of Beta--Binomials}
+%Although we have specified the two models for the data, we do not know which observation was generated by which model.
+Clearly, not all individuals are expected to exhibit an immune response to a stimulation.
+Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}.
+We capture this uncertainty with a mixture framework of the two competing beta--binomial models.
+The likelihood for the mixture is given by:
 
 \section*{Results}
-\subsection*{Simulations Show MIMOSA Outperforms Fisher's Exact Test and is Robust to Departures from Model Assumptions}
-We ran simulation studies to assess the performance of the constrained and unconstrained models, as described in the Methods.
-We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}).
-Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).
-
-Since the constrained model relies on Monte--Carlo integration, which can be computationally costly, to estimate the normalizing constant in the likelihood calculation, we examined whether the unconstrained model could be used to accurately fit data generated from the constrained model.
-We found that the unconstrained model performed as well as the constrained model when fitting data generated from the constrained model (Figure~\ref{fig:simulations}).
-
-In order to assess the sensitivity of the model to deviations from model assumptions, we fit the unconstrained model to simulated data where the proportions were drawn from truncated normal distributions (Figure~\ref{fig:simulations_trunc}).
-Even in these circumstances the unconstrained MIMOSA model outperformed Fisher's exact test and in fact performed about as well as the constrained model fit to constrained data.
-
-\subsection*{MIMOSA Outperforms Fisher's Exact Test in Real--World Vaccine Trial Data from HVTN054}
-
-We tested our method on ICS data from HVTN054, a safety and immunogenicity vaccine trial for a replication defective Adenovirus vector HIV vaccine.
-The data set consisted of 48 individuals who received two doses of vaccine or placebo, and ICS time points for these individuals were available at day 0 and day 28 after vaccination.
-Responders for each cytokine and stimulation combination were classified using Fisher's exact test and the constrained MIMOSA model at the 1\% FDR level, then the response rate within the control and treatment groups were calculated for both methods.
-Figure~\ref{fig:positivityrates} shows the response rates for a subset of the ICS data (Gag2 and Pol2 stimulations, for the IL2 and IFNg cytokines, at days 0 and 28, in both CD4 and CD8 T--cell subpopulations).
-On day 0, we found that the response rates from the MIMOSA model were zero, and equal to response rates computed from Fisher's exact test across all treatment and control groups.
-The same was true for the response rate in the control groups at day 28.
-In the treatment groups, at day 28, the response rates for the MIMOSA model were  generally higher or equal to those for Fisher's exact test, demonstrating the increased sensitivity and specificity of our method.
-
-The MIMOSA framework generates predictions for individual samples.
-To better understand the differences between the predictions from the MIMOSA model and Fisher's exact test, we examined the model fit from Gag2 stimulated CD4+ T--cells at day 28 producing IFNg (Figure~\ref{fig:icsdata}).
-We see that two individuals are predicted as responders by the MIMOSA model which are not detected by Fisher's exact test.
-Both individuals have a larger proportion of IFNg producing antigen specific T--cells in antigen stimulated than unstimulated samples, as measure by the MAP or ML estimates of the proportions (Figure~\ref{fig:icsdata}).
-
 \subsection*{Testing Multifunctional Response via the Multinomial-Dirichlet model}

### 2012-04-13 MIMOSA.tex
Updated fluidigm figures
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
+\subsection{Study 073}
+% latex table generated in R 2.16.0 by xtable 1.7-0 package
+% Wed Apr 11 15:01:22 2012
+\begin{table}[ht]
+\begin{center}
+\caption{Cytokine-specific response rates for ENV1 stimulated CD4+ T--cells on visits 2 through 14.1. No response is expected on visit 2 (day 0) for any cytokine.}\label{tab:positivityrates2}
+\begin{tabular}{rrrrrrrrrrrr}
+  \hline
+&Visit & 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 & 14.1 \\ 
+  \hline
+\multirow{2}{*}{IL2}&BB & 0.00 & 0.17 & 0.25 & 0.00 & 0.55 & 0.11 & 0.32 & 0.17 & 0.11 & 0.53 \\ 
+ &Fisher & 0.01 & 0.17 & 0.25 & 0.00 & 0.55 & 0.15 & 0.30 & 0.20 & 0.13 & 0.47 \\ 
+   \hline
+  \hline
+&Visit & 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 \\ 
+  \hline
+\multirow{2}{*}{TNF}&BB & 0.05 & 0.00 & 0.21 & 0.00 & 0.57 & 0.12 & 0.39 & 0.44 & 0.25 \\ 
+  &Fisher & 0.07 & 0.00 & 0.26 & 0.00 & 0.55 & 0.14 & 0.38 & 0.42 & 0.25 \\ 
+   \hline
+  \hline
+ &Visit& 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 & 14.1 \\ 
+  \hline
+\multirow{2}{*}{IFN}&BB & 0.00 & 0.43 & 0.31 & 0.29 & 0.63 & 0.28 & 0.39 & 0.34 & 0.17 & 0.56 \\ 
+  &Fisher & 0.00 & 0.29 & 0.29 & 0.43 & 0.60 & 0.26 & 0.37 & 0.32 & 0.16 & 0.44 \\ 
+   \hline
+\end{tabular}
+\end{center}
+\end{table}
 \subsection{Fluidigm}
@@
 \subsection{Parameter estimation}
-The multivariate model is fit using the EM algorithm. 
-The marginal likelihood of the non--responding component, given in eq~\eqref{NRmd} can be rewritten as:
-\begin{align}
-\begin{split}
-l(\bar{\alpha}_u|D)=\sum_i \left[ \log \Gamma (\sum_j \alpha_{uj}) - \log \Gamma (\sum_j \alpha_{uj}+n_{uij}+n_{sij})+ \sum_j (\log \Gamma(\alpha_{uj}+n_{uij}+n_{sij}) \right.\\ \left.
+The multivariate model is fit using the EM algorithm. We estimate the model hyper--parameters numerically ($\bar{\alpha}_u,\bar{\alpha}_s$) using Newton's method. 
+%The marginal likelihood of the non--responding component, given in eq~\eqref{NRmd} can be rewritten as:
+%\begin{align}
+%\begin{split}
+%l(\bar{\alpha}_u|D)=\sum_i \left[ \log \Gamma (\sum_j \alpha_{uj}) - \log \Gamma (\sum_j \alpha_{uj}+n_{uij}+n_{sij})+ \sum_j (\log \Gamma(\alpha_{uj}+n_{uij}+n_{sij}) \right.\\ \left.
 -\log\Gamma(\alpha_{uj}))+ \log\Gamma(N_{si}+1) +\log\Gamma(N_{ui}+1)-\right.\\ \left.\left[\sum_j(\log\Gamma(n_{sij}+1)+\log\Gamma(n_{uij}+1))\right]\right]\label{mdloglike}
-\end{split}
-\end{align}
+%\end{split}
+%\end{align}
 
-The gradient is given by:
+The gradient of the log of the marginal likelihood in equation~\eqref{NRmd} is:
 \begin{align}
@@
 
-and the Hessian is:
+and the Hessian:
 \begin{align}
@@
 
-The marginal likelihood of the responding component in eq~\ref{eq:postmult} can be rewritten as:
-\begin{align}
-\begin{split}
-\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
--\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
--\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
-\end{split}
-\end{align}
-with gradient:
+The gradient of the log of the marginal likelihood of the responding component in eq~\ref{eq:postmult} is:
+%\begin{align}
+%\begin{split}
+%\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
+%-\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
+%-\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
+%\end{split}
+%\end{align}
+%with gradient:
 \begin{align}
@@
 \subsubsection*{Mixture Model Complexity}
-We may wish to detect any of $2^3 = 8$ different possible scenarios where the proportion of events in corresponding cells of the contingency tables are either equal or unequal between stimulated and unstimulated conditions.
+In the two marker case described here, we may wish to test for any of 8 different possible scenarios where the proportion of events in corresponding cells of the contingency tables are either equal or unequal between stimulated and unstimulated conditions.
 Such a model would have 8 components and 55 parameters.
-However, if we recognize that the models can be nested, i.e.
-that parameters can be shared across components with similar outcomes, then the number of parameters can be reduced to 19, and further to 15 if we only consider components where any one cell of the tables differs between stimulated and unstimulated conditions.
-This is outlined in Table~\ref{tab:nesting}.
+However, if we recognize that the models can be nested, i.e., that parameters can be shared across components with similar outcomes, then the number of parameters can be reduced to 19, and further to 15 if we only consider components where any one cell of the tables differs between stimulated and unstimulated conditions.
+This number of parameters required for modelling different numbers of components is shown in Table~\ref{tab:nesting}.
 \begin{table}
@@
 \subsubsection*{Model simplifications for K>2 cytokines}
-In the case of multiple cytokines, there is a combinatorial explosion in the number of parameters of the resulting model.
+In the case of more than two cytokines, there is a combinatorial explosion in the number of parameters of the resulting model.
 We propose several alternative solutions to reducing the number of parameters in general, when the dimensionality of the Multinomial Dirichlet increases.
-If we model $N$ cytokines simultaneously, this results in $2^N$ possible combinations.
+If we model $N$ cytokines simultaneously, this results in $2^N$ possible marker combinations (i.e.
+each marker is either expressed, or not).
 We let $K$ be the number of possible cytokine combinations being modelled, with hyperparameter vector $\bar{\alpha}_0=\{\alpha_{00},\ldots,\alpha_{0K-1}\}$, and $\bar{\alpha}_s=\{\alpha_{s0},\ldots,\alpha_{sK-1}\}$ for the unstimulated and stimulated distributions, respectively.
@@
 \subsubsection*{Fully shared model hyperparameters}
-In the most simplistic case, we set up a two mixture component model (responders and non--responders), with a goal of identifying observations with no response (equal to the null), or some response in any of the possible combinations of measured cytokines.
-For ICS data, the first component of the hyperparameter vector represents cells not expressing any cytokine combination, and is usually much larger than the remaining components.
-As a first approximation, we can set the $K-1$ remaining parameters to be all equal $\{\alpha_{01}=\{\alpha_{01},\ldots,\alpha_{0K-1}\}$ and thus all variation across cells expressing different cytokine combinations in the unstimulated samples are captured by a single parameter.
-The same simplification is made for modeling the distribution of stimulated samples.
-As stated above, such a model can only be fit to two mixture components (the null component and the responder component), but requires estimation of only four parameters.
-In practice, such a model would likely have a vary high false positive and false negative rate.
+In the most simplistic case, we set up a two mixture component model (responders and non--responders).
+This formulation would ignore most of the marker combinations and only explicitly test for all combinations equal to the null, or all combinations different from the null.
+
+For ICS data, the first component of the hyper--parameter vector represents cells not expressing any cytokine combination, and is usually much larger than the remaining components.
+As a first approximation, we can set the $K-1$ remaining parameters to be all equal $\{\alpha_{01}=\{\alpha_{01},\ldots,\alpha_{0K-1}\}$ (e.g.
+shared across all possible combinations) and thus all variation across cells expressing different cytokine combinations in the stimulated and unstimulated samples are captured by four parameters.
+As stated above, such a model can only be fit to two mixture components (the null component and the responder component), but requires a total of only five parameters.
+In practice, such a model would likely have a vary high false negative rate, since observations with differences in only a few marker combinations would be absorbed into the null component.
 
 \subsubsection*{Second order hyperparameters as functions of first--order hyperparameters}
-Moving towards a more realistic case, we can define first--order hyperparameters as those elements of the $\alpha$ vectors corresponding to cells expressing individual cytokines (i.e.
+A more realistic scenario would be to define \textit{first--order} hyper--parameters as those elements of the $\alpha$ vectors corresponding to cells expressing individual cytokines (i.e.
 cytokine A only, or cytokine B only).
-Second--order and higher parameters are defined as cells expressing pairs, or triplets of cytokines, and so forth  (i.e.
-A and B, B and C, C and D).
-Higher order parameters can be expressed as functions of the first--order parameters (i.e.
+Second--order and higher parameters are defined for cells expressing pairs, triplets, and higher order combinations of cytokines (i.e.
+A and B, or B and C, or A,B and C).
+Higher order parameters could be expressed as functions of the first--order parameters (i.e.
 cells expressing A and B can be parameterized as $\alpha_{0ab} = \frac{\alpha_{0a}+\alpha_{0b}}{2}$ or $\alpha_{0ab}=\alpha_{0a}\alpha_{0b}$).
+
 This parameterization can be extended to models with more than two mixture components, allowing us to identify different types of responders.
@@
 \subsection{ICS example with two markers}
+We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
+As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
+We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
+
+The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations.
+We compared multivariate MIMOSA against Fisher's exact test.
+Fisher's exact test was used to test whether the number of cells expressing a specific combination of cytokines (i.e TNFa+/IL4-, TNFa+/IL4+, and so forth) was significantly higher in the stimulated samples than in the unstimulated samples.
+In this comparison, Fisher's exact test detected only one of the 8 responders detected by the MIMOSA model in the Env3 stimulated, day 28 data, and detected two false positives in the day zero data.
+Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.
+
 \section{Discussion}
-\section{Appendix}
+The variety of single--cell assays being adopted by the immunology community is increasing.
+Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data.
+Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.
+Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}.
+Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
+
+The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model cell counts in non--responding and responding individuals.
+Information is shared across non--responders and across responders through an empirical--Bayes prior, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}).
+The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample, our simulations show that we can relax this assumption to the unconstrained case where $p_s\ne p_u$ without compromising the observed false positive or false negative rates (Figures~\ref{fig:simulations} and \ref{fig:simulations_trunc}), while saving on computation time.
+
+\todo[inline]{Fix references to non-existent figure}
+The analysis of real--world ICS data from a vaccine trial demonstrated that the MIMOSA model outperformed Fisher's exact test in detecting individuals who were responders and non--responders to vaccine across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:positivityrates}).
+Importantly, the MIMOSA model was fit within each antigen stimulation $\times$ cytokine $\times$ time point combination but the model was blinded to placebo and vaccine treatment.
+Despite this, MIMOSA demonstrated a lower false positive rate at day zero (increased specificity), where no response is expected, than Fisher's exact test (Figure~\ref{fig:positivityrates} Gag2, IL2, CD4+ T--cells, day 0, treatment group T1).
+Similarly, none of the placebo treated samples were identified as responders by the MIMOSA model (Figure~\ref{fig:positivityrates} treatment group C).
+MIMOSA also exhibited an increased sensitivity in real--world data compared to Fisher's exact test, as demonstrated by the higher response rates for 7 of the 8 stimulation/cytokine/cell type combinations at day 28 to which the model was fit, and where a response is expected to be observed.
+Although it is not possible to validate the responders identified at day 28 for all cases, the conclusion of increased sensitivity is consistent with both the performance of the model in simulations, and with the increased specificity observed on day 0 and in the placebo treated samples.
+
+\todo[inline]{Fix references to non--existent figure}
+We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
+For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test.
+These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions.
+This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model.
+This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.
+
+Polyfunctionality (identifying cells coexpressing multiple proteins, cytokines, or genes, is important in single cells assays since it allows the identification of functionally distinct subsets of cells in heterogeneous cell populations~\cite{Milush:2009bz}.
+In vaccine trials, these polyfunctional cell subsets may be correlated with vaccine response or efficacy.
+Existing methods for identifying polyfunctional cytokine profiles from ICS data either test individual combinations of cytokines separately, perform global tests at the expense of decreased power when only specific differences are of interest, sometimes  empirical in nature, and generally don't share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
+As others have pointed out, in order to have the most power to detect a true difference, the statistical test selected to detect differences in various combinations of cytokines should take into consideration the cytokine combinations that are of interest~\cite{Nason:2006dx}.
+A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different.
+This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.
+
+We show a proof of principle example with two components to identify responder and non--responder observations for the four--dimensional (two cytokine) case.
+Sharing of parameters across nested models in the two--cytokine case allows us to reduce the number of parameters.
+The model is easily extended.
+First, to multiple components to identify not just responders and non--responders, but responders with respect to each combination of cytokines.
+Second, extensions to multiple cytokines are straightforward, only requiring increasing the dimensionality of the Multinomial--Dirichlet distribution.
+With the increase in dimensionality there is a combinatorial explosion in the number of combinations of cytokines, and the resulting number of parameters in the model.
+We present several possible approaches to reducing the number of parameters in the model by sharing the hyperparamters across different combinations of cytokines.
+A number of these strategies can be generalized to models with multiple components that allow us to identify observations with differences in specific cytokine combinations between treatment and control samples.
+
+\section{Conclusions}
+We have developed a combinatorial mixture model framework for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays.
+The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
 
@@
 The likelihood for the mixture is given by:
-
-\section*{Results}
-\subsection*{Testing Multifunctional Response via the Multinomial-Dirichlet model}
-We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
-As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
-We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
-
-The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations.
-We compared multivariate MIMOSA against Fisher's exact test.
-Fisher's exact test was used to test whether the number of cells expressing a specific combination of cytokines (i.e TNFa+/IL4-, TNFa+/IL4+, and so forth) was significantly higher in the stimulated samples than in the unstimulated samples.
-In this comparison, Fisher's exact test detected only one of the 8 responders detected by the MIMOSA model in the Env3 stimulated, day 28 data, and detected two false positives in the day zero data.
-Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.
 
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=6in]{Figures/PositivityRates} 
-   \caption{Comparison of MIMOSA and Fisher's exact test for calling responders in ICS data.
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=6in]{Figures/PositivityRates} 
+%   \caption{Comparison of MIMOSA and Fisher's exact test for calling responders in ICS data.
 Response rates for IL2 and IFNg cytokine positivity in Gag2 and Pol2 stimulated samples at days 0 and 28 in the CD4 and CD8 T--cell subpopulations of the HVTN054 ICS data.
@@
 In the control groups, response rates are equal (and zero) between Fisher's exact test and the MIMOSA model at day 0 and at day 28, as expected.}
-   \label{fig:positivityrates}
-\end{figure}
+%   \label{fig:positivityrates}
+%\end{figure}
 
@@
    \centering
-   \includegraphics[width=6in]{Figures/HVTN054LikelihoodAndVolcanoplots} 
-   \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28.
+   \includegraphics[width=6in]{Figures/FluidigmMIMOSA.pdf} 
+  % \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28.
 A significant difference between control and stimulated samples is called at the 1\% FDR threshold (red) for the Beta--binomial model, and at 1\% for FDR adjusted p--values from Fisher's exact test (green).
@@
 The effect of shrinking the MAP estimates towards zero can be seen in the volcano plot.}
-   \label{fig:icsdata}
+%   \label{fig:icsdata}
+\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
+The posterior probability of response times the sign of the change in expression is shown in the first panel (red indicates a significant decrease, green a significant increase, relative to the control).
+Columns and rows are clustered based on these signed posterior probabilities.
+The middle panel shows the $log_2$ odds ratio of the proportion of cells expressing a gene in the stimulated vs.
+control samples.
+Rows and columns are ordered as in the left panel for comparison.
+The right panel shows the difference in the proportion of cells expressing each gene in the stimulated vs.
+control samples.
+Ordering of the rows and columns is preserved as in the first panel.
+The traces show the deviations of each cell from zero.}
+\label{fig:fluidigm}
 \end{figure}
@@
 
-\section*{Discussion}
-The variety of single--cell assays being adopted by the immunology community is increasing.
-Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data.
-Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.
-Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}.
-Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
-
-The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model cell counts in non--responding and responding individuals.
-Information is shared across non--responders and across responders through an empirical--Bayes prior, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}).
-The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample, our simulations show that we can relax this assumption to the unconstrained case where $p_s\ne p_u$ without compromising the observed false positive or false negative rates (Figures~\ref{fig:simulations} and \ref{fig:simulations_trunc}), while saving on computation time.
-
-The analysis of real--world ICS data from a vaccine trial demonstrated that the MIMOSA model outperformed Fisher's exact test in detecting individuals who were responders and non--responders to vaccine across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:positivityrates}).
-Importantly, the MIMOSA model was fit within each antigen stimulation $\times$ cytokine $\times$ time point combination but the model was blinded to placebo and vaccine treatment.
-Despite this, MIMOSA demonstrated a lower false positive rate at day zero (increased specificity), where no response is expected, than Fisher's exact test (Figure~\ref{fig:positivityrates} Gag2, IL2, CD4+ T--cells, day 0, treatment group T1).
-Similarly, none of the placebo treated samples were identified as responders by the MIMOSA model (Figure~\ref{fig:positivityrates} treatment group C).
-MIMOSA also exhibited an increased sensitivity in real--world data compared to Fisher's exact test, as demonstrated by the higher response rates for 7 of the 8 stimulation/cytokine/cell type combinations at day 28 to which the model was fit, and where a response is expected to be observed.
-Although it is not possible to validate the responders identified at day 28 for all cases, the conclusion of increased sensitivity is consistent with both the performance of the model in simulations, and with the increased specificity observed on day 0 and in the placebo treated samples.
-
-We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
-For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test.
-These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions.
-This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model.
-This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.
-
-Polyfunctionality (identifying cells coexpressing multiple proteins, cytokines, or genes, is important in single cells assays since it allows the identification of functionally distinct subsets of cells in heterogeneous cell populations~\cite{Milush:2009bz}.
-In vaccine trials, these polyfunctional cell subsets may be correlated with vaccine response or efficacy.
-Existing methods for identifying polyfunctional cytokine profiles from ICS data either test individual combinations of cytokines separately, perform global tests at the expense of decreased power when only specific differences are of interest, sometimes  empirical in nature, and generally don't share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
-As others have pointed out, in order to have the most power to detect a true difference, the statistical test selected to detect differences in various combinations of cytokines should take into consideration the cytokine combinations that are of interest~\cite{Nason:2006dx}.
-A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different.
-This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.
-
-We show a proof of principle example with two components to identify responder and non--responder observations for the four--dimensional (two cytokine) case.
-Sharing of parameters across nested models in the two--cytokine case allows us to reduce the number of parameters.
-The model is easily extended.
-First, to multiple components to identify not just responders and non--responders, but responders with respect to each combination of cytokines.
-Second, extensions to multiple cytokines are straightforward, only requiring increasing the dimensionality of the Multinomial--Dirichlet distribution.
-With the increase in dimensionality there is a combinatorial explosion in the number of combinations of cytokines, and the resulting number of parameters in the model.
-We present several possible approaches to reducing the number of parameters in the model by sharing the hyperparamters across different combinations of cytokines.
-A number of these strategies can be generalized to models with multiple components that allow us to identify observations with differences in specific cytokine combinations between treatment and control samples.
-
-\section*{Conclusions}
-We have developed a combinatorial mixture model framework for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays.
-The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
 \renewcommand{\thesection}{S.\arabic{section}}

### 2012-04-26 MIMOSA.tex
Some changes to the model section

@@
 
-\usepackage{fontspec,xltxtra,xunicode}
-\defaultfontfeatures{Mapping=tex-text}
-\setromanfont[Mapping=tex-text]{Times New Roman}
-\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Gill Sans}
-\setmonofont[Scale=MatchLowercase]{Andale Mono}
+%\usepackage{fontspec,xltxtra,xunicode}
+%\defaultfontfeatures{Mapping=tex-text}
+%\setromanfont[Mapping=tex-text]{Times New Roman}
+%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Gill Sans}
+%\setmonofont[Scale=MatchLowercase]{Andale Mono}
 
-\title{Combinatorial Mixture Models for Single Cell Assays}
+\title{Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies}
 \author[1]{Greg Finak}
@@
 \section{Introduction}
-Single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
+In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
+Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers).
+Recently, DVS Sciences Inc.
+introduced a new cytometry technology named cytometry Time-Of-Flight (CyTOF) that uses rare isotopes instead of fluorochromes (as in standard FCM), as antibody tags to label single cells.
+CyTOF drastically extends the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell.
+Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize cell poly-functionality.
+A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6.
+While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability.
+This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data.
+Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general.
+For this reason, single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
 These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood.
@@
 
-A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine.
+A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine.
 Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells.
 While not all T--cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
-After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future\cite{McKinstry:2010ei}.
-These antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells.
-The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that was present in the original vaccine.
+After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}.
+These antigen--specific T--cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T--cells.
+The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
 Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
@@
 
-The ICS assay in flow cytometry is just one example of the applications of single--cell technologies to immunology.
+%The ICS assay in flow cytometry is just one example of the applications of single--cell technologies to immunology.
 Single--cell gene expression technologies such as Fluidigm are enabling researchers to measure the expression of 96 genes in 96 single cells from a homogeneous population sorted by flow cytometry\cite{Jang:2011uj}.
@@
 These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
+In the context of single-cell gene expression, very few methods have been proposed.
+To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements.
+However, the proportion of single cells expressing individual genes also carries information and should be evaluated.
 
@@
 The framework developed in this paper addresses these issues explicitly.
-We present a combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells.
+We present a multinomial-Dirichlet combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells.
 The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.
-Importantly, our approach shares information across observations by means of prior distributions on the responding, stimulated samples and the non--responding samples thereby increasing sensitivity and specificity to detect responding cell populations.
-The cell counts from the stimulated and unstimulated experiments are modelled independently, and different combinations of markers are represented as different mixture components in the framework.
+Importantly, our approach shares information across individuals (or subjects) by means of a Dirichlet prior distribution placed on the unknown cell population proportions of the mutinomial likelihood, which help increase sensitivity and specificity to detect rare antigen specific responding cell populations.
+The cell counts from the stimulated and unstimulated experiments are modeled independently, and different combinations of markers are represented as different mixture components in the framework.
 This approach allows us to omit certain combinations of markers that are not observable (i.e., two cytokines may never be co--expressed), or to explicitly represent combinations that are of interest (i.e., we can explicitly represent and test for a difference in a pair of cytokines in an experiment measuring many cytokines).
@@
 \section{Data structure and notation} 
-We analyzed two data sets from two different immunological single--cell assays typically used in vaccine trials, one flow cytometry data set, and one single--cell gene expression data set.
+In this paper we consider two different immunological single-cell assays typically used in vaccine trials, one flow cytometry data set, and one single-cell gene expression data set.
 
-\textit{Flow cytometry}: This ICS data is from a phase--I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post--vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
-The statistical analysis of responder and non--responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
-The goal of this data set was to assess response rates of CD4 and CD8 T--cell populations to different antigens.
+\textit{Flow cytometry}: This ICS data is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
+The statistical analysis of responder and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
+The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.
 
@@
 In the remainder of the paper, we use the following notation to describe our model.
-We observe cell counts from $P$ subjects in two conditions (stimulated and unstimulated).
+We observe cell counts from $I$ individuals in two conditions (stimulated and unstimulated).
 Each cell can either be positive or negative for a marker.
-We let $i$ index the subjects and $k$ index the cytokines.
+We let $i$ index the individual and $k$ index the measured markers, \textit{i.e.} cytokine for the ICS data and genes for the ICS data.
 The subscripts $s$,$u$ refer to the cell counts from stimulated and unstimulated conditions, respectively.
-The total number of cells is denoted by $N$, and given in our model, while $n$, the number of positive cells is part of the observe data.
+For a given individual, $p$, the total number of cells is denoted by $N_p$, while $n_p$, the number of positive cells is part of the observe data.
 Thus for a given cytokine, $N_{si}$ denotes the total number of cells collected for subject $i$ in the stimulated condition and $n_{si}$ denotes how many of those cells were cytokine positive.
@@
 \end{table}
-We consider a marker to be differentially expressed between conditions if the number of cells in the stimulated samples is significantly different from the number of cells in the control sample.
+We consider a marker to be differentially expressed between conditions if the proportion of cells in the stimulated samples is different from the number of cells in the control sample.
 The model is described below.
@@
 \subsection{Beta--Binomial Model for One Marker}
-The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as  binomial random variables.
+The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as binomial random variables, as follows,
 \begin{align}
@@
 
-Where $p_{si}$, $p_{ui}$ are the unobserved proportions.
-Model~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control, and thus no difference between the proportion of cytokine positive cells in stimulated and control samples.
-Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in cytokine response between the stimulation and the control, and thus a difference in the proportion of cells between stimulated and control samples.
-In the case of no response, the counts for the stimulated and unstimulated samples are independent draws from the same  binomial distribution.
+where $p_{si}$, $p_{ui}$ are the unobserved proportions for the stimulated and un-stimulated samples.
+Under the null hypothesis, \textit{i.e.} Equation~\ref{eq:null}, there is no difference between the stimulation and control samples, and the proportions are equal.
+Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in proportions between the stimulation and control samples.
 
-\subsection{Prior}
-Our model shares information across responders and non--responders, respectively. This is done by placing Beta priors on the $p_{si}$, $p_{ui}$, across subjects:
+\subsection{Priors}
+Our model shares information across all individuals using an exchangeable Beta priors on the unknown proportions, as follows, 
+ \begin{align}
+(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior-null}\\
+(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u) \label{eq:prior-alternative}
+ \end{align}
+where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.
+$p_{si}\ne p_{ui}$ (or $p_{si}> p_{ui}$), and 0 otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals.
+Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models.
+Under the null model, the stimulated condition should behave just like the unstimulated one, and thus
+the stimulated response rates (\textit{i.e.} $p_{si}$'s) should follow the same prior distribution.
+Finally, we assume that the $z_i$'s are independent and identically distributed as Bernoulli with probability $w$.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can easily be estimated via an Empirical-Bayes fashion using and Expectation-Maximization algorithm as described in Section \ref{sec:estimation}.
 
- \begin{align}
- p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s);&
- p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior}
- \end{align}
+\subsection{Parameter estimation}
+The log complete data likelihood can be written as 
+\begin{equation}
+l(\theta|y)=\sum_i z_i l_0(\alpha_u, \beta_u|y_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|y_i)
+\end{equation}
+where $l_0$ and $l_1$ are the marginal likelihoods for the null and alternative models obtained after integrating out the prior for $\alpha_u, \beta_u, \alpha_s$, and $\beta_s$.
 
-Additional sharing of information is done through the $\alpha_u, \beta_u$ hyper--parameters, which are common for the null and alternative models. 
-%If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models.
-Given this formulation, the posterior probability of the data given that it is generated by model \eqref{eq:null}, is:
-\subsection{Posetrior}
 \subsubsection{Null Model}

### 2012-05-02 MIMOSA.tex
Some more changes to the method section.

@@
 In the remainder of the paper, we use the following notation to describe our model. 
-We observe cell counts from $I$ individuals in two conditions (stimulated and unstimulated).
+From this point on, we assume that we observe cell counts from $I$ individuals in two conditions: stimulated and un-stimulated.
 Each cell can either be positive or negative for a marker.
-We let $i$ index the individual and $k$ index the measured markers, \textit{i.e.} cytokine for the ICS data and genes for the ICS data.
-The subscripts $s$,$u$ refer to the cell counts from stimulated and unstimulated conditions, respectively.
-For a given individual, $p$, the total number of cells is denoted by $N_p$, while $n_p$, the number of positive cells is part of the observe data.
-Thus for a given cytokine, $N_{si}$ denotes the total number of cells collected for subject $i$ in the stimulated condition and $n_{si}$ denotes how many of those cells were cytokine positive.
-Similarly, $N_{ui}$, $n_{ui}$ represent the total, and cytokine positive cell counts in the unstimulated condition for the same subject.
-For each cytokine and each stimulation, given the total cell counts, the observed data is a matrix of size $2 \times P$, where the $i$'th row is a subject, $\mathbf{y}_i = <n_{si},n_{ui}>$.
+Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations.
+We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples.
+We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individuals $i$ in each sample.
+For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{sk}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{uk}=\{n_{uik}: k=1,\dots,2^K\}$.
+Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
 
 \section{Differential expression with one marker}
-The data for a single marker can be represented in contingency table form:
+In practice, stimulation studies like the ones presented here, are often analyzed one marker at a times to avoid being under powered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
+As a consequence, we first consider the one marker case, \textit{i.e.} $K=1$.
+In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
 
@@
 \parbox{0.8\linewidth}{
-\caption{2 x 2 contingency table of counts for cytokine positive and cytokine negative events between stimulated and unstimulated conditions}\label{tab:twobytwo}
+\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given individual $i$.}\label{tab:twobytwo}
 \centering
@@
 \multicolumn{1}{l}{}&
-\multicolumn{2}{c}{Cytokine}\\
+\multicolumn{2}{c}{Marker}\\
  & Negative & Positive \\ 
@@
 \end{table}
-We consider a marker to be differentially expressed between conditions if the proportion of cells in the stimulated samples is different from the number of cells in the control sample.
-The model is described below.
+For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
+Individuals that show differential expression for a given marker will be called responders for that marker.
+In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described below.
 
 \subsection{Beta--Binomial Model for One Marker}
-The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as binomial random variables, as follows,
+For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
 \begin{align}
-(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non--responder}\\
+(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non-responder}\\
 (n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
  \end{align}
-
-where $p_{si}$, $p_{ui}$ are the unobserved proportions for the stimulated and un-stimulated samples.
-Under the null hypothesis, \textit{i.e.} Equation~\ref{eq:null}, there is no difference between the stimulation and control samples, and the proportions are equal.
-Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in proportions between the stimulation and control samples.
+ where $p_{si}$, $p_{ui}$ are the unobserved proportions for the stimulated and un-stimulated paired samples.
+Under the null hypothesis, \textit{i.e.}~\eqref{eq:null}, there is no difference between the stimulation and control samples, and the proportions are equal.
+Under the alternative model given by \eqref{eq:alternate}, there is a difference in proportions between the two samples.
 
@@
 where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.
-$p_{si}\ne p_{ui}$ (or $p_{si}> p_{ui}$), and 0 otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals.
-Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models.
-Under the null model, the stimulated condition should behave just like the unstimulated one, and thus
-the stimulated response rates (\textit{i.e.} $p_{si}$'s) should follow the same prior distribution.
-Finally, we assume that the $z_i$'s are independent and identically distributed as Bernoulli with probability $w$.
-Treating the $z_i$'s as missing data, the unknown parameter vector $\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can easily be estimated via an Empirical-Bayes fashion using and Expectation-Maximization algorithm as described in Section \ref{sec:estimation}.
+$p_{si}\ne p_{ui}$, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals.
+Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present for the alternative model.
+Under the null model, the stimulated condition should behave just like the unstimulated one, and thus the stimulated response rates (\textit{i.e.} $p_{si}$'s) should follow the same prior distribution.
+Finally, we assume that the $z_i$'s are independent and identically distributed as Bernoulli with probability $w$, where $w$ represents the proportion of responders.
+It follows that marginally, i.e.
+after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can easily be estimated via an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}.
+We will also explore a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ wis assumed to be drawn from a uniform distribution between 0 and 1.
+In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{sec:estimation}.
 
 \subsection{Parameter estimation}
-The log complete data likelihood can be written as 
-\begin{equation}
-l(\theta|y)=\sum_i z_i l_0(\alpha_u, \beta_u|y_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|y_i)
-\end{equation}
-where $l_0$ and $l_1$ are the marginal likelihoods for the null and alternative models obtained after integrating out the prior for $\alpha_u, \beta_u, \alpha_s$, and $\beta_s$.
-
-\subsubsection{Null Model}
-Given the above formulation, the posterior for the null model is Beta--binomial, with marginal likelihood (integrating out $p_u$):
-
+\label{sec:estimation}
+Whether we use an EM or an MCMC algorithm for parameter estimation, we make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p$'s for the null and alternative models, to simplify our calculations.
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Appendix A), and are given by,
  \begin{align}
-  	\mathrm{L}(\alpha_u,\beta_u|\mathbf{y})
+  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
 	&=\prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
-%	\intertext{Integrating out the proportions gives the marginal log--likelihood.}
-%	\begin{split}
-%	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N_{si}}{n_{si}}}+\log{\binom{N_{ui}}{n_{ui}}}+\right.\\
-%	\left.\log{\left(\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLL}
-%	\end{split}
  \end{align} 
- \subsubsection{Alternative Model - Unconstrained}
- If the data is generated by model~\eqref{eq:alternate} then we have  $p_s\ne p_u$, and the marginal likelihood of the alternative is:
+and
 \begin{align}
 	\begin{split}
-\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) =\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\label{model2:unconstrained}
+\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
+=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}.\label{model2:unconstrained}
 \end{split}
@@
 
-\subsubsection{Alternative Model - Constrained}
-For ICS data, however, we are generally interested in the one--sided case where $p_s>p_u$, which is consistent with a positive response to stimulation (i.e., more cells are producing the marker after stimulation).
-In this constrained case, integrating out $p_s$, $p_u$ leads to an additional factor, $C$, in the marginal likelihood, which is just the ratio of normalizing constants for the prior and the likelihood with constraints:
-\begin{align}
-\begin{split}
-	 C=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
-\label{eq:model2post}
-\end{split}
-%	\intertext{with marginal log--likelihood:}
-%\begin{split}
-%\mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N_{ui}}{n_{ui}}+ \log\binom{N_{si}}{n_{si}}+ \log\left(\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\right)+\right.
-\\ \left.
-\log\left(\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)\right) + \right.
-\\ \left.
-\log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right) \right.
-\right.
-\\  \left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)}\left(\mathrm{I_{1-p_{ui}}}(N_{si}-n_{si}+\beta_s,n_{si}+\alpha_s)\right)d p_{ui}\right] \\ -  \left.
-\log\left[\hspace{1em}\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right) \right.
-\right.
-\\  \left.
-\left.\vphantom{\int\left( \frac{1}{\mathrm{B}(\alpha,\beta)}\right)} \left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}\right] \right\}\label{eq:model2MLL}
-%\end{split}
-\end{align}
+Assuming that the missing data $z_i$'s are known, we define the complete log-likelihood as follows,
+\begin{equation}
+l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w)
+\label{eq:cll}
+\end{equation}
+where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.
 
-%The ratio of integrals in \eqref{eq:model2post} accounts for the different normalizing constants due to the constraints $p_{si}>p_{ui}$ on the prior and the posterior distributions.
-We call this the \emph{constrained} model.
+\subsubsection{EM algoritm}
+Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$, and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
+\[
+\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{\tilde{w}\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+(1-\tilde{w})\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
+\]
 
-The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$, is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional numerical integration for calculating the ratio of normalizing constants.
+Replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}, we then optimize the complete-data log-likelihood over $\boldsymbol{\theta}$. Straightforward calculations lead to 
+$\tilde w = \sum_i{\tilde{z_i}}/P$, but unfortunately no closed form solutions exist for the remaining parameters.
+We used numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.
+Starting from some initial values, the EM algorithms iterates between the E and M steps until convergence.
+In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components.
+We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.
 
-%Without constraints, ($p_s\ne p_u$), the ratio of integrals for the normalizing constant in \eqref{eq:model2post} is dropped. We call this the \emph{unconstrained} model.
-
-\subsection{The Mixture Model Framework}
-
-The two models (null and alternative), are represented as two components in our mixture model framework.
-The observed data is thus a mixture of observations drawn from one or the other of these components (i.e., each subject is either a responder, or a non--responder).
-Setting  $\theta_1=\{\alpha_u,\beta_u\},\theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, the likelihood for the mixture model is:
-
-%\begin{align}
-%f(y_i|\alpha_{u},\beta{u},\alpha_{s},\beta_{s}) = (1-\pi)f_1(y_i|\alpha_u,\beta_u) + \pi f_2(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)
-%\end{align}
-
- \begin{align}
-\begin{split}
-%L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi_k|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi_1 Pr(y_i|\alpha_u,\beta_u) +\pi_2 Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) \right] ,\\ \sum\limits_{k=1}\limits^2\pi_k=1\\
-L(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi|\mathbf{y})=\prod\limits_{i=1}\limits^P\left[ \pi f_1(y_i|\alpha_u,\beta_u) +(1-\pi)f_2(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) \right] 
-\end{split}
-\end{align}
-%,\\ \sum\limits_{k=1}\limits^2\pi_k=1
-%Where $\theta_1=\{\alpha_u,\beta_u\}, \theta_2=\{\alpha_u,\beta_u,\alpha_s,\beta_s\}$, $\pi_1$ is the fraction of observations exhibiting no response to stimulation, $\pi_2$ the fraction of observations exhibiting a response to stimulation, and $f_1 = Pr(y_i|\alpha_u,\beta_u), f_2 = Pr(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s)$ from \eqref{eq:model1post} and \eqref{eq:model2post}, above.
-The terms $f_1$,$f_2$ are the marginal likelihoods of the null and alternative models, respectively, $\pi$ is the proportion of responders.
-
-The component membership of each subject is unobserved and  treated as missing data. They are modelled as random variables  $\mathbf{z}_i = \left\{z_{i1},(1-z_{i1})\right\}$, where:
-
-$$
-z_{ik} = \left\{ \begin{array}{rl}
-1 &\mbox{ if observation $i$ is from the $k$'th model (component)} \\
-0&\mbox{ otherwise}
-\end{array} \right.
-$$
-
-Each $\mathbf{z}_{i}$ follows an independent multinomial distribution with one trial and  parameters $\boldsymbol{\pi}=\left\{\pi_1,1-\pi_1\right\}$.
-They are interpreted as the posterior probabilities of response.
-Given the $z_i$'s the complete data likelihood is written as:
-
-\begin{align}
-\begin{split}
-\mathbf{L_c}(\theta_k,\pi_k|\mathbf{y},\mathbf{z}) = \prod\limits_{i=1}\limits^P\prod\limits_{k=1}\limits^2\left[\pi_k f_k(y_i|\theta_k)\right]^{z_{ik}}\label{eq:cll}
-%\mathcal{L}_c(\alpha_s,\beta_s,\alpha_u,\beta_u,\pi|\mathbf{y},\mathbf{z})=\sum\limits_{i=1}\limits^P\sum\limits_{k=1}\limits^2{z_{ik}\left[ \log\pi_k+\log f_k(y_i|\theta_k)\right]}\label{eq:CDLL}
-\end{split}
-\end{align}
-
-\subsection{Parameter estimation}
-In the from of~\eqref{eq:cll}, we use the expectation--maximization (EM) algorithm~\cite{Dempster:1977ul} to fit the model.
-
-\subsubsection{E--step}
-Given the model parameters $\boldsymbol{\Psi}=\left\{\alpha_u,\beta_u,\alpha_s,\beta_s,\pi_k\right\}$, and the data $\mathbf{y}$, we estimate the unobserved component memberships, $\mathbf{Z_i}$  by computing the conditional expectation of the $\mathbf{Z_i}$'s,  $\mathbb{E}\boldsymbol{_\Psi}(\mathbf{Z_i}|\mathbf{y}_i)$:
-\begin{align}
-\tilde z_{ik} &= \frac{\pi_k f_k(\mathbf{y}_i|\theta_k)}{\sum\limits_{k=1}\limits^{2}\pi_kf_k(\mathbf{y}_i|\theta_k)}
-\end{align}
-
-\subsubsection{M--step}
-Given the $\tilde{z}_{ik}$, we update the estimates of the model parameters to maximize the conditional expectation of the complete--data log--likelihood. The mixing proportions are given by:
-\begin{align}
-\hat\pi_k = \frac{ \sum_i^P \tilde z_{ik}}{n}
-\end{align}
-
-There is no closed form for the model hyper--parameter estimates, $\alpha_u,\beta_u,\alpha_s,\beta_s$.
-They are estimated numerically using R's \textit{optim} function.
-To simplify the optimization, they are re--parameterized as $\mu_u=\frac{\alpha_u}{\alpha_u+\beta_u}$ and $S=\alpha_u+\beta_u$ (likewise for the $\alpha_s,\beta_s$), corresponding to the mean and sample size of the prior distributions.
-
-\subsubsection{Initialization}
-We initialize the $z_{ik}$'s using Fisher's exact test to assign each observation to either the null or alternative model components.
-We then use the $\hat{z}_i$'s to initialize the hyper--parameters to their method--of--moments estimates:
-\begin{align}
-\hat{\alpha} = \hat{\mu}\left(\frac{\hat{\mu}(1-\hat{\mu})}{\hat{\sigma}^2} -1\right)\\
-\hat{\beta} =  (1-\hat{\mu}\left(\frac{\hat{\mu}(1-\hat{\mu})}{\hat{\sigma}^2} -1\right)
-\end{align}
-
-Where $\hat{\mu}$ and $\hat{\sigma}^2$ are the sample mean and sample variance estimates, given the $z_{ik}$'s.
+\subsubsection{MCMC algorithm}
+Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand}.
+All updates were done via Metropolis-Hastings sampling except for the $z_i$'s that were done via Gibbs samplings.
+All full conditionals are given in supplementary material.
+We used the method of Raftery and Lewis \citep{Raftery1,Raftery2} to determine the number of iterations, based on a short pilot run of the sampler.
+For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
+Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.
 
@@
 
-\section{Differential expression across combinatorial marker combinations}
+\section{Differential expression across marker combinations}
 \subsection{Model}
-%\subsection{Generalization to Multiple Cytokines and Polyfunctionality with the Multinomial Dirichlet}
-The model can be generalized to handle multiple cytokines in a single stimulation, in order to assess polyfunctional cytokine responses of T--cells.
-Without loss of generality, for two markers we use the Multinomial--Dirichlet family of distributions to model counts of events in two \textit{different} 2x2 contingency tables.
-Here we consider only the \emph{unconstrained} case ($p_s\ne p_u$).
-The observed data can be represented as shown in Table~\ref{tab:multdir}.
+Our beta-binomial model described in section \ref{sec:} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
+As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{sk}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{uk}=\{n_{uik}: k=1,\dots,2^K\}$.
+Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
+
 \begin{table}[h]
@@
 \parbox{0.8\linewidth}{
-\caption{Contingency tables for counts of cells expressing two cytokines between stimulated and unstimulated conditions.
-$n_{\{s,u\}j}$ denotes observed counts for stimulated or unstimulated table cell $j$, and individual $i$  }\label{tab:multdir}
+\caption{Contingency tables for counts of cells expressing two markers between stimulated and unstimulated conditions.
+$n_{\{s,u\}k}$ denotes observed counts for stimulated or unstimulated table cell $k$, and individual $i$  }\label{tab:multdir}
 \begin{minipage}[b]{0.5\linewidth}
@@
 \multicolumn{1}{l}{}&
-\multicolumn{2}{c}{Cytokine A}\\
+\multicolumn{2}{c}{Marker A}\\
  & Negative & Positive \\ 
- \multicolumn{1}{l}{Cytokine B}&&\\
+ \multicolumn{1}{l}{Marker B}&&\\
   \hline
@@
 
-The vector of observed counts for individual $i$ in the stimulated or unstimulated sample is denoted: $\bar{n}_{\{s,u\}i} = \{n_{\{s,u\}ij}\} ; j\in\{1\ldots 4\}$, and $j$ indexes the cells of the appropriate contingency table shown in~Table~\ref{tab:multdir}.
-The counts are modelled as draws from different multinomial distributions:
+In this context, for a given individual,  the beta distribution is replaced by a multinomial distribution, as follows,
 
+\[
+ (\mathbf{n}_{ui}|\mathbf{p}_{ui},) \sim \mathcal{M}(N_{ui},\mathbf{p}_{ui}) \quad\mathrm{and}\quad (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(N_{si},\mathbf{p}_{si})
+\]
+%; \text{if $i$ is a non--responder}\\
+% &(\mathbf{n}_{ui}|\mathbf{p}_{ui}) \sim \mathcal{M}(\mathbf{p}_{ui},N_{ui}); (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(\mathbf{p}_{si},N_{si});\text{if $i$ is a responder}
+%\end{align}
+where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown vectors of proportions of cells. 
+As stated, the number of parameters to be estimated is $2(2^K-1)$, which grows exponentially with the number of makers $K$.
+In addition, we now have many possible patterns of expression across the different marker combinations.
+In fact, if we assume that each proportion can be equal (\textit{i.e.} $p_u=p_s$) or different (\textit{i.e.} $p_u\ne p_s$) the number of possible patterns is $2^{2^{K}}$, which grows super-exponentionally with $K$.
+For example, if the data considered have two markers resulting in four possible combinations, we have 16 possible patterns of expression, as follows.
+\begin{align*}
+p_{s1}&= p_{u1},\ p_{s2}= p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 0)}\\
+p_{s1}&\ne p_{u1},\ p_{s2}= p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 1)}\\
+p_{s1}&= p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 1)}\\
+p_{s1}&= p_{u1},\ p_{s2}= p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 1)}\\
+p_{s1}&= p_{u1},\ p_{s2}= p_{u2},\ p_{s3}= p_{u3},\ p_{s4}\ne p_{u4} \quad \text{(DE of degree 1)}\\
+p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 2)}\\
+p_{s1}&\ne p_{u1},\ p_{s2}= p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 2)}\\
+&\vdots\\
+p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 3)}\\
+p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}= p_{u3},\ p_{s4}\ne p_{u4} \quad \text{(DE of degree 3)}\\
+&\vdots\\
+p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}\ne p_{u4} \quad \text{(DE of degree 4)}.
+\end{align*}
+This \textit{combinatorial explosion} of possible patterns renders the model selection and parameter estimation difficult as one would need to explore $2^{2^K}$ possible models. 
+In practice, one could reduce the total number of patterns to a more manageable level, by eliminating marker combinations that known not to be co-expressed.
+An other possibility would be to only retain the null pattern
+and the pattern of highest degree (\textit{i.e.} all proportions are different), or retain the null pattern and all patterns of degree 1 (\textit{i.e.} one proportion is different).
+In our illustration here, we have opted for the former and thus have two patterns of expression.
+
+\subsection{Prior}
+As with the one-dimensional case, we share information across subjects using an exchangeable prior on the unknown proportions. We use conjugate Dirichlet priors, as follows,
 \begin{align}
- &(\bar{n}_{ui}|p_{ui},) \sim \mathcal{M}(\bar{p}_{ui},N_{ui}); (\bar{n}_{si}|p_{si}) \sim \mathcal{M}(\bar{p}_{ui},N_{si}); \text{if $i$ is a non--responder}\\
- &(\bar{n}_{ui}|p_{ui}) \sim \mathcal{M}(\bar{p}_{ui},N_{ui}); (\bar{n}_{si}|p_{si}) \sim \mathcal{M}(\bar{p}_{si},N_{si});\text{if $i$ is a responder}
+\mathbf{p}_{si} \sim \mathrm{Dir}(\mathbf{\alpha}_s) ; \mathbf{p}_{ui} \sim \mathrm{Dir}(\mathbf{\alpha}_u)
 \end{align}
-where the $N_{\{s,u\}i}=\sum\limits_{j=1}\limits^{J} n_{\{s,u\}ij}$.
-\subsection{Prior}
-As with the one--dimensional case, we share information across subjects. We put Dirichlet priors on the proportions.
-\begin{align}
-\bar{p}_{si} \sim \mathrm{Dir}(\bar{\alpha}_s) ; \bar{p}_{ui} \sim \mathrm{Dir}(\bar{\alpha}_u)
-\end{align}
-
+where $\mathbf{p}_{si}$ would satisfy the constrain given by the expression pattern. That is under the null pattern we have $\mathbf{p}_{si}=\mathbf{p}_{ui}$
 For the null component, the marginal likelihood of the resulting multinomial--dirichlet is given by: 

### 2012-05-04 MIMOSA.tex
Cleaned up the methods section

@@
 Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers).
-Recently, DVS Sciences Inc.
-introduced a new cytometry technology named cytometry Time-Of-Flight (CyTOF) that uses rare isotopes instead of fluorochromes (as in standard FCM), as antibody tags to label single cells.
-CyTOF drastically extends the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell.
+More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell.
 Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize cell poly-functionality.
@@
 
-Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}.
+Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}.
 These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
@@
 
-\textit{Flow cytometry}: This ICS data is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
-The statistical analysis of responder and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
+\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
+The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
 The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.
 
-\textit{Fluidigm single--cell gene expression}: This is a single--cell gene expression data set of flow--sorted CD8 T--cells from sixteen individuals.
-The T--cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single--cell level and compared to paired, unstimulated controls.
+\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8 T-cells from sixteen individuals.
+The T-cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single-cell level and compared to paired, unstimulated controls.
+\footnote{Add a bit more info here, how many genes, tetramer sorted, etc}.
 
@@
 We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples.
-We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individuals $i$ in each sample.
-For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{sk}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{uk}=\{n_{uik}: k=1,\dots,2^K\}$.
+We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individual $i$ in each sample.
+For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
 Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
@@
 \section{Differential expression with one marker}
-In practice, stimulation studies like the ones presented here, are often analyzed one marker at a times to avoid being under powered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
-As a consequence, we first consider the one marker case, \textit{i.e.} $K=1$.
+Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
+As a consequence, we first consider the one marker case where cell counts are marginalized and each marker is analyzed separately and $K=1$.
 In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
@@
 Individuals that show differential expression for a given marker will be called responders for that marker.
-In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described below.
+In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.
 
-\subsection{Beta--Binomial Model for One Marker}
+\subsection{Beta-Binomial Model for One Marker}
 For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
-\begin{align}
-(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non-responder}\\
-(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
- \end{align}
- where $p_{si}$, $p_{ui}$ are the unobserved proportions for the stimulated and un-stimulated paired samples.
-Under the null hypothesis, \textit{i.e.}~\eqref{eq:null}, there is no difference between the stimulation and control samples, and the proportions are equal.
-Under the alternative model given by \eqref{eq:alternate}, there is a difference in proportions between the two samples.
+
+\begin{equation}
+(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
+\end{equation}
+where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
+\begin{equation}
+{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:beta_prior}
+\end{equation}
+Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal.
+Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.
 
 \subsection{Priors}
-Our model shares information across all individuals using an exchangeable Beta priors on the unknown proportions, as follows, 
+Our model shares information across all individuals using exchangeable Beta priors on the unknown proportions, as follows, 
  \begin{align}
@@
 where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.
-$p_{si}\ne p_{ui}$, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals.
-Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present for the alternative model.
-Under the null model, the stimulated condition should behave just like the unstimulated one, and thus the stimulated response rates (\textit{i.e.} $p_{si}$'s) should follow the same prior distribution.
-Finally, we assume that the $z_i$'s are independent and identically distributed as Bernoulli with probability $w$, where $w$ represents the proportion of responders.
-It follows that marginally, i.e.
-after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$.
-Treating the $z_i$'s as missing data, the unknown parameter vector $\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can easily be estimated via an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}.
-We will also explore a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ wis assumed to be drawn from a uniform distribution between 0 and 1.
+${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals.
+Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model.
+Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the proportion of responders.
+It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}.
+As an alternative, we will also explore a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
 In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{sec:estimation}.
@@
 \label{sec:estimation}
-Whether we use an EM or an MCMC algorithm for parameter estimation, we make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p$'s for the null and alternative models, to simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Appendix A), and are given by,
+Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, to simplify our calculations.
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Supplementary material), and are given by,
  \begin{align}
   	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
-	&=\prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
+	&=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
  \end{align} 
@@
 \end{align}
-
-Assuming that the missing data $z_i$'s are known, we define the complete log-likelihood as follows,
+Assuming that the missing data, the $z_i$'s, are known, we define the complete log-likelihood as follows,
 \begin{equation}
@@
 \subsubsection{EM algoritm}
-Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$, and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
+Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
 \[
-\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{\tilde{w}\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+(1-\tilde{w})\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
+\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
 \] 
-
-Replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}, we then optimize the complete-data log-likelihood over $\boldsymbol{\theta}$. Straightforward calculations lead to 
-$\tilde w = \sum_i{\tilde{z_i}}/P$, but unfortunately no closed form solutions exist for the remaining parameters.
-We used numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.
+The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to 
+$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters.
+We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.
 Starting from some initial values, the EM algorithms iterates between the E and M steps until convergence.
@@
 Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand}.
-All updates were done via Metropolis-Hastings sampling except for the $z_i$'s that were done via Gibbs samplings.
-All full conditionals are given in supplementary material.
+All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
+Details about the algorithms are given in Supplementary material.
 We used the method of Raftery and Lewis \citep{Raftery1,Raftery2} to determine the number of iterations, based on a short pilot run of the sampler.
@@
 \section{Differential expression across marker combinations}
+Our beta-binomial model described in section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
+As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+%Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
 \subsection{Model}
-Our beta-binomial model described in section \ref{sec:} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
-As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{sk}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{uk}=\{n_{uik}: k=1,\dots,2^K\}$.
-Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
 
-\begin{table}[h]
-\centering
-\parbox{0.8\linewidth}{
-\caption{Contingency tables for counts of cells expressing two markers between stimulated and unstimulated conditions.
-$n_{\{s,u\}k}$ denotes observed counts for stimulated or unstimulated table cell $k$, and individual $i$  }\label{tab:multdir}
-\begin{minipage}[b]{0.5\linewidth}
-\centering
-\begin{tabular}{rrr}
-\multicolumn{3}{c}{Stimulated}\\
-  \hline
-\multicolumn{1}{l}{}&
-\multicolumn{2}{c}{Marker A}\\
- & Negative & Positive \\ 
- \multicolumn{1}{l}{Marker B}&&\\
-  \hline
-Negative &   $n_{si1}$ &   $n_{si2}$ \\ 
-Positive &   $n_{si3}$ &   $n_{si4}$ \\ 
-   \hline
-\end{tabular}
-\end{minipage}
-\begin{minipage}[b]{0.5\linewidth}
-\centering
-\begin{tabular}{rrr}
-\multicolumn{3}{c}{Unstimulated}\\
-  \hline
-\multicolumn{1}{l}{}&
-\multicolumn{2}{c}{Cytokine A}\\
- & Negative & Positive \\ 
- \multicolumn{1}{l}{Cytokine B}&&\\
-  \hline
-Negative &   $n_{ui1}$ &   $n_{ui2}$ \\ 
-Positive &   $n_{ui3}$ &   $n_{ui4}$ \\ 
-   \hline
-\end{tabular}
-\end{minipage}
-}
-\end{table}
-
-In this context, for a given individual,  the beta distribution is replaced by a multinomial distribution, as follows,
-
-\[
- (\mathbf{n}_{ui}|\mathbf{p}_{ui},) \sim \mathcal{M}(N_{ui},\mathbf{p}_{ui}) \quad\mathrm{and}\quad (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(N_{si},\mathbf{p}_{si})
-\]
+In our multivariate model, the beta distribution is replaced by a multinomial distribution, as follows,
+\begin{equation}
+ (\mathbf{n}_{ui}|\mathbf{p}_{ui},) \sim \mathcal{M}(N_{ui},\mathbf{p}_{ui})\quad\text{and}\quad (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(N_{si},\mathbf{p}_{si})\label{eq:mult_likeliehood}
+ \end{equation}
 %; \text{if $i$ is a non--responder}\\
@@
 %\end{align} 
-where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown vectors of proportions of cells. 
-As stated, the number of parameters to be estimated is $2(2^K-1)$, which grows exponentially with the number of makers $K$.
-In addition, we now have many possible patterns of expression across the different marker combinations.
-In fact, if we assume that each proportion can be equal (\textit{i.e.} $p_u=p_s$) or different (\textit{i.e.} $p_u\ne p_s$) the number of possible patterns is $2^{2^{K}}$, which grows super-exponentionally with $K$.
-For example, if the data considered have two markers resulting in four possible combinations, we have 16 possible patterns of expression, as follows.
-\begin{align*}
-p_{s1}&= p_{u1},\ p_{s2}= p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 0)}\\
-p_{s1}&\ne p_{u1},\ p_{s2}= p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 1)}\\
-p_{s1}&= p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 1)}\\
-p_{s1}&= p_{u1},\ p_{s2}= p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 1)}\\
-p_{s1}&= p_{u1},\ p_{s2}= p_{u2},\ p_{s3}= p_{u3},\ p_{s4}\ne p_{u4} \quad \text{(DE of degree 1)}\\
-p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}= p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 2)}\\
-p_{s1}&\ne p_{u1},\ p_{s2}= p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 2)}\\
-&\vdots\\
-p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}= p_{u4} \quad \text{(DE of degree 3)}\\
-p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}= p_{u3},\ p_{s4}\ne p_{u4} \quad \text{(DE of degree 3)}\\
-&\vdots\\
-p_{s1}&\ne p_{u1},\ p_{s2}\ne p_{u2},\ p_{s3}\ne p_{u3},\ p_{s4}\ne p_{u4} \quad \text{(DE of degree 4)}.
-\end{align*}
-This \textit{combinatorial explosion} of possible patterns renders the model selection and parameter estimation difficult as one would need to explore $2^{2^K}$ possible models. 
-In practice, one could reduce the total number of patterns to a more manageable level, by eliminating marker combinations that known not to be co-expressed.
-An other possibility would be to only retain the null pattern
-and the pattern of highest degree (\textit{i.e.} all proportions are different), or retain the null pattern and all patterns of degree 1 (\textit{i.e.} one proportion is different).
-In our illustration here, we have opted for the former and thus have two patterns of expression.
+where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown proportions for the un-stimulated and stimulated samples.
 
 \subsection{Prior}
-As with the one-dimensional case, we share information across subjects using an exchangeable prior on the unknown proportions. We use conjugate Dirichlet priors, as follows,
+As in the one-marker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
 \begin{align}
-\mathbf{p}_{si} \sim \mathrm{Dir}(\mathbf{\alpha}_s) ; \mathbf{p}_{ui} \sim \mathrm{Dir}(\mathbf{\alpha}_u)
+(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u)\\\nonumber
+(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s)\label{eq:dir_prior}
 \end{align}
-where $\mathbf{p}_{si}$ would satisfy the constrain given by the expression pattern. That is under the null pattern we have $\mathbf{p}_{si}=\mathbf{p}_{ui}$
-For the null component, the marginal likelihood of the resulting multinomial--dirichlet is given by: 
-\begin{align}
-\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u) &= \prod_{i=0}^P\frac{ \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\label{NRmd}
-\end{align}
-Where $\mathrm{B_j}$ is the $\mathrm{j}$--dimensional Beta function: $\frac{\prod_{j=1}^J\Gamma(\alpha_j)}{\Gamma(\sum\alpha_j)}$.
-
-Cells can express different combinations of markers.
-Since we now consider multiple markers simultaneously, the different marker combinations can be represented as separate components in the model.
-We examine the different possibilities below.
-
-\subsection{The simplest model: all marker combinations are different}
-The simplest case is where all possible marker combinations are expressed at different levels between the stimulated and unstimulated samples.
-In this case, we require only one additional component representing the alternative model, where $p_{sj} \ne p_{uj}$ for all $j$.
-An observation is assigned either to the null component (if there is no difference in any of the cell proportions), or to the alternative component, implying a difference in all the cell proportions.
-The marginal likelihood is given by:
-\begin{align}
-\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u,\bar{\alpha}_s) &= \prod_{i=0}^P\frac{  \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}) \mathrm{B_j}(\bar{\alpha}_{s}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_s)\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\label{eq:postmult}
-\end{align}
-
-However, it is possible to have a situation where only some combinations of markers exhibit differences between stimulation and control.
-These can be represented explicitly by additional components in the model.
-Without loss of generality, if only some $p_j$ are different between stimulated and unstimulated samples, the appropriate mixture components can be represented by simply substituting the relevant $\alpha_j$ in the calculation of the likelihood eq~\eqref{eq:postmult}.
+where the indicator variable $z_i$ is as defined in \eqref{eq:beta_prior}, i.e.
+$z_i\sim\mathrm{Be}(w)$ where $w$ is the proportion of responders.
+As in the beta-binomial case both an EM and MCMC algorithms can be used for parameter estimation.
+When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}_{\{u,s\}}$ and $w$ as for the beta-binomial model.
 
 \subsection{Parameter estimation}
-The multivariate model is fit using the EM algorithm. We estimate the model hyper--parameters numerically ($\bar{\alpha}_u,\bar{\alpha}_s$) using Newton's method. 
-%The marginal likelihood of the non--responding component, given in eq~\eqref{NRmd} can be rewritten as:
-%\begin{align}
-%\begin{split}
-%l(\bar{\alpha}_u|D)=\sum_i \left[ \log \Gamma (\sum_j \alpha_{uj}) - \log \Gamma (\sum_j \alpha_{uj}+n_{uij}+n_{sij})+ \sum_j (\log \Gamma(\alpha_{uj}+n_{uij}+n_{sij}) \right.\\ \left.
--\log\Gamma(\alpha_{uj}))+ \log\Gamma(N_{si}+1) +\log\Gamma(N_{ui}+1)-\right.\\ \left.\left[\sum_j(\log\Gamma(n_{sij}+1)+\log\Gamma(n_{uij}+1))\right]\right]\label{mdloglike}
-%\end{split}
-%\end{align}
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (See supplementary material).
+For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
+\begin{align}
+\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
+\end{align}
+where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$.
+Similarly the marginal likelihood for the alternative model is given by
+\begin{align}
+\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
+\end{align}
+The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.
+In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2.
+The EM algorithms becomes very dependent on the initial values, and can even fail to converge when good initial values are provided.
+Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large.
+More details about both algorithms are given in Supplementary material.
 
-The gradient of the log of the marginal likelihood in equation~\eqref{NRmd} is:
-\begin{align}
-\frac{\partial l}{ \partial \alpha_{uj}} = \sum_i \Psi(\sum_j \alpha_{uj}) - \Psi (\sum_j \alpha_{uj}+n_{uij}+n_{sij}) + \Psi(\alpha_{uj}+n_{uij}+n_{sij}) - \Psi(\alpha_{uj})\label{mdgrad}
-\end{align}
+\subsection{Parsimonious multivariate model representation}
+In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model.
+The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$.
+As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters.
+Here, we propose alternative model parametrizations that could be used to address this problem by reducing the number of required parameters.
+A simple solution would be to assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
+As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others.
+As a more flexible solution, we propose to define $K$ main effect parameters, denoted $\{\theta_k:k=1,\dots,K\}$, for the $K$ markers and represent co-expression parameters as a function of these main effects.
+So a combination of dregree $l$, with positive markers, $k_1$, $k_2$, \dots, $k_l$ would have a hyperparameter defined as $\tau_l(\theta_{k_1}\theta_{k_2}\cdots\theta_{k_l})^{1/l}$, which is the geometric mean of the corresponding main marker effects scaled by a degree effect, $\tau_l$.
+Note that the scaling degree effect would be the same for any combinations with the same number of positive markers.
+The rationale being that as the degree of polyfunctionality increases, the number of cells might decrease (or increase in some instances) and the corresponding parameters should be adjusted by a scaling factor to allow for such variation.
+The choice of the geometric mean versus an arithmetic mean is to account for the fact that when a particular combination of degree 1 is empty, \textit{i.e.} the corresponding $\theta$ value is close to zero, it is unlikely that other combination involving that maker would be large.
+Our specification would account for this since the product would be near zero if one of the $\theta$ is near zero.
+Finally, the null combination would have a single parameter equal to $\theta_0$.
 
-and the Hessian:
-\begin{align}
-j&\ne k; &\frac{\partial^2 l}{\partial \alpha_{uj}\alpha_{uk}} & = \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})\\
-j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{uj})
-\end{align}
-
-The gradient of the log of the marginal likelihood of the responding component in eq~\ref{eq:postmult} is:
-%\begin{align}
-%\begin{split}
-%\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
-%-\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
-%-\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
-%\end{split}
-%\end{align}
-%with gradient:
-\begin{align}
-\frac{\partial l}{\partial\alpha_{\phi j}} = \sum_i \Psi(\alpha_{\phi j}+n_{\phi ij})-\Psi(\alpha_{\phi j})+\Psi(\sum_j \alpha_{\phi j}) - \Psi(\sum_j \alpha_{\phi j}+n_{\phi ij}): (\phi \in {s,u})
-\end{align}
-and Hessian:
-\begin{align}
-\frac{\partial^2l}{\partial \alpha_{\phi j}^2}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}+n_{\phi ij}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
-\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}&\overset{k\ne j}{=}\sum_i\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
-\end{align}
-
-\subsubsection*{Mixture Model Complexity}
-In the two marker case described here, we may wish to test for any of 8 different possible scenarios where the proportion of events in corresponding cells of the contingency tables are either equal or unequal between stimulated and unstimulated conditions.
-Such a model would have 8 components and 55 parameters.
-However, if we recognize that the models can be nested, i.e., that parameters can be shared across components with similar outcomes, then the number of parameters can be reduced to 19, and further to 15 if we only consider components where any one cell of the tables differs between stimulated and unstimulated conditions.
-This number of parameters required for modelling different numbers of components is shown in Table~\ref{tab:nesting}.
-\begin{table}
-\centering
-\parbox{0.8\linewidth}{\caption{Nesting of models and parameter counts.
-Each row is a model component.
-The three columns correspond to cells two, three, and four of the contingency tables shown in Table~\ref{tab:multdir}.
-An open circle at a position indicates that the component models $p_{sj}=p_{uj}$, and a filled circle indicates that the component models $p_{sj}\ne p_{uj}$.
-The number of additional parameters that need to be estimated by including each additional component in the mixture model is in the fourth column (number of parameters for proportions + number of parameters for component weights).}\label{tab:nesting}
-\centering
-\begin{tabular}{ccccc}
-\multicolumn{3}{c}{Cell of Table}\\
-\hline
-cell 2&cell 3&cell 4&\# of parameters\\
-\hline
-$\circ$&$\circ$&$\circ$&6+1\\
-$\circ$&$\circ$&$\bullet$&2+1\\
-$\circ$&$\bullet$&$\circ$&2+1\\
-$\bullet$&$\circ$&$\circ$&2+1\\
-$\circ$&$\bullet$&$\bullet$&0+1\\
-$\bullet$&$\bullet$&$\circ$&0+1\\
-$\bullet$&$\circ$&$\bullet$&0+1\\
-$\bullet$&$\bullet$&$\bullet$&0\\
-\end{tabular}
-}
-\end{table}
-
-\subsubsection*{Model simplifications for K>2 cytokines}
-In the case of more than two cytokines, there is a combinatorial explosion in the number of parameters of the resulting model.
-We propose several alternative solutions to reducing the number of parameters in general, when the dimensionality of the Multinomial Dirichlet increases.
-If we model $N$ cytokines simultaneously, this results in $2^N$ possible marker combinations (i.e.
-each marker is either expressed, or not).
-We let $K$ be the number of possible cytokine combinations being modelled, with hyperparameter vector $\bar{\alpha}_0=\{\alpha_{00},\ldots,\alpha_{0K-1}\}$, and $\bar{\alpha}_s=\{\alpha_{s0},\ldots,\alpha_{sK-1}\}$ for the unstimulated and stimulated distributions, respectively.
-We formulate several strategies for reducing the number of parameters required to fit MIMOSA to high dimensional data by sharing hyperparameters across some of the $K$ combinations.
-
-\subsubsection*{Fully shared model hyperparameters}
-In the most simplistic case, we set up a two mixture component model (responders and non--responders).
-This formulation would ignore most of the marker combinations and only explicitly test for all combinations equal to the null, or all combinations different from the null.
-
-For ICS data, the first component of the hyper--parameter vector represents cells not expressing any cytokine combination, and is usually much larger than the remaining components.
-As a first approximation, we can set the $K-1$ remaining parameters to be all equal $\{\alpha_{01}=\{\alpha_{01},\ldots,\alpha_{0K-1}\}$ (e.g.
-shared across all possible combinations) and thus all variation across cells expressing different cytokine combinations in the stimulated and unstimulated samples are captured by four parameters.
-As stated above, such a model can only be fit to two mixture components (the null component and the responder component), but requires a total of only five parameters.
-In practice, such a model would likely have a vary high false negative rate, since observations with differences in only a few marker combinations would be absorbed into the null component.
-
-\subsubsection*{Second order hyperparameters as functions of first--order hyperparameters}
-A more realistic scenario would be to define \textit{first--order} hyper--parameters as those elements of the $\alpha$ vectors corresponding to cells expressing individual cytokines (i.e.
-cytokine A only, or cytokine B only).
-Second--order and higher parameters are defined for cells expressing pairs, triplets, and higher order combinations of cytokines (i.e.
-A and B, or B and C, or A,B and C).
-Higher order parameters could be expressed as functions of the first--order parameters (i.e.
-cells expressing A and B can be parameterized as $\alpha_{0ab} = \frac{\alpha_{0a}+\alpha_{0b}}{2}$ or $\alpha_{0ab}=\alpha_{0a}\alpha_{0b}$).
-
-This parameterization can be extended to models with more than two mixture components, allowing us to identify different types of responders.
-For example there may be observations where there are differences in the proportion of cells expressing combination A and B, but also observations where there are differences in the proportion of cells expressing A only, or B only.
-Distinguishing between these possibilities can increase the power of the model.
+Assuming that the $\theta$'s and $\tau$'s are stimulation specific, this parametrization would result in $4K+3$ parameters, and would scale linearly with the number of markers providing a good alternative between the fully parametrize model and the model with constant rate for moderate to large values of $K$.
+In this case, we would recommend the use of an MCMC algorithm.
+The $\theta$'s would be given the same exponential priors than the unconstrained model, \textit{i.e.} exponentional with mean 100.
+As for the $\tau's$ we would use an exponential distribution with mean 1\footnote{check with real data to see if that makes sense}.
 

### 2012-06-18 MIMOSA.tex
Update figures with new data and start rewrite of results
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 Cell counts from the different experimental conditions are modelled independently, while sharing information across responding and non--responding observations through empirical Bayes priors in order to increase the sensitivity and specificity of positivity calls.
-We compare our method against Fisher's exact test and show how it can be extended to model multivariate (multiple markers) cellular responses.
-In simulations and in HIV vaccine trial data we find that our method has higher sensitivity and specificity than Fisher's exact test for positivity calls.
+We compare our method against Fisher's exact test, a likelihood ratio test, and ranked log fold changes, and show how it can be extended to model multivariate (multiple markers) cellular responses.
+In simulations and applied to real data sets, we find that our method has higher sensitivity and specificity than Fisher's exact test and alternative methods.
 % simultaneously, sharing information across observations using empirical Bayes methods.
@@
 \subsubsection{MCMC algorithm}
-Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand}.
+Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand:1996wc}.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
 Details about the algorithms are given in Supplementary material.
-We used the method of Raftery and Lewis \citep{Raftery1,Raftery2} to determine the number of iterations, based on a short pilot run of the sampler.
+We used the method of Raftery and Lewis \citep{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
 For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
@@
 The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells.
+We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.
 
 \subsection{ICS}
-\subsubsection*{MIMOSA Outperforms Fisher's Exact Test in Real--World Vaccine Trial Data from HVTN054}
+\subsubsection*{MIMOSA Outperforms Competing Methods on Vaccine Trial Data from Study HVTN065}
 
-We tested our method on ICS data from HVTN054, a safety and immunogenicity vaccine trial for a replication defective Adenovirus vector HIV vaccine.
-The data set consisted of 48 individuals who received two doses of vaccine or placebo, and ICS time points for these individuals were available at day 0 and day 28 after vaccination.
-Responders for each cytokine and stimulation combination were classified using Fisher's exact test and the constrained MIMOSA model at the 1\% FDR level, then the response rates for \textit{any cytokine} being positive within the control and treatment groups were calculated for both methods.
-Table~\ref{tab:positivityrates} tabulates the response rates for a subset of the ICS data (Gag1, Gag2, Env1, Env2, and Pol2 stimulations, for any cytokine positive, at days 0 and 28, in both CD4 and CD8 T--cell subpopulations).
+We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 98 individuals.
+We examined time points from day 0 and day 182 (the primary endpoint).We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples.
+An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.
 
-% latex table generated in R 2.16.0 by xtable 1.7-0 package
-% Wed Apr  4 09:53:11 2012
-\begin{table}[ht]
-\begin{center}
-\caption{Response rates for Pol2, Gag1, Gag2, Env1, and Env2 stimulated CD4+ and CD8+ T--cells on Day 0 and Day 28.
-C, T1, T2 are the Control, and increasing dose treatment groups, respectively.
-No response is expected on Day 0 in any group, while a response is expected on Day 28 in the treatment groups.}\label{tab:positivityrates}
-\begin{tabular}{rrrrrrrrr}
-  \hline
- &&\multicolumn{3}{c}{Day 0}&\multicolumn{3}{c}{Day 28}\\
-  \cmidrule{3-8}
-&Model & C  & T1 & T2  &   C  & T1  & T2  \\ 
-  \cmidrule{2-8}
-\multirow{2}{*}{CD8+ Pol2}&Beta--binomial & 0.00 & \textbf{0.00} & 0.00 &   0.00 & \textbf{0.89} & 0.57 \\ 
-& Fisher & 0.00 & 0.05 & 0.00 &   0.00 & 0.83 & 0.57 \\ 
-   \hline
-\multirow{2}{*}{CD4+ Pol2}&Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &   0.00 & \textbf{0.89} & \textbf{0.71} \\ 
-&Fisher & 0.00 & 0.05 & 0.00  & 0.00 & 0.72 & 0.50 \\ 
- \hline
-\multirow{2}{*}{CD8+ Gag2}& Beta-binomial & 0.00 & 0.00 & 0.00  & 0.00 & 0.39 & \textbf{0.21} \\ 
-& Fisher & 0.00 & 0.00 & 0.00  & 0.00 & 0.39 & 0.14 \\ 
-   \hline
-\multirow{2}{*}{CD4+ Gag2}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &  0.00 & \textbf{0.44} & \textbf{0.50} \\ 
- & Fisher & 0.00 & 0.05 & 0.00 &  0.00 & 0.33 & 0.43 \\ 
-   \hline
-\multirow{2}{*}{ CD8+ Env1}& Beta-binomial  & 0.00 & 0.00 & 0.00 & 0.29 & \textbf{0.71} & 0.86 \\ 
-   & Fisher & 0.00 & 0.00 & 0.00 &  \textbf{0.00} & 0.65 & 0.86 \\ 
-   \hline
-\multirow{2}{*}{  CD4+ Env1}& Beta-binomial   & 0.00 & 0.00 & 0.00 &  0.14 & \textbf{1.00} & \textbf{0.86} \\ 
- & Fisher & 0.00 & 0.00 & 0.00 &  0.14 & 0.94 & 0.79 \\ 
-   \hline
-\multirow{2}{*}{CD8+ Env2}  & Beta-binomial & 0.00 & 0.05 & 0.08  & 0.29 & 0.83 & \textbf{0.57} \\ 
-  & Fisher & 0.00 & \textbf{0.00} & 0.08 & \textbf{0.14} & 0.83 & 0.43 \\ 
-   \hline
-\multirow{2}{*}{CD4+ Env2}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00 &  0.00 & \textbf{0.78} & 0.64 \\ 
-  & Fisher & 0.00 & 0.05 & 0.00 &  0.00 & 0.72 & 0.64 \\ 
-   \hline
-\multirow{2}{*}{CD8+ Gag1}& Beta-binomial & 0.00 & \textbf{0.00} & 0.00  & 0.00 & \textbf{0.82} & 0.79 \\ 
- & Fisher & 0.00 & 0.05 & 0.00  & 0.00 & 0.76 & 0.79 \\ 
-   \hline
-\multirow{2}{*}{CD4+ Gag1}& Beta-binomial & 0.00 & 0.00 & 0.00  & 0.00 & 0.71 & 0.79 \\ 
- & Fisher & 0.00 & 0.00 & 0.00  & 0.00 & 0.71 & 0.79 \\ 
-   \hline
-
-\end{tabular}
-\end{center}
-\end{table}
-
-On day zero, where no response is expected, response rates from the beta--binomial model were either smaller than (5 cases), or no greater (23 cases) than those from Fisher's exact test, with the exception of CD8+ Env2 stimulated T--cells, where Fisher's exact test (0\% response rate) outperformed the beta--binomial model (5\% response rate).
-On day 28, where a response is expected in the treatment groups, the beta--binomial model had a higher response rate in 12 cases and was no different from Fisher's exact test in the remaining eight cases.
-In the control group, the beta--binomial model was equal to Fisher's exact test in eight cases, and performed slightly worse than Fisher's exact test in two cases (Env2 and Env1 stimulated CD8+ cells).
-%On day zero, we found that the response rates from the MIMOSA model were zero, and equal to response rates computed from Fisher's exact test across all treatment and control groups.
-The same was true for the response rate in the control groups at day 28.
-In the treatment groups, at day 28, the response rates for the MIMOSA model were  generally higher or equal to those for Fisher's exact test, demonstrating the increased sensitivity and specificity of our method.
-
-%The MIMOSA framework generates predictions for individual samples.
-To better understand the differences between the predictions from the MIMOSA model and Fisher's exact test, we examined the model fit from Gag2 stimulated CD4+ T--cells at day 28 producing IFNg (Figure~\ref{fig:icsdata}).
-We see that two individuals are predicted as responders by the MIMOSA model which are not detected by Fisher's exact test.
-Both individuals have a larger proportion of IFNg producing antigen specific T--cells in antigen stimulated than unstimulated samples, as measure by the MAP or ML estimates of the proportions (Figure~\ref{fig:icsdata}).
-
-\subsection{Study 073}
-% latex table generated in R 2.16.0 by xtable 1.7-0 package
-% Wed Apr 11 15:01:22 2012
-\begin{table}[ht]
-\begin{center}
-\caption{Cytokine-specific response rates for ENV1 stimulated CD4+ T--cells on visits 2 through 14.1. No response is expected on visit 2 (day 0) for any cytokine.}\label{tab:positivityrates2}
-\begin{tabular}{rrrrrrrrrrrr}
-  \hline
-&Visit & 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 & 14.1 \\ 
-  \hline
-\multirow{2}{*}{IL2}&BB & 0.00 & 0.17 & 0.25 & 0.00 & 0.55 & 0.11 & 0.32 & 0.17 & 0.11 & 0.53 \\ 
- &Fisher & 0.01 & 0.17 & 0.25 & 0.00 & 0.55 & 0.15 & 0.30 & 0.20 & 0.13 & 0.47 \\ 
-   \hline
-  \hline
-&Visit & 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 \\ 
-  \hline
-\multirow{2}{*}{TNF}&BB & 0.05 & 0.00 & 0.21 & 0.00 & 0.57 & 0.12 & 0.39 & 0.44 & 0.25 \\ 
-  &Fisher & 0.07 & 0.00 & 0.26 & 0.00 & 0.55 & 0.14 & 0.38 & 0.42 & 0.25 \\ 
-   \hline
-  \hline
- &Visit& 2 & 5 & 6 & 8 & 9 & 11 & 12 & 13 & 14 & 14.1 \\ 
-  \hline
-\multirow{2}{*}{IFN}&BB & 0.00 & 0.43 & 0.31 & 0.29 & 0.63 & 0.28 & 0.39 & 0.34 & 0.17 & 0.56 \\ 
-  &Fisher & 0.00 & 0.29 & 0.29 & 0.43 & 0.60 & 0.26 & 0.37 & 0.32 & 0.16 & 0.44 \\ 
-   \hline
-\end{tabular}
-\end{center}
-\end{table}
 \subsection{Fluidigm}
+We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples.
+The unconstrained MIMOSA model was fit separately to for each stimulation, and the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions were calculated and heat maps generated (Figure~\ref{fit:fluidigm} A-C).
 
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
-Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
-The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test.
+Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from real data, we simulated data from the constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
+The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change.
 (Figure~\ref{fig:simulations}).
-The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions.
+The nominal vs observed false discovery rate was also examined to assess the model fit.
 
-We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}).
-Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).
-
-Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly,  we examined whether the unconstrained model could be used to accurately fit data generated from the constrained model.
-We found that the unconstrained model performed as well as the constrained model when fitting data generated from the constrained model (Figure~\ref{fig:simulations}).
+For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, B).
+The false discovery rate for MIMOSA more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations} C, D ).
+%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
 
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN054 data set (Figure~\ref{fig:simulations_trunc}).
+The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained model fit to constrained data.
@@
 
-\subsection{Parsimonious multivariate model representation}
-In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model.
+%\subsection{Parsimonious multivariate model representation}
+%In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model.
 The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$.
@@
 Finally, the null combination would have a single parameter equal to $\theta_0$.
-
-Assuming that the $\theta$'s and $\tau$'s are stimulation specific, this parametrization would result in $4K+3$ parameters, and would scale linearly with the number of markers providing a good alternative between the fully parametrize model and the model with constant rate for moderate to large values of $K$.
+%
+%Assuming that the $\theta$'s and $\tau$'s are stimulation specific, this parametrization would result in $4K+3$ parameters, and would scale linearly with the number of markers providing a good alternative between the fully parametrize model and the model with constant rate for moderate to large values of $K$.
 In this case, we would recommend the use of an MCMC algorithm.
@@
 As for the $\tau's$ we would use an exponential distribution with mean 1\footnote{check with real data to see if that makes sense}.
-
-\subsection{ICS example with two markers}
-We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
+%
+%\subsection{ICS example with two markers}
+%We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
 As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
 We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
-
-The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations.
+%
+%The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations.
 We compared multivariate MIMOSA against Fisher's exact test.
@@
 Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.
-
+%
+%
 \section{Discussion}
@@
 \todo[inline]{Fix references to non--existent figure}
-We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
+%We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
 For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test.
@@
 %
-%\begin{table}[ht]
-%\begin{center}
-%\begin{minipage}{0.9\linewidth}
-%\caption{Confusion matrices comparing responders and non--responders identified by the MIMOSA models and Fisher's exact test.
-Top left: Fisher's exact test vs the multivariate MIMOSA model for any combination of cytokines.Top right: Fisher's exact test vs MIMOSA for the univariate combination of IL4 and not TNFa.
-Bottom left: Fisher's exact test vs univariate MIMOSA for TNFa and not IL4.
-Bottom right: Fisher's exact test vs univariate MIMOSA for IL4 and TNFa.
-False discovery rate thresholds were set at 1\% for each cytokine combination (except in the multivariate model, where it was 1\% for each observation).
-}\label{tab:MDFisher}
-%\end{minipage}
-%\begin{minipage}{0.49\linewidth}
-%\center
-%\begin{tabular}{rrr}
-%  \hline
-%  \multicolumn{3}{c}{Any Combination}\\
-%  \hline
-%&\multicolumn{2}{c}{MIMOSA}\\
-%\cmidrule(r){2-3}
-% & NR & R \\ 
-%\cmidrule(r){1-1}
-%Fisher's Exact Test&&\\
-%  \hline
-%NR &  32 &   2 \\ 
-%  R &   0 &   4 \\ 
-%   \hline
-%\end{tabular}
-%\end{minipage}
-%\begin{minipage}{0.49\linewidth}
-%\center
-%\begin{tabular}{rrr}
-%\hline
-%\multicolumn{3}{c}{IL4 and not TNFa}\\
-%  \hline
-%&\multicolumn{2}{c}{MIMOSA}\\
-%\cmidrule(r){2-3}
-% & NR & R \\ 
-%\cmidrule(r){1-1}
-%Fisher's Exact Test&&\\
-%  \hline
-%NR &  38 &   0 \\ 
-%  R &   0 &   0 \\ 
-%   \hline
-%\end{tabular}
-%\end{minipage}
-%
-%\begin{minipage}{0.49\linewidth}
-%\center
-%\begin{tabular}{rrr}
-%  \hline
-%  \multicolumn{3}{c}{TNFa and not IL4}\\
-%  \hline
-%&\multicolumn{2}{c}{MIMOSA}\\
-%\cmidrule(r){2-3}
-% & NR & R \\ 
-%\cmidrule(r){1-1}
-%Fisher's Exact Test&&\\
-%  \hline
-%NR &  36 &   2 \\ 
-%  R &   0 &   0 \\ 
-%   \hline
-%\end{tabular}
-%\end{minipage}
-%\begin{minipage}{0.49\linewidth}
-%\center
-%\begin{tabular}{rrr}
-%  \hline
-%  \multicolumn{3}{c}{IL4 and TNFa}\\
-%  \hline
-%&\multicolumn{2}{c}{MIMOSA}\\
-%\cmidrule(r){2-3}
-% & NR & R \\ 
-%\cmidrule(r){1-1}
-%Fisher's Exact Test&&\\
-%  \hline
-%NR &  33 &   1 \\ 
-%  R &   0 &   4 \\ 
-%   \hline
-%\end{tabular}
-%\end{minipage}
-%\end{center}
-%\end{table}
-%
-%
+
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
    \centering
-   \includegraphics[width=4in]{Figures/simulations_all} 
-   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data.
+%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
+%   \includegraphics[width=.75\columnwidth]{Figures/12} 
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
+    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/12}};
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+    \node at (0,-0.25) [font=\tiny\sffamily] {C} ;
+    \node at (4.6,-0.25) [font=\tiny\sffamily] {D} ;
+\end{tikzpicture}
+   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065.
+Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg.
+ROC and FDR plots of other cytokine combinations can be found in the Supplementary Information.}
+\label{fig:HVTN065}
+\end{figure}
+
+%   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data.
 Data were simulated from a constrained model with hyper--parameters estimated from a real data set of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial.
@@
 The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta--binomial model are shown in the fourth panel.}
-   \label{fig:simulations}
+%   \label{fig:simulations}
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+\node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot3.pdf}};
+\node at (0,3) [font=\tiny\sffamily] {A} ;
+\node at (4.5,3) [font=\tiny\sffamily] {B} ;
+\node at (8.95,3) [font=\tiny\sffamily] {C} ;
+\end{tikzpicture}
+\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
+The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a significant decrease, green a significant increase, relative to the control).
+Columns and rows are clustered based on these signed posterior probabilities.
+B) The $log_2$ ratio of the proportion of cells expressing a gene in the stimulated vs.
+control samples.
+Rows and columns are ordered as in A) for comparison.
+C) The difference in the proportion of cells expressing each gene in the stimulated vs.
+control samples.
+Ordering of the rows and columns is preserved as in A).
+The traces show the deviations of each cell from zero.}
+\label{fig:fluidigm}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
+\node[anchor=south west, inner sep=0] at (6.5,0) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
+\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+\node at (0,4.5) [font=\tiny\sffamily] {A} ;
+\node at (6.25,4.5) [font=\tiny\sffamily] {B} ;
+\node at (0,0) [font=\tiny\sffamily] {C} ;
+\node at (6.25,0) [font=\tiny\sffamily] {D} ;
+\end{tikzpicture}
+\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
+The posterior difference in proportions between stimulated and unstimulated samples fitting the C) marginalized counts D) multivariate combinations.
+No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.}
+\label{fig:polyfunctionality}
 \end{figure}
@@
    \centering
-   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
-   \caption{Sensitivity to departures from model assumptions.
-We generated data from variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions.
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
+    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
+   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
+   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
+   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
+   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
+    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
+    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
+    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
+\end{tikzpicture}
+   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row).
+Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data.
+A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change.
+B) Boxplots for the unconstrained model.
+C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test.
+C) ROC curves for a representative simulated data set (N=100,000 counts).
+D) AUC  boxplots, E) observed vs.
+nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data.
+}\label{fig:simulations}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/Multivariate_Sims_AUCBOX.pdf}};
+    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
+   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
+    \node at (0,4.2) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,4.2) [font=\tiny\sffamily] {B} ;
+    \node at (9,4.2) [font=\tiny\sffamily] {C} ;
+\end{tikzpicture}
+   \caption{Multivariate simulations from a two--sided model.
+Ten, five-dimensional data sets were simulated from a two--sided model with an effect size of 2.5e-3 and -2.5e-3 in two of the five dimensions.
+Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value, combining Fisher's exact test run marginally on each of the five dimensions.
+A) Boxplots of AUCs (area under the curve) for the ten simulations.
+B) Observed and nominal false discovery rate for each method.
+C)ROC curves for the competing methods.}
+   \label{fig:mvsimulations}
+\end{figure}
+
+%
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
+%   \caption{Sensitivity to departures from model assumptions.
+We generated data from a variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions.
 The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity).
@@
 The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
-   \label{fig:simulations_trunc}
-\end{figure}
+%   \label{fig:simulations_trunc}
+%\end{figure}
 
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=6in]{Figures/FluidigmMIMOSA.pdf} 
-  % \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28.
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=6in]{Figures/FluidigmMIMOSA.pdf} 
+%  % \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28.
 A significant difference between control and stimulated samples is called at the 1\% FDR threshold (red) for the Beta--binomial model, and at 1\% for FDR adjusted p--values from Fisher's exact test (green).
@@
 The effect of shrinking the MAP estimates towards zero can be seen in the volcano plot.}
-%   \label{fig:icsdata}
-\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
+%%   \label{fig:icsdata}
+%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
 The posterior probability of response times the sign of the change in expression is shown in the first panel (red indicates a significant decrease, green a significant increase, relative to the control).
@@
 The traces show the deviations of each cell from zero.}
-\label{fig:fluidigm}
-\end{figure}
+%\label{fig:fluidigm}
+%\end{figure}
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=6in]{Figures/Polyfunctionality_ggplot2} 
-   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--2 peptide stimulated CD4+ T--cells.}
-   \label{fig:polyfunctionality}
-\end{figure}
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=6in]{Figures/Polyfunctionality_Env1} 
-   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--1 peptide stimulated CD4+ T--cells.}
-   \label{fig:polyfunctionalityenv1}
-\end{figure}
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=6in]{Figures/Polyfunctionality_ggplot2} 
+%   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--2 peptide stimulated CD4+ T--cells.}
+%   \label{fig:polyfunctionality}
+%\end{figure}
+%
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=6in]{Figures/Polyfunctionality_Env1} 
+%   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--1 peptide stimulated CD4+ T--cells.}
+%   \label{fig:polyfunctionalityenv1}
+%\end{figure}
+%
 
@@
 \section*{Supplementary Information}
-\section{Vaccine Trial ICS Dataset Description}
+\section{HVTN065 Vaccine Trial ICS Dataset Description}
 \label{supp:data}
-HVTN054 is a phase 1 (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
+HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B).
+CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay.
+Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays.
+Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools.
+Results of the trial have been published~\citep{Goonetilleke:2006jk}.
+
+%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
 The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.
@@
 
-\section{Data Import, Preprocessing, and Gating}
-The ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
+%\section{Data Import, Preprocessing, and Gating}
+%The ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
 Data were preprocessed using the flowJo--defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T--cells for each sample and stimulation~\cite{Hahne:2009vv}.
 
-\section{Statistical Analysis in the Published Trial}
-\label{supp:statpublished}
-The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
+%\section{Statistical Analysis in the Published Trial}
+%\label{supp:statpublished}
+%The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
 In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine--positive T-cells in the antigen--stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine--positive T-cells for the negative control (unstimulated) sample from the same individual.

### 2012-06-28 MIMOSA.tex
Results and discussion
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \author[2]{SC De Rosa}
+\author[3]{Mario Roederer}
 \author[1]{Raphael Gottardo}
@@
 \affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
-\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, Wa}
+\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
+\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
+
 \date{\today}
@@
 
-\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
+\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
 The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
@@
 
-We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 98 individuals.
-We examined time points from day 0 and day 182 (the primary endpoint).We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples.
+We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 120 individuals (98 vaccines and 20 placebo recipients).
+We analyzed the vaccinees (n=98) at the day 0 and day 182 (the primary endpoint) time points as a pooled data set.
+Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity).
+We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples.
 An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.
 
+The MMOSA model had higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e.
+ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs.
+0.923 for Fisher's exact test vs.
+0.827 for LRT vs.
+0.775 for ranked fold change).
+When the magnitude of response was larger (GAG-1-PTEG stimulation), MIMOSA performed as well as Fisher's exact test, and better than other competing methods (Figure~\ref{fig:HVTN065} C) (AUC=0.982 for MIMOSA vs.
+0.978 for Fisher's exact test vs.
+0.877 for LRT vs.
+0.784 for ranked fold change).
+Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D).
+These results are consistent for other cytokines (see supplementary material).
+
 \subsection{Fluidigm}
+\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single--Cell Data}
 We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples.
-The unconstrained MIMOSA model was fit separately to for each stimulation, and the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions were calculated and heat maps generated (Figure~\ref{fit:fluidigm} A-C).
+The unconstrained MIMOSA model was fit to each stimulation.
+From the model we calculated the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions.
+The data are presented in (Figure~\ref{fig:fluidigm} A-C).
+Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--subject variability (Figure~\ref{fig:fluidigm}).
+These patterns are preserved in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), as well as in the posterior differences and log--odds (Figure~\ref{fig:fluidigm} B,C).
 
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
-Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from real data, we simulated data from the constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$.
-The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change.
-(Figure~\ref{fig:simulations}).
-The nominal vs observed false discovery rate was also examined to assess the model fit.
+Using hyper parameters estimated from the constrained model fit to data from Gag1--stimulated, CD4--positive, IL2--expressing T--cells from day 28 post--vaccination of the HVTN054 trial~\cite{Peiperl:2010ej}.
+We simulated data from this constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, with ten independent realizations of data for each $N$.
+Constrained MIMOSA was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change.
+This procedure was repeated for the unconstrained model fit to unconstrained data (Figure~\ref{fig:simulations} A-D).
+The nominal vs observed false discovery rate was also examined to assess the model fit (Figure~\ref{fig:simulations} E-F).
 
-For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, B).
-The false discovery rate for MIMOSA more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations} C, D ).
+For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, D).
+Additionally, the false discovery rate for MIMOSA more closely reflected the nominal false discovery rate compared to Fisher's exact test (Figure~\ref{fig:simulations} E, F).
 %Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}).
-Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained model fit to constrained data.
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}).
+Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
 
@@
 %
+
+\subsubsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
+We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at coexpression of pairs of genes in single cells across the four stimulations.
+In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
+Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation.
+The typical approach to analyzing polyfunctional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.
+There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C).
+In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D).
+Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.
+
+Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we performed simulations (five--dimensional data) to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p--values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
+These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data.
+
 \section{Discussion}
@@
 
-The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model cell counts in non--responding and responding individuals.
-Information is shared across non--responders and across responders through an empirical--Bayes prior, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}).
-The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample, our simulations show that we can relax this assumption to the unconstrained case where $p_s\ne p_u$ without compromising the observed false positive or false negative rates (Figures~\ref{fig:simulations} and \ref{fig:simulations_trunc}), while saving on computation time.
+The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model counts in experimental units (i.e.
+individuals) across multiple conditions (i.e.
+vaccine responders and non--responders).
+Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}).
+The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
 
-\todo[inline]{Fix references to non-existent figure}
-The analysis of real--world ICS data from a vaccine trial demonstrated that the MIMOSA model outperformed Fisher's exact test in detecting individuals who were responders and non--responders to vaccine across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:positivityrates}).
-Importantly, the MIMOSA model was fit within each antigen stimulation $\times$ cytokine $\times$ time point combination but the model was blinded to placebo and vaccine treatment.
-Despite this, MIMOSA demonstrated a lower false positive rate at day zero (increased specificity), where no response is expected, than Fisher's exact test (Figure~\ref{fig:positivityrates} Gag2, IL2, CD4+ T--cells, day 0, treatment group T1).
-Similarly, none of the placebo treated samples were identified as responders by the MIMOSA model (Figure~\ref{fig:positivityrates} treatment group C).
-MIMOSA also exhibited an increased sensitivity in real--world data compared to Fisher's exact test, as demonstrated by the higher response rates for 7 of the 8 stimulation/cytokine/cell type combinations at day 28 to which the model was fit, and where a response is expected to be observed.
-Although it is not possible to validate the responders identified at day 28 for all cases, the conclusion of increased sensitivity is consistent with both the performance of the model in simulations, and with the increased specificity observed on day 0 and in the placebo treated samples.
+Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}).
+Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination but the model was blinded to vaccine time point.
+Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C).
+Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182.
+None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison.
+Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).
 
-\todo[inline]{Fix references to non--existent figure}
 %We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
@@
 
-Polyfunctionality (identifying cells coexpressing multiple proteins, cytokines, or genes, is important in single cells assays since it allows the identification of functionally distinct subsets of cells in heterogeneous cell populations~\cite{Milush:2009bz}.
-In vaccine trials, these polyfunctional cell subsets may be correlated with vaccine response or efficacy.
-Existing methods for identifying polyfunctional cytokine profiles from ICS data either test individual combinations of cytokines separately, perform global tests at the expense of decreased power when only specific differences are of interest, sometimes  empirical in nature, and generally don't share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
-As others have pointed out, in order to have the most power to detect a true difference, the statistical test selected to detect differences in various combinations of cytokines should take into consideration the cytokine combinations that are of interest~\cite{Nason:2006dx}.
+Measuring differences in polyfunctional cell populations (i.e.
+identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in single cell immunological assays since they allow the identification of functionally distinct cell subsets in heterogeneous populations~\cite{Milush:2009bz}.
+In vaccine trials, these polyfunctional cell subsets may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design.
+Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e.
+mixtures of responders and non--responders).
+Sometimes tests are empirical in nature (i.e.
+thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
+As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}.
 A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different.
@@
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg.
-ROC and FDR plots of other cytokine combinations can be found in the Supplementary Information.}
+ROC and FDR plots of other cytokine combinations can be found in the Supplementary material.}
 \label{fig:HVTN065}
@@
 
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
+    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
+   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
+   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
+   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
+   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
+    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
+    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
+    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
+\end{tikzpicture}
+   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row).
+Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data.
+A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change.
+B) Boxplots for the unconstrained model.
+C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test.
+C) ROC curves for a representative simulated data set (N=100,000 counts).
+D) AUC  boxplots, E) observed vs.
+nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data.
+}\label{fig:simulations}
+\end{figure}
+
 \begin{figure}[htbp]
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
-    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
-   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
-   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
-    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
-    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
-    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
-\end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row).
-Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data.
-A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change.
-B) Boxplots for the unconstrained model.
-C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test.
-C) ROC curves for a representative simulated data set (N=100,000 counts).
-D) AUC  boxplots, E) observed vs.
-nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data.
-}\label{fig:simulations}
-\end{figure}
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/Multivariate_Sims_AUCBOX.pdf}};
@@
    \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
-    \node at (0,4.2) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,4.2) [font=\tiny\sffamily] {B} ;
-    \node at (9,4.2) [font=\tiny\sffamily] {C} ;
+    \node at (0,2.1) [font=\tiny\sffamily] {A} ;
+    \node at (4.5,2.1) [font=\tiny\sffamily] {B} ;
+    \node at (9,2.1) [font=\tiny\sffamily] {C} ;
 \end{tikzpicture}
    \caption{Multivariate simulations from a two--sided model.
-Ten, five-dimensional data sets were simulated from a two--sided model with an effect size of 2.5e-3 and -2.5e-3 in two of the five dimensions.
+Ten, five-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the five dimensions.
 Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value, combining Fisher's exact test run marginally on each of the five dimensions.

### 2012-06-29 MIMOSA.tex
Discussion and conclusion. working on SI
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
-Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}.
+Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that.
+Added ref for the HVTN054 trial and the ICS standardization paper from Horton})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej}.
 These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
@@
 
-\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:data} for details)~\cite{Peiperl:2010ej}.
-The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
+\textit{Flow cytometry}:
+The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see supplementary material~\ref{supp:statpublished}).
+The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T--cell subsets.
+Here, we analyzed the vaccinees (n=98) at two time points as a pooled data set (day 0 where no response is expected, and at day 182, the primary immunological endpoint).
+
+Simulation studies were based on hyper parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}.
+The statistical analysis of ICS data in the published trial is described in the original manuscript and outlined in the supplementary information (supplementary material~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
 The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.
 
-\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8 T-cells from sixteen individuals.
-The T-cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single-cell level and compared to paired, unstimulated controls.
-\footnote{Add a bit more info here, how many genes, tetramer sorted, etc}.
+\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8--positivie T-cells from sixteen individuals.
+T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post--stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
+The expression from the simulated samples  was compared to paired, unstimulated controls.~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al.
+could you fill this in?}.
 
@@
 \subsection{Beta-Binomial Model for One Marker}
+\label{sec:DE}
 For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
@@
 Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, to simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Supplementary material), and are given by,
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (supplementary material), and are given by,
  \begin{align}
@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
-Details about the algorithms are given in Supplementary material.
+Details about the algorithms are given in supplementary material.
 We used the method of Raftery and Lewis \citep{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
@@
 Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D).
-These results are consistent for other cytokines (see supplementary material).
+These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figures~\ref{fig:HVTN065Results} and~\ref{fig:HVTN065ResultsCont}).
 
@@
 Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--subject variability (Figure~\ref{fig:fluidigm}).
-These patterns are preserved in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), as well as in the posterior differences and log--odds (Figure~\ref{fig:fluidigm} B,C).
+These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions(Figure~\ref{fig:fluidigm} B,C).
 
@@
 \subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (See supplementary material).
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see supplementary material).
 For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large.
-More details about both algorithms are given in Supplementary material.
+More details about both algorithms are given in supplementary material.
 
@@
 Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we performed simulations (five--dimensional data) to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p--values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
-These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data.
+These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data (Figure~\ref{fig:mvsimulations} B).
 
@@
 Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}).
-Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination but the model was blinded to vaccine time point.
+Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination, the model is naive to vaccine time point.
 Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C).
@@
 
-%We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).
-For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test.
-These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions.
-This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model.
-This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.
+Although ICS is the motivating example for the MIMOSA model, it can be applied any type of single--cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
+To date, the typical approach to Fluidigm single cell gene expression analysis has been focused on identifying differences in the continuous part of the signal, corresponding to the $C_t$ values and the ``missing'' data from cells that don't express a given gene has generally been ignored, or used as part of ad--hoc pre-filtering\cite{Flatz:2011jb}.
+The ability of MIMOSA to identify stimulation--specific expression patterns in single--cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).
+MIMOSA explicitly models and performs inference on the observed positive cell counts between different conditions.
+Additionally, extending the model to take into account multiple  conditions is straightforward.
 
-Measuring differences in polyfunctional cell populations (i.e.
-identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in single cell immunological assays since they allow the identification of functionally distinct cell subsets in heterogeneous populations~\cite{Milush:2009bz}.
-In vaccine trials, these polyfunctional cell subsets may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design.
-Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e.
+Detecting differences in polyfunctional cell populations (i.e.
+identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in vaccine research, since it allows the identification of more precisely defined, more homogeneous cell popualtions~\cite{Milush:2009bz}.
+These polyfunctional cell populations may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design.
+What's more, changes in these polyfunctional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
+Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines, or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e.
 mixtures of responders and non--responders).
-Sometimes tests are empirical in nature (i.e.
+Sometimes these tests are entirely empirical in nature (i.e.
 thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
 As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}.
-A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different.
-This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.
-
-We show a proof of principle example with two components to identify responder and non--responder observations for the four--dimensional (two cytokine) case.
-Sharing of parameters across nested models in the two--cytokine case allows us to reduce the number of parameters.
-The model is easily extended.
-First, to multiple components to identify not just responders and non--responders, but responders with respect to each combination of cytokines.
-Second, extensions to multiple cytokines are straightforward, only requiring increasing the dimensionality of the Multinomial--Dirichlet distribution.
-With the increase in dimensionality there is a combinatorial explosion in the number of combinations of cytokines, and the resulting number of parameters in the model.
-We present several possible approaches to reducing the number of parameters in the model by sharing the hyperparamters across different combinations of cytokines.
-A number of these strategies can be generalized to models with multiple components that allow us to identify observations with differences in specific cytokine combinations between treatment and control samples.
+Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single--cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).
 
 \section{Conclusions}
-We have developed a combinatorial mixture model framework for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays.
+We have developed a mixture model framework that we call MIMOSA .for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays that frequently arise in the context of vaccine trials and immunological studies.
+Our model can be applied in a univariate or multivariate fashion, and we have shown that it has greater sensitivity and specificity than other typical approaches used to analyze the type of count--based data derived from single cell assays, including Fisher's exact test, the non-Bayesian likelihood ratio test, or ad--hoc rank based approaches.
 The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
@@
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg.
-ROC and FDR plots of other cytokine combinations can be found in the Supplementary material.}
+ROC and FDR plots of other cytokine combinations can be found in the supplementary material.}
 \label{fig:HVTN065}
@@
 Ordering of the rows and columns is preserved as in A).
-The traces show the deviations of each cell from zero.}
+The traces show the deviations of each cell from zero.
+Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).}
+
 \label{fig:fluidigm}
@@
 }\label{fig:simulations}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_AUC.pdf}};
+    \node[anchor=south west, inner sep=0] at (4.5,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_ROC.pdf}};
+   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_FDR.pdf}};
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
+ \end{tikzpicture}
+   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated.
+Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.}
+   \label{fig:simulations_trunc}
 \end{figure}
@@
 \end{figure}
-
+\clearpage
 %
@@
 \renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
-\setcounter{section}{0}
+\setcounter{section}{1}
 \setcounter{subsection}{0}
+\renewcommand{\figurename}{Supplementary Figure}
+\setcounter{figure}{0}
 %\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}
+
 \section*{Supplementary Information}
-\section{HVTN065 Vaccine Trial ICS Dataset Description}
-\label{supp:data}
+\subsection{HVTN065 and HVTN054 Vaccine Trial ICS Data Description}
+\label{supp:statpublished}
 HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B).
@@
 
-%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
+HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
 The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.
@@
 The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}.
+The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.
+
+\subsection{HVTN065 Results for Other Cytokines}
+\label{supp:HVTN065Results}
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
+%   \includegraphics[width=.75\columnwidth]{Figures/12} 
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/1}};
+    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/3}};
+    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/5}};
+    \node[anchor=south west, inner sep=0] at (8,-3.75){\includegraphics[width=.5\columnwidth]{Figures/6}};
+    \node[anchor=south west, inner sep=0] at (0,-7.5){\includegraphics[width=.5\columnwidth]{Figures/7}};
+    \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
+    \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
+    \node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
+    \node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
+    \node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};
+
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
+    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
+    \node at (8,-0.5) [font=\tiny\sffamily] {D} ;
+    \node at (0,-4) [font=\tiny\sffamily] {E} ;
+    \node at (8,-4) [font=\tiny\sffamily] {F} ;
+    \node at (0,-7.75) [font=\tiny\sffamily] {G} ;
+    \node at (8,-7.75) [font=\tiny\sffamily] {H} ;
+    \node at (0,-11.75) [font=\tiny\sffamily] {I} ;
+    \node at (8,-11.75) [font=\tiny\sffamily] {J} ;
+    \end{tikzpicture}
+   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
+\label{fig:HVTN065Results}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
+%   \includegraphics[width=.75\columnwidth]{Figures/12} 
+\begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=.5\columnwidth]{Figures/14}};
+    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/15}};
+    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/16}};
+
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
+    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
+    \end{tikzpicture}
+   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
+\label{fig:HVTN065ResultsCont}
+\end{figure}
 

### 2012-07-02 MIMOSA.tex
update si
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.
+In the supplementary material we describe modifications for an alternative constrained model where $p_s>p_u$ (supplementary material~\ref{supp:constrained}).
 
@@
 \subsubsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
-We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at coexpression of pairs of genes in single cells across the four stimulations.
+We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at co--expression of pairs of genes in single cells across the four stimulations.
 In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
@@
 
+\subsection{Constrained Beta--binomial model}
+\label{supp:constrained}
+We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$.
+In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in equation~\eqref{model2:unconstrained}.
+The marginal likelihood for the constrained model becomes $\mathrm{L}_1\times \mathrm{C}$, where $\mathrm{C}$ is a ratio of normalizing constants for the likelihood and the prior.
+
+\begin{align}
+\begin{split}
+	 \mathrm{C}=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
+\label{eq:model2post}
+\end{split}
+\end{align}
+
+The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$ is the regularized incomplete Beta function, which is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional integral that can be computed via Monte--Carlo integration (i.e.
+in our EM implementation).
+However, this can be computationally costly in the MCMC framework.
+In the latter case, we can compute the integral exactly if $\alpha_s$ and $\beta_s$ are integers by using the identity:
+\begin{align}
+I_{1-p_u}(\beta_s,\alpha_s) = \sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^j(p_u)^{\beta_s+\alpha_s-j}
+\label{eq:IBident}
+\end{align}
+Typically, in ICS data $\beta_s>>\alpha_s$, leading to relatively few terms in the sum in equation~\ref{eq:IBident}.  This leads to an exact expression for the normalizing constant:
+$$
+foo
+$$
+
 \subsection{HVTN065 Results for Other Cytokines}

### 2012-07-14 MIMOSA.tex
Modified the abstract and intro.

@@
 \begin{abstract}
-Immunological endpoints in vaccine trials are measured through a variety of assays that  provide single--cell measurements of multiple genes and proteins in specific immunological cell populations.
-%intracellular or cell surface proteins, or mRNA expression levels of genes in single cells from specific immunological cell populations.
-Using single--cell data, we consider the problem of identifying subjects where these cell populations exhibit differential responses under different experimental conditions.  
-%While these measurements are usually continuous, they are often discretized and analyzed as counts of single cells. 
-For example, in the intracellular cytokine staining assay from flow cytometry, individual cells are classified as either positive or negative for a marker based on a predetermined threshold.
-%One such assay is the intracellular cytokine staining assay that
-The assay is used to assess an individual's immune response to a vaccine by measuring the number of antigen--specific cells  producing different cytokines in different T--cell subpopulations in response to different antigen stimulations.
-Individuals whose T--cells exhibit increased production of a cytokine in response to stimulation are termed ``positive'' for that cytokine, and multiple such ``positivity calls'' are used to identify vaccine responders.
-%Cells are classified as cytokine positive or negative and the resulting The cell counts are used to identify experimental conditions leading to an increase in cytokine producing cells.
-The ICS assay is generally used to identify ``vaccine responders''.
-These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline.
-The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern.
-The standard approach to analyze such data, Fisher's exact test, is problematic for two reasons.
-It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test.
-Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.
-Here we present a framework based on mixtures of Beta--binomial or Dirichlet--Multinomial distributions for analyzing count data derived from such single--cell assays.
-Our method models cellular responses in a marker--specific manner, treating the responding and non--responding observations as separate components in the model.
-Cell counts from the different experimental conditions are modelled independently, while sharing information across responding and non--responding observations through empirical Bayes priors in order to increase the sensitivity and specificity of positivity calls.
-We compare our method against Fisher's exact test, a likelihood ratio test, and ranked log fold changes, and show how it can be extended to model multivariate (multiple markers) cellular responses.
-In simulations and applied to real data sets, we find that our method has higher sensitivity and specificity than Fisher's exact test and alternative methods.
-% simultaneously, sharing information across observations using empirical Bayes methods.
-The model represents the responder and non-responder observations as separate components in the model, stimulated and unstimulated cell counts are modelled independently.
-Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test.
+Most biological samples (\textit{e.g.} blood, tissues) are composed of many small and functionally distinct cell subsets that can only be measured accurately using single cell assays.
+The characterization of these small cell subsets via single cell assays is important and cannot be done with assays that measure cell mixtures.
+For this reason, an increasingly number of studies rely on single cell assays to provide single-cell measurements of multiple genes and proteins.
+A common problem in the analysis of such data is to identify, for a given individual, markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combinations) in the cell subset(s) of interest.
+Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays.
+Our model allows inference of differential expression to be subject specific, while borrowing strength across individuals through common prior distributions.
+We propose two approaches for parameter estimation, one based on an Expectation-Maximization algorithm and one based on Markov chain Monte Carlo.
+We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
+Using several experimental immunological based assays measuring proteins and genes at the single-cell level, we find that our method has higher sensitivity and specificity than Fisher's exact test and alternative methods.
+Using a simulation study we also show that our model is robust from departures from model assumptions.
+Finally, we also demonstrate how our framework can be extended to testing differential expression of marker combinations using a Dirichlet-Multinomial model.
 \end{abstract}
@@
 \section{Introduction}
+Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
+This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
+For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
 In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
 Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers).
-More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell.
-Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize cell poly-functionality.
+More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell (\textbf{cite CYTOF}).
+Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells.
 A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6.
-While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability.
+While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references).
 This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data.
 Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general.
-For this reason, single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
-These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood.
-These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences.
-Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
+These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
+During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
+For some assays (\textit{e.g.} flow cytometry) such thresholds are set based on prior biological knowledge while for others thresholds are actually given by the assay directly.
+This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single cell level.
+After this thresholding step, we obtain a matrix of Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers.
+Using this matrix, one can form $2^K$ putative cell subsets; when $K$ is large there is a combinatorial explosion of the number of subsets and many of these might be small or even empty.
+A common statistical problem is then, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.} before and after vaccination).
 A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine.
-Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells.
-While not all T--cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
-After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}.
-These antigen--specific T--cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T--cells.
-The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cels via antigen presenting cells.
+While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
+After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}.
+These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells.
+The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
+Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected.
 Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
-These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences.
-Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw, Bendall:2011wf}.
-
-%The ICS assay in flow cytometry is just one example of the applications of single--cell technologies to immunology.
-Single--cell gene expression technologies such as Fluidigm are enabling researchers to measure the expression of 96 genes in 96 single cells from a homogeneous population sorted by flow cytometry\cite{Jang:2011uj}.
-This technology has been used to identify signatures of immune cell sub--populations that correlate with different types of vaccine~\cite{Flatz:2011jb}.
-To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements.
-However, the proportion of single cells expressing individual genes also carries information and should be evaluated.
-
-Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that.
+These counts are compared between antigen stimulated and unstimulated samples from a individual to identify significant differences.
+individuals who generate a response after stimulation are called responders whereas individual that do not show any differences are called non-responders.
+In many single-cell studies, the size of the functionally distinct subpopulations (\textit{i.e.} the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
+Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that.
 Added ref for the HVTN054 trial and the ICS standardization paper from Horton})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej}.
-These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
-In the context of single-cell gene expression, very few methods have been proposed.
-To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements.
-However, the proportion of single cells expressing individual genes also carries information and should be evaluated.
-
-%on a per--cytokine basis to identify observations where the number of cytokine producing cells is greater in the stimulated than the control data Problems with Fisher's exact when the test assumptions are not met have been reviewed in the literature~\cite{Proschan:2009ks}.
-In the context of single--cell assay data the assumption of fixed margins is not satisfied since the observations for treatment and control conditions arise from independent experiments, thus the p--values from Fisher's exact test are incorrect.
-More importantly, Fisher's exact test does not share information across observations, resulting in low power when counts are small.
+All of these methods analyze each individual separately, and no information is shared across individuals even though one could expect some similarities across responders in terms of their response.
+In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
 
 The framework developed in this paper addresses these issues explicitly.
-We present a multinomial-Dirichlet combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells.
-The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells.
-Importantly, our approach shares information across individuals (or subjects) by means of a Dirichlet prior distribution placed on the unknown cell population proportions of the mutinomial likelihood, which help increase sensitivity and specificity to detect rare antigen specific responding cell populations.
-The cell counts from the stimulated and unstimulated experiments are modeled independently, and different combinations of markers are represented as different mixture components in the framework.
-This approach allows us to omit certain combinations of markers that are not observable (i.e., two cytokines may never be co--expressed), or to explicitly represent combinations that are of interest (i.e., we can explicitly represent and test for a difference in a pair of cytokines in an experiment measuring many cytokines).
-This is a flexible approach that avoids the some of the drawbacks of global tests when only a few of the many measured markers show differences~\cite{Nason:2006dx}.
+In our model, cell counts are modeled via a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by mean of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
+In order to discriminate between responders and non-responders, the prior is written as a mixture of two Dirichlet distributions where the hyperparameters for each mixture component are shared across individuals.
+This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single cell assays, and increase sensitivity and specificity to detect responders.
+Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect some effect biolgical changes spread out across multiple cell subsets~\cite{Nason:2006dx}.
 
 \section{Data structure and notation} 
-In this paper we consider two different immunological single-cell assays typically used in vaccine trials, one flow cytometry data set, and one single-cell gene expression data set.
+In this paper we consider two different immunological single-cell assays, one flow cytometry data set and one single-cell gene expression data set.
 
@@
 The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see supplementary material~\ref{supp:statpublished}).
-The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T--cell subsets.
-Here, we analyzed the vaccinees (n=98) at two time points as a pooled data set (day 0 where no response is expected, and at day 182, the primary immunological endpoint).
+The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets.
+Here, we analyze a subset of the data consisting of 98 individual from the vaccinee group at two time points, day 0 and day 182.
+Samples on day 0 were taken just before vaccination and no response is expected there.
+The corresponding samples can be used as negative controls.
+Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least.
 
-Simulation studies were based on hyper parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}.
-The statistical analysis of ICS data in the published trial is described in the original manuscript and outlined in the supplementary information (supplementary material~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
-The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.
-
-\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8--positivie T-cells from sixteen individuals.
+\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen individuals.
 T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post--stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
@@
 Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
-As a consequence, we first consider the one marker case where cell counts are marginalized and each marker is analyzed separately and $K=1$.
+As a consequence, we first consider the one marker case where cell counts are marginalized and each marker is analyzed separately, \textit{i.e.} $K=1$.
 In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
@@
 
-\subsection{Beta-Binomial Model for One Marker}
+\subsection{Beta-binomial model}
 \label{sec:DE}
@@
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.
-In the supplementary material we describe modifications for an alternative constrained model where $p_s>p_u$ (supplementary material~\ref{supp:constrained}).
+In some study, such as the ICS trial used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
+This alternative parametrization is described in Supplementary Web material.
 
@@
 Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}.
-As an alternative, we will also explore a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
+As an alternative, we also describe a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
 In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{sec:estimation}.
@@
 where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.
+In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed form but can easily be estimated.
+This is described in Supplementary Web material.
 
@@
 \section{Results}
-The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation.
+In this section, we apply our MIMOSA model to the data described in Section, and present the results of a simulation study based on the ICS data.
+%The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation.
 The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells.
@@
 
-We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 120 individuals (98 vaccines and 20 placebo recipients).
-We analyzed the vaccinees (n=98) at the day 0 and day 182 (the primary endpoint) time points as a pooled data set.
-Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity).
-We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples.
+Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
+We examined the CD4+ T-cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples.
 An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.
 
-The MMOSA model had higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e.
+The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e.
 ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs.
@@
 \subsection{Fluidigm}
-\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single--Cell Data}
-We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples.
+\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single-Cell Data}
+We applied the MIMOSA model to Fluidigm single-cell gene expression data from CD8+ T-cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples.
 The unconstrained MIMOSA model was fit to each stimulation.
@@
 The data are presented in (Figure~\ref{fig:fluidigm} A-C).
-Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--subject variability (Figure~\ref{fig:fluidigm}).
-These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions(Figure~\ref{fig:fluidigm} B,C).
+Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--individual variability (Figure~\ref{fig:fluidigm}).
+These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions (Figure~\ref{fig:fluidigm} B,C).
 
 \subsection{Simulation Studies}
+
+Simulation studies were based on hyper parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}.
+The statistical analysis of ICS data in the published trial is described in the original manuscript and outlined in the supplementary information (supplementary material~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
+The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.
+
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
@@
 \subsection{Prior}
-As in the one-marker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
+As in the one-marker case, we share information across individuals using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
 \begin{align}

### 2012-07-16 MIMOSA.tex
updated refs
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers).
-More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell (\textbf{cite CYTOF}).
+More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}.
 Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells.

### 2012-07-19 MIMOSA.tex
Updated 065 results and simulations
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
     \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
-    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/12}};
+    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/1}};
     \node at (0,3.4) [font=\tiny\sffamily] {A} ;
@@
    \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065.
-Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg.
+Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) ENV-1-PTEG and expressing IL2.
 ROC and FDR plots of other cytokine combinations can be found in the supplementary material.}
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
-    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
     \node at (0,3.4) [font=\tiny\sffamily] {A} ;
     \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
-    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
-    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
-    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
 \end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row).
-Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data.
-A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change.
-B) Boxplots for the unconstrained model.
-C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test.
-C) ROC curves for a representative simulated data set (N=100,000 counts).
-D) AUC  boxplots, E) observed vs.
-nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data.
-}\label{fig:simulations}
+   \caption{Comparison of positivity detection methods on data simulated from the one--sided model.
+Ten simulations were generated at an $N$ of 5,000 total counts using hyper--parameter estimates from real ICS data (IFNg expressing CD4+ T--cells stimulated with ENV-1-PTEG from HVTN065) with a five--fold effect size between responder and non--responder components.
+A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
+Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and ranked log fold--change.
+Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in the Supplementary Material.}\label{fig:simulations}
 \end{figure}
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/1}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/12}};
     \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/3}};

### 2012-07-19 MIMOSA.tex
Updated multivariate simulations
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc}).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
@@
 vaccine responders and non--responders).
-Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}).
+Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}).
 The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_AUC.pdf}};
-    \node[anchor=south west, inner sep=0] at (4.5,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_FDR.pdf}};
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
- \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated.
-Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.}
-   \label{fig:simulations_trunc}
-\end{figure}
-
 \begin{figure}[htbp]
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/Multivariate_Sims_AUCBOX.pdf}};
-    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
-    \node at (0,2.1) [font=\tiny\sffamily] {A} ;
-    \node at (4.5,2.1) [font=\tiny\sffamily] {B} ;
-    \node at (9,2.1) [font=\tiny\sffamily] {C} ;
+    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
+   \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
+    \node at (0,3.5) [font=\tiny\sffamily] {A} ;
+    \node at (6,3.5) [font=\tiny\sffamily] {B} ;
 \end{tikzpicture}
    \caption{Multivariate simulations from a two--sided model.
-Ten, five-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the five dimensions.
-Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value, combining Fisher's exact test run marginally on each of the five dimensions.
-A) Boxplots of AUCs (area under the curve) for the ten simulations.
-B) Observed and nominal false discovery rate for each method.
-C)ROC curves for the competing methods.}
+Ten, eight-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
+Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value (combining Fisher's exact test run marginally on each of the five dimensions).
+A) Average ROC curves for the competing methods over 10 simulations.
+B) Average observed and nominal false discovery rate for each method over 10 simulations.}
    \label{fig:mvsimulations}
@@
 
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+ \end{tikzpicture}
+   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated.
+Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
+A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations at N=5,000.}
+   \label{fig:simulations_trunc}
+\end{figure}
+
 %\section{Data Import, Preprocessing, and Gating}

### 2012-07-20 MIMOSA.tex
Some changes to manuscript.

--- 2012-07-16 MIMOSA.tex
@@
 \begin{abstract}
-Most biological samples (\textit{e.g.} blood, tissues) are composed of many small and functionally distinct cell subsets that can only be measured accurately using single cell assays.
-The characterization of these small cell subsets via single cell assays is important and cannot be done with assays that measure cell mixtures.
-For this reason, an increasingly number of studies rely on single cell assays to provide single-cell measurements of multiple genes and proteins.
-A common problem in the analysis of such data is to identify, for a given individual, markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combinations) in the cell subset(s) of interest.
+Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single cell assays.
+The characterization of these small cell subsets is crucial to decipher system level biological changes.
+For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples.
+A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
 Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays.
-Our model allows inference of differential expression to be subject specific, while borrowing strength across individuals through common prior distributions.
-We propose two approaches for parameter estimation, one based on an Expectation-Maximization algorithm and one based on Markov chain Monte Carlo.
+Our model allows the inference to be subject specific, while borrowing strength across subjects\footnote{Should we use subject, indivuduals or something else?
+What about animal studies?} through common prior distributions.
+We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
 We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
-Using several experimental immunological based assays measuring proteins and genes at the single-cell level, we find that our method has higher sensitivity and specificity than Fisher's exact test and alternative methods.
-Using a simulation study we also show that our model is robust from departures from model assumptions.
-Finally, we also demonstrate how our framework can be extended to testing differential expression of marker combinations using a Dirichlet-Multinomial model.
+Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods.
+Using a simulation study we show that our framework is also robust to model misspecification.
+Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model.
 \end{abstract}
@@
 Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers).
-More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}.
+More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell (\textbf{cite CYTOF}).
 Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells.
@@
 During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
-For some assays (\textit{e.g.} flow cytometry) such thresholds are set based on prior biological knowledge while for others thresholds are actually given by the assay directly.
+For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology.
 This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single cell level.
-After this thresholding step, we obtain a matrix of Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers.
-Using this matrix, one can form $2^K$ putative cell subsets; when $K$ is large there is a combinatorial explosion of the number of subsets and many of these might be small or even empty.
-A common statistical problem is then, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.} before and after vaccination).
+After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers.
+Using this matrix, one can form $2^K$ putative cell subsets.
+When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
+A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
 A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine.
-Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cels via antigen presenting cells.
+Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
@@
 Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
-These counts are compared between antigen stimulated and unstimulated samples from a individual to identify significant differences.
-individuals who generate a response after stimulation are called responders whereas individual that do not show any differences are called non-responders.
-In many single-cell studies, the size of the functionally distinct subpopulations (\textit{i.e.} the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
+These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences.
+individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}.
+In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
+
 Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that.
 Added ref for the HVTN054 trial and the ICS standardization paper from Horton})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej}.
-All of these methods analyze each individual separately, and no information is shared across individuals even though one could expect some similarities across responders in terms of their response.
-In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.
+All of these methods analyze each individual separately, and no information is shared across individuals even though one could expect some similarities across responders (\textit{resp.} non-responders).
+In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
+Let's update this sentence later}.
 
@@
 In our model, cell counts are modeled via a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by mean of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
-In order to discriminate between responders and non-responders, the prior is written as a mixture of two Dirichlet distributions where the hyperparameters for each mixture component are shared across individuals.
-This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single cell assays, and increase sensitivity and specificity to detect responders.
-Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect some effect biolgical changes spread out across multiple cell subsets~\cite{Nason:2006dx}.
+In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals.
+This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell\footnote{Let's replace all single cell by single-cell} assays, and increase sensitivity and specificity when detecting responders.
+Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\cite{Nason:2006dx}.
 
 \section{Data structure and notation} 
-In this paper we consider two different immunological single-cell assays, one flow cytometry data set and one single-cell gene expression data set.
+\label{sec:data}
+In this paper we consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.
 
@@
 Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
-As a consequence, we first consider the one marker case where cell counts are marginalized and each marker is analyzed separately, \textit{i.e.} $K=1$.
+As a consequence, we first consider the one marker case (\textit{i.e.} $K=1$) where marginal cell counts are analyzed separately.
 In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
@@
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.
-In some study, such as the ICS trial used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
-This alternative parametrization is described in Supplementary Web material.
+In some study, such as the ICS studies used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
+This alternative parametrization is described in Web Supplementary material\footnote{let's make sure we use the proper format for referencing the supplementary material}, and we refer to it as the one-sided model.
 
@@
 Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model.
-Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the proportion of responders.
+Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the (unknown) proportion of responders.
 It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$.
@@
 \label{sec:estimation}
-Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, to simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (supplementary material), and are given by,
+Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations.
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Supplementary material), and are given by,
  \begin{align}
@@
 \end{align}
-Assuming that the missing data, the $z_i$'s, are known, we define the complete log-likelihood as follows,
+Assuming that the missing data, the $z_i$'s, are known, we define the complete data log-likelihood as follows,
 \begin{equation}
@@
 In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed form but can easily be estimated.
-This is described in Supplementary Web material.
+All calculations for the one-sided case are described in Supplementary Web material.
 
@@
 We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.
-Starting from some initial values, the EM algorithms iterates between the E and M steps until convergence.
+Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence.
 In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components.
@@
 \section{Results}
-In this section, we apply our MIMOSA model to the data described in Section, and present the results of a simulation study based on the ICS data.
+In this section, we apply our MIMOSA model to the data described in Section \ref{sec:data}, and present the results of a simulation study based on the ICS data.
 %The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation.
@@
 \subsection{ICS}
-\subsubsection*{MIMOSA Outperforms Competing Methods on Vaccine Trial Data from Study HVTN065}
+\subsubsection*{HVTN065}
 
+Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
-We examined the CD4+ T-cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples.
-An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.
-
-The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e.
+We examined the CD4+ T-cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples\footnote{Were these stimulations described in the data section}.
+The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders when the magnitude of response was small (i.e.
 ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs.
@@
 Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D).
-These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figures~\ref{fig:HVTN065Results} and~\ref{fig:HVTN065ResultsCont}).
+These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figures~\ref{fig:HVTN065Results} and~\ref{fig:HVTN065ResultsCont}) \footnote{Perhaps we could only keep the Env stimulation}.
 
-\subsection{Fluidigm}
-\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single-Cell Data}
+\subsubsection*{HVTN054}
+
+\subsection{Single-cell gene expression}
 We applied the MIMOSA model to Fluidigm single-cell gene expression data from CD8+ T-cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. 
-The unconstrained MIMOSA model was fit to each stimulation.
-From the model we calculated the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions.
-The data are presented in (Figure~\ref{fig:fluidigm} A-C).
-Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--individual variability (Figure~\ref{fig:fluidigm}).
+This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation.
+In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation.
+The results presented in Figure~\ref{fig:fluidigm} A-C show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm}).
 These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions (Figure~\ref{fig:fluidigm} B,C).
+A similar analysis using a two-sided Fisher's exact test (Figure xx) does not reveal any stimulation specific changes of expression.\footnote{Perhaps it would be good to compare the number of differentially expressed genes}.
 
@@
 
-Simulation studies were based on hyper parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}.
+%Simulation studies were based on hyper-parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}.
 The statistical analysis of ICS data in the published trial is described in the original manuscript and outlined in the supplementary information (supplementary material~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
-Using hyper parameters estimated from the constrained model fit to data from Gag1--stimulated, CD4--positive, IL2--expressing T--cells from day 28 post--vaccination of the HVTN054 trial~\cite{Peiperl:2010ej}.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to data from Gag1-stimulated, CD4-positive, IL2-expressing T-cells from day 28 post--vaccination of the HVTN054 trial~\cite{Peiperl:2010ej}.
 We simulated data from this constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, with ten independent realizations of data for each $N$.
-Constrained MIMOSA was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change.
-This procedure was repeated for the unconstrained model fit to unconstrained data (Figure~\ref{fig:simulations} A-D).
-The nominal vs observed false discovery rate was also examined to assess the model fit (Figure~\ref{fig:simulations} E-F).
+The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change.
+This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D).
+In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} E-F).
 
-For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, D).
-Additionally, the false discovery rate for MIMOSA more closely reflected the nominal false discovery rate compared to Fisher's exact test (Figure~\ref{fig:simulations} E, F).
+For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, D).
+Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test (Figure~\ref{fig:simulations} E, F).
 %Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}).
-Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065\footnote{Why do we change data now?} data set (Figure~\ref{fig:simulations_trunc}).
+Even under these departures from the model assumptions, MIMOSA outperformed other approaches.
 
 \section{Differential expression across marker combinations}
-Our beta-binomial model described in section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
-As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+Our beta-binomial model described in Section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
+As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.\footnote{At some point we should say that such multivariate models are not appropriate for one sided models.}
 %Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
@@
 \subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see supplementary material).
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Supplementary Material).
 For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2.
-The EM algorithms becomes very dependent on the initial values, and can even fail to converge when good initial values are provided.
+It becomes highly dependent on the initial values and can even fail to converge when good initial values are provided.
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large.
@@
 
-\subsubsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
-We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at co--expression of pairs of genes in single cells across the four stimulations.
+\subsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
+As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
 In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
 Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation.
-The typical approach to analyzing polyfunctional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible poly-functional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
 That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.
@@
 
-Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we performed simulations (five--dimensional data) to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p--values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
+Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we could not look at more than two makers at once.
+Therefore, we performed simulations in eight dimension to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p-values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
 These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data (Figure~\ref{fig:mvsimulations} B).
@@
 \section{Discussion}
-The variety of single--cell assays being adopted by the immunology community is increasing.
-Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data.
+The variety of single-cell assays being adopted by the immunology community is increasing.
+Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single-cell count data.
 Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.

### 2012-07-20 MIMOSA.tex

@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065\footnote{Why do we change data now?} data set (Figure~\ref{fig:simulations_trunc}).
-Even under these departures from the model assumptions, MIMOSA outperformed other approaches.
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc}).
+Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
 
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
-    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
     \node at (0,3.4) [font=\tiny\sffamily] {A} ;
     \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
-    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
-    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
-    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
 \end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row).
-Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data.
-A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change.
-B) Boxplots for the unconstrained model.
-C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test.
-C) ROC curves for a representative simulated data set (N=100,000 counts).
-D) AUC  boxplots, E) observed vs.
-nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data.
-}\label{fig:simulations}
-\end{figure}
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_AUC.pdf}};
-    \node[anchor=south west, inner sep=0] at (4.5,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_FDR.pdf}};
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
- \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated.
-Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.}
-   \label{fig:simulations_trunc}
+   \caption{Comparison of positivity detection methods on data simulated from the one--sided model.
+Ten simulations were generated at an $N$ of 5,000 total counts using hyper--parameter estimates from real ICS data (IFNg expressing CD4+ T--cells stimulated with ENV-1-PTEG from HVTN065) with a five--fold effect size between responder and non--responder components.
+A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
+Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and ranked log fold--change.
+Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in the Supplementary Material.}\label{fig:simulations}
 \end{figure}
@@
 \label{fig:HVTN065ResultsCont}
+\end{figure}
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+ \end{tikzpicture}
+   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated.
+Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
+A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations at N=5,000.}
+   \label{fig:simulations_trunc}
 \end{figure}

### 2012-07-20 MIMOSA.tex
Updated fluidigm plots
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}.
-Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells.
-A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6.
+Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
+A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
 While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references).
@@
 \node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
-\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot3.pdf}};
+\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
 \node at (0,3) [font=\tiny\sffamily] {A} ;
@@
 \caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
-The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a significant decrease, green a significant increase, relative to the control).
+The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
 Columns and rows are clustered based on these signed posterior probabilities.
-B) The $log_2$ ratio of the proportion of cells expressing a gene in the stimulated vs.
+B) The posterior differences in proportion of cells expressing a gene in the stimulated vs.
 control samples.
 Rows and columns are ordered as in A) for comparison.
-C) The difference in the proportion of cells expressing each gene in the stimulated vs.
-control samples.
-Ordering of the rows and columns is preserved as in A).
 The traces show the deviations of each cell from zero.
-Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).}
-
+Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
+C) Genes selected from Fisher's exact test at the 10\% FDR level.
+Clustering of the maximum likelihood estimates of difference in proportions of cells expressing a gene in stimulated vs.
+control samples with significant genes selected using Fisher's exact test.}
 \label{fig:fluidigm}
@@
 
-HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
+%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
 The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.

### 2012-07-23 MIMOSA.tex
Updated figures
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D).
-These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figures~\ref{fig:HVTN065Results} and~\ref{fig:HVTN065ResultsCont}) \footnote{Perhaps we could only keep the Env stimulation}.
+These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}) \footnote{Perhaps we could only keep the Env stimulation}.
 
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
-    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.5\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
     \node at (0,3.4) [font=\tiny\sffamily] {A} ;
@@
 \node[anchor=south west, inner sep=0] at (6.5,0) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
-\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
-\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
 \node at (0,4.5) [font=\tiny\sffamily] {A} ;
 \node at (6.25,4.5) [font=\tiny\sffamily] {B} ;
-\node at (0,0) [font=\tiny\sffamily] {C} ;
-\node at (6.25,0) [font=\tiny\sffamily] {D} ;
+%\node at (0,0) [font=\tiny\sffamily] {C} ;
+%\node at (6.25,0) [font=\tiny\sffamily] {D} ;
 \end{tikzpicture}
 \caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
-The posterior difference in proportions between stimulated and unstimulated samples fitting the C) marginalized counts D) multivariate combinations.
-No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.}
+No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
+Sample names highlighted in red identify those where MIMOSA did not detect a difference.}
 \label{fig:polyfunctionality}
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/12}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/16}};
     \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/3}};
@@
     \node[anchor=south west, inner sep=0] at (0,-7.5){\includegraphics[width=.5\columnwidth]{Figures/7}};
-    \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
-    \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
-    \node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
-    \node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
-    \node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};
+   % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
+   % \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
+    %\node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
+    %\node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
+    %\node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};
 
@@
     \node at (0,-4) [font=\tiny\sffamily] {E} ;
-    \node at (8,-4) [font=\tiny\sffamily] {F} ;
-    \node at (0,-7.75) [font=\tiny\sffamily] {G} ;
-    \node at (8,-7.75) [font=\tiny\sffamily] {H} ;
-    \node at (0,-11.75) [font=\tiny\sffamily] {I} ;
-    \node at (8,-11.75) [font=\tiny\sffamily] {J} ;
+  %  \node at (8,-4) [font=\tiny\sffamily] {F} ;
+%    \node at (0,-7.75) [font=\tiny\sffamily] {G} ;
+  %  \node at (8,-7.75) [font=\tiny\sffamily] {H} ;
+   % \node at (0,-11.75) [font=\tiny\sffamily] {I} ;
+   % \node at (8,-11.75) [font=\tiny\sffamily] {J} ;
     \end{tikzpicture}
-   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
+   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
 \label{fig:HVTN065Results}
@@
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
-%   \includegraphics[width=.75\columnwidth]{Figures/12} 
-\begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=.5\columnwidth]{Figures/14}};
-    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/15}};
-    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/16}};
-
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
-    \end{tikzpicture}
-   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
-\label{fig:HVTN065ResultsCont}
-\end{figure}
+%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
+%%   \includegraphics[width=.75\columnwidth]{Figures/12} 
+%\begin{tikzpicture}
+%    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=.5\columnwidth]{Figures/14}};
+%    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/15}};
+%    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/16}};
+%
+%
+%    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
+%    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
+%    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
+%    \end{tikzpicture}
+%   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
+%\label{fig:HVTN065ResultsCont}
+%\end{figure}
 
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
-    \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};        \node[anchor=south west,inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
     \node at (0,3.4) [font=\tiny\sffamily] {A} ;
     \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
+        \node at (0,0) [font=\tiny\sffamily] {C} ;
+    \node at (4.6,0) [font=\tiny\sffamily] {D} ;
  \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated.
-Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
-A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations at N=5,000.}
+   \caption{Unconstrained MIMOSA model fit to two--sided data (A-B) or to data where model assumptions are violated (C-D).
+Two--sided data were simulated from a standard model.
+A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations at N=5,000.
+Unconstrained MIMOSA fit to two--sided data simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
+C) Average ROC curves from 10 simulations and D) average observed vs nominal FDR from 10 simulations at N=5,000.
+}
    \label{fig:simulations_trunc}

### 2012-07-25 MIMOSA.tex
Rewrote the conclusion.

@@
 \subsection{ICS}
-\subsubsection*{HVTN065}
-
-Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
+Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change\footnote{Why ranked?}, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
-We examined the CD4+ T-cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples\footnote{Were these stimulations described in the data section}.
-The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders when the magnitude of response was small (i.e.
-ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs.
-0.923 for Fisher's exact test vs.
-0.827 for LRT vs.
-0.775 for ranked fold change).
-When the magnitude of response was larger (GAG-1-PTEG stimulation), MIMOSA performed as well as Fisher's exact test, and better than other competing methods (Figure~\ref{fig:HVTN065} C) (AUC=0.982 for MIMOSA vs.
-0.978 for Fisher's exact test vs.
-0.877 for LRT vs.
-0.784 for ranked fold change).
-Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D).
+We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?}.
+The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
+In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B).
 These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}) \footnote{Perhaps we could only keep the Env stimulation}.
 
-\subsubsection*{HVTN054}
-
 \subsection{Single-cell gene expression}
-We applied the MIMOSA model to Fluidigm single-cell gene expression data from CD8+ T-cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. 
+We applied the MIMOSA model to Fluidigm single-cell gene expression.
 This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation.
@@
 The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.
-In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2.
-It becomes highly dependent on the initial values and can even fail to converge when good initial values are provided.
-Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large.
-More details about both algorithms are given in supplementary material.
+In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than three.
+It becomes highly dependent on the initial values and can fail to converge in many instances.
+Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
+More details about both algorithms are given in Web Supplementary material.
 
@@
 \section{Discussion}
-The variety of single-cell assays being adopted by the immunology community is increasing.
-Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single-cell count data.
-Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.
-Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}.
-Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
+Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
+Single cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single cell).
+As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.
+Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small.
+Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
+In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single cell assay.
 
-The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model counts in experimental units (i.e.
-individuals) across multiple conditions (i.e.
-vaccine responders and non--responders).
-Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}).
-The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
+The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental individuals across multiple conditions (\textit{i.e.} vaccine responders and non-responders).
+Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}).
+The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
+This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.
 
-Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}).
+%Importantly, the analysis of real-world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}).
 Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination, the model is naive to vaccine time point.
-Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C).
+Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre-vaccination) and 182 (post-vaccination) than the current standard approach (Fisher's exact test), or non-Bayesian alternatives such as the likelihood ratio test or ranking by log-fold change (Figure~\ref{fig:HVTN065} A,C).
 Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182.
@@
 
-Although ICS is the motivating example for the MIMOSA model, it can be applied any type of single--cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
-To date, the typical approach to Fluidigm single cell gene expression analysis has been focused on identifying differences in the continuous part of the signal, corresponding to the $C_t$ values and the ``missing'' data from cells that don't express a given gene has generally been ignored, or used as part of ad--hoc pre-filtering\cite{Flatz:2011jb}.
-The ability of MIMOSA to identify stimulation--specific expression patterns in single--cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).
-MIMOSA explicitly models and performs inference on the observed positive cell counts between different conditions.
-Additionally, extending the model to take into account multiple  conditions is straightforward.
+Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
+In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or simply using this information for pre-filtering \cite{Flatz:2011jb}.
+The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).
 
-Detecting differences in polyfunctional cell populations (i.e.
-identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in vaccine research, since it allows the identification of more precisely defined, more homogeneous cell popualtions~\cite{Milush:2009bz}.
-These polyfunctional cell populations may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design.
-What's more, changes in these polyfunctional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
-Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines, or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e.
-mixtures of responders and non--responders).
-Sometimes these tests are entirely empirical in nature (i.e.
-thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}.
-As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}.
-Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single--cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).
+Detecting differences in poly-functional cell populations (\textit{i.e.} identifying changes in cell populations that co-express multiple proteins, cytokines, or genes), is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\cite{Milush:2009bz}.
+In the context of HIV, these poly-functional cell populations have been shown to be correlated with good outcome such as vaccine protection and long-term disease non-progression (ADD REFERENCES TO THE PAPERS STEVE SUGGESTED IN THE GRANT).
+In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected.
+It  follows that if a specific cell subset expressing multiple markers is being differentially expressed, differential expression based on the marginal cell counts should also be detected.
+As such, identifying poly-functional cytokine profiles from ICS data can be done in an iterative way.
+First univariate tests on marginal populations are performed and then specific cell subsets expressing the positive markers detected are then tested.
+Even in this situation, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that needs to be tested, and a multivariate approach might be preferable.
+In this case, as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\cite{Nason:2006dx}.
 
-\section{Conclusions}
-We have developed a mixture model framework that we call MIMOSA .for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays that frequently arise in the context of vaccine trials and immunological studies.
-Our model can be applied in a univariate or multivariate fashion, and we have shown that it has greater sensitivity and specificity than other typical approaches used to analyze the type of count--based data derived from single cell assays, including Fisher's exact test, the non-Bayesian likelihood ratio test, or ad--hoc rank based approaches.
-The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
+When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
+In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression.
+Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).
+Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
+In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model.
+The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$.
+As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters.
+A solution would be to explore alternative model parametrizations that could be used to reduce the number of required parameters.
+For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
+As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others.
+More exploratory work will need to be done in this area once high dimensional single cell level data with large number of samples become available.
 
+All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
+
+\section{Supplementary Materials}
+Web Appendices, Tables, and Figures (limit to the types of material actually in question) referenced in Sections (give section numbers) are available in the attached Web-based supplementary material.
+% This will be replaced at submission time
+%at the Biometrics website on Wiley Online Library.
+\section{Acknowledgments}
 %\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}

### 2012-07-25 MIMOSA.tex
Updated FDR curves for simulations and updated results text.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
-\title{Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies}
+\title{Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
 \author[1]{Greg Finak}
@@
 \begin{abstract}
-Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single cell assays.
+Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single--cell assays.
 The characterization of these small cell subsets is crucial to decipher system level biological changes.
@@
 Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays.
-Our model allows the inference to be subject specific, while borrowing strength across subjects\footnote{Should we use subject, indivuduals or something else?
-What about animal studies?} through common prior distributions.
+Our model allows the inference specific to each observation, while borrowing strength across observations through common prior distributions.
 We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
@@
 In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
-Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers).
+Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single--cell measurement of multiple surface and intracellular markers (up to 17 markers).
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}.
@@
 While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references).
-This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data.
-Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general.
+This heterogeneity may carry important information, thus single--cell expression data should not be analyzed in the same fashion as cell-population level data.
+Special treatment of single--cell level data, which preserves information about population heterogeneity, is warranted in general.
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
@@
 For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology.
-This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single cell level.
+This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single--cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers.
@@
 
-Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that.
-Added ref for the HVTN054 trial and the ICS standardization paper from Horton})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej}.
-All of these methods analyze each individual separately, and no information is shared across individuals even though one could expect some similarities across responders (\textit{resp.} non-responders).
-In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
+Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}.
+All of these methods test observations separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders).
+%In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
 Let's update this sentence later}.
@@
 The framework developed in this paper addresses these issues explicitly.
-In our model, cell counts are modeled via a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by mean of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
+In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
 In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals.
-This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell\footnote{Let's replace all single cell by single-cell} assays, and increase sensitivity and specificity when detecting responders.
+This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
 Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\cite{Nason:2006dx}.
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{sec:data}, and present the results of a simulation study based on the ICS data.
+The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold--change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.
+
 %The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation.
@@
 \subsection{ICS}
-Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, ranked log fold change\footnote{Why ranked?}, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
+Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold--change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
-We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?}.
-The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
+We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?
+Didn't look at CD8's}.
+The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold--change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
 In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B).
-These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}) \footnote{Perhaps we could only keep the Env stimulation}.
+These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}).
 
@@
 In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation.
-The results presented in Figure~\ref{fig:fluidigm} A-C show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm}).
-These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions (Figure~\ref{fig:fluidigm} B,C).
-A similar analysis using a two-sided Fisher's exact test (Figure xx) does not reveal any stimulation specific changes of expression.\footnote{Perhaps it would be good to compare the number of differentially expressed genes}.
+The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm} A,B).
+These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
+A similar analysis using a two-sided Fisher's exact test and clustering the signed q--values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation--specific patterns.
+Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
 
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to data from Gag1-stimulated, CD4-positive, IL2-expressing T-cells from day 28 post--vaccination of the HVTN054 trial~\cite{Peiperl:2010ej}.
-We simulated data from this constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, with ten independent realizations of data for each $N$.
-The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV--1--stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial~.
+We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000, and 50,000 events, with ten independent realizations of data for each $N$.
+The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold--change.
 This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D).
-In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} E-F).
+In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} A,B).
 
-For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, D).
-Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test (Figure~\ref{fig:simulations} E, F).
+For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Supplementary Figure~\ref{fig:simulations_trunc} A).
+Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Supplementary Figure~\ref{fig:simulations_trunc} B).
 %Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc}).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc} C,D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
@@
 Our beta-binomial model described in Section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
-As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.\footnote{At some point we should say that such multivariate models are not appropriate for one sided models.}
+As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
 %Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
@@
 
-\subsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
+\subsection{Polyfunctionality in Fluidigm Single--Cell Gene Expression Data}
 As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
@@
 Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible poly-functional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
 That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.
@@
 Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
-Single cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single cell).
-As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important.
+Single--cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single--cell).
+As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single--cell level is becoming increasingly important.
 Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small.
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
-In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single cell assay.
+In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single--cell assay.
 
@@
 As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others.
-More exploratory work will need to be done in this area once high dimensional single cell level data with large number of samples become available.
+More exploratory work will need to be done in this area once high dimensional single--cell level data with large number of samples become available.
 
@@
 \end{tikzpicture}
-\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
+\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array.
 The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
@@
 Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
-C) Genes selected from Fisher's exact test at the 10\% FDR level.
-Clustering of the maximum likelihood estimates of difference in proportions of cells expressing a gene in stimulated vs.
-control samples with significant genes selected using Fisher's exact test.}
+C) Clustering of the signed q--values from Fisher's exact test.
+Genes selected from Fisher's exact test at the 10\% FDR level.
+}
 \label{fig:fluidigm}
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
-    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.5\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
     \node at (0,3.4) [font=\tiny\sffamily] {A} ;
@@
 A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
-Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and ranked log fold--change.
+Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and log fold--change.
 Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in the Supplementary Material.}\label{fig:simulations}
@@
 %%   \label{fig:icsdata}
-%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array.
+%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array.
 The posterior probability of response times the sign of the change in expression is shown in the first panel (red indicates a significant decrease, green a significant increase, relative to the control).
@@
 \section*{Supplementary Information}
-\subsection{HVTN065 and HVTN054 Vaccine Trial ICS Data Description}
+\subsection{HVTN065 Vaccine Trial ICS Data Description}
 \label{supp:statpublished}
@@
 $$
-foo
+f
 $$
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/16}};
+    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/15}};
     \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/3}};

### 2012-07-26 MIMOSA.tex
Updated figure layouts so they work with the biometrics template. Converted to biometrics template.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
-\documentclass[11pt]{article}
-\usepackage{geometry,authblk,multicol,amsmath}
-\usepackage[square,numbers,sort&compress]{natbib}
-\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
+%\documentclass[11pt]{article}
+\documentclass[useAMS,usenatbib]{biom}
+\def\bSig\mathbf{\Sigma}
+\newcommand{\VS}{V\&S}
+\newcommand{\tr}{\mbox{tr}}
+%\usepackage{geometry,authblk,multicol,amsmath}
+\usepackage{amsmath,tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+
+%\usepackage[square,numbers,sort&compress]{natbib}
+%\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
+%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
 %\geometry{landscape}                % Activate for for rotated page geometry
 %\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}
+%\usepackage{graphicx}
 \usepackage{amssymb}
@@
 
-\title{Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
-\author[1]{Greg Finak}
-\author[2]{SC De Rosa}
-\author[3]{Mario Roederer}
-\author[1]{Raphael Gottardo}
+\title[MIMOSA: Combinatorial Mixture Models or Single--Cell Assays]{Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
+%\author[1]{Greg Finak}
+%\author[2]{SC De Rosa}
+%\author[3]{Mario Roederer}
+%\author[1]{Raphael Gottardo}
 
-\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
-\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
-\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
+%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
+%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
+%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
+\author{Greg Finak$^{1,*}$\email{gfinak@fhcrc.org},
+SC De Rosa$^{2,**}$\email{sderosa@fhcrc.org},
+Mario Roederer$^{3,***}$\email{marior@mail.nih.gov},
+Raphael Gottardo$^{1,****}$\email{rgotard@fhcrc.org}\\
+$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
+$^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
+$^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
 
-\date{\today}
+%\date{\today}
 
 \begin{document}
-\maketitle
+%\maketitle
+\date{{\it Received August} 2012.}
+
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
+\volume{ }
+\pubyear{ }
+\artmonth{ }
+
+\doi{12.1234/j.12345-1234.2012.12345.x}
+
+\label{firstpage}
 
@@
 
+\begin{keywords}
+Mixture modelling, hierarchical modelling, bayesian modelling, single--cell assays, immunology
+\end{keywords}
+
+\maketitle
+
 \section{Introduction}
+\label{s:intro}
 Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
@@
 A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
-While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references).
+While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
 This heterogeneity may carry important information, thus single--cell expression data should not be analyzed in the same fashion as cell-population level data. 
@@
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
-These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
@@
 A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
+
 A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine.
@@
  While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. 
-After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}.
+ After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}.
  These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells.
+
 The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
@@
 
-Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}.
+Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}.
 All of these methods test observations separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders).
@@
 This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
-Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\cite{Nason:2006dx}.
+Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
 
 \section{Data structure and notation} 
-\label{sec:data}
+\label{s:data}
 In this paper we consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.
@@
 Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. 
+The data is described in Web Appendix A.
 
@@
 \section{Differential expression with one marker}
+\label{s:DEone}
 Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
@@
 
-\begin{table}[ht]
+\begin{table}
 \centering
@@
 \end{table}
+
 For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
@@
 \subsection{Beta-binomial model}
-\label{sec:DE}
+\label{s:DE}
 For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
 
-\begin{equation}
+\begin{equation*}
 (n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
-\end{equation}
+\end{equation*}
 where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
-\begin{equation}
-{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:beta_prior}
-\end{equation}
+\begin{equation*}
+{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
+\end{equation*}
 Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal.
@@
 In some study, such as the ICS studies used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
-This alternative parametrization is described in Web Supplementary material\footnote{let's make sure we use the proper format for referencing the supplementary material}, and we refer to it as the one-sided model.
+This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.
 
 \subsection{Priors}
+\label{ss:priors}
 Our model shares information across all individuals using exchangeable Beta priors on the unknown proportions, as follows, 
- \begin{align}
-(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior-null}\\
-(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u) \label{eq:prior-alternative}
- \end{align}
+ \begin{align*}
+(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
+(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u)
+ \end{align*}
 where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.
@@
 It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$.
-Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\citep{Dempster:1977ul} algorithm as described in Section \ref{s:estimation}.
 As an alternative, we also describe a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
-In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{sec:estimation}.
+In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{s:estimation}.
 
 \subsection{Parameter estimation}
-\label{sec:estimation}
+\label{s:estimation}
 Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Supplementary material), and are given by,
- \begin{align}
-  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
-	&=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
- \end{align} 
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Appendix XX), and are given by,
+ \begin{align*}
+ \begin{split}
+  	&\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
+	=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
+	\end{split}
+ \end{align*}
+
 and
@@
 	\begin{split}
-\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
-=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}.\label{model2:unconstrained}
+&\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
+=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
+\label{model2:unconstrained}
 \end{split}
@@
 \begin{equation}
-l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w)
-\label{eq:cll}
+\begin{split}
+l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w)\label{eq:cll}
+\end{split}
 \end{equation}
@@
 In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed form but can easily be estimated.
-All calculations for the one-sided case are described in Supplementary Web material.
+All calculations for the one-sided case are described in Web Appendix XX.
 
-\subsubsection{EM algoritm}
+\subsection{EM algoritm}
 Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
 \[
-\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
+\begin{split}
+&\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\\ &\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
+\end{split}
 \] 
@@
 
-\subsubsection{MCMC algorithm}
-Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand:1996wc}.
+\subsection{MCMC algorithm}
+Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
-Details about the algorithms are given in supplementary material.
-We used the method of Raftery and Lewis \citep{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
+Details about the algorithms are given in Web Appendix XX.
+We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
 For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
@@
 \section{Results}
-In this section, we apply our MIMOSA model to the data described in Section \ref{sec:data}, and present the results of a simulation study based on the ICS data.
+\label{s:results}
+In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
 The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold--change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.
@@
 In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B).
-These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figure~\ref{fig:HVTN065Results}).
+These results are consistent for other cytokines (see Web Figure ~\ref{webfig:HVTN065Results}).
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto, node distance=0cm]
+\node at (0,0) (A){
+    \begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=.6\columnwidth]{Figures/2}};
+    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {A} ;
+    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
+    \end{scope}
+\end{tikzpicture}
+ };
+ \node [below=of A] (B) {
+ \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=.6\columnwidth]{Figures/1}};
+    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {C} ;
+    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
+    \end{scope}
+    \end{tikzpicture}
+ };
+\end{tikzpicture}
+   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065.
+Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) ENV-1-PTEG and expressing IL2.
+ROC and FDR plots of other cytokine combinations can be found in Web Figure ~\ref{webfig:HVTN065Results}.}
+\label{fig:HVTN065}
+\end{figure}
 
@@
 Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
+
+\begin{figure}
+\centering
+\begin{tikzpicture} [auto,node distance=0cm]
+\node at (0,0) (A) {
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+\node at (0,1) [font=\tiny\sffamily] {A} ;
+\end{scope}
+\end{tikzpicture}
+};
+\node [right=of A] (B) {
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+\end{tikzpicture}
+};
+\node [right=of B] (C) {
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
+\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+\node at (0,1) [font=\tiny\sffamily] {C} ;
+\end{scope}
+\end{tikzpicture}
+};
+\end{tikzpicture}
+\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array.
+The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
+Columns and rows are clustered based on these signed posterior probabilities.
+B) The posterior differences in proportion of cells expressing a gene in the stimulated vs.
+control samples.
+Rows and columns are ordered as in A) for comparison.
+The traces show the deviations of each cell from zero.
+Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
+C) Clustering of the signed q--values from Fisher's exact test.
+Genes selected from Fisher's exact test at the 10\% FDR level.
+}
+\label{fig:fluidigm}
+\end{figure}
 
@@
 
-For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Supplementary Figure~\ref{fig:simulations_trunc} A).
-Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Supplementary Figure~\ref{fig:simulations_trunc} B).
+For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Web Figure~\ref{webfig:simulations_trunc} panel A).
+Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Web Figure~\ref{webfig:simulations_trunc} panel B).
 %Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
 
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+   \begin{tikzpicture} [auto, node distance=0cm]
+   \node at (0,0) (A) {
+\begin{tikzpicture}
+    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
+    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {A} ;
+        \end{scope}
+    \end{tikzpicture}
+    };
+    \node [right=of A] (B) {
+    \begin{tikzpicture}
+    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
+    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {B} ;
+    \end{scope}
+    \end{tikzpicture}
+    };
+\end{tikzpicture}
+   \caption{Comparison of positivity detection methods on data simulated from the one--sided model.
+Ten simulations were generated at an $N$ of 5,000 total counts using hyper--parameter estimates from real ICS data (IFNg expressing CD4+ T--cells stimulated with ENV-1-PTEG from HVTN065) with a five--fold effect size between responder and non--responder components.
+A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
+Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and log fold--change.
+Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure~\ref{webfig:simulations_trunc}.}\label{fig:simulations}
+\end{figure}
+
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc} C,D).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (see Web Figure~\ref{webfig:simulations_trunc} panels C and D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
@@
 \section{Differential expression across marker combinations}
-Our beta-binomial model described in Section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
+\label{s:demarkercombos}
+Our beta-binomial model described in Section \ref{s:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
 As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
@@
 In our multivariate model, the beta distribution is replaced by a multinomial distribution, as follows,
-\begin{equation}
+\begin{equation*}
  (\mathbf{n}_{ui}|\mathbf{p}_{ui},) \sim \mathcal{M}(N_{ui},\mathbf{p}_{ui})\quad\text{and}\quad (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(N_{si},\mathbf{p}_{si})\label{eq:mult_likeliehood}
- \end{equation}
+ \end{equation*}
 %; \text{if $i$ is a non--responder}\\
@@
 As in the one-marker case, we share information across individuals using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
-\begin{align}
+\begin{align*}
 (\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u)\\\nonumber
-(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s)\label{eq:dir_prior}
-\end{align}
-where the indicator variable $z_i$ is as defined in \eqref{eq:beta_prior}, i.e.
+(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s)%\label{eq:dir_prior}
+\end{align*}
+where the indicator variable $z_i$ is as defined in \ref{ss:priors}, i.e.
 $z_i\sim\mathrm{Be}(w)$ where $w$ is the proportion of responders.
@@
 \subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Supplementary Material).
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Web Appendix XX).
 For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
-\begin{align}
+\begin{align*}
 \mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
-\end{align}
+\end{align*}
 where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$.
 Similarly the marginal likelihood for the alternative model is given by
-\begin{align}
-\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
-\end{align}
+\begin{align*}
+\begin{split}
+\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) =& \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \\ &\frac{N_{ui}!}{\prod_k n_{uik}!}.%\label{eq:postmult}
+\end{split}
+\end{align*}
 The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.
@@
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
-More details about both algorithms are given in Web Supplementary material.
+More details about both algorithms are given in Web Appendix XX.
 
@@
 
+\begin{figure}
+\centering
+\begin{tikzpicture} [auto, node distance=0cm]
+\node at (0,0) (A) {
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
+\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+\node at (0,1) [font=\tiny\sffamily] {A} ;
+\end{scope}
+\end{tikzpicture}
+};
+\node [right=of A] (B){
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
+\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+\node at (0,1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+\end{tikzpicture}
+};
+%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+%\node at (0,0) [font=\tiny\sffamily] {C} ;
+%\node at (6.25,0) [font=\tiny\sffamily] {D} ;
+\end{tikzpicture}
+\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
+No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
+Sample names highlighted in red identify those where MIMOSA did not detect a difference.}
+\label{fig:polyfunctionality}
+\end{figure}
+
 Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we could not look at more than two makers at once.
@@
 
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto, node distance=0cm]
+\node at (0,0) (A) {
+\begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
+\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {A} ;
+\end{scope}
+\end{tikzpicture}
+};
+\node [right=of A] (B) {
+\begin{tikzpicture}
+   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
+   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+       \node at (0,1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+\end{tikzpicture}
+};
+\end{tikzpicture}
+   \caption{Multivariate simulations from a two--sided model.
+Ten, eight-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
+Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value (combining Fisher's exact test run marginally on each of the five dimensions).
+A) Average ROC curves for the competing methods over 10 simulations.
+B) Average observed and nominal false discovery rate for each method over 10 simulations.}
+   \label{fig:mvsimulations}
+\end{figure}
+
 \section{Discussion}
+\label{s:discussion}
 Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
@@
 Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small.
-Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
+Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
 In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single--cell assay.
@@
 The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental individuals across multiple conditions (\textit{i.e.} vaccine responders and non-responders).
-Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure~\ref{fig:simulations_trunc}).
+Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Web Figure~\ref{webfig:simulations_trunc}).
 The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
@@
 Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
-In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or simply using this information for pre-filtering \cite{Flatz:2011jb}.
+In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or simply using this information for pre-filtering~\citep{Flatz:2011jb}.
 The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).
 
-Detecting differences in poly-functional cell populations (\textit{i.e.} identifying changes in cell populations that co-express multiple proteins, cytokines, or genes), is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\cite{Milush:2009bz}.
+Detecting differences in poly-functional cell populations (\textit{i.e.} identifying changes in cell populations that co-express multiple proteins, cytokines, or genes), is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\citep{Milush:2009bz}.
 In the context of HIV, these poly-functional cell populations have been shown to be correlated with good outcome such as vaccine protection and long-term disease non-progression (ADD REFERENCES TO THE PAPERS STEVE SUGGESTED IN THE GRANT).
@@
 Even in this situation, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that needs to be tested, and a multivariate approach might be preferable.
-In this case, as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\cite{Nason:2006dx}.
+In this case, as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\citep{Nason:2006dx}.
 
@@
 
-\section{Supplementary Materials}
-Web Appendices, Tables, and Figures (limit to the types of material actually in question) referenced in Sections (give section numbers) are available in the attached Web-based supplementary material.
-% This will be replaced at submission time
-%at the Biometrics website on Wiley Online Library.
-\section{Acknowledgments}
 %\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
@@
 %
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
-%   \includegraphics[width=.75\columnwidth]{Figures/12} 
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
-    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/1}};
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (0,-0.25) [font=\tiny\sffamily] {C} ;
-    \node at (4.6,-0.25) [font=\tiny\sffamily] {D} ;
-\end{tikzpicture}
-   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065.
-Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) ENV-1-PTEG and expressing IL2.
-ROC and FDR plots of other cytokine combinations can be found in the supplementary material.}
-\label{fig:HVTN065}
-\end{figure}
-
 %   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data.
@@
 
-\begin{figure}[htbp]
-\centering
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
-\node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
-\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
-\node at (0,3) [font=\tiny\sffamily] {A} ;
-\node at (4.5,3) [font=\tiny\sffamily] {B} ;
-\node at (8.95,3) [font=\tiny\sffamily] {C} ;
-\end{tikzpicture}
-\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array.
-The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
-Columns and rows are clustered based on these signed posterior probabilities.
-B) The posterior differences in proportion of cells expressing a gene in the stimulated vs.
-control samples.
-Rows and columns are ordered as in A) for comparison.
-The traces show the deviations of each cell from zero.
-Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
-C) Clustering of the signed q--values from Fisher's exact test.
-Genes selected from Fisher's exact test at the 10\% FDR level.
-}
-\label{fig:fluidigm}
-\end{figure}
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
-    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-\end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one--sided model.
-Ten simulations were generated at an $N$ of 5,000 total counts using hyper--parameter estimates from real ICS data (IFNg expressing CD4+ T--cells stimulated with ENV-1-PTEG from HVTN065) with a five--fold effect size between responder and non--responder components.
-A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
-Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and log fold--change.
-Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in the Supplementary Material.}\label{fig:simulations}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
-\node[anchor=south west, inner sep=0] at (6.5,0) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
-%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
-%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
-\node at (0,4.5) [font=\tiny\sffamily] {A} ;
-\node at (6.25,4.5) [font=\tiny\sffamily] {B} ;
-%\node at (0,0) [font=\tiny\sffamily] {C} ;
-%\node at (6.25,0) [font=\tiny\sffamily] {D} ;
-\end{tikzpicture}
-\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
-No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
-Sample names highlighted in red identify those where MIMOSA did not detect a difference.}
-\label{fig:polyfunctionality}
-\end{figure}
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
-   \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
-    \node at (0,3.5) [font=\tiny\sffamily] {A} ;
-    \node at (6,3.5) [font=\tiny\sffamily] {B} ;
-\end{tikzpicture}
-   \caption{Multivariate simulations from a two--sided model.
-Ten, eight-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
-Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value (combining Fisher's exact test run marginally on each of the five dimensions).
-A) Average ROC curves for the competing methods over 10 simulations.
-B) Average observed and nominal false discovery rate for each method over 10 simulations.}
-   \label{fig:mvsimulations}
-\end{figure}
-\clearpage
 %
@@
 
-\renewcommand{\thesection}{S.\arabic{section}}
-\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
-\setcounter{section}{1}
-\setcounter{subsection}{0}
-\renewcommand{\figurename}{Supplementary Figure}
-\setcounter{figure}{0}
-%\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}
+\backmatter
 
-\section*{Supplementary Information}
-\subsection{HVTN065 Vaccine Trial ICS Data Description}
+\section*{Supplementary Materials}
+
+Web Appendices A and B, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone} are available in the attached Web-based supplementary material.\vspace*{-8pt}
+% This will be replaced at submission time
+%at the Biometrics website on Wiley Online Library.
+
+\section*{Acknowledgments}
+We wish to acknowledge ...
+\appendix
+
+\section{}
+\subsection{Web Appendix A: HVTN065 Vaccine Trial ICS Data Description}
 \label{supp:statpublished}
@@
 
-\subsection{Constrained Beta--binomial model}
+\subsection{Web Appendix B: Constrained Beta--binomial model}
 \label{supp:constrained}
@@
 
-\begin{align}
+\begin{align*}
 \begin{split}
 	 \mathrm{C}=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
-\label{eq:model2post}
+%\label{eq:model2post}
 \end{split}
-\end{align}
+\end{align*}
 
@@
 
-\subsection{HVTN065 Results for Other Cytokines}
-\label{supp:HVTN065Results}
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%\renewcommand{\thesection}{S.\arabic{section}}
+%\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
+%\setcounter{section}{1}
+%\setcounter{subsection}{0}
+%\setcounter{figure}{0}
+%\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}
+\subsection{Web Figure A: HVTN065 Results for Other Cytokines}
+\begin{figure} %  figure placement: here, top, bottom, or page
    \centering
@@
 %   \includegraphics[width=.75\columnwidth]{Figures/12} 
+\begin{tikzpicture} [auto, node distance=0cm]
+\node at (0,0) (A){
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.5\columnwidth]{Figures/15}};
-    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/3}};
-    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/5}};
-    \node[anchor=south west, inner sep=0] at (8,-3.75){\includegraphics[width=.5\columnwidth]{Figures/6}};
-    \node[anchor=south west, inner sep=0] at (0,-7.5){\includegraphics[width=.5\columnwidth]{Figures/7}};
+    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=.5\columnwidth]{Figures/15}};
+    \begin{scope}[x={(a.south east)},y={(a.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {A} ;
+        \end{scope}
+        \end{tikzpicture}
+ };
+ \node [right=of A] (B) {
+ \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
+        \begin{scope}[x={(b.south east)},y={(b.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {B} ;
+                \end{scope}
+        \end{tikzpicture}
+ };
+ \node [below=of A] (C) {
+ \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
+        \begin{scope}[x={(c.south east)},y={(c.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {C} ;
+        \end{scope}
+        \end{tikzpicture}
+};
+\node [below=of B] (D) {
+\begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
+        \begin{scope}[x={(d.south east)},y={(d.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {D} ;
+                \end{scope}
+        \end{tikzpicture}
+};
+\node [below=of C] (E) {
+\begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
+        \begin{scope}[x={(d.south east)},y={(d.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {E} ;
+                \end{scope}
+        \end{tikzpicture}
+ };
    % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
@@
 
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
-    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
-    \node at (8,-0.5) [font=\tiny\sffamily] {D} ;
-    \node at (0,-4) [font=\tiny\sffamily] {E} ;
   %  \node at (8,-4) [font=\tiny\sffamily] {F} ;
@@
    \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
-\label{fig:HVTN065Results}
+\label{webfig:HVTN065Results}
 \end{figure}
@@
 %\end{figure}
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+\subsection{Web Figure B: Other Simulation Results}
+\begin{figure} %  figure placement: here, top, bottom, or page
    \centering
+\begin{tikzpicture} [auto,node distance=0cm]
+\node at (0,0) (A) {
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
-    \node[anchor=south west, inner sep=0] at (6,0) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};        \node[anchor=south west,inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
-    \node[anchor=south west, inner sep=0] at (6,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
-    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
-        \node at (0,0) [font=\tiny\sffamily] {C} ;
-    \node at (4.6,0) [font=\tiny\sffamily] {D} ;
+    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+    \begin{scope} [x={(a.south east)},y={(a.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {A} ;
+    \end{scope}
+    \end{tikzpicture}
+    };
+    \node [right=of A] (B) {
+    \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
+        \begin{scope} [x={(b.south east)},y={(b.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {B} ;
+        \end{scope}
+    \end{tikzpicture}
+    };
+    \node [below=of A] (C) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {C} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+     \node [below=of B] (D) {
+     \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+        \begin{scope} [x={(d.south east)},y={(d.north west)}]
+            \node at (0,1) [font=\tiny\sffamily] {D} ;
+\end{scope}
+    \end{tikzpicture}
+    };
  \end{tikzpicture}
@@
 }
-   \label{fig:simulations_trunc}
+   \label{webfig:simulations_trunc}
 \end{figure}
@@
 %\subsection{Derivation of the Beta--Binomial Model for $p_s=p_u$ and $p_s>p_u$}
-%\label{sec:derivation}
+%\label{s:derivation}
 %
@@
 We model this situation explicitly using a mixture--model framework.
-\clearpage
-\bibliographystyle{unsrtnat}
+\bibliographystyle{mn2e}
 \bibliography{MIMOSA}
+\label{lastpage}
 

### 2012-07-26 MIMOSA.tex
Set to one-column layout. split some equations over multiple lines so they fit on one or two column layout.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>
Added simulations for N=1000 in the appendix. Added refs
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 %\documentclass[11pt]{article}
-\documentclass[useAMS,usenatbib]{biom}
+\documentclass[useAMS,referee,usenatbib]{biom}
 \def\bSig\mathbf{\Sigma}
@@
 Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Appendix XX), and are given by,
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Appendix C), and are given by,
  \begin{align*}
@@
 In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed form but can easily be estimated.
-All calculations for the one-sided case are described in Web Appendix XX.
+All calculations for the one-sided case are described in Web Appendix B.
 
@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
-Details about the algorithms are given in Web Appendix XX.
+Details about the algorithms are given in Web Appendix D.
 We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
@@
 \subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Web Appendix XX).
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Web Appendix E).
 For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
-More details about both algorithms are given in Web Appendix XX.
+More details about both algorithms are given in Web Appendix D.
 
@@
 Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
-Single--cell assays will become even more routine once sequencing at the single-cell level becomes practical (Ref to RNA-Seq single--cell).
+Single--cell assays will become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}.
 As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single--cell level is becoming increasingly important.
@@
 Detecting differences in poly-functional cell populations (\textit{i.e.} identifying changes in cell populations that co-express multiple proteins, cytokines, or genes), is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\citep{Milush:2009bz}.
-In the context of HIV, these poly-functional cell populations have been shown to be correlated with good outcome such as vaccine protection and long-term disease non-progression (ADD REFERENCES TO THE PAPERS STEVE SUGGESTED IN THE GRANT).
+In the context of HIV, these poly-functional cell populations have been shown to be correlated with good outcome such as vaccine protection and long-term disease non-progression~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
 In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected.
@@
 
-Web Appendices A and B, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone} are available in the attached Web-based supplementary material.\vspace*{-8pt}
+Web Appendices A, B, C, D, and E, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:estimation},  and \ref{s:demarkercombos} are available in the attached Web-based supplementary material.\vspace*{-8pt}
 % This will be replaced at submission time
@@
 $$
+
+\subsection{Web Appendix C: Derivation of Marginal Likelihoods for Beta--Binomial Model}
+
+\subsection{Web Appendix D: The MCMC Algorithm}
+
+\subsection{Web Appendix E: Marginal Likelihoods for the Multivariate Model}
 
@@
 %\end{figure}
-\subsection{Web Figure B: Other Simulation Results}
+\subsection{Web Figure B: Additional Simulation Results}
 \begin{figure} %  figure placement: here, top, bottom, or page
@@
     };
- \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to two--sided data (A-B) or to data where model assumptions are violated (C-D).
+  \node [below=of C] (E) {
+\begin{tikzpicture}
+    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
+    \begin{scope} [x={(e.south east)},y={(e.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {E} ;
+    \end{scope}
+    \end{tikzpicture}
+    };
+    \node [right=of E] (F) {
+    \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
+        \begin{scope} [x={(f.south east)},y={(f.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {F} ;
+        \end{scope}
+    \end{tikzpicture}
+    };
+    \node [below=of E] (G) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
+         \begin{scope} [x={(g.south east)},y={(g.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {G} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+     \node [below=of F] (H) {
+     \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
+        \begin{scope} [x={(h.south east)},y={(h.north west)}]
+            \node at (0,1) [font=\tiny\sffamily] {H} ;
+\end{scope}
+    \end{tikzpicture}
+    };
+ \end{tikzpicture}
+   \caption{Unconstrained MIMOSA model fit to two--sided data (A,B,E,F) or to data where model assumptions are violated (C,D,G,H).
 Two--sided data were simulated from a standard model.
-A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations at N=5,000.
+A) Average ROC curves and B) average observed vs nominal FDR from 10 simulations with N=5,000, and E-F) N=1,000.
 Unconstrained MIMOSA fit to two--sided data simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
-C) Average ROC curves from 10 simulations and D) average observed vs nominal FDR from 10 simulations at N=5,000.
+C) Average ROC curves and D) average observed vs nominal FDR from 10 simulations at N=5,000, and G-H) N=1,000.
 }

### 2012-07-26 MIMOSA.tex
Split Web Supplementary material into another document
Also corrected a bit of text describing the polyfunctionalilty results

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \textit{Flow cytometry}:
-The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see supplementary material~\ref{supp:statpublished}).
+The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see Web Appendix A). 
 The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
@@
  \begin{split}
-  	&\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
-	=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
+  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
+	=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
 	\end{split}
@@
 	\begin{split}
-&\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
-=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
+\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
+=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
 \label{model2:unconstrained}
@@
 The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold--change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
-In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B).
-These results are consistent for other cytokines (see Web Figure ~\ref{webfig:HVTN065Results}).
+In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).
 
@@
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) ENV-1-PTEG and expressing IL2.
-ROC and FDR plots of other cytokine combinations can be found in Web Figure ~\ref{webfig:HVTN065Results}.}
+ROC and FDR plots of other cytokine combinations can be found in Web Figure A.}
 \label{fig:HVTN065}
@@
 
-For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Web Figure~\ref{webfig:simulations_trunc} panel A).
-Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Web Figure~\ref{webfig:simulations_trunc} panel B).
+For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Web Figure B, panel A).
+Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Web Figure B panel B).
 %Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
@@
 Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and log fold--change.
-Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure~\ref{webfig:simulations_trunc}.}\label{fig:simulations}
+Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.}\label{fig:simulations}
 \end{figure}
@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (see Web Figure~\ref{webfig:simulations_trunc} panels C and D).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (see Web Figure B panels C and D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
@@
 In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
-Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation.
+Only CCL5 positive cells express BIRC3, and the its expression increases upon stimulation.
 The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
 That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.
-There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C).
-In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D).
+When marginalizing over these cell populations, no difference is apparent in any of the samples.
+In contrast, multivariate MIMOSA tests all polyfunctional cell subpopulations simultaneously, identifying significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels).
 Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.
@@
 The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental individuals across multiple conditions (\textit{i.e.} vaccine responders and non-responders).
-Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Web Figure~\ref{webfig:simulations_trunc}).
+Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Web Figure B).
 The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
@@
 We wish to acknowledge ...
-\appendix
-
-\section{}
-\subsection{Web Appendix A: HVTN065 Vaccine Trial ICS Data Description}
-\label{supp:statpublished}
-HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B).
-CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay.
-Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays.
-Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools.
-Results of the trial have been published~\citep{Goonetilleke:2006jk}.
-
-%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
-The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.
-Vaccine was given at two increasing doses, as well as a placebo.
-T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}.
-The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}.
-The sample size consisted of 20 vaccine and four placebo recipients.
-The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}.
-The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.
-
-\subsection{Web Appendix B: Constrained Beta--binomial model}
-\label{supp:constrained}
-We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$.
-In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in equation~\eqref{model2:unconstrained}.
-The marginal likelihood for the constrained model becomes $\mathrm{L}_1\times \mathrm{C}$, where $\mathrm{C}$ is a ratio of normalizing constants for the likelihood and the prior.
-
-\begin{align*}
-\begin{split}
-	 \mathrm{C}=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
-%\label{eq:model2post}
-\end{split}
-\end{align*}
-
-The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$ is the regularized incomplete Beta function, which is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional integral that can be computed via Monte--Carlo integration (i.e.
-in our EM implementation).
-However, this can be computationally costly in the MCMC framework.
-In the latter case, we can compute the integral exactly if $\alpha_s$ and $\beta_s$ are integers by using the identity:
-\begin{align}
-I_{1-p_u}(\beta_s,\alpha_s) = \sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^j(p_u)^{\beta_s+\alpha_s-j}
-\label{eq:IBident}
-\end{align}
-Typically, in ICS data $\beta_s>>\alpha_s$, leading to relatively few terms in the sum in equation~\ref{eq:IBident}.  This leads to an exact expression for the normalizing constant:
-$$
-f
-$$
-
-\subsection{Web Appendix C: Derivation of Marginal Likelihoods for Beta--Binomial Model}
-
-\subsection{Web Appendix D: The MCMC Algorithm}
-
-\subsection{Web Appendix E: Marginal Likelihoods for the Multivariate Model}
-
-%\renewcommand{\thesection}{S.\arabic{section}}
-%\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
-%\setcounter{section}{1}
-%\setcounter{subsection}{0}
-%\setcounter{figure}{0}
-%\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}
-\subsection{Web Figure A: HVTN065 Results for Other Cytokines}
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
-%   \includegraphics[width=.75\columnwidth]{Figures/12} 
-\begin{tikzpicture} [auto, node distance=0cm]
-\node at (0,0) (A){
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=.5\columnwidth]{Figures/15}};
-    \begin{scope}[x={(a.south east)},y={(a.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {A} ;
-        \end{scope}
-        \end{tikzpicture}
- };
- \node [right=of A] (B) {
- \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
-        \begin{scope}[x={(b.south east)},y={(b.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {B} ;
-                \end{scope}
-        \end{tikzpicture}
- };
- \node [below=of A] (C) {
- \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
-        \begin{scope}[x={(c.south east)},y={(c.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {C} ;
-        \end{scope}
-        \end{tikzpicture}
-};
-\node [below=of B] (D) {
-\begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
-        \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {D} ;
-                \end{scope}
-        \end{tikzpicture}
-};
-\node [below=of C] (E) {
-\begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
-        \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {E} ;
-                \end{scope}
-        \end{tikzpicture}
- };
-   % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
-   % \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
-    %\node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
-    %\node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
-    %\node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};
-
-  %  \node at (8,-4) [font=\tiny\sffamily] {F} ;
-%    \node at (0,-7.75) [font=\tiny\sffamily] {G} ;
-  %  \node at (8,-7.75) [font=\tiny\sffamily] {H} ;
-   % \node at (0,-11.75) [font=\tiny\sffamily] {I} ;
-   % \node at (8,-11.75) [font=\tiny\sffamily] {J} ;
-    \end{tikzpicture}
-   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
-\label{webfig:HVTN065Results}
-\end{figure}
-
-%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-%   \centering
-%%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
-%%   \includegraphics[width=.75\columnwidth]{Figures/12} 
-%\begin{tikzpicture}
-%    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=.5\columnwidth]{Figures/14}};
-%    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/15}};
-%    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/16}};
-%
-%
-%    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-%    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
-%    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
-%    \end{tikzpicture}
-%   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
-%\label{fig:HVTN065ResultsCont}
-%\end{figure}
-\subsection{Web Figure B: Additional Simulation Results}
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture} [auto,node distance=0cm]
-\node at (0,0) (A) {
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
-    \begin{scope} [x={(a.south east)},y={(a.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {A} ;
-    \end{scope}
-    \end{tikzpicture}
-    };
-    \node [right=of A] (B) {
-    \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
-        \begin{scope} [x={(b.south east)},y={(b.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {B} ;
-        \end{scope}
-    \end{tikzpicture}
-    };
-    \node [below=of A] (C) {
-    \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
-         \begin{scope} [x={(c.south east)},y={(c.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {C} ;
-	\end{scope}
-     \end{tikzpicture}
-     };
-     \node [below=of B] (D) {
-     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
-        \begin{scope} [x={(d.south east)},y={(d.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {D} ;
-\end{scope}
-    \end{tikzpicture}
-    };
-  \node [below=of C] (E) {
-\begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
-    \begin{scope} [x={(e.south east)},y={(e.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {E} ;
-    \end{scope}
-    \end{tikzpicture}
-    };
-    \node [right=of E] (F) {
-    \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
-        \begin{scope} [x={(f.south east)},y={(f.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {F} ;
-        \end{scope}
-    \end{tikzpicture}
-    };
-    \node [below=of E] (G) {
-    \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
-         \begin{scope} [x={(g.south east)},y={(g.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {G} ;
-	\end{scope}
-     \end{tikzpicture}
-     };
-     \node [below=of F] (H) {
-     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
-        \begin{scope} [x={(h.south east)},y={(h.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {H} ;
-\end{scope}
-    \end{tikzpicture}
-    };
- \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to two--sided data (A,B,E,F) or to data where model assumptions are violated (C,D,G,H).
-Two--sided data were simulated from a standard model.
-A) Average ROC curves and B) average observed vs nominal FDR from 10 simulations with N=5,000, and E-F) N=1,000.
-Unconstrained MIMOSA fit to two--sided data simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
-C) Average ROC curves and D) average observed vs nominal FDR from 10 simulations at N=5,000, and G-H) N=1,000.
-}
-   \label{webfig:simulations_trunc}
-\end{figure}
 

### 2012-07-26 WebSupplementaryMaterialMIMOSA.tex
Split Web Supplementary material into another document
Also corrected a bit of text describing the polyfunctionalilty results

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\documentclass{article}
\title{Web-based Supplementary Materials for Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies by Greg Finak, SC De Rosa, Mario Roederer and Raphael Gottardo}

\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{multirow,natbib}
\usepackage{url}
\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
\usepackage{amsmath,tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\date{}
\begin{document}
\maketitle
\appendix
%\renewcommand{\thesection}{\Alph{section}}
\renewcommand{\thesubsection}{Web Appendix \Alph{subsection}:}
%\setcounter{section}{1}
\setcounter{subsection}{0}
%\setcounter{figure}{0}
\renewcommand{\figurename}{\textbf{Web Figure}}
\renewcommand{\thefigure}{\textbf{\Alph{figure}}}
\setcounter{figure}{0}
%\todo[inline]{Note that the Supplementary Information section needs corrections to notation and overall}

\section*{}
\subsection{HVTN065 Vaccine Trial ICS Data Description}
\label{supp:statpublished}
HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B). CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay. Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools. Results of the trial have been published~\citep{Goonetilleke:2006jk}.

%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}. The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}. Vaccine was given at two increasing doses, as well as a placebo. T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}. The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}. The sample size consisted of 20 vaccine and four placebo recipients. The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}. The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.

\subsection{Constrained Beta--binomial model}
\label{supp:constrained}
We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$. In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in equation~1. The marginal likelihood for the constrained model becomes $\mathrm{L}_1\times \mathrm{C}$, where $\mathrm{C}$ is a ratio of normalizing constants for the likelihood and the prior.

\begin{align*}
\begin{split}
	 \mathrm{C}=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
%\label{eq:model2post}
\end{split}
\end{align*}

The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$ is the regularized incomplete Beta function, which is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional integral that can be computed via Monte--Carlo integration (i.e. in our EM implementation). However, this can be computationally costly in the MCMC framework. In the latter case, we can compute the integral exactly if $\alpha_s$ and $\beta_s$ are integers by using the identity:
\begin{align}
I_{1-p_u}(\beta_s,\alpha_s) = \sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^j(p_u)^{\beta_s+\alpha_s-j}
\label{eq:IBident}
\end{align}
Typically, in ICS data $\beta_s>>\alpha_s$, leading to relatively few terms in the sum in equation~\ref{eq:IBident}.  This leads to an exact expression for the normalizing constant:
$$
f
$$

\subsection{Derivation of Marginal Likelihoods for Beta--Binomial Model}

\subsection{The MCMC Algorithm}

\subsection{Marginal Likelihoods for the Multivariate Model}

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=.5\columnwidth]{Figures/15}};
    \begin{scope}[x={(a.south east)},y={(a.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {A} ;
        \end{scope}
        \end{tikzpicture}
 };
 \node [right=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
        \begin{scope}[x={(b.south east)},y={(b.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {B} ;
                \end{scope}
        \end{tikzpicture}
 };
 \node [below=of A] (C) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
        \begin{scope}[x={(c.south east)},y={(c.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {C} ;
        \end{scope}
        \end{tikzpicture}
};
\node [below=of B] (D) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
        \begin{scope}[x={(d.south east)},y={(d.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {D} ;
                \end{scope}
        \end{tikzpicture}
};
\node [below=of C] (E) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
        \begin{scope}[x={(d.south east)},y={(d.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {E} ;
                \end{scope}
        \end{tikzpicture}
 };
   % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
   % \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
    %\node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
    %\node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
    %\node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};

  %  \node at (8,-4) [font=\tiny\sffamily] {F} ;
%    \node at (0,-7.75) [font=\tiny\sffamily] {G} ;
  %  \node at (8,-7.75) [font=\tiny\sffamily] {H} ;
   % \node at (0,-11.75) [font=\tiny\sffamily] {I} ;
   % \node at (8,-11.75) [font=\tiny\sffamily] {J} ;
    \end{tikzpicture}
   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
\label{webfig:HVTN065Results}
\end{figure}

%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%%   \includegraphics[width=.75\columnwidth]{Figures/12} 
%\begin{tikzpicture}
%    \node[anchor=south west, inner sep=0] at (0,0){\includegraphics[width=.5\columnwidth]{Figures/14}};
%    \node[anchor=south west, inner sep=0] at (8,0){\includegraphics[width=.5\columnwidth]{Figures/15}};
%    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.5\columnwidth]{Figures/16}};
%
%
%    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
%    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
%    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
%    \end{tikzpicture}
%   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG and GAG-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
%\label{fig:HVTN065ResultsCont}
%\end{figure}

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto,node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
    \begin{scope} [x={(a.south east)},y={(a.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {A} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of A] (B) {
    \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
        \begin{scope} [x={(b.south east)},y={(b.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {B} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [below=of A] (C) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {C} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [below=of B] (D) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
        \begin{scope} [x={(d.south east)},y={(d.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {D} ;
\end{scope}
    \end{tikzpicture}
    };
  \node [below=of C] (E) {
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
    \begin{scope} [x={(e.south east)},y={(e.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {E} ;
    \end{scope}
    \end{tikzpicture}
    };
    \node [right=of E] (F) {
    \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
        \begin{scope} [x={(f.south east)},y={(f.north west)}]
        \node at (0,1) [font=\tiny\sffamily] {F} ;
        \end{scope}
    \end{tikzpicture}
    };
    \node [below=of E] (G) {
    \begin{tikzpicture}
     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
         \begin{scope} [x={(g.south east)},y={(g.north west)}]
                 \node at (0,1) [font=\tiny\sffamily] {G} ;
	\end{scope}
     \end{tikzpicture}
     };
     \node [below=of F] (H) {
     \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
        \begin{scope} [x={(h.south east)},y={(h.north west)}]
            \node at (0,1) [font=\tiny\sffamily] {H} ;
\end{scope}
    \end{tikzpicture}
    };
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to two--sided data (A,B,E,F) or to data where model assumptions are violated (C,D,G,H). Two--sided data were simulated from a standard model. A) Average ROC curves and B) average observed vs nominal FDR from 10 simulations with N=5,000, and E-F) N=1,000. Unconstrained MIMOSA fit to two--sided data simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution. C) Average ROC curves and D) average observed vs nominal FDR from 10 simulations at N=5,000, and G-H) N=1,000. }
   \label{webfig:simulations_trunc}
\end{figure}
\bibliographystyle{mn2e}
\bibliography{MIMOSA}

\end{document}

### 2012-07-27 MIMOSA.tex
Update the manuscript and supp document.

@@
 
-\title[MIMOSA: Combinatorial Mixture Models or Single--Cell Assays]{Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
+\title[MIMOSA: Combinatorial Mixture Models or Single-Cell Assays]{Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
 %\author[1]{Greg Finak}
@@
 
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
+\pagerange{\pageref{firstpage}-\pageref{lastpage}} 
 \volume{ }
@@
 \begin{abstract}
-Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single--cell assays.
+Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays.
 The characterization of these small cell subsets is crucial to decipher system level biological changes.
@@
 \begin{keywords}
-Mixture modelling, hierarchical modelling, bayesian modelling, single--cell assays, immunology
+Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
 \end{keywords}
@@
 In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
-Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single--cell measurement of multiple surface and intracellular markers (up to 17 markers).
+Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers).
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
@@
 While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
-This heterogeneity may carry important information, thus single--cell expression data should not be analyzed in the same fashion as cell-population level data. 
-Special treatment of single--cell level data, which preserves information about population heterogeneity, is warranted in general. 
+This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data. 
+Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general. 
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
@@
 For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. 
-This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single--cell level.
+This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
  After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
@@
 The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
-Here, we analyze a subset of the data consisting of 98 individual from the vaccinee group at two time points, day 0 and day 182.
+Here, we analyze a subset of the data consisting of 98 individual from the vaccine group at two time points, day 0 and day 182.
+For ease of presentation we restricted ourselves to the CD4+ T-cell subsets.
 Samples on day 0 were taken just before vaccination and no response is expected there.
 The corresponding samples can be used as negative controls.
-Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. 
-The data is described in Web Appendix A.
+Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. More details can be found in Web Appendix A.
 
 \textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen individuals.
-T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post--stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
-The expression from the simulated samples  was compared to paired, unstimulated controls.~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al.
+T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
+The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al.
 could you fill this in?}.
@@
 For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
-Individuals that show differential expression for a given marker will be called responders for that marker.
+Individuals that show differential expression will be called responders for that marker.
 In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.
@@
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.
-In some study, such as the ICS studies used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
-This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.
+In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
+This alternative parametrization is described in Web Appendix C and we refer to it as the one-sided model.
 
@@
 Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Appendix C), and are given by,
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix B), and are given by,
  \begin{align*}
@@
 where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.
-In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed form but can easily be estimated.
-All calculations for the one-sided case are described in Web Appendix B.
+In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated.
+All calculations for the one-sided case are described in Web Appendix C.
 
@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
-Details about the algorithms are given in Web Appendix D.
+Details about the algorithms are given in Web Appendix B.
 We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold--change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.
+The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.
 
-%The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation.
-The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells.
+%The constrained model was applied to an ICS data set from a real-world vaccine trial in order to identify responders to antigen stimulation.
+The unconstrained model was applied to Fluidigm single-cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single-cells.
 We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.
@@
 \subsection{ICS}
-Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold--change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
+Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
@@
 Didn't look at CD8's}.
-The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold--change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
-In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).
+The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
+In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B).
+These results are consistent for other cytokines (see Web Figure A).
 
@@
 \end{tikzpicture}
-   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065.
-Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) ENV-1-PTEG and expressing IL2.
+   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from HVTN065.
+Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFNg or C-D) ENV-1-PTEG and expressing IL2.
 ROC and FDR plots of other cytokine combinations can be found in Web Figure A.}
@@
 These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
-A similar analysis using a two-sided Fisher's exact test and clustering the signed q--values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation--specific patterns.
+A similar analysis using a two-sided Fisher's exact test and clustering the signed q-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
 Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
@@
 \end{tikzpicture}
-\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array.
+\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
 The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
@@
 Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
-C) Clustering of the signed q--values from Fisher's exact test.
+C) Clustering of the signed q-values from Fisher's exact test.
 Genes selected from Fisher's exact test at the 10\% FDR level.
@@
 
-We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV--1--stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial~.
+We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV-1-stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial~.
 We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000, and 50,000 events, with ten independent realizations of data for each $N$.
-The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold--change.
+The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
 This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D).
@@
 Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Web Figure B panel B).
-%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
+%Since the constrained model relies on Monte-Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
 
@@
 \end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one--sided model.
-Ten simulations were generated at an $N$ of 5,000 total counts using hyper--parameter estimates from real ICS data (IFNg expressing CD4+ T--cells stimulated with ENV-1-PTEG from HVTN065) with a five--fold effect size between responder and non--responder components.
+   \caption{Comparison of positivity detection methods on data simulated from the one-sided model.
+Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFNg expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
 A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
-Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and log fold--change.
-Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.}\label{fig:simulations}
+Curves are shown for MIMOSA (one-sided), Fisher's exact test, the likelihood ratio test, and log fold-change.
+Results for the two-sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.}\label{fig:simulations}
 \end{figure}
@@
  \end{equation*}
-%; \text{if $i$ is a non--responder}\\
+%; \text{if $i$ is a non-responder}\\
 % &(\mathbf{n}_{ui}|\mathbf{p}_{ui}) \sim \mathcal{M}(\mathbf{p}_{ui},N_{ui}); (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(\mathbf{p}_{si},N_{si});\text{if $i$ is a responder}
@@
 \end{align*}
-where the indicator variable $z_i$ is as defined in \ref{ss:priors}, i.e.
+where the indicator variable $z_i$ is as defined in Section \ref{ss:priors}, i.e.
 $z_i\sim\mathrm{Be}(w)$ where $w$ is the proportion of responders.
@@
 The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.
+The EM algorithm is now initialize with the multivariate Fisher's exact test\footnote{Greg is this correct?}.
 In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than three.
-It becomes highly dependent on the initial values and can fail to converge in many instances.
+It becomes more dependent on the initial values and can fail to converge in many instances.
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
-More details about both algorithms are given in Web Appendix D.
+More details about our multivariate MCMC algorithm is given in Web Appendix D.
 
@@
 %\subsection{ICS example with two markers}
-%We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
-As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
-We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
+%We applied the Multinomial-Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
+As a proof of principle we limited the model to the two-component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
+We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T-cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
 %
@@
 
-\subsection{Polyfunctionality in Fluidigm Single--Cell Gene Expression Data}
+\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
 As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
 In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
-Only CCL5 positive cells express BIRC3, and the its expression increases upon stimulation.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations)\footnote{We have not discussed that yet.
+Perhaps we could add a sentence in the previous result section to discuss it and refer to the figure in supplementary material} would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
 That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.
@@
 \end{tikzpicture}
-   \caption{Multivariate simulations from a two--sided model.
-Ten, eight-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
-Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value (combining Fisher's exact test run marginally on each of the five dimensions).
+   \caption{Multivariate simulations from a two-sided model.
+Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
+Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p-value (combining Fisher's exact test run marginally on each of the five dimensions).
 A) Average ROC curves for the competing methods over 10 simulations.
@@
 Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
-Single--cell assays will become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}.
-As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single--cell level is becoming increasingly important.
+Single-cell assays will become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}.
+As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important.
 Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small.
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
-In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single--cell assay.
+In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single-cell assay.
 
@@
 None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison.
-Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).
+Additionally, MIMOSA was found to be a better fit to real-world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).
 
@@
 As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others.
-More exploratory work will need to be done in this area once high dimensional single--cell level data with large number of samples become available.
+More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.
 
 All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
+An R package will soon be released as part of the Bioconductor project\footnote{Add reference}.
 
-%\subsection*{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
-%Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold.
+%\subsection*{Statistical Analysis of Responder and Non-responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
+%Positivity calls for vaccine responders and non-responders depend upon the selection of an appropriate threshold.
 Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
 Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
-In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
+In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one-sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
 %
@@
 We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples).
-For a given cytokine, we let $n_s$ be the number of cytokine--positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine--positive and total number of cells in the unstimulated sample, respectively.
+For a given cytokine, we let $n_s$ be the number of cytokine-positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine-positive and total number of cells in the unstimulated sample, respectively.
 Note that usually, $N_u \ne N_s$.
@@
 
-%\subsection{The Mixture of Beta--Binomials}
+%\subsection{The Mixture of Beta-Binomials}
 %Although we have specified the two models for the data, we do not know which observation was generated by which model.
@@
 Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}.
-We capture this uncertainty with a mixture framework of the two competing beta--binomial models.
+We capture this uncertainty with a mixture framework of the two competing beta-binomial models.
 The likelihood for the mixture is given by:
 
-%The classification of responder and non--responder individuals was compared between the multivariate MIMOSA model, univariate MIMOSA models applied to each cytokine combination, and Fisher's exact test applied to each cytokine combination.
+%The classification of responder and non-responder individuals was compared between the multivariate MIMOSA model, univariate MIMOSA models applied to each cytokine combination, and Fisher's exact test applied to each cytokine combination.
 
@@
 The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).
-In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4--fold and 2.5--fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples.
-These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7--fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
+In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4-fold and 2.5-fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples.
+These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7-fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
 %
-%   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data.
-Data were simulated from a constrained model with hyper--parameters estimated from a real data set of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial.
+%   \caption{Performance of the constrained and unconstrained Beta-binomial mixture model vs Fisher's exact test on simulated data.
+Data were simulated from a constrained model with hyper-parameters estimated from a real data set of Gag1 stimulated, CD4+, IL2 expressing T-cells on day 28 from the HVTN054 trial.
 For total cell counts from 10,000 to 150,000, we simulated ten data sets each of 500 observations with a response rate of 40\%.
-The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta--binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells.
+The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta-binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells.
 The beta distributions for the estimated and true hyper parameters are shown in the second panel for one simulated data set, with N=150,000 events.
-The ROC curve for Fisher's exact test and the constrained and unconstrained Beta--binomial model for the same simulation are shown in the third panel.
-The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta--binomial model are shown in the fourth panel.}
+The ROC curve for Fisher's exact test and the constrained and unconstrained Beta-binomial model for the same simulation are shown in the third panel.
+The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta-binomial model are shown in the fourth panel.}
 %   \label{fig:simulations}
@@
 We generated data from a variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions.
-The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity).
+The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta-prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity).
 The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel.
@@
 %   \caption{Comparison of MIMOSA and Fisher's exact test for calling responders in ICS data.
-Response rates for IL2 and IFNg cytokine positivity in Gag2 and Pol2 stimulated samples at days 0 and 28 in the CD4 and CD8 T--cell subpopulations of the HVTN054 ICS data.
+Response rates for IL2 and IFNg cytokine positivity in Gag2 and Pol2 stimulated samples at days 0 and 28 in the CD4 and CD8 T-cell subpopulations of the HVTN054 ICS data.
 In the treatment groups, response rates for the MIMOSA model are greater than or equal to those for Fisher's exact test at day 28 but are equal (and zero) at day 0 as expected.
@@
 %  % \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28.
-A significant difference between control and stimulated samples is called at the 1\% FDR threshold (red) for the Beta--binomial model, and at 1\% for FDR adjusted p--values from Fisher's exact test (green).
+A significant difference between control and stimulated samples is called at the 1\% FDR threshold (red) for the Beta-binomial model, and at 1\% for FDR adjusted p-values from Fisher's exact test (green).
 The likelihood surface shows the observed counts from stimulated and unstimulated samples.
@@
 %%   \label{fig:icsdata}
-%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array.
+%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
 The posterior probability of response times the sign of the change in expression is shown in the first panel (red indicates a significant decrease, green a significant increase, relative to the control).
@@
 %   \includegraphics[width=6in]{Figures/Polyfunctionality_ggplot2} 
-%   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--2 peptide stimulated CD4+ T--cells.}
+%   \caption{Posterior probabilities from the four-dimensional, Multivariate-Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env-2 peptide stimulated CD4+ T-cells.}
 %   \label{fig:polyfunctionality}
@@
 %   \includegraphics[width=6in]{Figures/Polyfunctionality_Env1} 
-%   \caption{Posterior probabilities from the four--dimensional, Multivariate--Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env--1 peptide stimulated CD4+ T--cells.}
+%   \caption{Posterior probabilities from the four-dimensional, Multivariate-Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env-1 peptide stimulated CD4+ T-cells.}
 %   \label{fig:polyfunctionalityenv1}
@@
 %The ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
-Data were preprocessed using the flowJo--defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T--cells for each sample and stimulation~\cite{Hahne:2009vv}.
+Data were preprocessed using the flowJo-defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T-cells for each sample and stimulation~\cite{Hahne:2009vv}.
 
@@
 %\label{supp:statpublished}
-%The methodology for statistical analysis and calling responders and non--responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
-In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine--positive T-cells in the antigen--stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine--positive T-cells for the negative control (unstimulated) sample from the same individual.
-In the original trial, significance was measured via one--sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual.
-In order to control the false positive rate for responder calls, made based on positivity for multiple cytokines, a discrete Bonferroni adjustment for multiple comparisons (across cytokines) was applied, and stimulations with an adjusted p--value below $\alpha = 0.00001$ were called positive~\cite{Horton:2007tsa}.
+%The methodology for statistical analysis and calling responders and non-responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
+In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine-positive T-cells in the antigen-stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine-positive T-cells for the negative control (unstimulated) sample from the same individual.
+In the original trial, significance was measured via one-sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual.
+In order to control the false positive rate for responder calls, made based on positivity for multiple cytokines, a discrete Bonferroni adjustment for multiple comparisons (across cytokines) was applied, and stimulations with an adjusted p-value below $\alpha = 0.00001$ were called positive~\cite{Horton:2007tsa}.
 
-%\subsection{Derivation of the Beta--Binomial Model for $p_s=p_u$ and $p_s>p_u$}
+%\subsection{Derivation of the Beta-Binomial Model for $p_s=p_u$ and $p_s>p_u$}
 %\label{s:derivation}
 %
-%We derive the posterior predictive distribution and marginal log--likelihood for the null model ($p_s=p_u$) and the alternative model ($p_s>p_u$) below: patient indices, $i$ on the $\left<N_s,n_s,N_u,n_u\right>$ are omitted for clarity.
+%We derive the posterior predictive distribution and marginal log-likelihood for the null model ($p_s=p_u$) and the alternative model ($p_s>p_u$) below: patient indices, $i$ on the $\left<N_s,n_s,N_u,n_u\right>$ are omitted for clarity.
 %For model \eqref{eq:null}, ($p_s=p_u, \alpha_s=\alpha_u, \beta_s=\beta_u$), the posterior predictive distribution of the data given the model is:
@@
 %	\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
-%	\intertext{with marginal log--likelihood:}
+%	\intertext{with marginal log-likelihood:}
 %	\begin{split}
@@
 %\end{align}
-%The integral in equation~\eqref{eq:model2postsupp} is computed numerically. The marginal log--likelihood of model~\eqref{eq:alternate} is then:
+%The integral in equation~\eqref{eq:model2postsupp} is computed numerically. The marginal log-likelihood of model~\eqref{eq:alternate} is then:
 %\begin{equation}
@@
 Therefore, for a set of ICS data, any single observation, ($n^i_s,n^i_u$) could either be derived from model~\eqref{eq:null}, or from model~\eqref{eq:alternate}.
-We model this situation explicitly using a mixture--model framework.
+We model this situation explicitly using a mixture-model framework.
 \bibliographystyle{mn2e}

### 2012-07-27 WebSupplementaryMaterialMIMOSA.tex
Update the manuscript and supp document.

@@
 \documentclass{article}
-\title{Web-based Supplementary Materials for Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies by Greg Finak, SC De Rosa, Mario Roederer and Raphael Gottardo}
+\title{Web-based Supplementary Materials for Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Steve De Rosa, Mario Roederer and Raphael Gottardo}
 
+\usepackage[margin=1in]{geometry}
 \usepackage{amssymb}
@@
 \section*{}
-\subsection{HVTN065 Vaccine Trial ICS Data Description}
+\subsection{HVTN065 vaccine trial ICS data description}
 \label{supp:statpublished}
@@
 CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay.
-Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays.
+% No need to talk about this
+% Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. 
 Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools.
-Results of the trial have been published~\citep{Goonetilleke:2006jk}.
+Results of the trial have been published in \cite{Goonetilleke:2006jk}.
 
@@
 The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.
-
-\subsection{Constrained Beta--binomial model}
+\subsection{Computational details for the beta-binomial model}
+\noindent\textbf{Marginal likelihood derivations}\\
+\noindent\textbf{MCMC algorithm}\\
+In what follows, we use $(x|y)$ to denote the conditional distribution of $x$ given $y$.
+In particular, we use $(x|\cdots)$ to denote the distribution of $x$ conditional on everything else in the model.
+Our MCMC algorithms cycle through the following steps:
+\begin{enumerate}
+\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tune for each parameter using the approach of (Greg add the citation).
+\item Update $w$ by Gibbs sampling using the full conditional,
+\[
+(w|\cdots)\sim
+\]
+\item for each $i$, update $z_i$ by Gibbs sampling using the following full conditional,
+\[
+(z_i|\dots)\sim
+\]
+\end{enumerate}
+For each updated parameter, step 1 above involves the calculation of the following acceptance ratio
+\[
+add\ formula.
+\]
+where $\pi$ is the prior distribution of the corresponding parameter. In our case each parameter has the same exponential prior with mean $10^3$\footnote{Check that this is correct}.
+\subsection{Constrained beta--binomial model}
 \label{supp:constrained}
 We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$.
-In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in equation~1.
-The marginal likelihood for the constrained model becomes $\mathrm{L}_1\times \mathrm{C}$, where $\mathrm{C}$ is a ratio of normalizing constants for the likelihood and the prior.
+In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in the main manuscript by (1).
+Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)$ but requires computing \[
+Z(\alpha_u, \beta_u, \alpha_s, \beta_s)=\int_{0}^1{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}\int_{p_u}^1 p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}dp_sdp_u.
+\]
+Using this expression, the constrained alternative marginal likelihood can be written as 
+\begin{align}
+	\begin{split}
+\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
+=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.\\
+\label{model2:constrained}
+\end{split}
+\end{align}
 
+In general, there is no closed form expression for $Z(\cdot)$, and a numerical approximation must be used.
+Let us denote by $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ the approximation.
+A natural way to estimate $\tilde{Z}$ is to use Monte Carlo integration.
+Indeed, we can write
+\begin{equation}
+\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)=\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\int_{0}^1\frac{{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}}{\mathrm{B}(\alpha_u,\beta_u)}(1-F_{\alpha_s,\beta_s}(p_u))dp_u
+\label{equ:normZ}
+\end{equation}
+where $F_{\alpha_s,\beta_s}$ is the cumulative distribution function of a beta random variable with parameters $\alpha_s$ and $\beta_s$.
+Using this identity, it can be seen that $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ can be approximated by
+\[
+\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)\approx\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\sum_{k=1}^K(1-F_{\alpha_s,\beta_s}(X_k))
+\]
+where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha_s$, $\beta_s$ and $K$ is the number of terms used in the Monte Carlo approximation.
+This approximation works relatively well with our EM implementation and does not significantly increase the computing time.
+Unfortunately, the number of terms (\textit{i.e.} value of $I$) required for the approximation to be good might be large and computing such a normalizing constant at each iteration would significantly slow down our MCMC implementation.
+As it tuns out, a better approximation
+can be obtained when $\alpha_s$ and $\beta_s$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,
+\[
+F_{\alpha_s,\beta_s}(p_u)=\sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^jp_u^{\beta_s+\alpha_s-j}.
+\label{eq:IBident}
+\]
+Then using, this identity, we obtain
 \begin{align*}
-\begin{split}
-	 \mathrm{C}=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
-%\label{eq:model2post}
-\end{split}
+Z(\alpha_u, \beta_u, \alpha_s, \beta_s)&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\int_{0}^1(1-p_u)^{\beta_u-1+j}p_u^{\alpha_u-1+\beta_s+\alpha_s-j}dp_u\\
+&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\mathrm{B}(\beta_u+j)\mathrm{B}(\alpha_u+\beta_s+\alpha_s-j).
+\label{equ:normZ}
 \end{align*}
+Typically, in ICS data $\alpha_s$ is relatively small leading to relatively few terms in the sum.
+However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha_s$ and $\beta_s$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha_s$ is less than one) and can render mixing in the MCMC more difficult.
+In addition, even though the computation is exact and much faster for small values of $\alpha_s$, which is typically the case with ICS data, it is still more demanding than the unconstrained model.
+In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is less than that of the stimulation sample, $p_s$.
+Indeed, in the one-sided case, if $p_u>p_s$ the associated individual should be a non-responder and thus $z_i=0$.
+In our experience, this computational shortcut performs just as well as the true one-sided implementation while being computationally much less demanding.
 
-The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$ is the regularized incomplete Beta function, which is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional integral that can be computed via Monte--Carlo integration (i.e.
-in our EM implementation).
-However, this can be computationally costly in the MCMC framework.
-In the latter case, we can compute the integral exactly if $\alpha_s$ and $\beta_s$ are integers by using the identity:
-\begin{align}
-I_{1-p_u}(\beta_s,\alpha_s) = \sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^j(p_u)^{\beta_s+\alpha_s-j}
-\label{eq:IBident}
-\end{align}
-Typically, in ICS data $\beta_s>>\alpha_s$, leading to relatively few terms in the sum in equation~\ref{eq:IBident}.  This leads to an exact expression for the normalizing constant:
-$$
-f
-$$
+\subsection{Computational details for the Dirichlet-multinomial model}
+\noindent\textbf{Marginal likelihood derivations}\\
+Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.The derivations is described below,
 
-\subsection{Derivation of Marginal Likelihoods for Beta--Binomial Model}
-
-\subsection{The MCMC Algorithm}
-
-\subsection{Marginal Likelihoods for the Multivariate Model}
-
+\noindent\textbf{MCMC algorithm}\\
+GREG: Look at the what I did above, and do something similar here.
 \begin{figure} %  figure placement: here, top, bottom, or page

### 2012-07-27 MIMOSA.tex
Updated as per latest recommendations
Enlarged figure 2, put A and B on first Row, and C on second
Added two-sided simulations to figure 3 as C and D
Improved label for Web Figure B
Cleaned up a stale sentence on two-sided simulation.
Removed fisher's combined p-value from multivariate simulations
MArio's paper already referenced.
Clarified simulation results section referencing Web Figure B and Figure 3 more clearly now.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_plot1.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_plot2.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 };
-\node [right=of B] (C) {
+\node [below=of A] (C) {
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.4\columnwidth]{Figures/FluidigmFisher.pdf}};
 \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
@@
     };
+        \node [below=of A] (C) {
+    \begin{tikzpicture}
+    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_twosided_ROC_5000.pdf}};
+    \begin{scope}[x={(c.south east)},y={(c.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {C} ;
+    \end{scope}
     \end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one-sided model.
+    };
+    \node [right=of C] (D) {
+    \begin{tikzpicture}
+    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
+    \begin{scope}[x={(d.south east)},y={(d.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {D} ;
+    \end{scope}
+    \end{tikzpicture}
+    };
+
+\end{tikzpicture}
+   \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models.
 Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFNg expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
-A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000).
-Curves are shown for MIMOSA (one-sided), Fisher's exact test, the likelihood ratio test, and log fold-change.
-Results for the two-sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.}\label{fig:simulations}
+A) Average ROC curve over the 10 simulated data sets (N=5,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000), one--sided.
+C) Average ROC curves, two--sided model.
+D) Average observed and nominal FDR, two--sided model.
+Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.}\label{fig:simulations}
 \end{figure}
@@
 The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (see Web Figure B panels C and D).
-Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.
+Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
 
@@
 Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we could not look at more than two makers at once.
-Therefore, we performed simulations in eight dimension to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p-values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
+Therefore, we performed simulations in eight dimension to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
 These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data (Figure~\ref{fig:mvsimulations} B).
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
+   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
-Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p-value (combining Fisher's exact test run marginally on each of the five dimensions).
+Multivariate MIMOSA was compared against Fisher's exact test, and the likelihood ratio test.
 A) Average ROC curves for the competing methods over 10 simulations.
@@
 Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
-In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or simply using this information for pre-filtering~\citep{Flatz:2011jb}.
+In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or the information is used for pre-filtering~\citep{Flatz:2011jb}.
 The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).

### 2012-07-27 WebSupplementaryMaterialMIMOSA.tex
Updated as per latest recommendations
Enlarged figure 2, put A and B on first Row, and C on second
Added two-sided simulations to figure 3 as C and D
Improved label for Web Figure B
Cleaned up a stale sentence on two-sided simulation.
Removed fisher's combined p-value from multivariate simulations
MArio's paper already referenced.
Clarified simulation results section referencing Web Figure B and Figure 3 more clearly now.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \begin{tikzpicture} [auto,node distance=0cm]
+%\node at (0,0) (A) {
+%\begin{tikzpicture}
+%    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+%    \begin{scope} [x={(a.south east)},y={(a.north west)}]
+%        \node at (0,1) [font=\tiny\sffamily] {A} ;
+%    \end{scope}
+%    \end{tikzpicture}
+%    };
+%    \node [right=of A] (B) {
+%    \begin{tikzpicture}
+%    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
+%        \begin{scope} [x={(b.south east)},y={(b.north west)}]
+%        \node at (0,1) [font=\tiny\sffamily] {B} ;
+%        \end{scope}
+%    \end{tikzpicture}
+%    };
     \node at (0,0) (A) {
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
-    \begin{scope} [x={(a.south east)},y={(a.north west)}]
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
                  \node at (0,1) [font=\tiny\sffamily] {A} ;
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
-        \begin{scope} [x={(b.south east)},y={(b.north west)}]
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+        \begin{scope} [x={(d.south east)},y={(d.north west)}]
             \node at (0,1) [font=\tiny\sffamily] {B} ;
@@
 \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
-         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
+    \begin{scope} [x={(e.south east)},y={(e.north west)}]
         \node at (0,1) [font=\tiny\sffamily] {C} ;
@@
     };
-     \node [below=of B] (D) {
+    \node [right=of C] (D) {
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
-        \begin{scope} [x={(d.south east)},y={(d.north west)}]
+    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
+        \begin{scope} [x={(f.south east)},y={(f.north west)}]
         \node at (0,1) [font=\tiny\sffamily] {D} ;
@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
-    \begin{scope} [x={(e.south east)},y={(e.north west)}]
+     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
+         \begin{scope} [x={(g.south east)},y={(g.north west)}]
                  \node at (0,1) [font=\tiny\sffamily] {E} ;
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
-        \begin{scope} [x={(f.south east)},y={(f.north west)}]
+    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
+        \begin{scope} [x={(h.south east)},y={(h.north west)}]
             \node at (0,1) [font=\tiny\sffamily] {F} ;
@@
     };
-    \node [below=of E] (G) {
-    \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
-         \begin{scope} [x={(g.south east)},y={(g.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {G} ;
-	\end{scope}
  \end{tikzpicture}
-     };
-     \node [below=of F] (H) {
-     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
-        \begin{scope} [x={(h.south east)},y={(h.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {H} ;
-\end{scope}
-    \end{tikzpicture}
-    };
- \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to two--sided data (A,B,E,F) or to data where model assumptions are violated (C,D,G,H).
-Two--sided data were simulated from a standard model.
-A) Average ROC curves and B) average observed vs nominal FDR from 10 simulations with N=5,000, and E-F) N=1,000.
-Unconstrained MIMOSA fit to two--sided data simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.
-C) Average ROC curves and D) average observed vs nominal FDR from 10 simulations at N=5,000, and G-H) N=1,000.
-}
+   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two--sided data with small counts.
+Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.
+A) The average ROC from 10 simulation with N=5,000 events.
+B) The average observed and nominal FDR from 10 simulations with N=5,000 events.
+C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events.
+MIMOSA fit to two--sided data were simulated from the standard model with N=1,000 events.
+A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations.}
    \label{webfig:simulations_trunc}

### 2012-07-27 MIMOSA.tex
Continued previous changes
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV-1-stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial~.
-We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000, and 50,000 events, with ten independent realizations of data for each $N$.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV-1-stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial.
+We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000 events, with ten independent realizations of data for each $N$.
 The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
-This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A-D).
-In addition, the nominal \textit{vs.} observed FDR\footnote{Need to define this earlier} was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} A,B).
+This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C).
+In addition, the nominal \textit{vs.} observed FDR was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).
 
-For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A and Web Figure B, panel A).
-Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations} B and Web Figure B panel B).
+For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity evan at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E).
+Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F).
+Finally, even when MIMOSA was fit to data simulated from a model that violated model assumptions, it outperformed competing methods (Web Figure B, panels A-D).
 %Since the constrained model relies on Monte-Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
@@
 As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters.
-A solution would be to explore alternative model parametrizations that could be used to reduce the number of required parameters.
+A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters. 
 For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.

### 2012-07-27 WebSupplementaryMaterialMIMOSA.tex
Continued previous changes
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
     \begin{scope} [x={(e.south east)},y={(e.north west)}]
@@
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
         \begin{scope} [x={(f.south east)},y={(f.north west)}]
@@
     \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
+     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
          \begin{scope} [x={(g.south east)},y={(g.north west)}]
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
         \begin{scope} [x={(h.south east)},y={(h.north west)}]
@@
 MIMOSA fit to two--sided data were simulated from the standard model with N=1,000 events.
-A) Average ROC curves from 10 simulations and B) average observed vs nominal FDR from 10 simulations.}
+E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.}
    \label{webfig:simulations_trunc}

### 2012-07-27 MIMOSA.tex
Defined Observed False Discovery Rate and referenced Storey paper on FDR.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.
+The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate), defined as $\frac{\text{\# of false positives}}{\text{\# of observations called positive}}$ against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}.
 

### 2012-07-29 MIMOSA.tex
Minor changes.

@@
 
-\title[MIMOSA: Combinatorial Mixture Models or Single-Cell Assays]{Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+\title[MIMOSA: Mixture Models or Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
 %\author[1]{Greg Finak}
@@
 A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
-Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays.
-Our model allows the inference specific to each observation, while borrowing strength across observations through common prior distributions.
+Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays.
+Our model allows the inference to be individual specific, while borrowing strength across individuals through common prior distributions.
 We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
 We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
-Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods.
-Using a simulation study we show that our framework is also robust to model misspecification.
-Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model.
+Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
+Additional simulations show that our framework is also robust to model misspecification.
+Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
 \end{abstract}
@@
 \begin{keywords}
-Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
+Bayesian Inference, Expectation-Maximization, Flow cytometry, Gene expression, Hierarchical modeling, Immunology, Markov chain Monte Carlo
 \end{keywords}
@@
 In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
-Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers).
+Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers)\footnote{Reference Mario's paper}.
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
@@
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
-These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood, peripheral blood cell mononuclease (PBMC), or pre-sorted cell populations (\textit{e.g.} T-cells), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
@@
 These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences.
-individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}.
+Individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}.
 In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
@@
 Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}.
-All of these methods test observations separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders).
+All of these methods test individuals separately, and no information is shared across individuals\footnote{We should say at some point that here we use individuals but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders).
 %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
@@
 Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
+Our paper is organized as follows, ...
 

### 2012-07-30 MIMOSA.tex
Resized Figure 1. Enlarged subpanel labels (A,B,C etc)
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

--- 2012-07-27 MIMOSA.tex
@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=.6\columnwidth]{Figures/2}};
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/2}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {A} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
+    \node at (0,1) [font=\small\sffamily] {A} ;
+    \node at (0.5,1) [font=\small\sffamily] {B} ;
     \end{scope}
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=.6\columnwidth]{Figures/1}};
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/1}};
     \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {C} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
+    \node at (0,1) [font=\small\sffamily] {C} ;
+    \node at (0.5,1) [font=\small\sffamily] {D} ;
     \end{scope}
@@
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-\node at (0,1) [font=\tiny\sffamily] {A} ;
+\node at (0,1) [font=\small\sffamily] {A} ;
 \end{scope}
@@
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
+\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
 \end{scope}
@@
 \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-\node at (0,1) [font=\tiny\sffamily] {C} ;
+\node at (0,1) [font=\small\sffamily] {C} ;
 \end{scope}
@@
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {A} ;
+        \node at (0,1) [font=\small\sffamily] {A} ;
         \end{scope}
@@
     \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {B} ;
+    \node at (0,1) [font=\small\sffamily] {B} ;
     \end{scope}
@@
     \begin{scope}[x={(c.south east)},y={(c.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {C} ;
+    \node at (0,1) [font=\small\sffamily] {C} ;
     \end{scope}
@@
     \begin{scope}[x={(d.south east)},y={(d.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {D} ;
+    \node at (0,1) [font=\small\sffamily] {D} ;
     \end{scope}
@@
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (0,1) [font=\tiny\sffamily] {B} ;
+\node at (0,1) [font=\small\sffamily] {B} ;
 \end{scope}
@@
 %\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
-%\node at (0,0) [font=\tiny\sffamily] {C} ;
-%\node at (6.25,0) [font=\tiny\sffamily] {D} ;
+%\node at (0,0) [font=\small\sffamily] {C} ;
+%\node at (6.25,0) [font=\small\sffamily] {D} ;
 \end{tikzpicture}
@@
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {A} ;
+    \node at (0,1) [font=\small\sffamily] {A} ;
 \end{scope}
@@
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-       \node at (0,1) [font=\tiny\sffamily] {B} ;
+       \node at (0,1) [font=\small\sffamily] {B} ;
 \end{scope}

### 2012-07-30 MIMOSA.tex
Make some more changes to the manuscript.

--- 2012-07-29 MIMOSA.tex
@@
 Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays.
-Our model allows the inference to be individual specific, while borrowing strength across individuals through common prior distributions.
+Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions.
 We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
-We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
+We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
 Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
@@
 When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
-A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
+A common statistical problem is, for a given marker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
 
-A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine.
+A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine.
 Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
@@
 These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells.
-
 The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
@@
 Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
-These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences.
-Individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}.
+These counts are compared between antigen stimulated and unstimulated samples from an subject to identify significant differences.
+Subjects who generate a response after stimulation are called \textit{responders} whereas subject that do not show any differences are called \textit{non-responders}.
 In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
 
-Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}.
-All of these methods test individuals separately, and no information is shared across individuals\footnote{We should say at some point that here we use individuals but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders).
+Although there is no standard approach to analyzing single-cell assays, current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}\footnote{Can we match the citations with the methods used}.
+All of these methods test subjects separately, and no information is shared across subjects\footnote{We should say at some point that here we use subject but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders).
 %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
@@
 The framework developed in this paper addresses these issues explicitly.
-In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
-In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals.
-This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
+In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
+In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects.
+This sharing of information across subjects helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
 Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
-Our paper is organized as follows, ...
+Our paper is organized as follows, Section \ref{s:data} introduces the data and notations used in the paper.
+In Section \ref{s:DEone}, we present our model for testing differential marker expression in the univariate case.
+Section \ref{s:results} compares our approach to alternative methods and test the robustness of our model.
+In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using our single cell gene expression data.
+Finally, in Section \ref{s:discussion} we discuss our findings and future work.
 
@@
 \textit{Flow cytometry}:
-The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see Web Appendix A). 
+The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA\footnote{Define} vaccines in a prime-boost regimen with 120 subjects \citep{Goonetilleke:2006jk}.
 The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
-Here, we analyze a subset of the data consisting of 98 individual from the vaccine group at two time points, day 0 and day 182.
+Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points, day 0 and day 182. 
+Three cytokines (IFN$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (HIV Envelope peptide pool) matching part of the vaccine.
 For ease of presentation we restricted ourselves to the CD4+ T-cell subsets.
@@
 The corresponding samples can be used as negative controls.
-Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. More details can be found in Web Appendix A.
+Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.
 
-\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen individuals.
-T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
+\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects.
+T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
 The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al.
@@
 In the remainder of the paper, we use the following notation to describe our model.
-From this point on, we assume that we observe cell counts from $I$ individuals in two conditions: stimulated and un-stimulated.
+From this point on, we assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated.
 Each cell can either be positive or negative for a marker.
@@
 We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples.
-We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individual $i$ in each sample.
-For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample.
+For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
 Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
@@
 \label{s:DEone}
-Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
-As a consequence, we first consider the one marker case (\textit{i.e.} $K=1$) where marginal cell counts are analyzed separately.
-In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
+Datasets like the ones presented here are usually analyzed in in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
+Here, by univariate, we mean that we have only one positive cell subset.
+This cell subset can be defined as considering by the expression of one marker alone (marginalizing over all other measured markers) such as IL2$+$ (\textit{vs.} IL2$-$), or considering a specific positive marker combination (and marginalizing over everything else) such as IL2$+$ and/or INF$\gamma+$ (\textit{vs.} IL2$-$/INF$\gamma-$).
+Without loss of generality, we treat the univariate case as a one marker case (\textit{i.e.} $K=1$).
+In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
 
@@
 \parbox{0.8\linewidth}{
-\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given individual $i$.}\label{tab:twobytwo}
+\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
 \centering
@@
 
-For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
-Individuals that show differential expression will be called responders for that marker.
+For a given subject and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
+Subjects that show differential expression will be called responders for that marker.
 In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.
@@
 \label{s:DE}
-For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
+For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
 
@@
 \end{equation*}
-where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
+where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding subjects we consider two competing models:
 \begin{equation*}
@@
 Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal.
-Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.
+Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder.
 In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
@@
 \label{ss:priors}
-Our model shares information across all individuals using exchangeable Beta priors on the unknown proportions, as follows, 
+Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows, 
  \begin{align*}
@@
  \end{align*}
-where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.
-${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals.
+where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, i.e.
+${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects.
 Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model.
@@
 $\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters.
-We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.
+We use numerical optimization as implemented in R\footnote{Add citation to R}'s \textit{optim} function to estimate the remaining parameters.
 Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence.
@@
 We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
-For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
+For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities\footnote{Need to check this}.
 Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate), defined as $\frac{\text{\# of false positives}}{\text{\# of observations called positive}}$ against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}.
+The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}.
 
@@
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
-We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?
-Didn't look at CD8's}.
 The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
+At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positive being detected.
+Our comparison also show that ranking based on log-fold change alone is not reliable and should not be used.
 In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B).
-These results are consistent for other cytokines (see Web Figure A).
+Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study\footnote{reference}.
+These results are consistent for other cytokines and cytokine combinations (see Web Figure A).
 
@@
 In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation.
-The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm} A,B).
+The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A,B).
 These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
-A similar analysis using a two-sided Fisher's exact test and clustering the signed q-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
+A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
 Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ENV-1-stimulated, CD4-positive, IL2-expressing T-cells from the primary immunogenicity time point of the HVTN065 trial.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ICS data (IL2 univariate) from the primary immunogenicity time point of the HVTN065 trial.
 We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000 events, with ten independent realizations of data for each $N$.
@@
 Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F).
-Finally, even when MIMOSA was fit to data simulated from a model that violated model assumptions, it outperformed competing methods (Web Figure B, panels A-D).
 %Since the constrained model relies on Monte-Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (see Web Figure B panels C and D).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the ICS data set (see Web Figure B panels C and D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
@@
 \subsection{Prior}
-As in the one-marker case, we share information across individuals using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
+As in the one-marker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
 \begin{align*}
@@
 Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations)\footnote{We have not discussed that yet.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN$\gamma+$)\footnote{We have not discussed that yet.
 Perhaps we could add a sentence in the previous result section to discuss it and refer to the figure in supplementary material} would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
@@
 In contrast, multivariate MIMOSA tests all polyfunctional cell subpopulations simultaneously, identifying significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels).
-Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.
+Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.
 
@@
 As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important.
-Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite unreliable especially when the cell counts are small.
+Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite sub-optimal especially when the cell counts are small.
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
@@
 
-The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental individuals across multiple conditions (\textit{i.e.} vaccine responders and non-responders).
+The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.} vaccine responders and non-responders).
 Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Web Figure B).
@@
 The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$.
-As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters.
+As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of subjects to properly estimate all parameters.
 A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters. 
@@
 \section*{Acknowledgments}
-We wish to acknowledge ...
-
+GF and RG acknowledge funding from National Institute of Health grants R01 EB008400 and U01 AI068635-01, and a Bill \& Melinda Gates Foundation grant OPP38744 \footnote{Change this number later}. 
 %\section{Data Import, Preprocessing, and Gating}

### 2012-07-30 MIMOSA.tex

@@
 
-\title[MIMOSA: Mixture Models or Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+\title[MIMOSA: Combinatorial Mixture Models or Single-Cell Assays]{Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
 %\author[1]{Greg Finak}
@@
 Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
+Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays.
+Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions.
+We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
+We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
+Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
+Additional simulations show that our framework is also robust to model misspecification.
+Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
 \end{abstract}
@@
 \begin{keywords}
-Bayesian Inference, Expectation-Maximization, Flow cytometry, Gene expression, Hierarchical modeling, Immunology, Markov chain Monte Carlo
+Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
 \end{keywords}
@@
 In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
-Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers)\footnote{Reference Mario's paper}.
+Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers).
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
@@
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
-These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood, peripheral blood cell mononuclease (PBMC), or pre-sorted cell populations (\textit{e.g.} T-cells), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
@@
 Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
+
 These counts are compared between antigen stimulated and unstimulated samples from an subject to identify significant differences.

### 2012-07-30 MIMOSA.tex
Web supp changes.. added derivations and references
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

--- 2012-07-30 MIMOSA.tex
@@
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
-We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+?
-Didn't look at CD8's}.
+We examined the CD4+ T-cell cytokine responses.
 The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
@@
 All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
-An R package will soon be released as part of the Bioconductor project\footnote{Add reference}.
+An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
 
@@
 % 	\mathrm{Pr}(y_i|\alpha_u,\beta_u) &= \int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i,p_u|\alpha_u,\beta_u)dp_u\\
-%	&=\int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u\\
-%	\begin{split}
+%&=\int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u\\
+%\begin{split}
 %	=\int\limits_{p_u=0}\limits^{1}\binom{N_s+n_s}{n_s}p_u^{n_s}(1-p_u)^{N_s}\binom{N_u+n_u}{n_u}p_u^{n_u}(1-p_u)^{N_u}\cdot \\ 
@@
 % 	=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\cdot\\ \int\limits_{p_u=0}\limits^{1}p_u^{n_s+n_u+\alpha_u-1}(1-p_u)^{N_s+N_u+\beta_u-1}dp_u
-%	\end{split}\\
-%	\intertext{The integrand is the kernel of a beta distribution with parameters $\left(n_u+n_s+\alpha_u,N_u+N_s+\beta_u\right)$, giving a closed form expression for the  posterior predictive distribution of model \eqref{eq:null}:}
-%	\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
-%	\intertext{with marginal log-likelihood:}
-%	\begin{split}
-%	\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N^i_{s}+n^i_{s}}{n^i_{s}}}+\log{\binom{N^i_{u}+n^i_{u}}{n^i_{u}}}+\right.\\
-%	\left.\log{\left(\mathrm{B}(n^i_s+n^i_u+\alpha_u,N^i_s+N^i_u+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLLsupp}
-%	\end{split}
+%\end{split}\\
+%\intertext{The integrand is the kernel of a beta distribution with parameters $\left(n_u+n_s+\alpha_u,N_u+N_s+\beta_u\right)$, giving a closed form expression for the  posterior predictive distribution of model \eqref{eq:null}:}
+%\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
+%\intertext{with marginal log-likelihood:}
+%\begin{split}
+%\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N^i_{s}+n^i_{s}}{n^i_{s}}}+\log{\binom{N^i_{u}+n^i_{u}}{n^i_{u}}}+\right.\\
+%\left.\log{\left(\mathrm{B}(n^i_s+n^i_u+\alpha_u,N^i_s+N^i_u+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLLsupp}
+%\end{split}
 % \end{align} 
@@
 We model this situation explicitly using a mixture-model framework.
-\bibliographystyle{mn2e}
+\bibliographystyle{biom}
 \bibliography{MIMOSA}

### 2012-07-30 MIMOSA.tex
Updated the grant number.

--- 2012-07-30 MIMOSA.tex
@@
 \section*{Acknowledgments}
-GF and RG acknowledge funding from National Institute of Health grants R01 EB008400 and U01 AI068635-01, and a Bill \& Melinda Gates Foundation grant OPP38744 \footnote{Change this number later}. 
+GF and RG acknowledge funding from National Institute of Health grants R01 EB008400 and U01 AI068635-01, and a Bill \& Melinda Gates Foundation grant OPP1032317. 
 %\section{Data Import, Preprocessing, and Gating}

### 2012-07-30 WebSupplementaryMaterialMIMOSA.tex
Resized Figure 1. Enlarged subpanel labels (A,B,C etc)
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \noindent\textbf{Marginal likelihood derivations}\\
+The probability of the data, given that it is drawn from the non--responder components for the beta--binomial model is given by:
+\[
+\begin{split}
+\mathrm{Pr}(n_{s},n_{u}|\alpha_0,\beta_0) =&\int_0^1 \mathrm{Pr}(n_{s},n_{u}|p_u,\alpha_0,\beta_0)\mathrm{d}p_u  \text{ : if } z=0 \\
+=& \int_0^1Pr(n_s|p_u)Pr(n_u|p_u)Pr(p_u|\alpha_0,\beta_0) \mathrm{d}p_u\\
+=&\int_0^1 \mathrm{Bin}(N_s,p_u)\mathrm{Bin}(N_u,p_u)\mathrm{Beta}(\alpha_0,\beta_0 ) \mathrm{d}p_u\\
+=&\int_0^1\binom{N_s}{n_s}p_u^{n_s}(1-p_u)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}p_u^{\alpha_0-1}(1-p_u)^{\beta_0-1} \mathrm{d}p_u\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}\int_0^1 p_u^{n_s+n_u+\alpha_0-1}(1-p_u)^{N_s+N_u-n_s-n_u+\beta_0-1} \mathrm{d}p_u\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_0,N_s+N_u-n_s-n_u+\beta_0)}{\mathrm{B}(\alpha_0,\beta_0)}\\
+\end{split}
+\]
+giving the marginal likelihood function
+\[
+\mathrm{L_0}(\alpha_u,\beta_u|n_{si},n_{ui}) = \prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_0,N_{si}+N_{ui}-n_{si}-n_{ui}+\beta_0)}{\mathrm{B}(\alpha_0,\beta_0)}\\
+\]
+
+For the responder component, the $n_s$ are drawn from a binomial distribution with parameter $p_s$, giving
+\[
+\begin{split}
+\mathrm{Pr}(n_{s},n_{u}|\alpha_0,\beta_0,\alpha_s,\beta_s) =&\int_0^1\int_0^1 \mathrm{Pr}(n_{s},n_{u}|p_u,p_s,\alpha_0,\beta_0,\alpha_s,\beta_s)\mathrm{d}p_s\mathrm{d}p_u  \text{ : if } z=1 \\
+=& \int_0^1\int_0^1Pr(n_s|p_s)Pr(n_u|p_u)Pr(p_u|\alpha_0,\beta_0)Pr(p_s|\alpha_s,\beta_s)\mathrm{d}p_s \mathrm{d}p_u\\
+=&\int_0^1\int_0^1 \mathrm{Bin}(N_s,p_s)\mathrm{Bin}(N_u,p_u)\mathrm{Beta}(\alpha_0,\beta_0 )\mathrm{Beta}(\alpha_s,\beta_s )\mathrm{d}p_s \mathrm{d}p_u\\
+=&\int_0^1\int_0^1\binom{N_s}{n_s}p_s^{n_s}(1-p_s)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}p_u^{\alpha_0-1}(1-p_u)^{\beta_0-1} \cdot\\ 
+&\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1} \mathrm{d}p_s\mathrm{d}p_u\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\int_0^1 p_u^{n_u+\alpha_0-1}(1-p_u)^{N_u-n_u+\beta_0-1} \cdot\\
+&\int_0^1 p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s-n_s+\beta_s-1}\mathrm{d}p_s\mathrm{d}p_u \\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_u+\alpha_0,N_u-n_u+\beta_0)\mathrm{B}(n_s+\alpha_s,N_s-n_s+\beta_s)}{\mathrm{B}(\alpha_0,\beta_0)\mathrm{B}(\alpha_s,\beta_s)}\\
+\end{split}
+\]
+giving marginal likelihood
+\[
+\mathrm{L_0}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{si},n_{ui}) = \prod_{i=1}^P \binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{ui}+\alpha_0,N_{ui}-n_{ui}+\beta_0)\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_0,\beta_0)\mathrm{B}(\alpha_s,\beta_s)}
+\]
+
 \noindent\textbf{MCMC algorithm}\\
@@
 \begin{enumerate}
-\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tune for each parameter using the approach of (Greg add the citation).
+\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc,Raftery:1992vp,Raftery:1996ws}.
 \item Update $w$ by Gibbs sampling using the full conditional,
 \[
-(w|\cdots)\sim
+(w|\cdots)\sim \mathrm{Beta}(\sum_i z_i,\sum_i (1-z_i))
 \]
@@
 \[
-(z_i|\dots)\sim
+(z_i|\cdots)\sim\mathrm{B(1,p_i)}
+\]
+where,
+\[
+p_i=\frac{w\cdot\mathrm{L_1}}{w\cdot\mathrm{L_1}+(1-w)\cdot\mathrm{L_0}}
 \]
 \end{enumerate}
-For each updated parameter, step 1 above involves the calculation of the following acceptance ratio
+For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, with the example below given for $\alpha_u$,
 \[
-add\ formula.
+\begin{split}
+\frac{\mathrm{L_0}(\alpha^{\text{new}}_u|\cdots)\mathrm{L_1}(\alpha^{\text{new}}_u|\cdots)\pi(\alpha^{\text{new}}_u)}{\mathrm{L_0}(\alpha^{\text{old}}_u|\cdots)\mathrm{L_1}(\alpha^{\text{old}}_u|\cdots)\pi(\alpha^{\text{old}}_u)}\\
+\end{split}
 \]
-where $\pi$ is the prior distribution of the corresponding parameter. In our case each parameter has the same exponential prior with mean $10^3$\footnote{Check that this is correct}.
+where $\pi$ is the prior distribution of $\alpha_u$.
+The obvious changes in the above expression are made for the acceptance ratios of $\alpha_s,\beta_s,\beta_0$.
+In our case each parameter has the same exponential prior with mean $1,000$
 \subsection{Constrained beta--binomial model}
@@
 \]
-Then using, this identity, we obtain
+Then using this identity, we obtain
 \begin{align*}
@@
     \begin{scope}[x={(a.south east)},y={(a.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {A} ;
+        \node at (0,1) [font=\small\sffamily] {A} ;
         \end{scope}
@@
         \begin{scope}[x={(b.south east)},y={(b.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {B} ;
+        \node at (0,1) [font=\small\sffamily] {B} ;
                 \end{scope}
@@
         \begin{scope}[x={(c.south east)},y={(c.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {C} ;
+        \node at (0,1) [font=\small\sffamily] {C} ;
         \end{scope}
@@
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {D} ;
+        \node at (0,1) [font=\small\sffamily] {D} ;
                 \end{scope}
@@
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {E} ;
+        \node at (0,1) [font=\small\sffamily] {E} ;
                 \end{scope}
@@
 
-  %  \node at (8,-4) [font=\tiny\sffamily] {F} ;
-%    \node at (0,-7.75) [font=\tiny\sffamily] {G} ;
-  %  \node at (8,-7.75) [font=\tiny\sffamily] {H} ;
-   % \node at (0,-11.75) [font=\tiny\sffamily] {I} ;
-   % \node at (8,-11.75) [font=\tiny\sffamily] {J} ;
+  %  \node at (8,-4) [font=\small\sffamily] {F} ;
+%    \node at (0,-7.75) [font=\small\sffamily] {G} ;
+  %  \node at (8,-7.75) [font=\small\sffamily] {H} ;
+   % \node at (0,-11.75) [font=\small\sffamily] {I} ;
+   % \node at (8,-11.75) [font=\small\sffamily] {J} ;
     \end{tikzpicture}
@@
 %
-%    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
-%    \node at (8,3.4) [font=\tiny\sffamily] {B} ;
-%    \node at (0,-0.5) [font=\tiny\sffamily] {C} ;
+%    \node at (0,3.4) [font=\small\sffamily] {A} ;
+%    \node at (8,3.4) [font=\small\sffamily] {B} ;
+%    \node at (0,-0.5) [font=\small\sffamily] {C} ;
 %    \end{tikzpicture}
@@
 %    \begin{scope} [x={(a.south east)},y={(a.north west)}]
-%        \node at (0,1) [font=\tiny\sffamily] {A} ;
+%        \node at (0,1) [font=\small\sffamily] {A} ;
 %    \end{scope}
@@
 %        \begin{scope} [x={(b.south east)},y={(b.north west)}]
-%        \node at (0,1) [font=\tiny\sffamily] {B} ;
+%        \node at (0,1) [font=\small\sffamily] {B} ;
 %        \end{scope}
@@
          \begin{scope} [x={(c.south east)},y={(c.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {A} ;
+                 \node at (0,1) [font=\small\sffamily] {A} ;
 	\end{scope}
@@
         \begin{scope} [x={(d.south east)},y={(d.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {B} ;
+            \node at (0,1) [font=\small\sffamily] {B} ;
 \end{scope}
@@
     \begin{scope} [x={(e.south east)},y={(e.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {C} ;
+        \node at (0,1) [font=\small\sffamily] {C} ;
     \end{scope}
@@
         \begin{scope} [x={(f.south east)},y={(f.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {D} ;
+        \node at (0,1) [font=\small\sffamily] {D} ;
         \end{scope}
@@
          \begin{scope} [x={(g.south east)},y={(g.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {E} ;
+                 \node at (0,1) [font=\small\sffamily] {E} ;
 	\end{scope}
@@
         \begin{scope} [x={(h.south east)},y={(h.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {F} ;
+            \node at (0,1) [font=\small\sffamily] {F} ;
 \end{scope}
@@
 \end{figure}
-\bibliographystyle{mn2e}
+\bibliographystyle{biom}
 \bibliography{MIMOSA}

### 2012-07-30 MIMOSA.tex
Bunch of updates, fixed IFNg AND OR IL2 Figure for ENV-1-PTEG
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 A common statistical problem is, for a given marker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
+Note that we use the term 'subject' throughout the paper, but the approaches described are general and can be applied to animal studies, for example.
 
@@
 
-Although there is no standard approach to analyzing single-cell assays, current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}\footnote{Can we match the citations with the methods used}.
-All of these methods test subjects separately, and no information is shared across subjects\footnote{We should say at some point that here we use subject but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders).
+Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}
+All of these methods test subjects separately, and no information is shared across subjects even though one could expect some similarities across responders (\textit{resp.} non-responders).
 %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
@@
 \textit{Flow cytometry}:
-The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA\footnote{Define} vaccines in a prime-boost regimen with 120 subjects \citep{Goonetilleke:2006jk}.
+The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen with 120 subjects \citep{Goonetilleke:2006jk}.
 The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
@@
 $\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters.
-We use numerical optimization as implemented in R\footnote{Add citation to R}'s \textit{optim} function to estimate the remaining parameters.
+We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters~\citep{Ihaka:1996ud}.
 Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence.
@@
 We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
-For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities\footnote{Need to check this}.
+For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
 Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.
@@
 Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
-The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A.
+The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
 At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positive being detected.
 Our comparison also show that ranking based on log-fold change alone is not reliable and should not be used.
-In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B).
-Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study\footnote{reference}.
+In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F).
+Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
 These results are consistent for other cytokines and cytokine combinations (see Web Figure A).
@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/2}};
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/1}};
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1}};
     \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
  };
+  \node [below=of B] (E) {
+ \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15}};
+    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+    \node at (0,1) [font=\small\sffamily] {E} ;
+    \node at (0.5,1) [font=\small\sffamily] {F} ;
+    \end{scope}
+    \end{tikzpicture}
+ };
 \end{tikzpicture}
@@
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFNg or C-D) ENV-1-PTEG and expressing IL2.
+E-F) ENV-1-PTEG and expressing IFNg and/or IL2.
 ROC and FDR plots of other cytokine combinations can be found in Web Figure A.}
@@
 The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.
-The EM algorithm is now initialize with the multivariate Fisher's exact test\footnote{Greg is this correct?}.
+The EM algorithm is now initialized with the multivariate Fisher's exact test.
 In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than three.

### 2012-07-30 MIMOSA.tex
Derivations done. Just the MCMC for Dir-Multinomial left.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 %\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
-\author{Greg Finak$^{1,*}$\email{gfinak@fhcrc.org},
-SC De Rosa$^{2,**}$\email{sderosa@fhcrc.org},
-Mario Roederer$^{3,***}$\email{marior@mail.nih.gov},
-Raphael Gottardo$^{1,****}$\email{rgotard@fhcrc.org}\\
+\author{Greg Finak$^{1}$, Andrew McDavid$^{1}$, \ldots, SC De Rosa$^{2}$, Mario Roederer$^{3}$, and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
 $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
@@
 Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
-These results are consistent for other cytokines and cytokine combinations (see Web Figure A).
+These results are consistent for other cytokines and cytokine combinations (see Web Figure A)\footnote{Add a sentence discussing the AND/OR combination}.
 
@@
 Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN$\gamma+$)\footnote{We have not discussed that yet.
-Perhaps we could add a sentence in the previous result section to discuss it and refer to the figure in supplementary material} would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN$\gamma+$) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
 That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.

### 2012-07-30 WebSupplementaryMaterialMIMOSA.tex
Bunch of updates, fixed IFNg AND OR IL2 Figure for ENV-1-PTEG
Signed-off-by: Greg Finak <gfinak@fhcrc.org>
Derivations done. Just the MCMC for Dir-Multinomial left.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools.
-Results of the trial have been published in \cite{Goonetilleke:2006jk}.
+Results of the trial have been published in \cite{Goepfert:2011ci}.
 
@@
 \noindent\textbf{Marginal likelihood derivations}\\
-Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.The derivations is described below,
+Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.
+For a non--responder, the probability of the observed data (of a single observation) is given by:
 
+\[
+\begin{split}
+Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u)=&\int_0^1Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
+=&\int_0^1Pr(\mathbf{n}_s|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
+=&\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_u)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
+=&\int_0^1 \frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\prod_k p^{n_{sk}}_{uk} \frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\prod_k p^{n_{uk}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
+=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sk}+n_{uk}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
+=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_s+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
+\end{split}
+\]
+The product over all observations gives the marginal likelihood,
+\[
+\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
+\]
+
+For the responder component, again the $\mathbf{n}_s$ are drawn from an independent multinomial distribution, and the $\mathbf{p}_s$ from a dirichlet with hyperparameters $\boldsymbol{\alpha_s}$, giving,
+\[
+\begin{split}
+Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s)=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
+=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\int_0^1\prod_k p^{n_{uk}+\alpha_{uk}-1}_{uk}\int_0^1 \prod_k p^{n_{sk}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
+=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_u+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_s+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
+\end{split}
+\]
+with marginal likelihood:
+\[
+\mathrm{L_1}(\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P\frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
+\]
 \noindent\textbf{MCMC algorithm}\\
@@
 \begin{tikzpicture} [auto, node distance=0cm]
-\node at (0,0) (A){
+ \node at (0,) (A) {
  \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=.5\columnwidth]{Figures/15}};
-    \begin{scope}[x={(a.south east)},y={(a.north west)}]
+    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
+        \begin{scope}[x={(b.south east)},y={(b.north west)}]
         \node at (0,1) [font=\small\sffamily] {A} ;
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
-        \begin{scope}[x={(b.south east)},y={(b.north west)}]
+    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
+        \begin{scope}[x={(c.south east)},y={(c.north west)}]
         \node at (0,1) [font=\small\sffamily] {B} ;
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
-        \begin{scope}[x={(c.south east)},y={(c.north west)}]
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
+        \begin{scope}[x={(d.south east)},y={(d.north west)}]
         \node at (0,1) [font=\small\sffamily] {C} ;
@@
 };
-\node [below=of B] (D) {
-\begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
-        \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\small\sffamily] {D} ;
-                \end{scope}
-        \end{tikzpicture}
-};
-\node [below=of C] (E) {
+\node [right=of C] (D) {
 \begin{tikzpicture}
@@
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\small\sffamily] {E} ;
+        \node at (0,1) [font=\small\sffamily] {D} ;
                 \end{scope}

### 2012-07-30 WebSupplementaryMaterialMIMOSA.tex
multivariate mcmc described
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 \end{enumerate}
-For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, with the example below given for $\alpha_u$,
+For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, (e.g. $\alpha_u$, below),
 \[
@@
 \]
-The product over all observations gives the marginal likelihood,
+The product over all $P$ observations gives the marginal likelihood,
 \[
@@
 \noindent\textbf{MCMC algorithm}\\
-GREG: Look at the what I did above, and do something similar here.
+The MCMC algorithm for the dirichlet--multinomial model is analogous to the beta--binomial, above. The parameter vectors $\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u$ are updated component--wise:
+\begin{enumerate}
+\item  Update each $\alpha_{sk}$, $\alpha_{uk}$ using a Gaussian symmetric proposal distribution with the variance of each proposal tuned using the approach of~\cite{Gelman:2004tc}. 
+\item Update $w$ by Gibbs sampling using the full conditional,
+\[
+(w|\cdots)\sim\mathrm{Beta}(\sum_i z_i,\sum_i(1-z_i))
+\]
+\item For each $i$, update $z_i$ using the full conditional,
+\[
+(z_i|\cdots)\sim \mathrm{B}(1,p_i)
+\]
+where 
+\[
+p_i = \frac{w\cdot\mathrm{L_1}}{w\cdot\mathrm{L_1}+(1-w)\cdot\mathrm{L_0}}
+\]
+For each parameter component updated in step 1 above, compute the acceptance ratio (e.g. $\alpha_{uk}$, below):
+\[
+\frac{\mathrm{L_0}(\alpha^{\text{new}}_{uk}|\cdots)\mathrm{L_1}(\alpha^{\text{new}}_{uk}|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L_0}(\alpha^{\text{old}}_{uk}|\cdots)\mathrm{L_1}(\alpha^{\text{old}}_{uk}|\cdots)\pi(\alpha^{\text{old}}_{uk})}\\
+\]
+
+where $\pi$ is the prior distribution of the parameter. Again, we have used the same exponential prior with mean 1,000 for each parameter.
+\end{enumerate}
+
 \begin{figure} %  figure placement: here, top, bottom, or page

### 2012-07-31 MIMOSA.tex
Modified some of the equations in supplementary material and in the main text.

@@
  \begin{split}
-  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
-	=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
+  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y}_i)
+	=&\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
 	\end{split}
@@
 	\begin{split}
-\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
-=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
+\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i) 
+=&\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
 \label{model2:unconstrained}

### 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
Modified some of the equations in supplementary material and in the main text.

@@
 \section*{}
-\subsection{HVTN065 vaccine trial ICS data description}
-\label{supp:statpublished}
-HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B).
-CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay.
-% No need to talk about this
-% Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. 
-Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools.
-Results of the trial have been published in \cite{Goepfert:2011ci}.
-
-%HVTN054 is a phase 1 (safety and immunogenicity) trial of an adenoviral vector vaccine in individuals without prior immunity~\cite{Peiperl:2010ej}.
-The vaccine vector expressed Gag, Pol and Env proteins from multiple HIV clades~\cite{Peiperl:2010ej}.
-Vaccine was given at two increasing doses, as well as a placebo.
-T--cell responses to antigens in the vaccine were measured via the ICS assay~\cite{Peiperl:2010ej,Horton:2007tsa}.
-The cytokines measured were IFNg (Interferon--$\gamma$), IL2 (Interleukin--2), TNFa (Tumor necrosis factor--$\alpha$) and IL4 (Interleukin 4)~\cite{Horton:2007tsa}.
-The sample size consisted of 20 vaccine and four placebo recipients.
-The original statistical analysis of the positivity calls is described in the associated publication~\cite{Peiperl:2010ej}.
-The Gag stimulated, IL2 expressing, CD4+ T--cell data from day 28 was used to derive hyper--parameter estimates for the simulation studies.
 \subsection{Computational details for the beta-binomial model}
 \noindent\textbf{Marginal likelihood derivations}\\
-The probability of the data, given that it is drawn from the non--responder components for the beta--binomial model is given by:
+For a given subject $i$, the null marginal likelihood is obtained after integrating out the prior from the likelihood for model ${\cal M}_0$, as follows,
 \[
 \begin{split}
-\mathrm{Pr}(n_{s},n_{u}|\alpha_0,\beta_0) =&\int_0^1 \mathrm{Pr}(n_{s},n_{u}|p_u,\alpha_0,\beta_0)\mathrm{d}p_u  \text{ : if } z=0 \\
-=& \int_0^1Pr(n_s|p_u)Pr(n_u|p_u)Pr(p_u|\alpha_0,\beta_0) \mathrm{d}p_u\\
-=&\int_0^1 \mathrm{Bin}(N_s,p_u)\mathrm{Bin}(N_u,p_u)\mathrm{Beta}(\alpha_0,\beta_0 ) \mathrm{d}p_u\\
-=&\int_0^1\binom{N_s}{n_s}p_u^{n_s}(1-p_u)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}p_u^{\alpha_0-1}(1-p_u)^{\beta_0-1} \mathrm{d}p_u\\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}\int_0^1 p_u^{n_s+n_u+\alpha_0-1}(1-p_u)^{N_s+N_u-n_s-n_u+\beta_0-1} \mathrm{d}p_u\\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_0,N_s+N_u-n_s-n_u+\beta_0)}{\mathrm{B}(\alpha_0,\beta_0)}\\
+\mathrm{L_0}(\alpha_u,\beta_u|n_{si},n_{ui}) =&\int_0^1 \mathrm{Pr}(n_{si},n_{ui}|p_u)\pi(p_u|\alpha_u,\beta_u)\mathrm{d}p_u\\
+=& \int_0^1Pr(n_s|p_u)Pr(n_u|p_u)\pi(p_u|\alpha_u,\beta_u) \mathrm{d}p_u\\
+=&\int_0^1\binom{N_s}{n_s}p_u^{n_s}(1-p_u)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1} \mathrm{d}p_u\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\int_0^1 p_u^{n_s+n_u+\alpha_u-1}(1-p_u)^{N_s+N_u-n_s-n_u+\beta_u-1} \mathrm{d}p_u\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u-n_s-n_u+\beta_0)}{\mathrm{B}(\alpha_u,\beta_u)}.
 \end{split}
 \]
-giving the marginal likelihood function
-\[
-\mathrm{L_0}(\alpha_u,\beta_u|n_{si},n_{ui}) = \prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_0,N_{si}+N_{ui}-n_{si}-n_{ui}+\beta_0)}{\mathrm{B}(\alpha_0,\beta_0)}\\
-\]
 
-For the responder component, the $n_s$ are drawn from a binomial distribution with parameter $p_s$, giving
+\noindent Similarly, the alternative marginal likelihood for a given subject is defined as,
 \[
 \begin{split}
-\mathrm{Pr}(n_{s},n_{u}|\alpha_0,\beta_0,\alpha_s,\beta_s) =&\int_0^1\int_0^1 \mathrm{Pr}(n_{s},n_{u}|p_u,p_s,\alpha_0,\beta_0,\alpha_s,\beta_s)\mathrm{d}p_s\mathrm{d}p_u  \text{ : if } z=1 \\
-=& \int_0^1\int_0^1Pr(n_s|p_s)Pr(n_u|p_u)Pr(p_u|\alpha_0,\beta_0)Pr(p_s|\alpha_s,\beta_s)\mathrm{d}p_s \mathrm{d}p_u\\
-=&\int_0^1\int_0^1 \mathrm{Bin}(N_s,p_s)\mathrm{Bin}(N_u,p_u)\mathrm{Beta}(\alpha_0,\beta_0 )\mathrm{Beta}(\alpha_s,\beta_s )\mathrm{d}p_s \mathrm{d}p_u\\
-=&\int_0^1\int_0^1\binom{N_s}{n_s}p_s^{n_s}(1-p_s)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}p_u^{\alpha_0-1}(1-p_u)^{\beta_0-1} \cdot\\ 
+\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{si},n_{ui}) =&\int_0^1\int_0^1 \mathrm{Pr}(n_{si},n_{ui}|p_u,p_s)\pi(p_u,p_s|\alpha_u,\beta_u,\alpha_s,\beta_s)\mathrm{d}p_s\mathrm{d}p_u \\
+=&\int_0^1\int_0^1\binom{N_s}{n_s}p_s^{n_s}(1-p_s)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1} \cdot\\ 
 &\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1} \mathrm{d}p_s\mathrm{d}p_u\\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_0,\beta_0)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\int_0^1 p_u^{n_u+\alpha_0-1}(1-p_u)^{N_u-n_u+\beta_0-1} \cdot\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\int_0^1 p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u-n_u+\beta_u-1} \cdot\\
 &\int_0^1 p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s-n_s+\beta_s-1}\mathrm{d}p_s\mathrm{d}p_u \\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_u+\alpha_0,N_u-n_u+\beta_0)\mathrm{B}(n_s+\alpha_s,N_s-n_s+\beta_s)}{\mathrm{B}(\alpha_0,\beta_0)\mathrm{B}(\alpha_s,\beta_s)}\\
+=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_u+\alpha_u,N_u-n_u+\beta_u)\mathrm{B}(n_s+\alpha_s,N_s-n_s+\beta_s)}{\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)}.
 \end{split}
 \]
-giving marginal likelihood
-\[
-\mathrm{L_0}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{si},n_{ui}) = \prod_{i=1}^P \binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{ui}+\alpha_0,N_{ui}-n_{ui}+\beta_0)\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_0,\beta_0)\mathrm{B}(\alpha_s,\beta_s)}
-\]
+%and taking the product over subjects leads to
+%\[
+%\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_{s},\mathbf{n}_{u}) = \prod_{i=1}^P \binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)}
+%\]
 
@@
 In particular, we use $(x|\cdots)$ to denote the distribution of $x$ conditional on everything else in the model.
-Our MCMC algorithms cycle through the following steps:
+Our MCMC algorithms cycles through the following steps:
 \begin{enumerate}
@@
 \[
-(w|\cdots)\sim \mathrm{Beta}(\sum_i z_i,\sum_i (1-z_i))
+(w|\cdots)\sim \mathrm{Beta}(\sum_i z_i,\sum_i (1-z_i)).
 \]
@@
 \[
-(z_i|\cdots)\sim\mathrm{B(1,p_i)}
+(z_i|\cdots)\sim\mathrm{B}(1,p_i).
 \]
@@
 \[
-p_i=\frac{w\cdot\mathrm{L_1}}{w\cdot\mathrm{L_1}+(1-w)\cdot\mathrm{L_0}}
+p_i=\frac{w\cdot\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s| n_{ui},n_{si})}{w\cdot\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{ui},n_{si})+(1-w)\cdot\mathrm{L_0}(\alpha_u,\beta_u|n_{ui},n_{si})}.
 \]
@@
 \begin{split}
-\frac{\mathrm{L_0}(\alpha^{\text{new}}_u|\cdots)\mathrm{L_1}(\alpha^{\text{new}}_u|\cdots)\pi(\alpha^{\text{new}}_u)}{\mathrm{L_0}(\alpha^{\text{old}}_u|\cdots)\mathrm{L_1}(\alpha^{\text{old}}_u|\cdots)\pi(\alpha^{\text{old}}_u)}\\
+\frac{\mathrm{L}(\alpha^{\text{new}}_u,\beta_u,\alpha_s,\beta_s|\cdots)\pi(\alpha^{\text{new}}_u)}{\mathrm{L}(\alpha_u,\beta_u,\alpha_s,\beta_s|\cdots)\pi(\alpha_u)}\\
 \end{split}
 \]
-where $\pi$ is the prior distribution of $\alpha_u$.
-The obvious changes in the above expression are made for the acceptance ratios of $\alpha_s,\beta_s,\beta_0$.
+where $L$ is the complete marginal likelihood conditional on $\mathbf{z}$ defined as,
+\[
+\mathrm{L}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_u,\mathbf{n}_s,\mathbf{z})=\prod_i\mathrm{L}_{z_i}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_u,\mathbf{n}_s)
+\]
+and $\pi$ is the prior distribution of $\alpha_u$.
+The obvious changes in the above expression are made for the acceptance ratios of $\alpha_s,\beta_s,\beta_u$.
 In our case each parameter has the same exponential prior with mean $1,000$
@@
 Using this expression, the constrained alternative marginal likelihood can be written as 
-\begin{align}
-	\begin{split}
-\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
-=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.\\
+\[
+\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i) 
+=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.
 \label{model2:constrained}
-\end{split}
-\end{align}
-
-In general, there is no closed form expression for $Z(\cdot)$, and a numerical approximation must be used.
+\]
+In general, there is no closed-form expression for $Z(\cdot)$, and a numerical approximation must be used.
 Let us denote by $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ the approximation.
@@
 This approximation works relatively well with our EM implementation and does not significantly increase the computing time.
-Unfortunately, the number of terms (\textit{i.e.} value of $I$) required for the approximation to be good might be large and computing such a normalizing constant at each iteration would significantly slow down our MCMC implementation.
+Unfortunately, the number of terms (\textit{i.e.} value of $K$) required for the approximation to be good might be large and computing such a normalizing constant at each iteration would significantly slow down our MCMC implementation.
 As it tuns out, a better approximation
@@
 In addition, even though the computation is exact and much faster for small values of $\alpha_s$, which is typically the case with ICS data, it is still more demanding than the unconstrained model.
-In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is less than that of the stimulation sample, $p_s$.
+In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is greater than that of the stimulation sample, $p_s$.
 Indeed, in the one-sided case, if $p_u>p_s$ the associated individual should be a non-responder and thus $z_i=0$.

### 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
Updated multivariate part of the web supp to match the earlier part
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.
-For a non--responder, the probability of the observed data (of a single observation) is given by:
+For a non--responder, the null marginal likelihood for a given subject is:
 
@@
 \begin{split}
-Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u)=&\int_0^1Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
-=&\int_0^1Pr(\mathbf{n}_s|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
+Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u)=&\int_0^1Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
+=&\int_0^1Pr(\mathbf{n}_s|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
 =&\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_u)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
@@
 \]
-The product over all $P$ observations gives the marginal likelihood,
-\[
-\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
-\]
+%The product over all $P$ observations gives the marginal likelihood,
+%\[
+%\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
+%\]
 
-For the responder component, again the $\mathbf{n}_s$ are drawn from an independent multinomial distribution, and the $\mathbf{p}_s$ from a dirichlet with hyperparameters $\boldsymbol{\alpha_s}$, giving,
+Similarly, the alternative marginal likelihood for a given subject is:
 \[
@@
 \[
-p_i = \frac{w\cdot\mathrm{L_1}}{w\cdot\mathrm{L_1}+(1-w)\cdot\mathrm{L_0}}
+p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}_{u},|\mathbf{n}_s,\mathbf{n}_u)}
 \]
@@
 \[
-\frac{\mathrm{L_0}(\alpha^{\text{new}}_{uk}|\cdots)\mathrm{L_1}(\alpha^{\text{new}}_{uk}|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L_0}(\alpha^{\text{old}}_{uk}|\cdots)\mathrm{L_1}(\alpha^{\text{old}}_{uk}|\cdots)\pi(\alpha^{\text{old}}_{uk})}\\
+\frac{\mathrm{L}(\alpha^{\text{new}}_{uk},\boldsymbol{\theta}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha_{uk})}\\
 \]
 
-where $\pi$ is the prior distribution of the parameter. Again, we have used the same exponential prior with mean 1,000 for each parameter.
+where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\theta}_u = \{\alpha_{uj} : (j \in \mathbb{Z}) (1\leq j \leq K) (j\neq k)\}$.
+Again, we have used the same exponential prior with mean 1,000 for each parameter.
+$\mathrm{L}$ is the complete marginal likelihood conditional on $\mathbf{z}$, defined as
+\[
+\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u,\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)
+\]
+
 \end{enumerate}

### 2012-07-31 MIMOSA.tex
Fixed references to web appendices in the main paper
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
-This alternative parametrization is described in Web Appendix C and we refer to it as the one-sided model.
+This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.
 
@@
 Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix B), and are given by,
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix A), and are given by,
  \begin{align*}
@@
 In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated.
-All calculations for the one-sided case are described in Web Appendix C.
+All calculations for the one-sided case are described in Web Appendix B.
 
@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
-Details about the algorithms are given in Web Appendix B.
+Details about the algorithms are given in Web Appendix A.
 We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
@@
 \subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Web Appendix E).
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Web Appendix C).
 For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
-More details about our multivariate MCMC algorithm is given in Web Appendix D.
+More details about our multivariate MCMC algorithm is given in Web Appendix C.
 
@@
 
-Web Appendices A, B, C, D, and E, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:estimation},  and \ref{s:demarkercombos} are available in the attached Web-based supplementary material.\vspace*{-8pt}
+Web Appendices A, B, C, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:estimation},  and \ref{s:demarkercombos} are available in the attached Web-based supplementary material.\vspace*{-8pt}
 % This will be replaced at submission time

### 2012-07-31 MIMOSA.tex
Minor changes (equations, text, etc).

@@
 The characterization of these small cell subsets is crucial to decipher system level biological changes.
-For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples.
+For this reason, an increasing number of studies rely assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples.
 A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
-Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays.
-Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions.
-We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
-We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
-Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
-Additional simulations show that our framework is also robust to model misspecification.
-Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
 Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays.
@@
  After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
-Using this matrix, one can form $2^K$ putative cell subsets. 
+Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations. 
 When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
@@
 
-Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}
+Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
 All of these methods test subjects separately, and no information is shared across subjects even though one could expect some similarities across responders (\textit{resp.} non-responders).
@@
 Section \ref{s:results} compares our approach to alternative methods and test the robustness of our model.
-In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using our single cell gene expression data.
+In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using a single-cell gene expression data.
 Finally, in Section \ref{s:discussion} we discuss our findings and future work.
@@
 Here, by univariate, we mean that we have only one positive cell subset.
-This cell subset can be defined as considering by the expression of one marker alone (marginalizing over all other measured markers) such as IL2$+$ (\textit{vs.} IL2$-$), or considering a specific positive marker combination (and marginalizing over everything else) such as IL2$+$ and/or INF$\gamma+$ (\textit{vs.} IL2$-$/INF$\gamma-$).
+This cell subset can be defined by considering the expression of one marker alone (marginalizing over all other measured markers) such as IL2$+$ (\textit{vs.} IL2$-$), or considering a specific positive marker combination (and marginalizing over everything else) such as IL2$+$ and/or INF$\gamma+$ (\textit{vs.} IL2$-$/INF$\gamma-$).
 Without loss of generality, we treat the univariate case as a one marker case (\textit{i.e.} $K=1$).
@@
 \end{equation*}
-where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding subjects we consider two competing models:
+where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding subjects, we consider two competing models:
 \begin{equation*}
@@
 \label{s:estimation}
-Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations.
+Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative likelihoods, which greatly simplify our calculations.
 Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix A), and are given by,
@@
 
-\subsection{EM algoritm}
+\noindent\textbf{EM algoritm}\\
 Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
@@
 
-\subsection{MCMC algorithm}
+\noindent\textbf{MCMC algorithm}\\
 Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}.
@@
 A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
-Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
+Comparing the size of the gene sets at an FDR of 10\%, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
 
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ICS data (IL2 univariate) from the primary immunogenicity time point of the HVTN065 trial.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ICS data (IL2 univariate) from the primary immunogenicity time point of the ICS data.
 We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000 events, with ten independent realizations of data for each $N$.
@@
 
-For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity evan at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E).
+For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E).
 Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F).
@@
 \begin{align*}
-\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
+\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
 \end{align*}
@@
 Similarly the marginal likelihood for the alternative model is given by
-\begin{align*}
-\begin{split}
-\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) =& \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \\ &\frac{N_{ui}!}{\prod_k n_{uik}!}.%\label{eq:postmult}
-\end{split}
-\end{align*}
+\[
+\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_{si}|\mathbf{n}_{si},\mathbf{n}_{ui}) =\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \frac{N_{ui}!}{\prod_k n_{uik}!}.%\label{eq:postmult}
+\]
 The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.

### 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
Fixed a lot of notations and equations.

@@
 In our case each parameter has the same exponential prior with mean $1,000$
-\subsection{Constrained beta--binomial model}
+\subsection{Constrained beta-binomial model}
 \label{supp:constrained}
@@
 Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion. 
-For a non--responder, the null marginal likelihood for a given subject is:
-
+For a given subject, the null marginal likelihood is defined as
 \[
 \begin{split}
-Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u)=&\int_0^1Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
-=&\int_0^1Pr(\mathbf{n}_s|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
-=&\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_u)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
-=&\int_0^1 \frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\prod_k p^{n_{sk}}_{uk} \frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\prod_k p^{n_{uk}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
-=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sk}+n_{uk}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
-=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_s+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
+\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui})=&\idotsint\mathrm{Pr}(\mathbf{n}_{si},\mathbf{n}_{ui}|\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
+=&\idotsint \frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\prod_k p^{n_{sik}}_{uk} \frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\prod_k p^{n_{uik}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
+=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sik}+n_{uik}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
+=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
 \end{split}
@@
 
-Similarly, the alternative marginal likelihood for a given subject is:
+\noindent Similarly, the alternative marginal likelihood for a given subject is:
 \[
 \begin{split}
-Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s)=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
-=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\int_0^1\prod_k p^{n_{uk}+\alpha_{uk}-1}_{uk}\int_0^1 \prod_k p^{n_{sk}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
-=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_u+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_s+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
+\mathrm{L_1}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{ui},\mathbf{n}_{si})
+%=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
+=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\idotsint\prod_k p^{n_{uik}+\alpha_{uk}-1}_{uk}\idotsint \prod_k p^{n_{sik}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
+=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}.
 \end{split}
 \]
-with marginal likelihood:
-\[
-\mathrm{L_1}(\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P\frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
-\]
 \noindent\textbf{MCMC algorithm}\\
-The MCMC algorithm for the dirichlet--multinomial model is analogous to the beta--binomial, above. The parameter vectors $\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u$ are updated component--wise:
+The MCMC algorithm for the Dirichlet-multinomial model is analogous to the beta-binomial, above. The parameter vectors $\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u$ are updated component-wise:
 \begin{enumerate}
@@
 \[
-p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}_{u},|\mathbf{n}_s,\mathbf{n}_u)}
+p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui})}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui})+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}_{u}|\mathbf{n}_{si},\mathbf{n}_{ui})}
 \]
@@
 \[
-\frac{\mathrm{L}(\alpha^{\text{new}}_{uk},\boldsymbol{\theta}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha_{uk})}\\
+\frac{\mathrm{L}(\alpha^{\text{new}}_{uk},\boldsymbol{\alpha}_{u\{-k\}},\boldsymbol{\alpha}_s|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha_{uk})}\\
 \]
 
-where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\theta}_u = \{\alpha_{uj} : (j \in \mathbb{Z}) (1\leq j \leq K) (j\neq k)\}$.
+where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\alpha}_{u\{-k\}} = \{\alpha_{uj} :  j\neq k\}$.
 Again, we have used the same exponential prior with mean 1,000 for each parameter.
@@
 \[
-\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u,\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u)
+\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui},\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui}).
 \]
@@
  \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two--sided data with small counts.
+   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two-sided data with small counts.
 Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.
@@
 C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events.
-MIMOSA fit to two--sided data were simulated from the standard model with N=1,000 events.
+MIMOSA fit to two-sided data were simulated from the standard model with N=1,000 events.
 E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.}

### 2012-07-31 MIMOSA.tex
Fixed up labels on simulation figures (MCMC, EM one-sided, two-sided etc)
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 %\author[1]{Greg Finak}
-%\author[2]{SC De Rosa}
+%\author[2]{Steve De Rosa}
 %\author[3]{Mario Roederer}
@@
 %\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
-\author{Greg Finak$^{1}$, Andrew McDavid$^{1}$, \ldots, SC De Rosa$^{2}$, Mario Roederer$^{3}$, and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
+\author{Greg Finak$^{1}$, Andrew McDavid$^{1}$, \ldots, Steve De Rosa$^{1,2}$, Mario Roederer$^{3}$, and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
 $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
@@
 
-The framework developed in this paper addresses these issues explicitly.
+The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly.
 In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
@@
 Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
-These results are consistent for other cytokines and cytokine combinations (see Web Figure A)\footnote{Add a sentence discussing the AND/OR combination}.
+These results are consistent for other cytokines and cytokine combinations (see Web Figure A).
 

### 2012-07-31 MIMOSA.tex
Minor changes. Added "this figure appears in color …" to each figure caption as required.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 
-\title[MIMOSA: Mixture Models or Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
 %\author[1]{Greg Finak}
@@
 \begin{abstract}
-Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays.
+Blood and tissue are composed of many functionally distinct cell subsets.
+In immunological studies, these can only be measured accurately using single-cell assays.
 The characterization of these small cell subsets is crucial to decipher system level biological changes.
-For this reason, an increasing number of studies rely assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples.
+For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples.
 A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
@@
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
-These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
 During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
-For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. 
+For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology. 
 This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
@@
 A common statistical problem is, for a given marker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
-Note that we use the term 'subject' throughout the paper, but the approaches described are general and can be applied to animal studies, for example.
+Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (e.g.
+animal studies).
 
@@
 
-These counts are compared between antigen stimulated and unstimulated samples from an subject to identify significant differences.
+These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences.
 Subjects who generate a response after stimulation are called \textit{responders} whereas subject that do not show any differences are called \textit{non-responders}.
@@
 Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
-All of these methods test subjects separately, and no information is shared across subjects even though one could expect some similarities across responders (\textit{resp.} non-responders).
+All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders).
 %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
@@
 In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects.
-This sharing of information across subjects helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
+This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
 Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
@@
 In Section \ref{s:DEone}, we present our model for testing differential marker expression in the univariate case.
-Section \ref{s:results} compares our approach to alternative methods and test the robustness of our model.
+Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model.
 In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using a single-cell gene expression data.
@@
 \textit{Flow cytometry}:
-The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen with 120 subjects \citep{Goonetilleke:2006jk}.
+The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen with 120 subjects \citep{Goepfert:2011ci}.
 The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
@@
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder.
-In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
+In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
 This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.
@@
 E-F) ENV-1-PTEG and expressing IFNg and/or IL2.
-ROC and FDR plots of other cytokine combinations can be found in Web Figure A.}
+ROC and FDR plots of other cytokine combinations can be found in Web Figure A.This figure appears in color in the electronic version of this article.}
 \label{fig:HVTN065}
@@
 Genes selected from Fisher's exact test at the 10\% FDR level.
-}
+This figure appears in color in the electronic version of this article.}
 \label{fig:fluidigm}
@@
 We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to data from ICS data (IL2 univariate) from the primary immunogenicity time point of the ICS data.
+Using hyper parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
 We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000 events, with ten independent realizations of data for each $N$.
@@
 Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
-Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.}\label{fig:simulations}
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.
+This figure appears in color in the electronic version of this article.}\label{fig:simulations}
 \end{figure}
@@
 No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
-Sample names highlighted in red identify those where MIMOSA did not detect a difference.}
+Sample names highlighted in red identify those where MIMOSA did not detect a difference.
+This figure appears in color in the electronic version of this article.}
 \label{fig:polyfunctionality}
@@
 Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we could not look at more than two makers at once.
-Therefore, we performed simulations in eight dimension to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
+Therefore, we performed simulations in eight dimension to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x8 tables, as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
 These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data (Figure~\ref{fig:mvsimulations} B).
@@
 A) Average ROC curves for the competing methods over 10 simulations.
-B) Average observed and nominal false discovery rate for each method over 10 simulations.}
+B) Average observed and nominal false discovery rate for each method over 10 simulations.
+This figure appears in color in the electronic version of this article.}
    \label{fig:mvsimulations}
@@
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
-In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single-cell assay.
+In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single-cell assays.
 
@@
 
-All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).
+All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/gfinak/MIMOSA}).
 An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.

### 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
Minor changes. Added "this figure appears in color …" to each figure caption as required.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
     \end{tikzpicture}
-   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
+   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.
+This figure appears in color in the electronic version of this article.}
 \label{webfig:HVTN065Results}
@@
 MIMOSA fit to two-sided data were simulated from the standard model with N=1,000 events.
-E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.}
+E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.
+This figure appears in color in the electronic version of this article.}
    \label{webfig:simulations_trunc}

### 2012-07-31 MIMOSA.tex
Added .pdf extension to figure names in \includegraphics statements
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2}};
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1}};
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
     \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15}};
+    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
     \begin{scope} [x={(baz.south east)},y={(baz.north west)}]

### 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
Added .pdf extension to figure names in \includegraphics statements
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
+    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3.pdf}};
         \begin{scope}[x={(b.south east)},y={(b.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
+    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5.pdf}};
         \begin{scope}[x={(c.south east)},y={(c.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6.pdf}};
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
+    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7.pdf}};
         \begin{scope}[x={(d.south east)},y={(d.north west)}]

### 2012-09-11 MIMOSA.tex
Reformat for Biometrics

--- 2012-07-31 MIMOSA.tex
@@
 \newcommand{\VS}{V\&S}
+%\usepackage{natbib}
 \newcommand{\tr}{\mbox{tr}}
-%\usepackage{geometry,authblk,multicol,amsmath}
+%\usepackage[margin=1in]{geometry}
+%\usepackage{authblk}
+%\usepackage{multicol}
+%\usepackage{tikz}
 \usepackage{amsmath,tikz}
@@
 
-%\usepackage[square,numbers,sort&compress]{natbib}
+%\usepackage[round,comma]{natbib}
 %\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
@@
 \usepackage{amssymb}
-\usepackage{booktabs}
-\usepackage{multirow}
+%\usepackage{booktabs}
+%\usepackage{multirow}
 \usepackage{url}
@@
 
-\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+%\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
+
 %\author[1]{Greg Finak}
-%\author[2]{Steve De Rosa}
+%\author[1]{Andrew McDavid}
+%\author[3]{Pratip Chattopadhyay}
+%\author[3]{Maria Dominguez}
+%\author[1,2]{Steve De Rosa}
 %\author[3]{Mario Roederer}
 %\author[1]{Raphael Gottardo}
-
+%
 %\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
@@
 %\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
-\author{Greg Finak$^{1}$, Andrew McDavid$^{1}$, \ldots, Steve De Rosa$^{1,2}$, Mario Roederer$^{3}$, and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
+
+\author{Greg Finak$^{1}$, \and
+Andrew McDavid$^{1}$,
+Pratip Chattopadhyay$^{3}$,
+Maria Dominguez$^{3}$, 
+Steve De Rosa$^{1,2}$,
+Mario Roederer$^{3}$,
+\and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
 $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
@@
 %\date{\today}                                       
+\begin{document}
 
-\begin{document}
-%\maketitle
-\date{{\it Received August} 2012.}
+\date{{\it Received XXXXX} 2012.}
 
@@
 \label{firstpage}
-
+%\maketitle
 \begin{abstract}
@@
 For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples.
-A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
-Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays.
+A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
+Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays.
 Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions.
 We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
-We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
+We compare our method against frequentist approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
 Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
 Additional simulations show that our framework is also robust to model misspecification.
-Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
+Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
 \end{abstract}
@@
 \end{keywords}
-
 \maketitle
 
+\clearpage
 \section{Introduction}
@@
 In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
-Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers).
-More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
-Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
+Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
+More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e.
+proteins, genes, cytokines, etc.) per cell~\citep{Bendall:2011wf}.
+Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
 A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
@@
 For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
-These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
-During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
+During analysis, cell level biomarker fluorescence intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
 For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology. 
 This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
- After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
+ After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers. 
 Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations. 
 When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
-A common statistical problem is, for a given marker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
-Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (e.g.
-animal studies).
+A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
+Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (\textit{e.g.}, animal studies).
 
@@
  After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}.
-These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells.
+The antigen--specific T-cell subpopulations (i.e.
+the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells.
 The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Individual cells are labelled using fluorescently conjugated antibodies against phenotypic biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
@@
 These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences.
-Subjects who generate a response after stimulation are called \textit{responders} whereas subject that do not show any differences are called \textit{non-responders}.
+Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
 In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
@@
 Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
-All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders).
-%In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here?
-Let's update this sentence later}.
+All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (or non-responders).
 
@@
 In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects.
-This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders.
-Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
-Our paper is organized as follows, Section \ref{s:data} introduces the data and notations used in the paper.
-In Section \ref{s:DEone}, we present our model for testing differential marker expression in the univariate case.
+This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increases sensitivity and specificity when detecting responders.
+Because our framework is multivariate in nature, multiple cell subsets can be modelled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
+Our paper is organized as follows; Section \ref{s:data} introduces the data and notations used in the paper.
+In Section \ref{s:DEone}, we present our model for testing differential biomarker expression in the univariate case.
 Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model.
-In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using a single-cell gene expression data.
+In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present some results using a single-cell gene expression data.
 Finally, in Section \ref{s:discussion} we discuss our findings and future work.
 
-\section{Data structure and notation} 
+\section{Notation and Data} 
 \label{s:data}
-In this paper we consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.
+
+In the remainder of the paper, we use the following notation to describe our model.
+We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated.
+Each cell can either be positive or negative for a biomarker.
+Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations.
+We denote by $n_{sik}$ and $n_{uik}$, $k=1,\dots, 2^K, i=1,\dots,I$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples, respectively.
+We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample, respectively.
+For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
+
+We consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.
 
 \textit{Flow cytometry}:
-The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen with 120 subjects \citep{Goepfert:2011ci}.
-The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
-Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points, day 0 and day 182. 
-Three cytokines (IFN$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (HIV Envelope peptide pool) matching part of the vaccine.
+The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
+The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets. 
+Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points: day 0 and day 182. 
+Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine.
 For ease of presentation we restricted ourselves to the CD4+ T-cell subsets.
@@
 \textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects.
-T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
-The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al.
-could you fill this in?}.
+T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
+The expression from the simulated samples  was compared to paired, unstimulated controls.
 
-In the remainder of the paper, we use the following notation to describe our model.
-From this point on, we assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated.
-Each cell can either be positive or negative for a marker.
-Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations.
-We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples.
-We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample.
-For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
-Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
+Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e.
+antigen or condition) at a time vs.
+unstimulated.
+The issue of multiple antigens is handled through multiple testing correction.
 
-\section{Differential expression with one marker}
+\section{Differential expression with one biomarker}
 \label{s:DEone}
-Datasets like the ones presented here are usually analyzed in in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
+Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
 Here, by univariate, we mean that we have only one positive cell subset.
-This cell subset can be defined by considering the expression of one marker alone (marginalizing over all other measured markers) such as IL2$+$ (\textit{vs.} IL2$-$), or considering a specific positive marker combination (and marginalizing over everything else) such as IL2$+$ and/or INF$\gamma+$ (\textit{vs.} IL2$-$/INF$\gamma-$).
-Without loss of generality, we treat the univariate case as a one marker case (\textit{i.e.} $K=1$).
+This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$).
+Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$).
 In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
@@
 \parbox{0.8\linewidth}{
-\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
+\caption{2 x 2 contingency table of counts for biomarker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
 \centering
@@
 \multicolumn{1}{l}{}&
-\multicolumn{2}{c}{Marker}\\
+\multicolumn{2}{c}{Biomarker}\\
  & Negative & Positive \\ 
@@
 
-For a given subject and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
-Subjects that show differential expression will be called responders for that marker.
-In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.
+For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
+Subjects that show differential expression will be called responders for that biomarker.
+In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described in what follows.
 
@@
 \label{s:DE}
-For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
+For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows:
 
@@
 \end{equation*}
-where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding subjects, we consider two competing models:
+where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models:
 \begin{equation*}
@@
 \end{equation*}
-Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal.
+Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ).
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder.
 In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
-This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.
+This alternative parametrization is described in Web Appendix B, and we refer to it as the one-sided model.
 
@@
 \label{ss:priors}
-Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows, 
+Our model shares information across all subjects using exchangeable~\citep{Bernardo:1996tl} Beta priors on the unknown proportions, as follows: 
  \begin{align*}
 (p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
-(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u)
+(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u),
  \end{align*}
-where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, i.e.
-${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects.
-Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model.
-Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the (unknown) proportion of responders.
-It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$.
-Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\citep{Dempster:1977ul} algorithm as described in Section \ref{s:estimation}.
-As an alternative, we also describe a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
-In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{s:estimation}.
+where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects.
+Note that the parameters $\alpha_u \text{ and } \beta_u$ are explicitly shared across the two models, whereas $\alpha_s \text{ and } \beta_s$ are only present in the alternative model. 
+Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders. 
+It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p_{ui}$ and $p_{si}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
+As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha_u, \beta_u$, $\alpha_s,\text{ and }\beta_s$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
+In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.
 
@@
 \label{s:estimation}
-Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative likelihoods, which greatly simplify our calculations.
+In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p_{si}$ and $p_{ui}$ from the null and alternative likelihoods.
 Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix A), and are given by,
+
  \begin{align*}
@@
 \end{align}
-Assuming that the missing data, the $z_i$'s, are known, we define the complete data log-likelihood as follows,
+Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i, i=1,\dots,I$, are known, we define the complete data log-likelihood:
 \begin{equation}
 \begin{split}
-l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w)\label{eq:cll}
+l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
 \end{split}
@@
 
-\noindent\textbf{EM algoritm}\\
+\noindent\textbf{EM algorithm}\\
 Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
@@
 \noindent\textbf{MCMC algorithm}\\
-Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}.
-All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
+We generated realizations from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}.
+All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
 Details about the algorithms are given in Web Appendix A.
-We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
-For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
-Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.
+We used the method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
+For each dataset presented here, we calculated that no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
+To leave some margin, we used 2,000,000 iterations after 50,000 burn-in iterations for each dataset explored here.
 
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}.
-
-%The constrained model was applied to an ICS data set from a real-world vaccine trial in order to identify responders to antigen stimulation.
-The unconstrained model was applied to Fluidigm single-cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single-cells.
-We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.
+We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}, where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder by the model (or a competing method).
 
 \subsection{ICS}
-Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
-Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders).
+Using the ICS data, we performed an ROC (receiver operator characteristic) analysis to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
+We considered observations at the day 0 time point as true negatives, and observations at the day 182 time point as true positives (potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182 being treated as true positives).
 The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
-At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positive being detected.
-Our comparison also show that ranking based on log-fold change alone is not reliable and should not be used.
+At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
+Our comparisons also show that ranking based on log-fold change alone is not reliable and should not be used.
 In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F).
-Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
+Here we present the results based on IL2 and IFN-$\gamma$ alone and the subset IL2 and/or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
 These results are consistent for other cytokines and cytokine combinations (see Web Figure A).
@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2}};
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.ps}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-    \node at (0,1) [font=\small\sffamily] {A} ;
-    \node at (0.5,1) [font=\small\sffamily] {B} ;
+    \node at (0,1) [font=\tiny\sffamily] {A} ;
+    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
     \end{scope}
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1}};
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.ps}};
     \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\small\sffamily] {C} ;
-    \node at (0.5,1) [font=\small\sffamily] {D} ;
+    \node at (0,1) [font=\tiny\sffamily] {C} ;
+    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
     \end{scope}
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15}};
+    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.ps}};
     \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-    \node at (0,1) [font=\small\sffamily] {E} ;
-    \node at (0.5,1) [font=\small\sffamily] {F} ;
+    \node at (0,1) [font=\tiny\sffamily] {E} ;
+    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
     \end{scope}
@@
 \end{tikzpicture}
-   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from HVTN065.
-Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFNg or C-D) ENV-1-PTEG and expressing IL2.
-E-F) ENV-1-PTEG and expressing IFNg and/or IL2.
+   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.
+Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2.
+E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2.
 ROC and FDR plots of other cytokine combinations can be found in Web Figure A.This figure appears in color in the electronic version of this article.}
@@
 \subsection{Single-cell gene expression}
-We applied the MIMOSA model to Fluidigm single-cell gene expression.
-This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation.
-In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation.
-The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A,B).
-These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
+We applied the MIMOSA model to a Fluidigm single-cell gene expression data set.
+We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation.
+In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation.
+The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A,B).
+These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
 A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
-Comparing the size of the gene sets at an FDR of 10\%, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.
+For an FDR of 10\%, Fisher's exact test identified 47 significant genes, while MIMOSA identified 50 significant genes.
+Both methods identified 39 genes in common.
 
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.ps}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-\node at (0,1) [font=\small\sffamily] {A} ;
+\node at (0,1) [font=\tiny\sffamily] {A} ;
 \end{scope}
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.ps}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
+\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.4\columnwidth]{Figures/FluidigmFisher.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.ps}};
 \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-\node at (0,1) [font=\small\sffamily] {C} ;
+\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
 \end{scope}
@@
 
-%Simulation studies were based on hyper-parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}.
-The statistical analysis of ICS data in the published trial is described in the original manuscript and outlined in the supplementary information (supplementary material~\ref{supp:statpublished})~\cite{Peiperl:2010ej}.
-The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.
+We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
+For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
+We simulated data from this constrained model with 200 observations, a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events, with ten independent realizations of data for each $N$.
+We fit the one-sided MIMOSA model to this data.
+We evaluated the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
+We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C).
+In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).
 
-We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
-Using hyper parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
-We simulated data from this constrained model with 200 observations, a response rate of 60\%, an $N$ of 1,000, 5,000, 10,000 events, with ten independent realizations of data for each $N$.
-The one-sided MIMOSA model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
-This procedure was repeated for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C).
-In addition, the nominal \textit{vs.} observed FDR was also examined to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).
-
-For both the constrained and unconstrained simulations, MIMOSA out-performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E).
+For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E).
 Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F).
-%Since the constrained model relies on Monte-Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.
 
@@
 \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
+    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.ps}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-        \node at (0,1) [font=\small\sffamily] {A} ;
+        \node at (0,1) [font=\tiny\sffamily] {A} ;
         \end{scope}
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
+    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.ps}};
     \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\small\sffamily] {B} ;
+    \node at (0,1) [font=\tiny\sffamily] {B} ;
     \end{scope}
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_twosided_ROC_5000.pdf}};
+    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.ps}};
     \begin{scope}[x={(c.south east)},y={(c.north west)}]
-    \node at (0,1) [font=\small\sffamily] {C} ;
+    \node at (0,1) [font=\tiny\sffamily] {C} ;
     \end{scope}
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
+    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.ps}};
     \begin{scope}[x={(d.south east)},y={(d.north west)}]
-    \node at (0,1) [font=\small\sffamily] {D} ;
+    \node at (0,1) [font=\tiny\sffamily] {D} ;
     \end{scope}
@@
    \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models.
-Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFNg expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
+Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
 A) Average ROC curve over the 10 simulated data sets (N=5,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000), one--sided.
@@
 
-To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the ICS data set (see Web Figure B panels C and D).
+To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Web Figure B panels C and D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
 
-\section{Differential expression across marker combinations}
+\section{Differential expression across biomarker combinations}
 \label{s:demarkercombos}
-Our beta-binomial model described in Section \ref{s:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations.
-As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
-%Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
+Our beta-binomial model described in Section \ref{s:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple biomarker combinations.
+As described in the data section, we now have counts for each biomarker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+
 \subsection{Model}
 
-In our multivariate model, the beta distribution is replaced by a multinomial distribution, as follows,
+In our multivariate model, the beta distribution is replaced by a multinomial distribution, as follows:
 \begin{equation*}
@@
  \end{equation*}
-%; \text{if $i$ is a non-responder}\\
-% &(\mathbf{n}_{ui}|\mathbf{p}_{ui}) \sim \mathcal{M}(\mathbf{p}_{ui},N_{ui}); (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(\mathbf{p}_{si},N_{si});\text{if $i$ is a responder}
-%\end{align} 
-where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown proportions for the un-stimulated and stimulated samples.
+
+where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown proportions for the un-stimulated and stimulated samples, respectively.
 
 \subsection{Prior}
-As in the one-marker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
+As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, such that
 \begin{align*}
-(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u)\\\nonumber
-(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s)%\label{eq:dir_prior}
-\end{align*}
-where the indicator variable $z_i$ is as defined in Section \ref{ss:priors}, i.e.
-$z_i\sim\mathrm{Be}(w)$ where $w$ is the proportion of responders.
-As in the beta-binomial case both an EM and MCMC algorithms can be used for parameter estimation.
+(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\\\nonumber
+(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s),\end{align*}
+where the indicator variable $z_i$ is defined in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
+As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation.
 When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}_{\{u,s\}}$ and $w$ as for the beta-binomial model.
@@
 \begin{align*}
-\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
+\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!},
 \end{align*}
@@
 \[
-\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_{si}|\mathbf{n}_{si},\mathbf{n}_{ui}) =\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \frac{N_{ui}!}{\prod_k n_{uik}!}.%\label{eq:postmult}
+\begin{split}
+\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_{si}|\mathbf{n}_{si},\mathbf{n}_{ui}) =\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot\\ \frac{N_{si}!}{\prod_k n_{sik}!} \frac{N_{ui}!}{\prod_k n_{uik}!}.
+\end{split}
 \]
-The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger.
-The EM algorithm is now initialized with the multivariate Fisher's exact test.
-In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than three.
-It becomes more dependent on the initial values and can fail to converge in many instances.
+The estimation procedures (both EM and MCMC based) for the Dirichlet--multinomial distribution are the same as for the beta-binomial model except that the number of parameters to estimate is larger.
+We initialize the $z_i$ in the EM algorithm with the positivity calls from the multivariate Fisher's exact test.
+In our experience, the performance of the EM algorithm greatly deteriorates for $K>3$, and is more dependent on the initial values and can fail to converge in many instances.
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
@@
 
-%\subsection{Parsimonious multivariate model representation}
-%In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model.
-The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$.
-As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of individuals to properly estimate all parameters.
-Here, we propose alternative model parametrizations that could be used to address this problem by reducing the number of required parameters.
-A simple solution would be to assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
-As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others.
-As a more flexible solution, we propose to define $K$ main effect parameters, denoted $\{\theta_k:k=1,\dots,K\}$, for the $K$ markers and represent co-expression parameters as a function of these main effects.
-So a combination of dregree $l$, with positive markers, $k_1$, $k_2$, \dots, $k_l$ would have a hyperparameter defined as $\tau_l(\theta_{k_1}\theta_{k_2}\cdots\theta_{k_l})^{1/l}$, which is the geometric mean of the corresponding main marker effects scaled by a degree effect, $\tau_l$.
-Note that the scaling degree effect would be the same for any combinations with the same number of positive markers.
-The rationale being that as the degree of polyfunctionality increases, the number of cells might decrease (or increase in some instances) and the corresponding parameters should be adjusted by a scaling factor to allow for such variation.
-The choice of the geometric mean versus an arithmetic mean is to account for the fact that when a particular combination of degree 1 is empty, \textit{i.e.} the corresponding $\theta$ value is close to zero, it is unlikely that other combination involving that maker would be large.
-Our specification would account for this since the product would be near zero if one of the $\theta$ is near zero.
-Finally, the null combination would have a single parameter equal to $\theta_0$.
-%
-%Assuming that the $\theta$'s and $\tau$'s are stimulation specific, this parametrization would result in $4K+3$ parameters, and would scale linearly with the number of markers providing a good alternative between the fully parametrize model and the model with constant rate for moderate to large values of $K$.
-In this case, we would recommend the use of an MCMC algorithm.
-The $\theta$'s would be given the same exponential priors than the unconstrained model, \textit{i.e.} exponentional with mean 100.
-As for the $\tau's$ we would use an exponential distribution with mean 1\footnote{check with real data to see if that makes sense}.
-%
-%\subsection{ICS example with two markers}
-%We applied the Multinomial-Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles.
-As a proof of principle we limited the model to the two-component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations.
-We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T-cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}).
-%
-%The day 28 stimulations show that the posterior probability of response is generally higher for the multivariate MIMOSA model than for the univariate models, and the day zero stimulation (Figure~\ref{fig:polyfunctionality}) shows that the multivariate model has fewer false positives than the univariate MIMOSA model fit to distinct cytokine combinations.
-We compared multivariate MIMOSA against Fisher's exact test.
-Fisher's exact test was used to test whether the number of cells expressing a specific combination of cytokines (i.e TNFa+/IL4-, TNFa+/IL4+, and so forth) was significantly higher in the stimulated samples than in the unstimulated samples.
-In this comparison, Fisher's exact test detected only one of the 8 responders detected by the MIMOSA model in the Env3 stimulated, day 28 data, and detected two false positives in the day zero data.
-Fisher's exact test also failed to identify four of the 12 responders identified by the MIMOSA model in the day 28, Env1 stimulated data.
-%
-%
-
 \subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
-As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
+As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely BIRC3 and CCL5. For this example, $K=2$, and we have four possible combinations. 
 In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
 Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN$\gamma+$) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
-That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN-$\gamma+$) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+That is, the number of BIRC3-/CCL5+ cells decreases upon stimulation, while the number of BIRC3+/CCL5+ cells increases.
 When marginalizing over these cell populations, no difference is apparent in any of the samples.
-In contrast, multivariate MIMOSA tests all polyfunctional cell subpopulations simultaneously, identifying significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels).
+In contrast, the multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels).
 Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.
@@
 \centering
-\begin{tikzpicture} [auto, node distance=0cm]
+\begin{tikzpicture} [auto, node distance=0 cm]
 \node at (0,0) (A) {
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.ps}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-\node at (0,1) [font=\small\sffamily] {A} ;
+\node at (0,1) [font=\tiny\sffamily] {A} ;
 \end{scope}
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.ps}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (0,1) [font=\small\sffamily] {B} ;
+\node at (-0.02,1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 };
-%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
-%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
 %\node at (0,0) [font=\small\sffamily] {C} ;
@@
 
-Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we could not look at more than two makers at once.
-Therefore, we performed simulations in eight dimension to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x8 tables, as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C).
-These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data (Figure~\ref{fig:mvsimulations} B).
+Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once.
+Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Figure~\ref{fig:mvsimulations} A-C).
+These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Figure~\ref{fig:mvsimulations} B).
 
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.ps}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-    \node at (0,1) [font=\small\sffamily] {A} ;
+    \node at (0,1) [font=\tiny\sffamily] {A} ;
 \end{scope}
@@
 \begin{tikzpicture}
-   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.4\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
+   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.ps}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-       \node at (0,1) [font=\small\sffamily] {B} ;
+       \node at (0,1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 \label{s:discussion}
-Experimentalists have already access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
-Single-cell assays will become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}.
-As a consequence, the development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important.
-Current approaches for single-cell assays are for the most part simplistic (t-test, $\chi^2$ test, Fisher's exact test), and resulting inference can be quite sub-optimal especially when the cell counts are small.
+Experimentalists already have access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
+As single-cell assays become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}, the development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important.
+Current approaches for single-cell assays are for the most part simplistic such as the t-test, $\chi^2$ test, and Fisher's exact test, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
-In addition, most of these methods are univariate in nature and inappropriate for high dimensional next generation single-cell assays.
+In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.
 
-The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.} vaccine responders and non-responders).
+The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders).
 Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Web Figure B).
@@
 
-%Importantly, the analysis of real-world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}).
-Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination, the model is naive to vaccine time point.
-Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre-vaccination) and 182 (post-vaccination) than the current standard approach (Fisher's exact test), or non-Bayesian alternatives such as the likelihood ratio test or ranking by log-fold change (Figure~\ref{fig:HVTN065} A,C).
-Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182.
-None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison.
-Additionally, MIMOSA was found to be a better fit to real-world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).
-
 Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
-In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.} the gene is not expressed in the cell), or the information is used for pre-filtering~\citep{Flatz:2011jb}.
+In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.}, the gene is not expressed in the cell), or the information is used for pre-filtering~\citep{Flatz:2011jb}.
 The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).
 
-Detecting differences in poly-functional cell populations (\textit{i.e.} identifying changes in cell populations that co-express multiple proteins, cytokines, or genes), is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\citep{Milush:2009bz}.
-In the context of HIV, these poly-functional cell populations have been shown to be correlated with good outcome such as vaccine protection and long-term disease non-progression~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
-In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected.
-It  follows that if a specific cell subset expressing multiple markers is being differentially expressed, differential expression based on the marginal cell counts should also be detected. 
+Detecting differences in poly-functional cell populations (\textit{i.e.}, identifying changes in cell populations that co-express multiple proteins, cytokines, or genes) is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\citep{Milush:2009bz}.
+In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies (e.g.
+in Leishmania) poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
+In the ICS data used here, the stimulation is expected to increase only the number of antigen specific cells detected.
+Hence, if a specific cell subset expressing multiple biomarkers is being differentially expressed, differential expression based on the marginal cell counts should also be detected. 
 As such, identifying poly-functional cytokine profiles from ICS data can be done in an iterative way. 
-First univariate tests on marginal populations are performed and then specific cell subsets expressing the positive markers detected are then tested. 
-Even in this situation, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that needs to be tested, and a multivariate approach might be preferable.
-In this case, as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\citep{Nason:2006dx}.
+First, univariate tests on marginal populations are performed, and then specific cell subsets expressing the positive biomarkers detected are tested. 
+However, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that need to be tested, and a multivariate approach might be preferable. 
+In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\citep{Nason:2006dx}.
 
-When changes are two-sided, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
-In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential marker expression. 
-Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).
+For two--sided changes, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
+In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential biomarker expression. 
+Here, we have shown that MIMOSA has higher sensitivity and specificity than the competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data obtained from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).
 Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes. 
-In the case of more than two markers, there is a combinatorial explosion in the number of parameters involved in our multinomial-Dirichlet mixture model.
-The number of parameters to be estimated is $2^{K+1}+1$, which can become very large even for moderate values of $K$.
-As an example, $K=4$ leads to 33 parameters, and this would require one to have a large number of subjects to properly estimate all parameters.
+In the case of more than two biomarkers, the number of parameters to estimate for our Dirichlet--multinomial model is $2^{K+1}+1$, which is large even for moderate values of K.
+As an example, we would need both, a large number of subjects and a large number of events (cells) collected, to properly estimate the 33 parameters for $K=4$.
+
 A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters. 
-For example, one could assume that the hyper-parameters are constant across marker combinations, \textit{i.e.} $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
-As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain markers more than others. 
+For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
+As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others. 
 More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.
 
-All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub (\url{http://www.github.org/gfinak/MIMOSA}).
-An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
-
-%\subsection*{Statistical Analysis of Responder and Non-responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
-%Positivity calls for vaccine responders and non-responders depend upon the selection of an appropriate threshold.
-Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods.
-Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines.
-In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one-sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.
-%
-%Our approach to modelling an individual's response to vaccine using ICS data takes a Bayesian approach.
-We model all observations (individuals) simultaneously for each combination of cytokine and stimulation (including the unstimulated samples).
-For a given cytokine, we let $n_s$ be the number of cytokine-positive cells in the stimulated sample, $N_s$ the total number of cells in the stimulated sample, and $n_u, N_u$, the number of cytokine-positive and total number of cells in the unstimulated sample, respectively.
-Note that usually, $N_u \ne N_s$.
-The observed count data $\mathbf{y}$ is a matrix of size $4 \times P$, where $P$ is the number of participants.
-For the $i$'th individual $\mathbf{y}_i = \left<N_{si},n_{si},N_{ui},n_{ui}\right>$, and can be represented as the following contingency table:
-% latex table generated in R 2.14.0 by xtable 1.5-6 package
-% Wed Oct 12 09:41:50 2011
-
-%\subsection{The Mixture of Beta-Binomials}
-%Although we have specified the two models for the data, we do not know which observation was generated by which model.
-Clearly, not all individuals are expected to exhibit an immune response to a stimulation.
-Any individual observation, $y_i$, could either be generated by model~\eqref{eq:null} or by model~\eqref{eq:alternate}.
-We capture this uncertainty with a mixture framework of the two competing beta-binomial models.
-The likelihood for the mixture is given by:
-
-%The classification of responder and non-responder individuals was compared between the multivariate MIMOSA model, univariate MIMOSA models applied to each cytokine combination, and Fisher's exact test applied to each cytokine combination.
-
-%The multivariate MIMOSA model detected two additional responders than Fisher's exact test at the 1\% FDR level (Table~\ref{tab:MDFisher}, top left).
-The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).
-In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4-fold and 2.5-fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples.
-These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7-fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.
-%
-%   \caption{Performance of the constrained and unconstrained Beta-binomial mixture model vs Fisher's exact test on simulated data.
-Data were simulated from a constrained model with hyper-parameters estimated from a real data set of Gag1 stimulated, CD4+, IL2 expressing T-cells on day 28 from the HVTN054 trial.
-For total cell counts from 10,000 to 150,000, we simulated ten data sets each of 500 observations with a response rate of 40\%.
-The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta-binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells.
-The beta distributions for the estimated and true hyper parameters are shown in the second panel for one simulated data set, with N=150,000 events.
-The ROC curve for Fisher's exact test and the constrained and unconstrained Beta-binomial model for the same simulation are shown in the third panel.
-The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta-binomial model are shown in the fourth panel.}
-%   \label{fig:simulations}
-
-%
-%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-%   \centering
-%   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
-%   \caption{Sensitivity to departures from model assumptions.
-We generated data from a variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions.
-The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta-prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity).
-The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel.
-The beta distributions for the estimated and true hyper parameters  are shown in the second panel.
-The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
-%   \label{fig:simulations_trunc}
-%\end{figure}
-
-%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-%   \centering
-%   \includegraphics[width=6in]{Figures/PositivityRates} 
-%   \caption{Comparison of MIMOSA and Fisher's exact test for calling responders in ICS data.
-Response rates for IL2 and IFNg cytokine positivity in Gag2 and Pol2 stimulated samples at days 0 and 28 in the CD4 and CD8 T-cell subpopulations of the HVTN054 ICS data.
-In the treatment groups, response rates for the MIMOSA model are greater than or equal to those for Fisher's exact test at day 28 but are equal (and zero) at day 0 as expected.
-In the control groups, response rates are equal (and zero) between Fisher's exact test and the MIMOSA model at day 0 and at day 28, as expected.}
-%   \label{fig:positivityrates}
-%\end{figure}
-
-%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-%   \centering
-%   \includegraphics[width=6in]{Figures/FluidigmMIMOSA.pdf} 
-%  % \caption{Likelihood surface and volcano plot for IFNg producing, CD4+ T-cells in Gag2 stimulated vs control samples on day 28.
-A significant difference between control and stimulated samples is called at the 1\% FDR threshold (red) for the Beta-binomial model, and at 1\% for FDR adjusted p-values from Fisher's exact test (green).
-The likelihood surface shows the observed counts from stimulated and unstimulated samples.
-The volcano plot shows the difference between the proportion of cytokine positive cells in the stimulated and unstimulated samples, for the maximum likelihood estimates of the proportions (triangles) and for the MAP estimates (crosses).
-The effect of shrinking the MAP estimates towards zero can be seen in the volcano plot.}
-%%   \label{fig:icsdata}
-%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
-The posterior probability of response times the sign of the change in expression is shown in the first panel (red indicates a significant decrease, green a significant increase, relative to the control).
-Columns and rows are clustered based on these signed posterior probabilities.
-The middle panel shows the $log_2$ odds ratio of the proportion of cells expressing a gene in the stimulated vs.
-control samples.
-Rows and columns are ordered as in the left panel for comparison.
-The right panel shows the difference in the proportion of cells expressing each gene in the stimulated vs.
-control samples.
-Ordering of the rows and columns is preserved as in the first panel.
-The traces show the deviations of each cell from zero.}
-%\label{fig:fluidigm}
-%\end{figure}
-
-%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-%   \centering
-%   \includegraphics[width=6in]{Figures/Polyfunctionality_ggplot2} 
-%   \caption{Posterior probabilities from the four-dimensional, Multivariate-Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env-2 peptide stimulated CD4+ T-cells.}
-%   \label{fig:polyfunctionality}
-%\end{figure}
-%
-%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-%   \centering
-%   \includegraphics[width=6in]{Figures/Polyfunctionality_Env1} 
-%   \caption{Posterior probabilities from the four-dimensional, Multivariate-Dirichlet, MIMOSA model, compared against posterior probabilities from univariate MIMOSA for each combination of the cytokines IL4 and TNFa, on days zero and 28 for Env-1 peptide stimulated CD4+ T-cells.}
-%   \label{fig:polyfunctionalityenv1}
-%\end{figure}
-%
+All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub \\
+(\url{http://www.github.org/gfinak/MIMOSA}). An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
 
 \backmatter
+
+\section*{Acknowledgments}
+This work was supported by the Intramural Research Program of the  National Institute of Allergy and Infectious Diseases (NIAID) and the National Institutes of Health (NIH), and by grants R01 EB008400 and U01 AI068635-01 to RG, grants \#OPP38744, and \#OPP1032317 from the Bill \& Melinda Gates Foundation to VISC (Vaccine Immunology Statistical Center), grant \#OPP1032325 from the Bill \& Melinda Gates Foundation to the CAVD (Collaboration for AIDS Vaccine Discovery).
+Study HVTN065 was conducted by the HIV Vaccine Trials Network (HVTN), and supported by the National Institute of Allergy and Infectious Diseases (NIAID).
+The the HVTN Laboratory Program is supported by grant \#UM1AI068618.
+Funding was also provided by Public Health Service grant UM1 AI068618 from the NIH and the University of Washington Center for AIDS Research (CfAR), an NIH-funded program (P30 AI027757).
+We thank the James B.
+Pendleton Charitable Trust for their generous equipment donation.
 
@@
 
-\section*{Acknowledgments}
-GF and RG acknowledge funding from National Institute of Health grants R01 EB008400 and U01 AI068635-01, and a Bill \& Melinda Gates Foundation grant OPP1032317. 
-%\section{Data Import, Preprocessing, and Gating}
-%The ICS assay data was imported into R from the original flowJo workspaces (version 6, TreeStar Inc, Ashland, OR)  using the BioConductor tool, \textit{flowWorkspace} (v 1.1.6) and ncdfFlow (v 1.1.4).
-Data were preprocessed using the flowJo-defined compensation matrices and data transformations extracted from the workspace file, and gated using methods from the flowCore package (v 1.19.2) to extract counts of cytokine positive and negative T-cells for each sample and stimulation~\cite{Hahne:2009vv}.
-
-%\section{Statistical Analysis in the Published Trial}
-%\label{supp:statpublished}
-%The methodology for statistical analysis and calling responders and non-responders in the original vaccine trial is described in the original publication~\cite{Peiperl:2010ej}.
-In general, a participant is called a ``responder'' to an antigen stimulation if, for a given cytokine, the number of cytokine-positive T-cells in the antigen-stimulated sample is significantly greater (for some statistical measure of significance) than the number of cytokine-positive T-cells for the negative control (unstimulated) sample from the same individual.
-In the original trial, significance was measured via one-sided Fisher's exact test for each participant and cytokine, comparing stimulated against unstimulated samples from that individual.
-In order to control the false positive rate for responder calls, made based on positivity for multiple cytokines, a discrete Bonferroni adjustment for multiple comparisons (across cytokines) was applied, and stimulations with an adjusted p-value below $\alpha = 0.00001$ were called positive~\cite{Horton:2007tsa}.
-
-%\subsection{Derivation of the Beta-Binomial Model for $p_s=p_u$ and $p_s>p_u$}
-%\label{s:derivation}
-%
-%We derive the posterior predictive distribution and marginal log-likelihood for the null model ($p_s=p_u$) and the alternative model ($p_s>p_u$) below: patient indices, $i$ on the $\left<N_s,n_s,N_u,n_u\right>$ are omitted for clarity.
-%For model \eqref{eq:null}, ($p_s=p_u, \alpha_s=\alpha_u, \beta_s=\beta_u$), the posterior predictive distribution of the data given the model is:
-% \begin{align}
-% 	\mathrm{Pr}(y_i|\alpha_u,\beta_u) &= \int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i,p_u|\alpha_u,\beta_u)dp_u\\
-%&=\int\limits_{p_u=0}\limits^{1} \mathrm{Pr}(y_i|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u\\
-%\begin{split}
-%	=\int\limits_{p_u=0}\limits^{1}\binom{N_s+n_s}{n_s}p_u^{n_s}(1-p_u)^{N_s}\binom{N_u+n_u}{n_u}p_u^{n_u}(1-p_u)^{N_u}\cdot \\ 
-%	\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1}dp_u
-% \end{split}\\
-% \begin{split}
-% 	=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\cdot\\ \int\limits_{p_u=0}\limits^{1}p_u^{n_s+n_u+\alpha_u-1}(1-p_u)^{N_s+N_u+\beta_u-1}dp_u
-%\end{split}\\
-%\intertext{The integrand is the kernel of a beta distribution with parameters $\left(n_u+n_s+\alpha_u,N_u+N_s+\beta_u\right)$, giving a closed form expression for the  posterior predictive distribution of model \eqref{eq:null}:}
-%\mathrm{Pr}(y_i|\alpha_u,\beta_u)&=\binom{N_s+n_s}{n_s}\binom{N_u+n_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1postsupp}\\
-%\intertext{with marginal log-likelihood:}
-%\begin{split}
-%\mathcal{L}(\alpha_u,\beta_u|\mathbf{y})=\sum_{i=1}^P\left[\log{\binom{N^i_{s}+n^i_{s}}{n^i_{s}}}+\log{\binom{N^i_{u}+n^i_{u}}{n^i_{u}}}+\right.\\
-%\left.\log{\left(\mathrm{B}(n^i_s+n^i_u+\alpha_u,N^i_s+N^i_u+\beta_u)\right)}\right]-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)\label{eq:model1MLLsupp}
-%\end{split}
-% \end{align} 
-% 
-%For model \eqref{eq:alternate}, ($p_s>p_u$), the posterior predictive distribution of the data given the model is:
-%\begin{align}
-% 	\mathrm{Pr}(y_i|\alpha_u,\beta_u,\alpha_s,\beta_s) &= \int\limits_{p_u=0}\limits^1\int\limits_{p_s>p_u}\limits^{1} \mathrm{Pr}(y_i,p_u,p_s|\alpha_u,\beta_u,\alpha_s,\beta_s) dp_u  dp_s\label{eq:postpredmodel2}\\
-%	\intertext{Assuming independence of stimulated and unstimulated observations in \eqref{eq:postpredmodel2}:} 
-%	&=\int\limits_{p_u=0}\limits^{1}\int\limits_{p_s>p_u}\limits^1 \mathrm{Pr}(n_s|p_s)\mathrm{Pr}(n_u|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)\mathrm{Pr}(p_s|\alpha_s,\beta_s)dp_u dp_s\\
-%		&=\int\limits_{p_u=0}\limits^{1}\mathrm{Pr}(n_u|p_u)\mathrm{Pr}(p_u|\alpha_u,\beta_u)dp_u \int\limits_{p_s>p_u}\limits^1 \mathrm{Pr}(n_s|p_s)\mathrm{Pr}(p_s|\alpha_s,\beta_s)dp_s\\
-%		\begin{split}
-%	=\int\limits_{p_u=0}\limits^{1}\left[ \binom{N_u+n_u}{n_u}p_u^{n_u}(1-p_u)^{N_u} \frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1}d p_u\right]\cdot \\ \int\limits_{p_s>p_u}\limits^{1}\left[ \binom{N_s+n_s}{n_s}p_s^{n_s}(1-p_s)^{N_s} \frac{1}{\mathrm{B}(\alpha_s,\beta_s)}p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}d p_s\right]
-%	\end{split}\\
-%	\begin{split}
-%	 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
-%	 \int\limits_{p_u=0}\limits^{1}\left[p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1} d p_u\right]\int\limits_{p_s>p_u}\limits^{1}\left[p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s+\beta_s-1}d p_s\right]
-%	\end{split}
-%	\end{align}
-%	\begin{align}
-%	\begin{split}
-%	 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
-%	 \int\limits_{p_u=0}\limits^{1}\left[p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1}d p_u \right]\left(1-\int\limits^{p_u}\limits_{p_s=0}\left[p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s+\beta_s-1}d p_s\right]\right)
-%	\end{split}
-%	\intertext{The second integral can be expressed as the CDF of the beta distribution, also known as the regularized incomplete beta function, $\mathrm{I}_x(\alpha,\beta)$}
-%	\begin{split}
-%	 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_s+\alpha_s,N_s+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
-%	 \int\limits_{p_u=0}\limits^{1}\left[p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1} \right]\left(1-\mathrm{I_{p_s}}(n_s+\alpha_s,N_s+\beta_s)\right)d p_u
-%	\end{split}\\
-%	\intertext{giving the posterior predictive distribution of model \eqref{eq:alternate}}
-%	\begin{split}
-%		 =\binom{N_u+n_u}{n_u} \binom{N_s+n_s}{n_s}\frac{\mathrm{B}(n_u+\alpha_u,N_u+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_s+\alpha_s,N_s+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}\cdot\\
-%	 \int\limits_{p_u=0}\limits^{1}\left[\frac{1}{\mathrm{B}(n_u+\alpha_u,N_u+\beta_u)}p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u+\beta_u-1} \right]\left(1-\mathrm{I_{p_u}}(n_s+\alpha_s,N_s+\beta_s)\right)d p_u\label{eq:model2postsupp}
-%	\end{split}
-%\end{align}
-%The integral in equation~\eqref{eq:model2postsupp} is computed numerically. The marginal log-likelihood of model~\eqref{eq:alternate} is then:
-%\begin{equation}
-%		 \begin{split}
-%		 \mathcal{L}(\alpha_s,\alpha_u,\beta_s,\beta_u|\mathbf{y})=-P\log\left(\mathrm{B}(\alpha_u,\beta_u)\right)-P\log\left(\mathrm{B}+(\alpha_s,\beta_s)\right)+\\ \sum_{i=0}^P\left\{\log\binom{N^i_u+n^i_u}{n^i_u}+ \log\binom{N^i_s+n^i_s}{n^i_s}+ \log\left(\mathrm{B}(n^i_u+\alpha_u,N^i_u+\beta_u)\right)+ \log\left(\mathrm{B}(n^i_s+\alpha_s,N^i_s+\beta_s)\right)+\right.
-\\ \left.
-\log\left(\hspace{1em}\int\limits_{p_u=0}\limits^{1}\left[\frac{1}{\mathrm{B}(n^i_u+\alpha_u,N^i_u+\beta_u)}p_u^{n^i_u+\alpha_u-1}(1-p_u)^{N^i_u+\beta_u-1} \right]\left(1-\mathrm{I_{p_u}}(n^i_s+\alpha_s,N^i_s+\beta_s)\right)d p_u\right)\right\}\label{eq:model2MLLsupp}
-%\end{split}
-%\end{equation}
-%\subsection{The Beta-Binomial Mixture Model}
-%For a given stimulation, not all individuals are expected to exhibit an immune response to that stimulation.
-Therefore, for a set of ICS data, any single observation, ($n^i_s,n^i_u$) could either be derived from model~\eqref{eq:null}, or from model~\eqref{eq:alternate}.
-We model this situation explicitly using a mixture-model framework.
 \bibliographystyle{biom}
+%\bibliographystyle{unsrtnat}
 \bibliography{MIMOSA}

### 2012-09-11 WebSupplementaryMaterialMIMOSA.tex
Reformat for Biometrics

--- 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
@@
 \documentclass{article}
-\title{Web-based Supplementary Materials for Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Steve De Rosa, Mario Roederer and Raphael Gottardo}
+\title{Web-based Supplementary Materials for Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Andrew McDavid, Pratip Chattopadhyay, Maria Dominguez, Steve De Rosa, Mario Roederer and Raphael Gottardo}
 
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
+    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3.ps}};
         \begin{scope}[x={(b.south east)},y={(b.north west)}]
-        \node at (0,1) [font=\small\sffamily] {A} ;
+        \node at (0,1) [font=\tiny\sffamily] {A} ;
                 \end{scope}
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
+    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5.ps}};
         \begin{scope}[x={(c.south east)},y={(c.north west)}]
-        \node at (0,1) [font=\small\sffamily] {B} ;
+        \node at (0,1) [font=\tiny\sffamily] {B} ;
         \end{scope}
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6.ps}};
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\small\sffamily] {C} ;
+        \node at (0,1) [font=\tiny\sffamily] {C} ;
                 \end{scope}
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
+    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7.ps}};
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\small\sffamily] {D} ;
+        \node at (0,1) [font=\tiny\sffamily] {D} ;
                 \end{scope}
@@
     \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.ps}};
          \begin{scope} [x={(c.south east)},y={(c.north west)}]
-                 \node at (0,1) [font=\small\sffamily] {A} ;
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
 	\end{scope}
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.ps}};
         \begin{scope} [x={(d.south east)},y={(d.north west)}]
-            \node at (0,1) [font=\small\sffamily] {B} ;
+            \node at (0,1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.ps}};
     \begin{scope} [x={(e.south east)},y={(e.north west)}]
-        \node at (0,1) [font=\small\sffamily] {C} ;
+        \node at (0,1) [font=\tiny\sffamily] {C} ;
     \end{scope}
@@
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.ps}};
         \begin{scope} [x={(f.south east)},y={(f.north west)}]
-        \node at (0,1) [font=\small\sffamily] {D} ;
+        \node at (0,1) [font=\tiny\sffamily] {D} ;
         \end{scope}
@@
     \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
+     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.ps}};
          \begin{scope} [x={(g.south east)},y={(g.north west)}]
-                 \node at (0,1) [font=\small\sffamily] {E} ;
+                 \node at (0,1) [font=\tiny\sffamily] {E} ;
 	\end{scope}
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.ps}};
         \begin{scope} [x={(h.south east)},y={(h.north west)}]
-            \node at (0,1) [font=\small\sffamily] {F} ;
+            \node at (0,1) [font=\tiny\sffamily] {F} ;
 \end{scope}

### 2012-09-14 MIMOSA.tex
Final Revisions From Raphael

@@
 \begin{keywords}
-Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
+Expectations-Maximization, Markov Chain Monte Carlo, Marginal
+Likelihood, Bayesian Modeling, Hierarchical Modeling, Immunology, Flow
+Cytometry, Single-Cell Gene Expression
 \end{keywords}
@@
 The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against phenotypic biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected.
@@
 
-Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
+Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
 All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (or non-responders).
@@
 The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
-Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points: day 0 and day 182. 
+Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
 Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine.
@@
 unstimulated.
-The issue of multiple antigens is handled through multiple testing correction.
 
@@
 \label{ss:priors}
-Our model shares information across all subjects using exchangeable~\citep{Bernardo:1996tl} Beta priors on the unknown proportions, as follows: 
+Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows:
  \begin{align*}
 (p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
-(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u),
+(p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u,
+\beta_u) \quad\mathrm{and}\quad (p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s),
  \end{align*}
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}, where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder by the model (or a competing method).
+We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}, where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder.
 
@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.ps}};
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.ps}};
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
     \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.ps}};
+    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
     \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.ps}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.ps}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.ps}};
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
 \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
@@
 
-For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C and Web Figure B, panel E).
-Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D, and Web Figure B panel F).
+For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C).
+Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D).
 
@@
 \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.ps}};
+    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.ps}};
+    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
     \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.ps}};
+    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
     \begin{scope}[x={(c.south east)},y={(c.north west)}]
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.ps}};
+    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
     \begin{scope}[x={(d.south east)},y={(d.north west)}]
@@
 (\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\\\nonumber
-(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s),\end{align*}
+ (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s) \quad \text{and}\quad (\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\end{align*}
 where the indicator variable $z_i$ is defined in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.ps}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.ps}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.ps}};
+    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.ps}};
+   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 Current approaches for single-cell assays are for the most part simplistic such as the t-test, $\chi^2$ test, and Fisher's exact test, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
-Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.
+Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy,Gottardo:2006}.
 In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.

### 2012-09-14 WebSupplementaryMaterialMIMOSA.tex
Final Revisions From Raphael

--- 2012-07-31 WebSupplementaryMaterialMIMOSA.tex
@@
 \begin{enumerate}
-\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc,Raftery:1992vp,Raftery:1996ws}.
+\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc}.
 \item Update $w$ by Gibbs sampling using the full conditional,
@@
         \begin{scope}[x={(b.south east)},y={(b.north west)}]
-        \node at (0,1) [font=\small\sffamily] {A} ;
+        \node at (0,1) [font=\tiny\sffamily] {A} ;
                 \end{scope}
@@
         \begin{scope}[x={(c.south east)},y={(c.north west)}]
-        \node at (0,1) [font=\small\sffamily] {B} ;
+        \node at (0,1) [font=\tiny\sffamily] {B} ;
         \end{scope}
@@
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\small\sffamily] {C} ;
+        \node at (0,1) [font=\tiny\sffamily] {C} ;
                 \end{scope}
@@
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
-        \node at (0,1) [font=\small\sffamily] {D} ;
+        \node at (0,1) [font=\tiny\sffamily] {D} ;
                 \end{scope}
@@
          \begin{scope} [x={(c.south east)},y={(c.north west)}]
-                 \node at (0,1) [font=\small\sffamily] {A} ;
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
 	\end{scope}
@@
         \begin{scope} [x={(d.south east)},y={(d.north west)}]
-            \node at (0,1) [font=\small\sffamily] {B} ;
+            \node at (0,1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
     \begin{scope} [x={(e.south east)},y={(e.north west)}]
-        \node at (0,1) [font=\small\sffamily] {C} ;
+        \node at (0,1) [font=\tiny\sffamily] {C} ;
     \end{scope}
@@
         \begin{scope} [x={(f.south east)},y={(f.north west)}]
-        \node at (0,1) [font=\small\sffamily] {D} ;
+        \node at (0,1) [font=\tiny\sffamily] {D} ;
         \end{scope}
@@
     };
-    \node [below=of C] (E) {
-    \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
-         \begin{scope} [x={(g.south east)},y={(g.north west)}]
-                 \node at (0,1) [font=\small\sffamily] {E} ;
-	\end{scope}
+%     \node [below=of C] (E) {
+%     \begin{tikzpicture}
+%      \node[anchor=south west,inner sep=0] at (0,0) (g) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_ROC_1000.pdf}};
+%          \begin{scope} [x={(g.south east)},y={(g.north west)}]
+%                  \node at (0,1) [font=\tiny\sffamily] {E} ;
+% 	\end{scope}
+%      \end{tikzpicture}
+%      };
+%      \node [right=of E] (F) {
+%      \begin{tikzpicture}
+%     \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
+%         \begin{scope} [x={(h.south east)},y={(h.north west)}]
+%             \node at (0,1) [font=\tiny\sffamily] {F} ;
+% \end{scope}
+%     \end{tikzpicture}
+%     };
  \end{tikzpicture}
-     };
-     \node [right=of E] (F) {
-     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (h) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Twosided_FDR_1000.pdf}};
-        \begin{scope} [x={(h.south east)},y={(h.north west)}]
-            \node at (0,1) [font=\small\sffamily] {F} ;
-\end{scope}
-    \end{tikzpicture}
-    };
- \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to data from a model violating model assumptions and to two-sided data with small counts.
-Data was simulated from a model where proportions were sampled from a truncated normal distribution over $[0,1]$ rather than a Beta distribution.
-A) The average ROC from 10 simulation with N=5,000 events.
-B) The average observed and nominal FDR from 10 simulations with N=5,000 events.
-C) Average ROC for N=1,000 events D) Average observed and nominal FDR for N=1,000 events.
-MIMOSA fit to two-sided data were simulated from the standard model with N=1,000 events.
-E) Average ROC curves from 10 simulations and F) average observed vs nominal FDR from 10 simulations.
-This figure appears in color in the electronic version of this article.}
+   \caption{Unconstrained MIMOSA model fit to data from a model
+     violating model assumptions and to two-sided data with small
+     counts. Data was simulated from a model where proportions were
+     sampled from a truncated normal distribution over $[0,1]$ rather
+     than a Beta distribution.  A) The average ROC from 10 simulation
+     with N=5,000 events. B) The average observed and nominal FDR from
+     10 simulations with N=5,000 events. C) Average ROC for N=1,000
+     events D) Average observed and nominal FDR for N=1,000 events.
+% MIMOSA fit to two-sided data were simulated from the standard model
+% with N=1,000 events. E) Average ROC curves from 10 simulations and
+% F) average observed vs nominal FDR from 10 simulations. This figure
+% appears in color in the electronic version of this article.
+}
    \label{webfig:simulations_trunc}

### 2012-09-14 MIMOSA.tex
updated link to MIMOSA on github

@@
 All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub \\
-(\url{http://www.github.org/gfinak/MIMOSA}). An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
+(\url{http://www.github.org/RGLab/MIMOSA}). An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
 

### 2012-09-14 MIMOSA.tex
Revisions for submission to Biometrics
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
-Details about the algorithms are given in Web Appendix A.
-We used the method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
-For each dataset presented here, we calculated that no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
-To leave some margin, we used 2,000,000 iterations after 50,000 burn-in iterations for each dataset explored here.
+Details about the algorithms are given in Web Appendix A. We used the
+method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
+the number of iterations, based on a short pilot run of the
+sampler. For each dataset presented here, we calculated that no more
+than about 500,000 iterations with 50,000 burn-in iterations (thinning
+interval of 50) was sufficient to estimate standard posterior quantities. To leave some margin, we used 1,000,000 iterations after 50,000 burn-in iterations for each dataset explored here.
 

### 2012-09-14 MIMOSA.tex
updated iterations required for convergence
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

--- 2012-09-14 MIMOSA.tex
@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
-Details about the algorithms are given in Web Appendix A.
-We used the method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler.
-For each dataset presented here, we calculated that no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
-To leave some margin, we used 2,000,000 iterations after 50,000 burn-in iterations for each dataset explored here.
+Details about the algorithms are given in Web Appendix A. We used the
+method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
+the number of iterations, based on a short pilot run of the
+sampler. For each dataset presented here, we calculated that no more
+than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
+To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.
 

### 2012-12-03 MIMOSA.tex
Revised for Biostatistics

@@
 
-%\documentclass[11pt]{article}
-\documentclass[useAMS,referee,usenatbib]{biom}
+\documentclass[12pt,oupdraft]{biostatistics}
+%\usepackage{tikz}
+%\usetikzlibrary{calc}
+%\usetikzlibrary{positioning}
+%\usepackage{authblk}
+\usepackage{amsmath}
+
+%\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
+%\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=blue} 
+\usepackage{url}
 \def\bSig\mathbf{\Sigma}
 \newcommand{\VS}{V\&S}
-%\usepackage{natbib}
 \newcommand{\tr}{\mbox{tr}}
-%\usepackage[margin=1in]{geometry}
-%\usepackage{authblk}
-%\usepackage{multicol}
-%\usepackage{tikz}
-\usepackage{amsmath,tikz}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
 
@@
 %\usepackage{graphicx}
-\usepackage{amssymb}
+%\usepackage{amssymb}
 %\usepackage{booktabs}
 %\usepackage{multirow}
-\usepackage{url}
+%\usepackage{url}
 
@@
 
-%\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
-\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
-
+\title{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+%\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
+%
 %\author[1]{Greg Finak}
@@
 
-\author{Greg Finak$^{1}$, \and
-Andrew McDavid$^{1}$,
-Pratip Chattopadhyay$^{3}$,
-Maria Dominguez$^{3}$,
-Steve De Rosa$^{1,2}$,
-Mario Roederer$^{3}$,
-\and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
-$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
+ \author{GREG FINAK$^{1\ast}$, 
+ ANDREW MCDAVID$^{1}$,
+ PRATIP CHATTOPADHYAY$^{3}$,
+ MARIA DOMINQUEZ$^{3}$,
+ STEVE DE ROSA$^{1,2}$,
+ MARIO ROEDERER$^{3}$,\\
+ RAPHAEL GOTTARDO$^{1}$
+ \\[4pt]
+\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
  $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
  $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
+ \\[2pt]
+ {gfinak@fhcrc.org}}
 
-%\date{\today}
+\date{}
 \begin{document}
 
-\date{{\it Received XXXXX} 2012.}
+%\date{{\it Received XXXXX} 2012.}
 
-\pagerange{\pageref{firstpage}-\pageref{lastpage}}
-\volume{ }
-\pubyear{ }
-\artmonth{ }
+\maketitle
+\footnotetext{To whom correspondence should be addressed.}
 
-\doi{12.1234/j.12345-1234.2012.12345.x}
-
-\label{firstpage}
-%\maketitle
 \begin{abstract}
-Blood and tissue are composed of many functionally distinct cell subsets.
+{Blood and tissue are composed of many functionally distinct cell subsets.
 In immunological studies, these can only be measured accurately using single-cell assays.
@@
 Additional simulations show that our framework is also robust to model misspecification.
-Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
+Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
+Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
+Immunology, Flow Cytometry, Single-Cell Gene Expression, MIMOSA}
 \end{abstract}
-
-\begin{keywords}
-Expectations-Maximization, Markov Chain Monte Carlo, Marginal
-Likelihood, Bayesian Modeling, Hierarchical Modeling, Immunology, Flow
-Cytometry, Single-Cell Gene Expression
-\end{keywords}
-\maketitle
 
@@
 
-Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e.
+%Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e.
 antigen or condition) at a time vs.
@@
 Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
-Here, by univariate, we mean that we have only one positive cell subset.
+By univariate, we mean that we have only one positive cell subset.
 This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$).
@@
 
-\begin{table}
-\centering
-\parbox{0.8\linewidth}{
-\caption{2 x 2 contingency table of counts for biomarker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
-\centering
-\begin{tabular}{rrr}
-
-  \hline
-\multicolumn{1}{l}{}&
-\multicolumn{2}{c}{Biomarker}\\
- & Negative & Positive \\
-  \hline
-Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\
-Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\
-   \hline
-\end{tabular}
-}
-\end{table}
+%\begin{table}
+%\centering
+%\parbox{0.8\linewidth}{
+%\caption{2 x 2 contingency table of counts for biomarker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
+%\centering
+%\begin{tabular}{rrr}
+%
+%  \hline
+%\multicolumn{1}{l}{}&
+%\multicolumn{2}{c}{Biomarker}\\
+% & Negative & Positive \\
+%  \hline
+%Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\
+%Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\
+%   \hline
+%\end{tabular}
+%}
+%\end{table}
 
@@
 Subjects that show differential expression will be called responders for that biomarker.
-In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described in what follows.
+In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described below.
 
@@
 In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
-This alternative parametrization is described in Web Appendix B, and we refer to it as the one-sided model.
+This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.
 
@@
 In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p_{si}$ and $p_{ui}$ from the null and alternative likelihoods.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix A), and are given by,
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Appendix B, Supplementary Material), and are given by,
 
- \begin{align*}
- \begin{split}
+ \begin{equation*}
   	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y}_i)
-=&\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
-\end{split}
- \end{align*}
+=\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot \frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
+ \end{equation*}
 
 and
-\begin{align}
-\begin{split}
+\begin{equation}
 \mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
-=&\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
+=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
 \label{model2:unconstrained}
-\end{split}
-\end{align}
+\end{equation}
 Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i, i=1,\dots,I$, are known, we define the complete data log-likelihood:
@@
 \begin{split}
-l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
+l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
 \end{split}
@@
 In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated.
-All calculations for the one-sided case are described in Web Appendix B.
+All calculations for the one-sided case are described in Appendix A of the Supplementary Material.
 
@@
 \[
-\begin{split}
-&\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\\ &\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
-\end{split}
+\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
 \]
@@
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
-Details about the algorithms are given in Web Appendix A. We used the
+Details about the algorithms are given in Appendix B of the Supplementary Material. We used the
 method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
@@
 \subsection{ICS}
-Using the ICS data, we performed an ROC (receiver operator characteristic) analysis to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders.
-We considered observations at the day 0 time point as true negatives, and observations at the day 182 time point as true positives (potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182 being treated as true positives).
-The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
+In the ICS data we considered observations at the day 0 time point as true negatives, and observations at the day 182 time point as true positives (potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182).
+The MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
 At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
@@
 Here we present the results based on IL2 and IFN-$\gamma$ alone and the subset IL2 and/or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
-These results are consistent for other cytokines and cytokine combinations (see Web Figure A).
+These results are consistent for other cytokines and cytokine combinations (see Supplementary Figure 1).
 
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture} [auto, node distance=0cm]
-\node at (0,0) (A){
-    \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
-    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {A} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
-    \end{scope}
-\end{tikzpicture}
- };
- \node [below=of A] (B) {
- \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
-    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {C} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
-    \end{scope}
-    \end{tikzpicture}
- };
-  \node [below=of B] (E) {
- \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
-    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {E} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
-    \end{scope}
-    \end{tikzpicture}
- };
-\end{tikzpicture}
-   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.
+%\begin{figure} %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics{TIKZFig1.pdf}
+%%\begin{tikzpicture} [auto, node distance=0cm]
+%%\node at (0,0) (A){
+%%    \begin{tikzpicture}
+%%    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
+%%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {A} ;
+%%    \node at (0.5,1) [font=\small\sffamily] {B} ;
+%%    \end{scope}
+%%\end{tikzpicture}
+%% };
+%% \node [below=of A] (B) {
+%% \begin{tikzpicture}
+%%    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
+%%    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {C} ;
+%%    \node at (0.5,1) [font=\small\sffamily] {D} ;
+%%    \end{scope}
+%%    \end{tikzpicture}
+%% };
+%%  \node [below=of B] (E) {
+%% \begin{tikzpicture}
+%%    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
+%%    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {E} ;
+%%    \node at (0.5,1) [font=\small\sffamily] {F} ;
+%%    \end{scope}
+%%    \end{tikzpicture}
+%% };
+%%\end{tikzpicture}
+%   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2.
 E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2.
-ROC and FDR plots of other cytokine combinations can be found in Web Figure A.This figure appears in color in the electronic version of this article.}
-\label{fig:HVTN065}
-\end{figure}
+ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
+%\label{fig:HVTN065}
+%\end{figure}
+
+We verified the assumption of a common distribution for the $p_{iu}$
+across subjects by examining the empirical estimates of $p_{iu}$ and the
+density of the Beta distribution with parameters $\hat{\alpha}_u,
+\hat{\beta}_u$ fitted to ENV-1-PTEG stimulated
+CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 2).
 
@@
 
-\begin{figure}
-\centering
-\begin{tikzpicture} [auto,node distance=0cm]
-\node at (0,0) (A) {
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
-\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-\node at (0,1) [font=\tiny\sffamily] {A} ;
-\end{scope}
-\end{tikzpicture}
-};
-\node [right=of A] (B) {
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
-\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
-\end{scope}
-\end{tikzpicture}
-};
-\node [below=of A] (C) {
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
-\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
-\end{scope}
-\end{tikzpicture}
-};
-\end{tikzpicture}
-\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
+%\begin{figure}
+%\centering
+%\includegraphics{TIKZFig2.pdf}
+%%\begin{tikzpicture} [auto,node distance=0cm]
+%%\node at (0,0) (A) {
+%%\begin{tikzpicture}
+%%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+%%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%%\node at (0,1) [font=\small\sffamily] {A} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%\node [right=of A] (B) {
+%%\begin{tikzpicture}
+%%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+%%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%%\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%\node [below=of A] (C) {
+%%\begin{tikzpicture}
+%%\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
+%%\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+%%\node at (-0.05,1) [font=\small\sffamily] {C} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%\end{tikzpicture}
+%\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
 The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
@@
 This figure appears in color in the electronic version of this article.}
-\label{fig:fluidigm}
-\end{figure}
-
+%\label{fig:fluidigm}
+%\end{figure}
+%
 \subsection{Simulation Studies}
@@
 
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-   \begin{tikzpicture} [auto, node distance=0cm]
-   \node at (0,0) (A) {
-\begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
-    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-        \node at (0,1) [font=\tiny\sffamily] {A} ;
-        \end{scope}
-    \end{tikzpicture}
-    };
-    \node [right=of A] (B) {
-    \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
-    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {B} ;
-    \end{scope}
-    \end{tikzpicture}
-    };
-        \node [below=of A] (C) {
-    \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
-    \begin{scope}[x={(c.south east)},y={(c.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {C} ;
-    \end{scope}
-    \end{tikzpicture}
-    };
-    \node [right=of C] (D) {
-    \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
-    \begin{scope}[x={(d.south east)},y={(d.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {D} ;
-    \end{scope}
-    \end{tikzpicture}
-    };
-
-\end{tikzpicture}
-   \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models.
+%\begin{figure} %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics{TIKZFig3.pdf}
+%%   \begin{tikzpicture} [auto, node distance=0cm]
+%%   \node at (0,0) (A) {
+%%\begin{tikzpicture}
+%%    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
+%%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+%%        \node at (0,1) [font=\small\sffamily] {A} ;
+%%        \end{scope}
+%%    \end{tikzpicture}
+%%    };
+%%    \node [right=of A] (B) {
+%%    \begin{tikzpicture}
+%%    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
+%%    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {B} ;
+%%    \end{scope}
+%%    \end{tikzpicture}
+%%    };
+%%        \node [below=of A] (C) {
+%%    \begin{tikzpicture}
+%%    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+%%    \begin{scope}[x={(c.south east)},y={(c.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {C} ;
+%%    \end{scope}
+%%    \end{tikzpicture}
+%%    };
+%%    \node [right=of C] (D) {
+%%    \begin{tikzpicture}
+%%    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
+%%    \begin{scope}[x={(d.south east)},y={(d.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {D} ;
+%%    \end{scope}
+%%    \end{tikzpicture}
+%%    };
+%%
+%%\end{tikzpicture}
+%   \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models.
 Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
@@
 Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
-Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Web Figure B.
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Supplementary Figure 3.
 This figure appears in color in the electronic version of this article.}\label{fig:simulations}
-\end{figure}
+%\end{figure}
 
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Web Figure B panels C and D).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 3 panels C and D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
@@
 \subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Web Appendix C).
+Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Appendix C of the Supplementary Material).
 For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 \[
-\begin{split}
-\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_{si}|\mathbf{n}_{si},\mathbf{n}_{ui}) =\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot\\ \frac{N_{si}!}{\prod_k n_{sik}!} \frac{N_{ui}!}{\prod_k n_{uik}!}.
-\end{split}
+\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_{si}|\mathbf{n}_{si},\mathbf{n}_{ui}) =\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \frac{N_{ui}!}{\prod_k n_{uik}!}.
 \]
@@
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
-More details about our multivariate MCMC algorithm is given in Web Appendix C.
+More details about our multivariate MCMC algorithm is given in Appendix C of the Supplementary Material.
 
@@
 
-\begin{figure}
-\centering
-\begin{tikzpicture} [auto, node distance=0 cm]
-\node at (0,0) (A) {
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
-\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-\node at (0,1) [font=\tiny\sffamily] {A} ;
-\end{scope}
-\end{tikzpicture}
-};
-\node [right=of A] (B){
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
-\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (-0.02,1) [font=\tiny\sffamily] {B} ;
-\end{scope}
-\end{tikzpicture}
-};
-%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
-%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
-%\node at (0,0) [font=\small\sffamily] {C} ;
-%\node at (6.25,0) [font=\small\sffamily] {D} ;
-\end{tikzpicture}
-\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
+%\begin{figure}
+%\centering
+%\includegraphics{TIKZFig4.pdf}
+%%\begin{tikzpicture} [auto, node distance=0 cm]
+%%\node at (0,0) (A) {
+%%\begin{tikzpicture}
+%%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
+%%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%%\node at (0,1) [font=\small\sffamily] {A} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%\node [right=of A] (B){
+%%\begin{tikzpicture}
+%%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
+%%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%%\node at (-0.02,1) [font=\small\sffamily] {B} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+%%%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+%%%\node at (0,0) [font=\small\sffamily] {C} ;
+%%%\node at (6.25,0) [font=\small\sffamily] {D} ;
+%%\end{tikzpicture}
+%\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
 No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
@@
 This figure appears in color in the electronic version of this article.}
-\label{fig:polyfunctionality}
-\end{figure}
+%\label{fig:polyfunctionality}
+%\end{figure}
+%
 
@@
 
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture} [auto, node distance=0cm]
-\node at (0,0) (A) {
-\begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
-\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {A} ;
-\end{scope}
-\end{tikzpicture}
-};
-\node [right=of A] (B) {
-\begin{tikzpicture}
-   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
-   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-       \node at (0,1) [font=\tiny\sffamily] {B} ;
-\end{scope}
-\end{tikzpicture}
-};
-\end{tikzpicture}
-   \caption{Multivariate simulations from a two-sided model.
+%
+%\begin{figure} %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics{TIKZFig5.pdf}
+%%\begin{tikzpicture} [auto, node distance=0cm]
+%%\node at (0,0) (A) {
+%%\begin{tikzpicture}
+%%    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
+%%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%%    \node at (0,1) [font=\small\sffamily] {A} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%\node [right=of A] (B) {
+%%\begin{tikzpicture}
+%%   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
+%%   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%%       \node at (0,1) [font=\small\sffamily] {B} ;
+%%\end{scope}
+%%\end{tikzpicture}
+%%};
+%%\end{tikzpicture}
+%   \caption{Multivariate simulations from a two-sided model.
 Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
@@
 This figure appears in color in the electronic version of this article.}
-   \label{fig:mvsimulations}
-\end{figure}
+%   \label{fig:mvsimulations}
+%\end{figure}
 
@@
 \label{s:discussion}
-Experimentalists already have access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR, to name a few.
-As single-cell assays become even more routine once sequencing at the single-cell level becomes practical~\citep{Ramskold:2012gj}, the development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important.
-Current approaches for single-cell assays are for the most part simplistic such as the t-test, $\chi^2$ test, and Fisher's exact test, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
+Experimentalists already have access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR.
+The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays will become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}.
+Current approaches for single-cell assays such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy,Gottardo:2006}.
@@
 The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders).
-Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Web Figure B).
+Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 3).
 The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
@@
 
-\backmatter
+\section*{Supplementary Materials}
+
+Appendix A, B, and C and Supplementary Figures 1--3 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
 
 \section*{Acknowledgments}
-This work was supported by the Intramural Research Program of the  National Institute of Allergy and Infectious Diseases (NIAID) and the National Institutes of Health (NIH), and by grants R01 EB008400 and U01 AI068635-01 to RG, grants \#OPP38744, and \#OPP1032317 from the Bill \& Melinda Gates Foundation to VISC (Vaccine Immunology Statistical Center), grant \#OPP1032325 from the Bill \& Melinda Gates Foundation to the CAVD (Collaboration for AIDS Vaccine Discovery).
+This work was supported by the Intramural Research Program of the  National Institute of Allergy and Infectious Diseases (NIAID) and the National Institutes of Health (NIH) [R01 EB008400, U01 AI068635-01 to RG], grants from the Bill and Melinda Gates Foundation VISC (Vaccine Immunology Statistical Center) [grant numbers OPP38744,  OPP1032317], a grant from the Bill \& Melinda Gates Foundation to the CAVD (Collaboration for AIDS Vaccine Discovery) [grant number OPP1032325].
 Study HVTN065 was conducted by the HIV Vaccine Trials Network (HVTN), and supported by the National Institute of Allergy and Infectious Diseases (NIAID).
-The the HVTN Laboratory Program is supported by grant \#UM1AI068618.
-Funding was also provided by Public Health Service grant UM1 AI068618 from the NIH and the University of Washington Center for AIDS Research (CfAR), an NIH-funded program (P30 AI027757).
+The the HVTN Laboratory Program is supported by the NIH [grant  number UM1AI068618].
+Funding was also provided by the Public Health Service from the NIH and the University of Washington Center for AIDS Research (CfAR), an NIH-funded program [grant numbers UM1 AI068618, P30 AI027757].
 We thank the James B.
@@
 
-\section*{Supplementary Materials}
-
-Web Appendices A, B, C, and Web Figures A and B referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:estimation},  and \ref{s:demarkercombos} are available in the attached Web-based supplementary material.\vspace*{-8pt}
-% This will be replaced at submission time
-%at the Biometrics website on Wiley Online Library.
-
-\bibliographystyle{biom}
+\bibliographystyle{biorefs}
 %\bibliographystyle{unsrtnat}
 \bibliography{MIMOSA}
-\label{lastpage}
+\clearpage
+
+\begin{table}
+\centering
+\parbox{0.8\linewidth}{
+\caption{2 x 2 contingency table of counts for biomarker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
+\centering
+\begin{tabular}{rrr}
+
+  \hline
+\multicolumn{1}{l}{}&
+\multicolumn{2}{c}{Biomarker}\\
+ & Negative & Positive \\
+  \hline
+Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\
+Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\
+   \hline
+\end{tabular}
+}
+\end{table}
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics{TIKZFig1.eps}
+%\begin{tikzpicture} [auto, node distance=0cm]
+%\node at (0,0) (A){
+%    \begin{tikzpicture}
+%    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
+%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {A} ;
+%    \node at (0.5,1) [font=\small\sffamily] {B} ;
+%    \end{scope}
+%\end{tikzpicture}
+% };
+% \node [below=of A] (B) {
+% \begin{tikzpicture}
+%    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
+%    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {C} ;
+%    \node at (0.5,1) [font=\small\sffamily] {D} ;
+%    \end{scope}
+%    \end{tikzpicture}
+% };
+%  \node [below=of B] (E) {
+% \begin{tikzpicture}
+%    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
+%    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {E} ;
+%    \node at (0.5,1) [font=\small\sffamily] {F} ;
+%    \end{scope}
+%    \end{tikzpicture}
+% };
+%\end{tikzpicture}
+   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.
+Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2.
+E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2.
+ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
+\label{fig:HVTN065}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{TIKZFig2.eps}
+%\begin{tikzpicture} [auto,node distance=0cm]
+%\node at (0,0) (A) {
+%\begin{tikzpicture}
+%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%\node at (0,1) [font=\small\sffamily] {A} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\node [right=of A] (B) {
+%\begin{tikzpicture}
+%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\node [below=of A] (C) {
+%\begin{tikzpicture}
+%\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
+%\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+%\node at (-0.05,1) [font=\small\sffamily] {C} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\end{tikzpicture}
+\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
+The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
+Columns and rows are clustered based on these signed posterior probabilities.
+B) The posterior differences in proportion of cells expressing a gene in the stimulated vs.
+control samples.
+Rows and columns are ordered as in A) for comparison.
+The traces show the deviations of each cell from zero.
+Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
+C) Clustering of the signed q-values from Fisher's exact test.
+Genes selected from Fisher's exact test at the 10\% FDR level.
+This figure appears in color in the electronic version of this article.}
+\label{fig:fluidigm}
+\end{figure}
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics{TIKZFig3.eps}
+%   \begin{tikzpicture} [auto, node distance=0cm]
+%   \node at (0,0) (A) {
+%\begin{tikzpicture}
+%    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
+%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+%        \node at (0,1) [font=\small\sffamily] {A} ;
+%        \end{scope}
+%    \end{tikzpicture}
+%    };
+%    \node [right=of A] (B) {
+%    \begin{tikzpicture}
+%    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
+%    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {B} ;
+%    \end{scope}
+%    \end{tikzpicture}
+%    };
+%        \node [below=of A] (C) {
+%    \begin{tikzpicture}
+%    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+%    \begin{scope}[x={(c.south east)},y={(c.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {C} ;
+%    \end{scope}
+%    \end{tikzpicture}
+%    };
+%    \node [right=of C] (D) {
+%    \begin{tikzpicture}
+%    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.45\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
+%    \begin{scope}[x={(d.south east)},y={(d.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {D} ;
+%    \end{scope}
+%    \end{tikzpicture}
+%    };
+%
+%\end{tikzpicture}
+   \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models.
+Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
+A) Average ROC curve over the 10 simulated data sets (N=5,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000), one--sided.
+C) Average ROC curves, two--sided model.
+D) Average observed and nominal FDR, two--sided model.
+Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Supplementary Figure 3.
+This figure appears in color in the electronic version of this article.}\label{fig:simulations}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{TIKZFig4.eps}
+%\begin{tikzpicture} [auto, node distance=0 cm]
+%\node at (0,0) (A) {
+%\begin{tikzpicture}
+%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
+%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%\node at (0,1) [font=\small\sffamily] {A} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\node [right=of A] (B){
+%\begin{tikzpicture}
+%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
+%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%\node at (-0.02,1) [font=\small\sffamily] {B} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+%%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+%%\node at (0,0) [font=\small\sffamily] {C} ;
+%%\node at (6.25,0) [font=\small\sffamily] {D} ;
+%\end{tikzpicture}
+\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
+No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
+Sample names highlighted in red identify those where MIMOSA did not detect a difference.
+This figure appears in color in the electronic version of this article.}
+\label{fig:polyfunctionality}
+\end{figure}
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics{TIKZFig5.eps}
+%\begin{tikzpicture} [auto, node distance=0cm]
+%\node at (0,0) (A) {
+%\begin{tikzpicture}
+%    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
+%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {A} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\node [right=of A] (B) {
+%\begin{tikzpicture}
+%   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
+%   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%       \node at (0,1) [font=\small\sffamily] {B} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\end{tikzpicture}
+   \caption{Multivariate simulations from a two-sided model.
+Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
+Multivariate MIMOSA was compared against Fisher's exact test, and the likelihood ratio test.
+A) Average ROC curves for the competing methods over 10 simulations.
+B) Average observed and nominal false discovery rate for each method over 10 simulations.
+This figure appears in color in the electronic version of this article.}
+   \label{fig:mvsimulations}
+\end{figure}
 

### 2012-12-03 PosteriorUnstimProps_065.Rnw
Revised for Biostatistics

\documentclass{article}

\begin{document}
<<load>>=
require(MIMOSA)
require(ggplot2)
setwd("~/Documents/manuscripts/PositivityCallsHypothesisTesting/MIMOSA_Paper/")
hvtn<-read.csv("~/Documents/Projects/HVTNTrials/v065_ics_3cyt_counts.csv")
hvtn$fcsfile <- "FOO"
colnames(hvtn)[1] <- "ID"
hvtn$ID <- factor(hvtn$ID)
hvtn.data <- ICS(hvtn,extra.identifiers="visitno")
@

<<userfuncs>>=
merge.icsdata <- function(x,y,...){
    z <- rbind(x,y)
    z <- setpData(z,rbind(getpData(x),getpData(y)))
    z
}
@

<<ifng>>=
v12 <- extractData(ics=hvtn.data,control="negctrl",stim="ENV-1-PTEG",subset=c("12","cd3+/cd4+","IFN+"),subset.mapping=c("visitno","parent","cytokine"))
v2 <- extractData(ics=hvtn.data,control="negctrl",stim="ENV-1-PTEG",subset=c("2","cd3+/cd4+","IFN+"),subset.mapping=c("visitno","parent","cytokine"))
d <- merge.icsdata(v2,v12)
fit.env.ifn <- MIMOSA:::.fitMCMC(d,MDMix(d,initonly=TRUE),210000,10000,1,FAST=TRUE)
fit.env.ifn$params <- fit.env$getmcmc()

v12 <- extractData(ics=hvtn.data,control="negctrl",stim="GAG-1-PTEG",subset=c("12","cd3+/cd4+","IFN+"),subset.mapping=c("visitno","parent","cytokine"))
v2 <- extractData(ics=hvtn.data,control="negctrl",stim="GAG-1-PTEG",subset=c("2","cd3+/cd4+","IFN+"),subset.mapping=c("visitno","parent","cytokine"))
d <- merge.icsdata(v2,v12)
fit.gag.ifn <- MIMOSA:::.fitMCMC(d,MDMix(d,initonly=TRUE),210000,10000,1,FAST=TRUE)
fit.gag.ifn$params <- fit.gag$getmcmc()

pd.ifn <- data.frame(getpData.icsdata(fit.env.ifn),empirical.p=prop.table(as.matrix(fit.env.ifn$n.unstim),margin=1)[,2])
env.params.ifn <- apply(fit.env.ifn$params,2,function(x){x <- density(x);x$x[which.max(x$y)]})
density.env.ifn <- dbeta(seq(0,5e-4,l=1000),shape2=env.params.ifn[3],shape1=env.params.ifn[4])

d1 <- density(apply(fit.env.ifn$params,1,function(x){
    rbeta(100,shape2=x[3],shape1=x[2])
}))

#average density
ggplot(data=pd.ifn)+geom_histogram(aes(x=empirical.p,y=..density..),alpha=0.25,fill="gray",col="black",binwidth=2.25e-5)+theme_bw()+facet_wrap(~antigen)+geom_line(aes(x=x,y=y),data=data.frame(x=c(d1$x),y=c(d1$y),antigen=c("ENV-1-PTEG")[gl(1,512)]),col="blue",lty=4,lwd=1)+scale_x_continuous(expression(Empirical~~hat(p)[u]))+scale_y_continuous(name="Density")+theme(axis.text.y=element_text(size=8),axis.text.x  = element_text(size=8))

@

<<post_density_plots,fig.width=86/24.5,fig.height=110/24.5>>=

ggplot(data=pd.ifn)+geom_histogram(aes(x=empirical.p,y=..density..),alpha=0.25,fill="gray",col="black",binwidth=2.25e-5)+theme_bw()+facet_wrap(~antigen)+geom_line(aes(x=x,y=y),data=data.frame(x=seq(0,5e-4,l=1000),y=c(density.env.ifn),antigen=c("ENV-1-PTEG")[gl(1,1000)]),col="blue",lty=4,lwd=1)+scale_x_continuous(expression(Empirical~~hat(p)[u]))+scale_y_continuous(name="Density")+theme(axis.text.y=element_text(size=8),axis.text.x  = element_text(size=8))

dev.copy2pdf(file="Figures/post_density_plots.pdf")
@
\end{document}

### 2012-12-03 SupplementaryMaterialMIMOSA.tex
Revised for Biostatistics

--- 2012-09-14 WebSupplementaryMaterialMIMOSA.tex
@@
 \documentclass{article}
-\title{Web-based Supplementary Materials for Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Andrew McDavid, Pratip Chattopadhyay, Maria Dominguez, Steve De Rosa, Mario Roederer and Raphael Gottardo}
-
 \usepackage[margin=1in]{geometry}
 \usepackage{amssymb}
+\usepackage{authblk}
+
 \usepackage{booktabs}
@@
 \usetikzlibrary{positioning}
+\title{Mixture Models for Single-Cell Assays with Application to Vaccine Studies: Supplementary Materials}
+
+ \author{GREG FINAK$^{1\ast}$, 
+ ANDREW MCDAVID$^{1}$,
+ PRATIP CHATTOPADHYAY$^{3}$,
+ MARIA DOMINGUEZ$^{3}$,
+ STEVE DE ROSA$^{1,2}$,
+ MARIO ROEDERER$^{3}$, 
+ RAPHAEL GOTTARDO$^{1}$\\
+\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
+ $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
+ $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}}
 \date{}
@@
 %\renewcommand{\thesection}{\Alph{section}}
-\renewcommand{\thesubsection}{Web Appendix \Alph{subsection}:}
+\renewcommand{\thesubsection}{Appendix \Alph{subsection}:}
 %\setcounter{section}{1}
@@
 %\setcounter{figure}{0}
-\renewcommand{\figurename}{\textbf{Web Figure}}
-\renewcommand{\thefigure}{\textbf{\Alph{figure}}}
+\renewcommand{\figurename}{\textbf{Supplementary Figure}}
+\renewcommand{\thefigure}{\textbf{\arabic{figure}}}
 \setcounter{figure}{0}
@@
 \section*{}
+
+\subsection{Constrained beta-binomial model}
+\label{supp:constrained}
+We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$.
+In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in the main manuscript by (1).
+Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)$ but requires computing \[
+Z(\alpha_u, \beta_u, \alpha_s, \beta_s)=\int_{0}^1{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}\int_{p_u}^1 p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}dp_sdp_u.
+\]
+Using this expression, the constrained alternative marginal likelihood can be written as
+\[
+\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
+=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.
+\label{model2:constrained}
+\]
+In general, there is no closed-form expression for $Z(\cdot)$, and a numerical approximation must be used.
+Let us denote by $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ the approximation.
+A natural way to estimate $\tilde{Z}$ is to use Monte Carlo integration.
+Indeed, we can write
+\begin{equation}
+\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)=\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\int_{0}^1\frac{{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}}{\mathrm{B}(\alpha_u,\beta_u)}[1-F_{\alpha_s,\beta_s}(p_u)]dp_u
+\label{equ:normZ}
+\end{equation}
+where $F_{\alpha_s,\beta_s}$ is the cumulative distribution function of a beta random variable with parameters $\alpha_s$ and $\beta_s$.
+Using this identity, it can be seen that $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ can be approximated by
+\[
+\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)\approx\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\sum_{k=1}^K[1-F_{\alpha_s,\beta_s}(X_k)]
+\]
+where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha_s$, $\beta_s$ and $K$ is the number of terms used in the Monte Carlo approximation.
+This approximation works relatively well with our EM implementation and does not significantly increase the computing time.
+Unfortunately, the number of terms (\textit{i.e.} value of $K$) required for the approximation to be good might be large and computing such a normalizing constant at each iteration would significantly slow down our MCMC implementation.
+As it tuns out, a better approximation
+can be obtained when $\alpha_s$ and $\beta_s$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,
+\[
+F_{\alpha_s,\beta_s}(p_u)=\sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^jp_u^{\beta_s+\alpha_s-j}.
+\label{eq:IBident}
+\]
+Then using this identity, we obtain
+\begin{align*}
+Z(\alpha_u, \beta_u, \alpha_s, \beta_s)&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\int_{0}^1(1-p_u)^{\beta_u-1+j}p_u^{\alpha_u-1+\beta_s+\alpha_s-j}dp_u\\
+&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\mathrm{B}(\beta_u+j)\mathrm{B}(\alpha_u+\beta_s+\alpha_s-j).
+\label{equ:normZ}
+\end{align*}
+Typically, in ICS data $\alpha_s$ is relatively small leading to relatively few terms in the sum.
+However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha_s$ and $\beta_s$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha_s$ is less than one) and can render mixing in the MCMC more difficult.
+In addition, even though the computation is exact and much faster for small values of $\alpha_s$, which is typically the case with ICS data, it is still more demanding than the unconstrained model.
+In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is greater than that of the stimulation sample, $p_s$.
+Indeed, in the one-sided case, if $p_u>p_s$ the associated individual should be a non-responder and thus $z_i=0$.
+In our experience, this computational shortcut performs just as well as the true one-sided implementation while being computationally much less demanding.
+
 \subsection{Computational details for the beta-binomial model}
@@
 In our case each parameter has the same exponential prior with mean $1,000$
-\subsection{Constrained beta-binomial model}
-\label{supp:constrained}
-We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$.
-In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in the main manuscript by (1).
-Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)$ but requires computing \[
-Z(\alpha_u, \beta_u, \alpha_s, \beta_s)=\int_{0}^1{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}\int_{p_u}^1 p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}dp_sdp_u.
-\]
-Using this expression, the constrained alternative marginal likelihood can be written as 
-\[
-\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i) 
-=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.
-\label{model2:constrained}
-\]
-In general, there is no closed-form expression for $Z(\cdot)$, and a numerical approximation must be used.
-Let us denote by $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ the approximation.
-A natural way to estimate $\tilde{Z}$ is to use Monte Carlo integration.
-Indeed, we can write
-\begin{equation}
-\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)=\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\int_{0}^1\frac{{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}}{\mathrm{B}(\alpha_u,\beta_u)}(1-F_{\alpha_s,\beta_s}(p_u))dp_u
-\label{equ:normZ}
-\end{equation}
-where $F_{\alpha_s,\beta_s}$ is the cumulative distribution function of a beta random variable with parameters $\alpha_s$ and $\beta_s$.
-Using this identity, it can be seen that $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ can be approximated by
-\[
-\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)\approx\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\sum_{k=1}^K(1-F_{\alpha_s,\beta_s}(X_k))
-\]
-where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha_s$, $\beta_s$ and $K$ is the number of terms used in the Monte Carlo approximation.
-This approximation works relatively well with our EM implementation and does not significantly increase the computing time.
-Unfortunately, the number of terms (\textit{i.e.} value of $K$) required for the approximation to be good might be large and computing such a normalizing constant at each iteration would significantly slow down our MCMC implementation.
-As it tuns out, a better approximation
-can be obtained when $\alpha_s$ and $\beta_s$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,
-\[
-F_{\alpha_s,\beta_s}(p_u)=\sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^jp_u^{\beta_s+\alpha_s-j}.
-\label{eq:IBident}
-\]
-Then using this identity, we obtain
-\begin{align*}
-Z(\alpha_u, \beta_u, \alpha_s, \beta_s)&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\int_{0}^1(1-p_u)^{\beta_u-1+j}p_u^{\alpha_u-1+\beta_s+\alpha_s-j}dp_u\\
-&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\mathrm{B}(\beta_u+j)\mathrm{B}(\alpha_u+\beta_s+\alpha_s-j).
-\label{equ:normZ}
-\end{align*}
-Typically, in ICS data $\alpha_s$ is relatively small leading to relatively few terms in the sum.
-However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha_s$ and $\beta_s$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha_s$ is less than one) and can render mixing in the MCMC more difficult.
-In addition, even though the computation is exact and much faster for small values of $\alpha_s$, which is typically the case with ICS data, it is still more demanding than the unconstrained model.
-In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is greater than that of the stimulation sample, $p_s$.
-Indeed, in the one-sided case, if $p_u>p_s$ the associated individual should be a non-responder and thus $z_i=0$.
-In our experience, this computational shortcut performs just as well as the true one-sided implementation while being computationally much less demanding.
 
@@
 
+ \begin{figure}
+   \centering
+   \begin{tikzpicture} [auto,node distance=0cm]
+     \node at (0,0) (A) {
+       \begin{tikzpicture}
+         \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/post_density_plots.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+           \node at (0,1) [font=\tiny\sffamily] {A} ;
+         \end{scope}
+       \end{tikzpicture}
+      };
+    \end{tikzpicture}
+    \caption{Histogram of the empirical proportions of unstimulated cells
+      and overlaid posterior densities of the beta distribution with
+      $\alpha_u$ and $\beta_u$ estimated from the data for ENV-1-PTEG
+      stimulated, IFN--$\gamma$+, CD4+ T--cells, demonstrating that the assumption of a common distribution for $p_{iu}$ across subjects is reasonable. }
+   \label{fig:empiricalpdist}
+ \end{figure}
+
 \begin{figure} %  figure placement: here, top, bottom, or page
@@
 \end{figure}
-\bibliographystyle{biom}
+\clearpage
+
+\bibliographystyle{biorefs}
 \bibliography{MIMOSA}

### 2012-12-03 TIKZFig1.tex
Revised for Biostatistics

\documentclass{article}
\usepackage{tikz,amsmath,siunitx}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows,snakes,backgrounds,patterns,matrix,shapes,fit,calc,shadows,plotmarks}
\usepackage[graphics,tightpage,active]{preview}
\PreviewEnvironment{tikzpicture}
\PreviewEnvironment{equation}
\PreviewEnvironment{equation*}
\newlength{\imagewidth}
\newlength{\imagescale}
\pagestyle{empty}
\thispagestyle{empty}
\begin{document}
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
    \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/2.pdf}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
\end{tikzpicture}
 };
 \node [below=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/1.pdf}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
 };
  \node [below=of B] (E) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=\columnwidth]{Figures/15.pdf}};
    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {E} ;
    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
    \end{scope}
    \end{tikzpicture}
 };
\end{tikzpicture}
\end{document}

### 2012-12-03 TIKZFig2.tex
Revised for Biostatistics

--- 2012-12-03 TIKZFig1.tex
@@
 \begin{document}
-\begin{tikzpicture} [auto, node distance=0cm]
-\node at (0,0) (A){
+\begin{tikzpicture} [auto,node distance=0cm]
+\node at (0,0) (A) {
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/2.pdf}};
-    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
 \node at (0,1) [font=\tiny\sffamily] {A} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 };
- \node [below=of A] (B) {
+\node [right=of A] (B) {
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/1.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot2.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {C} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
+\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 };
-  \node [below=of B] (E) {
+\node [below=of A] (C) {
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=\columnwidth]{Figures/15.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmFisher.pdf}};
 \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {E} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
+\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
 \end{scope}

### 2012-12-03 TIKZFig3.tex
Revised for Biostatistics

--- 2012-12-03 TIKZFig1.tex
@@
   \begin{tikzpicture} [auto, node distance=0cm]
-\node at (0,0) (A){
+   \node at (0,0) (A) {
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/2.pdf}};
+    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
         \node at (0,1) [font=\tiny\sffamily] {A} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
         \end{scope}
@@
     };
- \node [below=of A] (B) {
+    \node [right=of A] (B) {
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/1.pdf}};
-    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {C} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
+    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
+    \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {B} ;
     \end{scope}
@@
     };
-  \node [below=of B] (E) {
+        \node [below=of A] (C) {
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=\columnwidth]{Figures/15.pdf}};
-    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-    \node at (0,1) [font=\tiny\sffamily] {E} ;
-    \node at (0.5,1) [font=\tiny\sffamily] {F} ;
+    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+    \begin{scope}[x={(c.south east)},y={(c.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {C} ;
     \end{scope}
@@
     };
+    \node [right=of C] (D) {
+    \begin{tikzpicture}
+    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
+    \begin{scope}[x={(d.south east)},y={(d.north west)}]
+    \node at (0,1) [font=\tiny\sffamily] {D} ;
+    \end{scope}
+    \end{tikzpicture}
+    };
+
 \end{tikzpicture}

### 2012-12-03 TIKZFig4.tex
Revised for Biostatistics

--- 2012-12-03 TIKZFig2.tex
@@
 \begin{document}
-\begin{tikzpicture} [auto,node distance=0cm]
+\begin{tikzpicture} [auto, node distance=0 cm]
 \node at (0,0) (A) {
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 };
-\node [right=of A] (B) {
+\node [right=of A] (B){
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
+\node at (-0.02,1) [font=\tiny\sffamily] {B} ;
 \end{scope}
@@
 };
-\node [below=of A] (C) {
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmFisher.pdf}};
-\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
-\end{scope}
-\end{tikzpicture}
-};
-\end{tikzpicture}
-\end{document}
+%\node [anchor = south west, inner sep=0] at (0,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_Marginalized_BIRC3_CCL5.pdf}};
+%\node [anchor=south west, inner sep=0] at (6.5,-4) {\includegraphics[width=0.4\columnwidth]{Fluidigm_Multivariate_BIRC3_CCL5.pdf}};
+%\node at (0,0) [font=\small\sffamily] {C} ;
+%\node at (6.25,0) [font=\small\sffamily] {D} ;
+\end{tikzpicture}\end{document}

### 2012-12-03 TIKZFig6.tex
Revised for Biostatistics

--- 2012-12-03 TIKZFig2.tex
@@
 \begin{document}
-\begin{tikzpicture} [auto,node distance=0cm]
+\begin{tikzpicture} [auto, node distance=0cm]
 \node at (0,0) (A) {
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot1.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
-\end{scope}
-\end{tikzpicture}
-};
-\node [below=of A] (C) {
-\begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmFisher.pdf}};
-\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-\node at (-0.05,1) [font=\tiny\sffamily] {C} ;
+       \node at (0,1) [font=\tiny\sffamily] {B} ;
 \end{scope}

### 2013-05-14 MIMOSA.tex
Updated notation from _{s,u} to ^{(s,u)}.

@@
 Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations.
-We denote by $n_{sik}$ and $n_{uik}$, $k=1,\dots, 2^K, i=1,\dots,I$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples, respectively.
-We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample, respectively.
-For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples, respectively.
+We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample, respectively.
+For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$.
 Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
@@
 \begin{equation*}
-(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
+(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}
 \end{equation*}
-where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models:
+where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively.
+In order to detect responding subjects, we consider two competing models:
 \begin{equation*}
-{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
+{\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}
 \end{equation*}
@@
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder.
-In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$.
+In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p^{(s)}>p^{(u)}$.
 This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.
@@
  \begin{align*}
-(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
-(p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u,
-\beta_u) \quad\mathrm{and}\quad (p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s),
+(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
+(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
+\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
  \end{align*}
-where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects.
-Note that the parameters $\alpha_u \text{ and } \beta_u$ are explicitly shared across the two models, whereas $\alpha_s \text{ and } \beta_s$ are only present in the alternative model.
+where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}$ are unknown hyper-parameters shared across all subjects.
+Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are explicitly shared across the two models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
 Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
-It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p_{ui}$ and $p_{si}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
-Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
-As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha_u, \beta_u$, $\alpha_s,\text{ and }\beta_s$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
+It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p^{(u)}_{i}$ and $p^{(s)}_{i}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
+As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
 In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.
@@
 \label{s:estimation}
-In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p_{si}$ and $p_{ui}$ from the null and alternative likelihoods.
+In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p^{(s)}_{i}$ and $p^{(u)}_{i}$ from the null and alternative likelihoods.
 Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Appendix B, Supplementary Material), and are given by,
@@
  \begin{equation*}
-  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y}_i)
-=\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot \frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
+  	\mathrm{L}_0(\alpha^{(u)},\beta^{(u)}|\mathbf{y}_i)
+=\binom{N^{(u)}_{i}}{n^{(u)}_{i}}\binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+n^{(u)}_{i}+\alpha^{(u)},N^{(s)}_{i}-n^{(s)}_{i}+N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}
  \end{equation*}
@@
 \begin{equation}
-\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
-=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
+\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
+=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot\frac{\mathrm{B}(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}
 \label{model2:unconstrained}
@@
 \begin{split}
-l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
+l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha^{(u)}, \beta^{(u)}|\mathbf{y}_i) +(1-z_i) l_1(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)}|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w),\label{eq:cll}
 \end{split}
 \end{equation}
-where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.
-In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated.
+where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ is the vector of parameters to be estimated.
+In the one-sided case, the alternative prior specification must satisfy the constraint $p^{(s)}>p^{(u)}$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated.
 All calculations for the one-sided case are described in Appendix A of the Supplementary Material.
@@
 \noindent\textbf{EM algorithm}\\
-Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
+Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
 \[
-\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
+\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.
 \]
@@
 In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components.
-We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.
+We then use the estimated $z_i$'s to estimate the $p^{(u)}_{i}$'s and $p^{(s)}_{i}$'s and use these to set the hyper-parameters to their method-of-moments estimates.
 
@@
 
-We verified the assumption of a common distribution for the $p_{iu}$
-across subjects by examining the empirical estimates of $p_{iu}$ and the
-density of the Beta distribution with parameters $\hat{\alpha}_u,
-\hat{\beta}_u$ fitted to ENV-1-PTEG stimulated
+We verified the assumption of a common distribution for the $p^{(u)}_{i}$
+across subjects by examining the empirical estimates of $p^{(u)}_{i}$ and the
+density of the Beta distribution with parameters $\hat{\alpha}^{(u)},
+\hat{\beta}^{(u)}$ fitted to ENV-1-PTEG stimulated
 CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 2).
@@
 
-We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta-binomial mixture models via simulations.
+We examined the performance of the constrained ($p^{(s)}>p^{(u)}$) and unconstrained ($p^{(s)} \ne p^{(u)}$) beta-binomial mixture models via simulations.
 For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
@@
 Our beta-binomial model described in Section \ref{s:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple biomarker combinations.
-As described in the data section, we now have counts for each biomarker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$.
+As described in the data section, we now have counts for each biomarker combination, denoted by  $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$.
 
@@
 \begin{equation*}
- (\mathbf{n}_{ui}|\mathbf{p}_{ui},) \sim \mathcal{M}(N_{ui},\mathbf{p}_{ui})\quad\text{and}\quad (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(N_{si},\mathbf{p}_{si})\label{eq:mult_likeliehood}
+ (\mathbf{n}^{(u)}_{i}|\mathbf{p}^{(u)}_{i},) \sim \mathcal{M}(N^{(u)}_{i},\mathbf{p}^{(u)}_{i})\quad\text{and}\quad (\mathbf{n}^{(s)}_{i}|\mathbf{p}^{(s)}_{i}) \sim \mathcal{M}(N^{(s)}_{i},\mathbf{p}^{(s)}_{i})\label{eq:mult_likeliehood}
  \end{equation*}
 
-where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown proportions for the un-stimulated and stimulated samples, respectively.
+where $N^{\{s,u\}}_{i}=\sum\limits_{k=1}\limits^{2^K} n^{\{s,u\}}_{ik}$ are the number of cells collected and $\mathbf{p}^{(u)}_{i}$ and $\mathbf{p}^{(s)}_{i}$ are the unknown proportions for the un-stimulated and stimulated samples, respectively.
 
@@
 \begin{align*}
-(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\\\nonumber
- (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s) \quad \text{and}\quad (\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u),\end{align*}
+(\mathbf{p}^{(u)}_{i}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\\\nonumber
+ (\mathbf{p}^{(s)}_{i}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}^{(s)}) \quad \text{and}\quad (\mathbf{p}^{(u)}_{i}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\end{align*}
 where the indicator variable $z_i$ is defined in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
 As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation.
-When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}_{\{u,s\}}$ and $w$ as for the beta-binomial model.
+When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}^{\{u,s\}}$ and $w$ as for the beta-binomial model.
 
@@
 \begin{align*}
-\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!},
+\mathrm{L}_0(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \cdot \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!},
 \end{align*}
@@
 \[
-\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_{si}|\mathbf{n}_{si},\mathbf{n}_{ui}) =\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \frac{N_{ui}!}{\prod_k n_{uik}!}.
+\mathrm{L}_1(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}_{i}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) =\frac{\mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}) \mathrm{B}(\boldsymbol{\alpha}^{(s)}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(s)})\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!}.
 \]
@@
 \subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
-As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely BIRC3 and CCL5. For this example, $K=2$, and we have four possible combinations.
-In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
+As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
+In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
 Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation.
@@
 Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 3).
-The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
+The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
 This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.
@@
 A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.
-For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha_{\{s,u\}k}=\alpha_{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
+For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha^{\{s,u\}}_{k}=\alpha^{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
 As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
@@
   \hline
-Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\
-Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\
+Stimulated &   $N^{(s)}_{i} - n^{(s)}_{i}$ &   $n^{(s)}_{i}$ \\
+Unstimulated &   $N^{(u)}_{i}-n^{(u)}_{i}$ &   $n^{(u)}_{i}$ \\
    \hline

### 2013-05-14 SupplementaryMaterialMIMOSA.tex
Updated notation from _{s,u} to ^{(s,u)}.

@@
 \label{supp:constrained}
-We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$.
+We can define a model where we constrain the stimulated proportions under the alternative model such that $p^{(s)}>p^{(u)}$.
 In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in the main manuscript by (1).
-Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)$ but requires computing \[
-Z(\alpha_u, \beta_u, \alpha_s, \beta_s)=\int_{0}^1{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}\int_{p_u}^1 p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1}dp_sdp_u.
+Due to the constraint, the normalizing constant of the prior under the alternative (model ${\cal M}_1$) is not given by $\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})$ but requires computing \[
+Z(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})=\int_{0}^1{p^{(u)}}^{\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1}\int_{p^{(u)}}^1 p^{(s)\alpha^{(s)}-1}(1-p^{(s)})^{\beta^{(s)}-1}dp^{(s)}dp^{(u)}.
 \]
@@
 \[
-\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}_i)
-=\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{Z(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u,n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{Z(\alpha_u,\beta_u,\alpha_s,\beta_s)}.
+\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
+=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\frac{Z(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)},n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{Z(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)})}.
 \label{model2:constrained}
@@
 In general, there is no closed-form expression for $Z(\cdot)$, and a numerical approximation must be used.
-Let us denote by $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ the approximation.
+Let us denote by $\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})$ the approximation.
 A natural way to estimate $\tilde{Z}$ is to use Monte Carlo integration.
@@
 \begin{equation}
-\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)=\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\int_{0}^1\frac{{p_u}^{\alpha_u-1}(1-p_u)^{\beta_u-1}}{\mathrm{B}(\alpha_u,\beta_u)}[1-F_{\alpha_s,\beta_s}(p_u)]dp_u
+\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})=\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})\int_{0}^1\frac{{p^{(u)}}^{\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1}}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}[1-F_{\alpha^{(s)},\beta^{(s)}}(p^{(u)})]dp^{(u)}
 \label{equ:normZ}
 \end{equation}
-where $F_{\alpha_s,\beta_s}$ is the cumulative distribution function of a beta random variable with parameters $\alpha_s$ and $\beta_s$.
-Using this identity, it can be seen that $\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)$ can be approximated by
+where $F_{\alpha^{(s)},\beta^{(s)}}$ is the cumulative distribution function of a beta random variable with parameters $\alpha^{(s)}$ and $\beta^{(s)}$.
+Using this identity, it can be seen that $\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})$ can be approximated by
 \[
-\tilde{Z}(\alpha_u, \beta_u, \alpha_s, \beta_s)\approx\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)\sum_{k=1}^K[1-F_{\alpha_s,\beta_s}(X_k)]
+\tilde{Z}(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})\approx\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})\sum_{k=1}^K[1-F_{\alpha^{(s)},\beta^{(s)}}(X_k)]
 \]
-where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha_s$, $\beta_s$ and $K$ is the number of terms used in the Monte Carlo approximation.
+where the $X_k$'s are \textit{iid} beta distributed random variables with parameters $\alpha^{(s)}$, $\beta^{(s)}$ and $K$ is the number of terms used in the Monte Carlo approximation.
 This approximation works relatively well with our EM implementation and does not significantly increase the computing time.
@@
 As it tuns out, a better approximation
-can be obtained when $\alpha_s$ and $\beta_s$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,
+can be obtained when $\alpha^{(s)}$ and $\beta^{(s)}$ are integers. In this case, the cdf function in \eqref{equ:normZ} can be calculated exactly using integration by parts, as follows,
 \[
-F_{\alpha_s,\beta_s}(p_u)=\sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^jp_u^{\beta_s+\alpha_s-j}.
+F_{\alpha^{(s)},\beta^{(s)}}(p^{(u)})=\sum_{j=\beta^{(s)}}^{\beta^{(s)}+\alpha^{(s)}-1} \frac{(\beta^{(s)}+\alpha^{(s)}-1)!}{j!(\beta^{(s)}+\alpha^{(s)}-j)!}(1-p^{(u)})^jp^{(u)\beta^{(s)}+\alpha^{(s)}-j}.
 \label{eq:IBident}
@@
 \begin{align*}
-Z(\alpha_u, \beta_u, \alpha_s, \beta_s)&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\int_{0}^1(1-p_u)^{\beta_u-1+j}p_u^{\alpha_u-1+\beta_s+\alpha_s-j}dp_u\\
-&=\mathrm{B}(\alpha_s,\beta_s)\sum_{j=\beta_s}^{\beta_s+\alpha_s-1}\frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}\mathrm{B}(\beta_u+j)\mathrm{B}(\alpha_u+\beta_s+\alpha_s-j).
+Z(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}, \beta^{(s)})&=\mathrm{B}(\alpha^{(s)},\beta^{(s)})\sum_{j=\beta^{(s)}}^{\beta^{(s)}+\alpha^{(s)}-1}\frac{(\beta^{(s)}+\alpha^{(s)}-1)!}{j!(\beta^{(s)}+\alpha^{(s)}-j)!}\int_{0}^1(1-p^{(u)})^{\beta^{(u)}-1+j}p^{(u)\alpha^{(u)}-1+\beta^{(s)}+\alpha^{(s)}-j}dp^{(u)}\\
+&=\mathrm{B}(\alpha^{(s)},\beta^{(s)})\sum_{j=\beta}^{(s)\beta^{(s)}+\alpha^{(s)}-1}\frac{(\beta^{(s)}+\alpha^{(s)}-1)!}{j!(\beta^{(s)}+\alpha^{(s)}-j)!}\mathrm{B}(\beta^{(u)}+j)\mathrm{B}(\alpha^{(u)}+\beta^{(s)}+\alpha^{(s)}-j).
 \label{equ:normZ}
 \end{align*}
-Typically, in ICS data $\alpha_s$ is relatively small leading to relatively few terms in the sum.
-However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha_s$ and $\beta_s$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha_s$ is less than one) and can render mixing in the MCMC more difficult.
-In addition, even though the computation is exact and much faster for small values of $\alpha_s$, which is typically the case with ICS data, it is still more demanding than the unconstrained model.
-In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p_u$, is greater than that of the stimulation sample, $p_s$.
-Indeed, in the one-sided case, if $p_u>p_s$ the associated individual should be a non-responder and thus $z_i=0$.
+Typically, in ICS data $\alpha^{(u)}$ is relatively small leading to relatively few terms in the sum.
+However, the use of this exact identity in our MCMC algorithm requires the use of discrete priors on $\alpha^{(s)}$ and $\beta^{(s)}$, which can be restrictive in terms of fit (\textit{e.g.},  if the true $\alpha^{(s)}$ is less than one) and can render mixing in the MCMC more difficult.
+In addition, even though the computation is exact and much faster for small values of $\alpha^{(s)}$, which is typically the case with ICS data, it is still more demanding than the unconstrained model.
+In our case, we have decided to use the unconstrained model and simply fix the $z_i$ to zero if the empirical proportion for the un-stimulated sample, $p^{(u)}$, is greater than that of the stimulation sample, $p^{(s)}$.
+Indeed, in the one-sided case, if $p^{(u)}>p^{(s)}$ the associated individual should be a non-responder and thus $z_i=0$.
 In our experience, this computational shortcut performs just as well as the true one-sided implementation while being computationally much less demanding.
@@
 \begin{split}
-\mathrm{L_0}(\alpha_u,\beta_u|n_{si},n_{ui}) =&\int_0^1 \mathrm{Pr}(n_{si},n_{ui}|p_u)\pi(p_u|\alpha_u,\beta_u)\mathrm{d}p_u\\
-=& \int_0^1Pr(n_s|p_u)Pr(n_u|p_u)\pi(p_u|\alpha_u,\beta_u) \mathrm{d}p_u\\
-=&\int_0^1\binom{N_s}{n_s}p_u^{n_s}(1-p_u)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1} \mathrm{d}p_u\\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\int_0^1 p_u^{n_s+n_u+\alpha_u-1}(1-p_u)^{N_s+N_u-n_s-n_u+\beta_u-1} \mathrm{d}p_u\\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_s+n_u+\alpha_u,N_s+N_u-n_s-n_u+\beta_0)}{\mathrm{B}(\alpha_u,\beta_u)}.
+\mathrm{L_0}(\alpha^{(u)},\beta^{(u)}|n^{(s)}_{i},n^{(u)}_{i}) =&\int_0^1 \mathrm{Pr}(n^{(s)}_{i},n^{(u)}_{i}|p^{(u)})\pi(p^{(u)}|\alpha^{(u)},\beta^{(u)})\mathrm{d}p^{(u)}\\
+=& \int_0^1Pr(n^{(s)}|p^{(u)})Pr(n^{(u)}|p^{(u)})\pi(p^{(u)}|\alpha^{(u)},\beta^{(u)}) \mathrm{d}p^{(u)}\\
+=&\int_0^1\binom{N^{(s)}}{n^{(s)}}p^{(s)n^{(s)}}(1-p^{(u)})^{N^{(s)}-n^{(s)}}\binom{N^{(u)}}{n^{(u)}}p^{(u)n^{(u)}}(1-p^{(u)})^{N^{(u)}-n^{(u)}}\cdot \\&\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}p^{(u)\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1} \mathrm{d}p^{(u)}\\
+=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\int_0^1 p^{(u)n^{(s)}+n^{(u)}+\alpha^{(u)}-1}(1-p^{(u)})^{N^{(s)}+N^{(u)}-n^{(s)}-n^{(u)}+\beta^{(u)}-1} \mathrm{d}p^{(u)}\\
+=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{\mathrm{B}(n^{(s)}+n^{(u)}+\alpha^{(u)},N^{(s)}+N^{(u)}-n^{(s)}-n^{(u)}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}.
 \end{split}
@@
 \begin{split}
-\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{si},n_{ui}) =&\int_0^1\int_0^1 \mathrm{Pr}(n_{si},n_{ui}|p_u,p_s)\pi(p_u,p_s|\alpha_u,\beta_u,\alpha_s,\beta_s)\mathrm{d}p_s\mathrm{d}p_u \\
-=&\int_0^1\int_0^1\binom{N_s}{n_s}p_s^{n_s}(1-p_s)^{N_s-n_s}\binom{N_u}{n_u}p_u^{n_u}(1-p_u)^{N_u-n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_u^{\alpha_u-1}(1-p_u)^{\beta_u-1} \cdot\\
-&\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}p_s^{\alpha_s-1}(1-p_s)^{\beta_s-1} \mathrm{d}p_s\mathrm{d}p_u\\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}\frac{1}{\mathrm{B}(\alpha_s,\beta_s)}\int_0^1 p_u^{n_u+\alpha_u-1}(1-p_u)^{N_u-n_u+\beta_u-1} \cdot\\
-&\int_0^1 p_s^{n_s+\alpha_s-1}(1-p_s)^{N_s-n_s+\beta_s-1}\mathrm{d}p_s\mathrm{d}p_u \\
-=&\binom{N_s}{n_s}\binom{N_u}{n_u}\frac{\mathrm{B}(n_u+\alpha_u,N_u-n_u+\beta_u)\mathrm{B}(n_s+\alpha_s,N_s-n_s+\beta_s)}{\mathrm{B}(\alpha_u,\beta_u)\mathrm{B}(\alpha_s,\beta_s)}.
+\mathrm{L_1}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|n^{(s)}_{i},n^{(u)}_{i}) =&\int_0^1\int_0^1 \mathrm{Pr}(n^{(s)}_{i},n^{(u)}_{i}|p^{(u)},p^{(s)})\pi(p^{(u)},p^{(s)}|\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)})\mathrm{d}p^{(s)}\mathrm{d}p^{(u)} \\
+=&\int_0^1\int_0^1\binom{N^{(s)}}{n^{(s)}}p^{(s)n^{(s)}}(1-p^{(s)})^{N^{(s)}-n^{(s)}}\binom{N^{(u)}}{n^{(u)}}p^{(u)n^{(u)}}(1-p^{(u)})^{N^{(u)}-n^{(u)}} \cdot\\  &\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}p^{(u)\alpha^{(u)}-1}(1-p^{(u)})^{\beta^{(u)}-1}
+\frac{1}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}p^{(s)\alpha^{(s)}-1}(1-p^{(s)})^{\beta^{(s)}-1} \mathrm{d}p^{(s)}\mathrm{d}p^{(u)}\\
+=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{1}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\frac{1}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}\int_0^1 p^{(u)n^{(u)}+\alpha^{(u)}-1}(1-p^{(u)})^{N^{(u)}-n^{(u)}+\beta^{(u)}-1} \cdot\\
+&\int_0^1 p^{(s)n^{(s)}+\alpha^{(s)}-1}(1-p^{(s)})^{N^{(s)}-n^{(s)}+\beta^{(s)}-1}\mathrm{d}p^{(s)}\mathrm{d}p^{(u)} \\
+=&\binom{N^{(s)}}{n^{(s)}}\binom{N^{(u)}}{n^{(u)}}\frac{\mathrm{B}(n^{(u)}+\alpha^{(u)},N^{(u)}-n^{(u)}+\beta^{(u)})\mathrm{B}(n^{(s)}+\alpha^{(s)},N^{(s)}-n^{(s)}+\beta^{(s)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})\mathrm{B}(\alpha^{(s)},\beta^{(s)})}.
 \end{split}
@@
 \begin{enumerate}
-\item Update each $\alpha_u, \beta_u, \alpha_s$ and $\beta_s$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc}.
+\item Update each $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)}$ and $\beta^{(s)}$ by Metropolis-Hastings using a Gaussian symmetric proposal where the variance of the proposal is tuned for each parameter using the approach of~\cite{Gelman:2004tc}.
 \item Update $w$ by Gibbs sampling using the full conditional,
@@
 \[
-p_i=\frac{w\cdot\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s| n_{ui},n_{si})}{w\cdot\mathrm{L_1}(\alpha_u,\beta_u,\alpha_s,\beta_s|n_{ui},n_{si})+(1-w)\cdot\mathrm{L_0}(\alpha_u,\beta_u|n_{ui},n_{si})}.
+p_i=\frac{w\cdot\mathrm{L_1}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}| n^{(u)}_{i},n^{(s)}_{i})}{w\cdot\mathrm{L_1}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|n^{(u)}_{i},n^{(s)}_{i})+(1-w)\cdot\mathrm{L_0}(\alpha^{(u)},\beta^{(u)}|n^{(u)}_{i},n^{(s)}_{i})}.
 \]
 \end{enumerate}
-For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, (e.g. $\alpha_u$, below),
+For each updated parameter, step 1 above involves the calculation of the following acceptance ratio, (e.g. $\alpha^{(u)}$, below),
 \[
 \begin{split}
-\frac{\mathrm{L}(\alpha^{\text{new}}_u,\beta_u,\alpha_s,\beta_s|\cdots)\pi(\alpha^{\text{new}}_u)}{\mathrm{L}(\alpha_u,\beta_u,\alpha_s,\beta_s|\cdots)\pi(\alpha_u)}\\
+\frac{\mathrm{L}(\alpha^{(u)\text{new}},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\cdots)\pi(\alpha^{(u)\text{new}})}{\mathrm{L}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\cdots)\pi(\alpha^{(u)})}\\
 \end{split}
@@
 \[
-\mathrm{L}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_u,\mathbf{n}_s,\mathbf{z})=\prod_i\mathrm{L}_{z_i}(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{n}_u,\mathbf{n}_s)
+\mathrm{L}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{n}^{(u)},\mathbf{n}^{(s)},\mathbf{z})=\prod_i\mathrm{L}_{z_i}(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{n}^{(u)},\mathbf{n}^{(s)})
 \]
-and $\pi$ is the prior distribution of $\alpha_u$.
-The obvious changes in the above expression are made for the acceptance ratios of $\alpha_s,\beta_s,\beta_u$.
+and $\pi$ is the prior distribution of $\alpha^{(u)}$.
+The obvious changes in the above expression are made for the acceptance ratios of $\alpha^{(s)},\beta^{(s)},\beta^{(u)}$.
 In our case each parameter has the same exponential prior with mean $1,000$
@@
 \begin{split}
-\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_{si},\mathbf{n}_{ui})=&\idotsint\mathrm{Pr}(\mathbf{n}_{si},\mathbf{n}_{ui}|\mathbf{p}_u)\pi(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
-=&\idotsint \frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\prod_k p^{n_{sik}}_{uk} \frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\prod_k p^{n_{uik}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
-=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sik}+n_{uik}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
-=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
+\mathrm{L_0}(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})=&\idotsint\mathrm{Pr}(\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}|\mathbf{p}^{(u)})\pi(\mathbf{p}^{(u)}|\boldsymbol{\alpha}^{(u)})\mathrm{d}\mathbf{p}^{(u)}\\
+=&\idotsint \frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\prod_k p^{(u)n^{(s)}_{ik}}_{k} \frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\prod_k p^{(u)n^{(u)}_{ik}}_{k}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})}\prod_k p^{(u)\alpha^{(u)}_{k}-1}_{k}\mathrm{d}\mathbf{p}^{(u)}\\
+=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})}\int_0^1 \prod_k p^{(u)n^{(s)}_{ik}+n^{(u)}_{ik}+\alpha^{(u)}_{k}-1}_{k}\mathrm{d}\mathbf{p}^{(u)}\\
+=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{\mathrm{B}(\mathbf{n}^{(s)}_{i}+\mathbf{n}^{(u)}+\boldsymbol{\alpha}^{(u)})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})}
 \end{split}
@@
 \begin{split}
-\mathrm{L_1}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{ui},\mathbf{n}_{si})
+\mathrm{L_1}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(u)}_{i},\mathbf{n}^{(s)}_{i})
 %=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
-=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\idotsint\prod_k p^{n_{uik}+\alpha_{uk}-1}_{uk}\idotsint \prod_k p^{n_{sik}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
-=&\frac{\mathbf{N}_s!}{\prod_k {n}_{sik}!}\frac{\mathbf{N}_u!}{\prod_k {n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}.
+=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})\mathrm{B}(\boldsymbol{\alpha}^{(s)})}\idotsint\prod_k p^{(u)n^{(u)}_{ik}+\alpha^{(u)}_{k}-1}_{k}\idotsint \prod_k p^{(s)n^{(s)}_{ik}+\alpha^{(s)}_{k}-1}_{k}\mathrm{d}\mathbf{p}^{(u)}\mathrm{d}\mathbf{p}^{(s)}\\
+=&\frac{\mathbf{N}^{(s)}!}{\prod_k {n}^{(s)}_{ik}!}\frac{\mathbf{N}^{(u)}!}{\prod_k {n}^{(u)}_{ik}!}\frac{\mathrm{B}(\mathbf{n}^{(u)}_{i}+\boldsymbol{\alpha}^{(u)})\mathrm{B}(\mathbf{n}^{(s)}_{i}+\boldsymbol{\alpha}^{(s)})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})\mathrm{B}(\boldsymbol{\alpha}^{(s)})}.
 \end{split}
@@
 \noindent\textbf{MCMC algorithm}\\
-The MCMC algorithm for the Dirichlet-multinomial model is analogous to the beta-binomial, above. The parameter vectors $\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u$ are updated component-wise:
+The MCMC algorithm for the Dirichlet-multinomial model is analogous to the beta-binomial, above. The parameter vectors $\boldsymbol{\alpha}^{(s)},\boldsymbol{\alpha}^{(u)}$ are updated component-wise:
 \begin{enumerate}
-\item  Update each $\alpha_{sk}$, $\alpha_{uk}$ using a Gaussian symmetric proposal distribution with the variance of each proposal tuned using the approach of~\cite{Gelman:2004tc}.
+\item  Update each $\alpha^{(s)}_{k}$, $\alpha^{(u)}_{k}$ using a Gaussian symmetric proposal distribution with the variance of each proposal tuned using the approach of~\cite{Gelman:2004tc}.
 \item Update $w$ by Gibbs sampling using the full conditional,
@@
 \[
-p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui})}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}_{u},\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui})+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}_{u}|\mathbf{n}_{si},\mathbf{n}_{ui})}
+p_i = \frac{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})}{w\cdot\mathrm{L_1}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})+(1-w)\cdot\mathrm{L_0}(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i})}
 \]
-For each parameter component updated in step 1 above, compute the acceptance ratio (e.g. $\alpha_{uk}$, below):
+For each parameter component updated in step 1 above, compute the acceptance ratio (e.g. $\alpha^{(u)}_{k}$, below):
 \[
-\frac{\mathrm{L}(\alpha^{\text{new}}_{uk},\boldsymbol{\alpha}_{u\{-k\}},\boldsymbol{\alpha}_s|\cdots)\pi(\alpha^{\text{new}}_{uk})}{\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\cdots)\pi(\alpha_{uk})}\\
+\frac{\mathrm{L}(\alpha^{(u)\text{new}}_{k},\boldsymbol{\alpha}^{(u)}_{\{-k\}},\boldsymbol{\alpha}^{(s)}|\cdots)\pi(\alpha^{(u)\text{new}}_{k})}{\mathrm{L}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\cdots)\pi(\alpha^{(u)}_{k})}\\
 \]
 
-where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\alpha}_{u\{-k\}} = \{\alpha_{uj} :  j\neq k\}$.
+where $\pi$ is the prior distribution of the parameter, and $\boldsymbol{\alpha}^{(u)}_{\{-k\}} = \{\alpha^{(u)}_{j} :  j\neq k\}$.
 Again, we have used the same exponential prior with mean 1,000 for each parameter.
@@
 \[
-\mathrm{L}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui},\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_{si},\mathbf{n}_{ui}).
+\mathrm{L}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i},\mathbf{z}) = \prod_i \mathrm{L}_{z_i}(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}).
 \]
@@
       and overlaid posterior densities of the beta distribution with
-      $\alpha_u$ and $\beta_u$ estimated from the data for ENV-1-PTEG
+      $\alpha^{(u)}$ and $\beta^{(u)}$ estimated from the data for ENV-1-PTEG
       stimulated, IFN--$\gamma$+, CD4+ T--cells, demonstrating that the assumption of a common distribution for $p_{iu}$ across subjects is reasonable. }

### 2013-05-15 TIKZFig1.tex
Updated figure 1

@@
     \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/2.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=\columnwidth]{Figures/revised_2.pdf}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/1.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=\columnwidth]{Figures/revised_1.pdf}};
     \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=\columnwidth]{Figures/15.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=\columnwidth]{Figures/revised_15.pdf}};
     \begin{scope} [x={(baz.south east)},y={(baz.north west)}]

### 2013-05-15 SupplementaryMaterialMIMOSA.tex
Updated figure S1

@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/revised_3.pdf}};
         \begin{scope}[x={(b.south east)},y={(b.north west)}]
@@
  \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/revised_5.pdf}};
         \begin{scope}[x={(c.south east)},y={(c.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/revised_6.pdf}};
         \begin{scope}[x={(d.south east)},y={(d.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/revised_7.pdf}};
         \begin{scope}[x={(d.south east)},y={(d.north west)}]

### 2013-05-15 TIKZFig2.tex
Updated Figure 2

@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_plot2.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmEmpiricalProps.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmFisher.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmCombinedFit.pdf}};
 \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
@@
 };
+\node [below=of B] (D) {
+\begin{tikzpicture}
+\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.5\columnwidth]{Figures/FluidigmFisher.pdf}};
+\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
+\node at (-0.05,1) [font=\tiny\sffamily] {D} ;
+\end{scope}
+\end{tikzpicture}
+};
+
 \end{tikzpicture}

### 2013-05-15 MIMOSA.tex
Updated figure 1
Update legend for Figure 2

@@
 \documentclass[12pt,oupdraft]{biostatistics}
-%\usepackage{tikz}
-%\usetikzlibrary{calc}
-%\usetikzlibrary{positioning}
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
 %\usepackage{authblk}
@@
 %    \begin{tikzpicture}
-%    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/2.pdf}};
+%    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/revised_2.pdf}};
 %    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
 % \begin{tikzpicture}
-%    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/1.pdf}};
+%    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/revised_1.pdf}};
 %    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
@@
 % \begin{tikzpicture}
-%    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/15.pdf}};
+%    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/revised_15.pdf}};
 %    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
@@
 %\end{tikzpicture}
-\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array.
-The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control).
+\caption{Signed posterior probability and difference in empirical  in proportions of single-cells expressing each gene on a 96x96 Fluidigm array.
+The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control) for a model fit to each stimulation and gene separately.
 Columns and rows are clustered based on these signed posterior probabilities.
-B) The posterior differences in proportion of cells expressing a gene in the stimulated vs.
+B) The empirical differences in proportion of cells expressing a gene in the stimulated vs.
 control samples.
@@
 Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).
-C) Clustering of the signed q-values from Fisher's exact test.
+C) Signed posterior probabilities for a model fit per gene but across all stimulations simultaneously.
+D) Clustering of the signed q-values from Fisher's exact test.
 Genes selected from Fisher's exact test at the 10\% FDR level.

### 2013-05-15 MIMOSA.tex
Update Figure 4

@@
 %\end{tikzpicture}
-\caption{Counts of cells expressing different combinations of BIRC3 and CCL5 genes in the A) unstimulated and B) stimulated conditions.
-No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 13 of 16 samples.
+\caption{Counts of cells expressing A) different combinations of CCR7 and GZMK genes in the unstimulated and stimulated conditions (+/+,+/-,-/+,-/-), and for B) the marginalized positive counts in stimulated and unstimulated conditions.
+No difference is observed from the marginalized counts, while multivariate MIMOSA detects a difference between stimulated and unstimulated conditions in 12 of 16 samples, while Fisher's test detects 9 of 16.
 Sample names highlighted in red identify those where MIMOSA did not detect a difference.

### 2013-05-15 TIKZFig4.tex
Update Figure 4

@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Unstim_Counts.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/revised_Fluidigm_Multivariate_CCR7_GZMK.pdf}};
 \begin{scope} [x={(foo.south east)},y={(foo.north west)}]
@@
 \begin{tikzpicture}
-\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Fluidigm_Multivariate_BIRC3_CCL5_Stim_Counts.pdf}};
+\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/revised_Fluidigm_Marginal_CCR7_GZMK.pdf}};
 \begin{scope} [x={(bar.south east)},y={(bar.north west)}]

### 2013-05-16 MIMOSA.tex
Updated Supp figures

@@
    \caption{Comparison of positivity detection methods on data simulated from the one-sided and two--sided models.
-Ten simulations were generated at an $N$ of 5,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
-A) Average ROC curve over the 10 simulated data sets (N=5,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000), one--sided.
+Ten simulations were generated at an $N$ of 50,000 total counts using hyper-parameter estimates from real ICS data (IFN-$\gamma$ expressing CD4+ T-cells stimulated with ENV-1-PTEG from HVTN065) with a five-fold effect size between responder and non-responder components.
+A) Average ROC curve over the 10 simulated data sets (N=50,000), one--sided B) Average observed and nominal false discovery rate over 10 simulated data sets (N=50,000), one--sided.
 C) Average ROC curves, two--sided model.
@@
 Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
-Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Supplementary Figure 3.
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N and I (number of subjects) are in Supplementary Figure 3.
 This figure appears in color in the electronic version of this article.}\label{fig:simulations}

### 2013-05-16 SupplementaryMaterialMIMOSA.tex
Updated Supp figures

@@
     \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_5000.pdf}};
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_50K.pdf}};
          \begin{scope} [x={(c.south east)},y={(c.north west)}]
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_5000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_50K.pdf}};
         \begin{scope} [x={(d.south east)},y={(d.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_1000.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_10K.pdf}};
     \begin{scope} [x={(e.south east)},y={(e.north west)}]
@@
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_1000.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_10K.pdf}};
         \begin{scope} [x={(f.south east)},y={(f.north west)}]

### 2013-05-16 TIKZFig3.tex
Updated Supp figures

@@
 \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Onesided_ROC_5000.pdf}};
+    \node [anchor=south west] at (0,0) (foo) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_OneSided_ROC_50K.pdf}};
     \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Onesided_FDR_5000.pdf}};
+    \node [anchor=south west] at (0,0) (bar) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_OneSided_FDR_50K.pdf}};
     \begin{scope}[x={(bar.south east)},y={(bar.north west)}]
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Twosided_ROC_5000.pdf}};
+    \node [anchor=south west] at (0,0) (c) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_TwoSided_ROC_50K.pdf}};
     \begin{scope}[x={(c.south east)},y={(c.north west)}]
@@
     \begin{tikzpicture}
-    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_Twosided_FDR_5000.pdf}};
+    \node [anchor=south west] at (0,0) (d) {\includegraphics[width=0.5\columnwidth]{Figures/Sim_TwoSided_FDR_50K.pdf}};
     \begin{scope}[x={(d.south east)},y={(d.north west)}]

### 2013-05-17 MIMOSA.tex
Update S4

@@
 Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
-Results for MIMOSA fit to a model violating model assumptions, as well as other values of N and I (number of subjects) are in Supplementary Figure 3.
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Supplementary Figure 3, and for varying values of I (number of observations) are in Supplementary Figure 4.
 This figure appears in color in the electronic version of this article.}\label{fig:simulations}

### 2013-05-17 SupplementaryMaterialMIMOSA.tex
Update S4

@@
      than a Beta distribution.  A) The average ROC from 10 simulation
-     with N=5,000 events. B) The average observed and nominal FDR from
-     10 simulations with N=5,000 events. C) Average ROC for N=1,000
-     events D) Average observed and nominal FDR for N=1,000 events.
-% MIMOSA fit to two-sided data were simulated from the standard model
-% with N=1,000 events. E) Average ROC curves from 10 simulations and
-% F) average observed vs nominal FDR from 10 simulations. This figure
-% appears in color in the electronic version of this article.
+     with N=50,000 events. B) The average observed and nominal FDR from
+     10 simulations with N=50,000 events. C) Average ROC for N=10,000
+     events D) Average observed and nominal FDR for N=10,000 events.
 }
    \label{webfig:simulations_trunc}
+\end{figure}
+\clearpage
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto,node distance=0cm]
+    \node at (0,0) (A) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/Sim_OneSided_ROC_varyNobs.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+     \node [below=of A] (B) {
+     \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/Sim_OneSided_FDR_varyNobs.pdf}};
+        \begin{scope} [x={(d.south east)},y={(d.north west)}]
+            \node at (0,1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+    \end{tikzpicture}
+    };
+  \end{tikzpicture}
+   \caption{One--sided MIMOSA model fit to simulated data with 50K events and varying values of I (number of observations).
+A) Average ROC curves from 10 simulations for 20, 50 and 100 observations.
+B) Average observed and nominal FDR from 10 simulations for 20, 50, and 100 observations.
+This figure appears in color in the electronic version of this article.
+}
+   \label{webfig:simulations_Nobs}
 \end{figure}

### 2013-05-17 MIMOSA.tex
Updated competing ROC in SI and varying number of observations in SI

@@
 Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
-A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected.
+A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
 Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
@@
 Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
-In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.
+In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.
 
@@
 Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.
+In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively) were collected for the stimulated and unstimulated samples.
+The effect sizes (differences in the proportions of stimulated and unstimulated cells) are on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, with a few samples exhibiting large numbers of positive events.
 
@@
 The expression from the simulated samples  was compared to paired, unstimulated controls.
+In this data set, we have approximately 90 single cells per subject per stimulation condition with 96 gene expression measurements per cell.
 

### 2013-05-17 SupplementaryMaterialMIMOSA.tex
Updated competing ROC in SI and varying number of observations in SI

@@
 \clearpage
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto,node distance=0cm]
+    \node at (0,0) (A) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/competeROC.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+     \node [below=of A] (B) {
+     \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/competeROCfdr.pdf}};
+        \begin{scope} [x={(d.south east)},y={(d.north west)}]
+            \node at (0,1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+    \end{tikzpicture}
+    };
+  \end{tikzpicture}
+   \caption{Effect of deviations of the assumptions of 100\% true responders at the post-vaccine time point on A) ROC and B) FDR curves.
+The true response rate is shown in the gray box above each plot.
+}
+   \label{webfig:simulations_Nobs}
+\end{figure}
+\clearpage
 

### 2013-05-18 MIMOSA.tex
Update revisions

@@
 For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples.
-A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
+A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after stimulation), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
 Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays.
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}, where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder.
+We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder.
 
@@
 
+Our model is fit to dichotomized data  (cells are either positive or negative).
+While this is the standard approach to ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information.
+Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal, though some have addressed both changes in frequency of expression and expression level\cite{McDavid2012}.
+Extensions of the MIMSOA model, incorporating the continuous MFI or gene expression level are warranted in the future.
+
 Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.

### 2013-05-18 MIMOSA.tex
Update revisions

@@
 When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
-A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).
+A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation).
 Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (\textit{e.g.}, animal studies).
@@
 Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
+The comparisons between stimulated and unstimulated samples are typically done within a time point (\textit{i.e.,} within pre--vaccination and within post--vaccination time points), rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw; additionally not all vaccine trials collect baseline (pre--vaccination) samples.
 In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.

### 2013-05-18 SupplementaryMaterialMIMOSA.tex
Update revisions

@@
 
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto,node distance=0cm]
+    \node at (0,0) (A) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/volcanoplots.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+    \end{tikzpicture}
+   \caption{Volcano plots of effect size vs posterior probability of response for each model fit to each cytokine in the ICS data set.
+Green points indicate subjects called responders for that cytokine and stimulation at a 1\% FDR threshold (adjusted across subjects within cytokine subset).
+Red points indicate non-responders.
+}
+   \label{webfig:simulations_Nobs}
+\end{figure}
+\clearpage
+
 \bibliographystyle{biorefs}

### 2013-05-18 MIMOSA.tex
Reordered AUC on Figures for ICS data

@@
 \subsection{ICS}
-In the ICS data we considered observations at the day 0 time point as true negatives, and observations at the day 182 time point as true positives (potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182).
+In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives.
 The MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
 At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
-Our comparisons also show that ranking based on log-fold change alone is not reliable and should not be used.
 In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F).
@@
 These results are consistent for other cytokines and cytokine combinations (see Supplementary Figure 1).
+We also evaluated,through simulation, what would happen to the ROC curves when these assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 6), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.
 

### 2013-05-22 MIMOSA.tex
Update figure 1 to look nicer

@@
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e.
-proteins, genes, cytokines, etc.) per cell~\citep{Bendall:2011wf}.
+DNA content, cell--surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
 Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
@@
 For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
-We simulated data from this constrained model with 200 observations, a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events, with ten independent realizations of data for each $N$.
+We simulated data from this constrained model with 200 observations (subjects), a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events (cells), with ten independent realizations of data for each $N$.
 We fit the one-sided MIMOSA model to this data.
-We evaluated the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
+We evaluated the sensitivity and specificity of the model's ability to correctly identify observations (subjects) from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
 We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C).
@@
 
-To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
+To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the event (cell) proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
 The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 3 panels C and D).

### 2013-05-23 MIMOSA.tex
More revisions to text

@@
 The antigen--specific T-cell subpopulations (i.e.
-the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells.
+the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.1\% for HIV vaccine studies like the one shown here).
 The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
@@
 Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
-The comparisons between stimulated and unstimulated samples are typically done within a time point (\textit{i.e.,} within pre--vaccination and within post--vaccination time points), rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw; additionally not all vaccine trials collect baseline (pre--vaccination) samples.
+Vaccine--specific response is determined when a subject has no response at a pre--vaccine time point and a response at a post--vaccine time point, and furthermore, when placebo recipients exhibit no response at either time point.
+The comparisons between stimulated and unstimulated samples are typically done within a time point rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw, and additionally because baseline (pre--vaccination) samples are not always available.
 In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.
@@
 The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly.
-In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
-In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects.
-This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increases sensitivity and specificity when detecting responders.
-Because our framework is multivariate in nature, multiple cell subsets can be modelled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
+In our model, cell counts are modeled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
+In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet) distributions where the hyper-parameters for each mixture component are shared across subjects.
+This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with the single-cell assays considered here, and increases sensitivity and specificity when detecting responders.
+Because our framework is multivariate in nature, multiple cell subsets can be modeled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
 Our paper is organized as follows; Section \ref{s:data} introduces the data and notations used in the paper.
@@
 Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.
-In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively) were collected for the stimulated and unstimulated samples.
+In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively).
 The effect sizes (differences in the proportions of stimulated and unstimulated cells) are on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, with a few samples exhibiting large numbers of positive events.

### 2013-05-23 MIMOSA.tex
More revisions

@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder.
+We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder.
 
@@
 In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives.
-The MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
+This assumption was required, since we do not have independent, external measurement of the true response for the day 182 vaccinees.
+Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
 At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
 In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F).
-Here we present the results based on IL2 and IFN-$\gamma$ alone and the subset IL2 and/or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
-These results are consistent for other cytokines and cytokine combinations (see Supplementary Figure 1).
-We also evaluated,through simulation, what would happen to the ROC curves when these assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 6), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.
+Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
+These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (see Supplementary Figure 1).
+We also evaluated,through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.
 
@@
 \hat{\beta}^{(u)}$ fitted to ENV-1-PTEG stimulated
-CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 2).
+CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 3).
+The effect of information sharing on the sensitivity and specificity of the model was also investigated by varying the number of subjects in simulations (Supplementary Figure 4).
+Our model performed well when as few as 20 subjects were included, although the observed false discovery rate was below the nominal rate with so few subjects.
 

### 2013-05-23 MIMOSA.tex
Update fluidigm

@@
 A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
-For an FDR of 10\%, Fisher's exact test identified 47 significant genes, while MIMOSA identified 50 significant genes.
-Both methods identified 39 genes in common.
+For an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes.
+The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA.
 

### 2013-05-24 MIMOSA.tex
update paper

@@
 In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation.
-The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A,B).
-These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
-A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns.
-For an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes.
+The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A).
+These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and closely match the empirical estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
+As a comparison, we also fit the model across all stimulations (Figure~\ref{fig:fluidigm} C).
+Although this leads to a sensible clustering by stimulation, the observed patterns do not agree as well with the empirical differences.
+A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation).
+At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes.
 The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA.
@@
 For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
-We simulated data from this constrained model with 200 observations (subjects), a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events (cells), with ten independent realizations of data for each $N$.
+We simulated data from this constrained model with 200 subjects, a response rate of 50\%, $N=$ 10,000, 25,000, and 50,000 cells, with ten independent realizations of data for each $N$.
 We fit the one-sided MIMOSA model to this data.
-We evaluated the sensitivity and specificity of the model's ability to correctly identify observations (subjects) from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
-We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C).
+We evaluated the sensitivity and specificity of the model's ability to correctly identify  subjects from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
+We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A, C).
 In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).
 
-For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even at small values of $N$ (Figure~\ref{fig:simulations} A and C).
+For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even fewer subjects and smaller cell counts (Supplementary Figures 4 A,B and 5 A,B).
 Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D).
@@
 To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the event (cell) proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 3 panels C and D).
+The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 5 panels C and D).
 Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
@@
 As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
-In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B).
-Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN-$\gamma+$) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
-That is, the number of BIRC3-/CCL5+ cells decreases upon stimulation, while the number of BIRC3+/CCL5+ cells increases.
-When marginalizing over these cell populations, no difference is apparent in any of the samples.
-In contrast, the multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels).
+In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B).
+The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing IL2$+$ or IFN-$\gamma+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, black labels).
+In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B).
 Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.
@@
 The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders).
-Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 3).
+Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5).
 The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
 This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.
+In some cases, the FDR is not well controlled by any method tested here (Supplementary Figure 1), volcano plots of effect size vs posterior probability of response for different cytokine subsets demonstrate these are subjects at a pre--vaccine time--point with a response at the assay--level (Supplementary Figure 6).
+Such biological variation is occasionally expected.
 
@@
 
-Appendix A, B, and C and Supplementary Figures 1--3 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
+Appendix A, B, and C and Supplementary Figures 1--6 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
 
@@
 D) Clustering of the signed q-values from Fisher's exact test.
-Genes selected from Fisher's exact test at the 10\% FDR level.
+All genes  were significant at, or below, the 10\% FDR level.
 This figure appears in color in the electronic version of this article.}
@@
 Curves are shown for MIMOSA, Fisher's exact test, the likelihood ratio test, and log fold-change.
-Results for MIMOSA fit to a model violating model assumptions, as well as other values of N are in Supplementary Figure 3, and for varying values of I (number of observations) are in Supplementary Figure 4.
+Results for MIMOSA fit to a model violating model assumptions, as well as other values of N (number of cells) and values of I (number of subjects) are in Supplementary Figures 3 and 4.
 This figure appears in color in the electronic version of this article.}\label{fig:simulations}

### 2013-05-24 SupplementaryMaterialMIMOSA.tex
More revisions
update paper

@@
 
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto,node distance=0cm]
+    \node at (0,0) (A) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/competeROC.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+     \node [below=of A] (B) {
+     \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/competeROCfdr.pdf}};
+        \begin{scope} [x={(d.south east)},y={(d.north west)}]
+            \node at (0,1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+    \end{tikzpicture}
+    };
+  \end{tikzpicture}
+   \caption{Effect of deviations of the assumptions of 100\% true responders at the post-vaccine time point on A) ROC and B) FDR curves.
+The true response rate is shown in the gray box above each plot.
+}
+   \label{webfig:simulations_competeroc}
+\end{figure}
 %\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
 \begin{tikzpicture} [auto,node distance=0cm]
+    \node at (0,0) (A) {
+    \begin{tikzpicture}
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/Sim_OneSided_ROC_varyNobs.pdf}};
+         \begin{scope} [x={(c.south east)},y={(c.north west)}]
+                 \node at (0,1) [font=\tiny\sffamily] {A} ;
+	\end{scope}
+     \end{tikzpicture}
+     };
+     \node [below=of A] (B) {
+     \begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/Sim_OneSided_FDR_varyNobs.pdf}};
+        \begin{scope} [x={(d.south east)},y={(d.north west)}]
+            \node at (0,1) [font=\tiny\sffamily] {B} ;
+\end{scope}
+    \end{tikzpicture}
+    };
+  \end{tikzpicture}
+   \caption{One--sided MIMOSA model fit to simulated data with 50K cells and varying values of I (number of observations).
+A) Average ROC curves from 10 simulations for 20, 50 and 100 observations.
+B) Average observed and nominal FDR from 10 simulations for 20, 50, and 100 observations.
+This figure appears in color in the electronic version of this article.
+}
+   \label{webfig:simulations_Nobs}
+\end{figure}
+\clearpage
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
+\begin{tikzpicture} [auto,node distance=0cm]
 %\node at (0,0) (A) {
@@
     \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_50K.pdf}};
+     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_10K.pdf}};
          \begin{scope} [x={(c.south east)},y={(c.north west)}]
@@
      \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_50K.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_10K.pdf}};
         \begin{scope} [x={(d.south east)},y={(d.north west)}]
@@
 \begin{tikzpicture}
-    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_10K.pdf}};
+    \node[anchor=south west,inner sep=0] at (0,0) (e) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_ROC_50K.pdf}};
     \begin{scope} [x={(e.south east)},y={(e.north west)}]
@@
     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_10K.pdf}};
+    \node[anchor=south west, inner sep=0] at (0,0) (f) {\includegraphics[width=0.4\columnwidth]{Figures/Sim_Truncated_FDR_50K.pdf}};
         \begin{scope} [x={(f.south east)},y={(f.north west)}]
@@
  \end{tikzpicture}
-   \caption{Unconstrained MIMOSA model fit to data from a model
-     violating model assumptions and to two-sided data with small
-     counts. Data was simulated from a model where proportions were
+   \caption{Unconstrained MIMOSA model fit to two-sided data with small
+     counts  and to  data from a model
+     violating model assumptions.  For two-sided data A) the average ROC from 10 simulation
+     with N=10,000 cells. B) the average observed and nominal FDR from
+     10 simulations with N=10,000 cells. Data was simulated from a model where proportions were
      sampled from a truncated normal distribution over $[0,1]$ rather
-     than a Beta distribution.  A) The average ROC from 10 simulation
-     with N=50,000 events. B) The average observed and nominal FDR from
-     10 simulations with N=50,000 events. C) Average ROC for N=10,000
-     events D) Average observed and nominal FDR for N=10,000 events.
+     than a Beta distribution. C) Average ROC for N=50,000
+     cells D) Average observed and nominal FDR for N=50,000 cells.
 }
    \label{webfig:simulations_trunc}
-\end{figure}
-\clearpage
-
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture} [auto,node distance=0cm]
-    \node at (0,0) (A) {
-    \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/Sim_OneSided_ROC_varyNobs.pdf}};
-         \begin{scope} [x={(c.south east)},y={(c.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {A} ;
-	\end{scope}
-     \end{tikzpicture}
-     };
-     \node [below=of A] (B) {
-     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/Sim_OneSided_FDR_varyNobs.pdf}};
-        \begin{scope} [x={(d.south east)},y={(d.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {B} ;
-\end{scope}
-    \end{tikzpicture}
-    };
-  \end{tikzpicture}
-   \caption{One--sided MIMOSA model fit to simulated data with 50K events and varying values of I (number of observations).
-A) Average ROC curves from 10 simulations for 20, 50 and 100 observations.
-B) Average observed and nominal FDR from 10 simulations for 20, 50, and 100 observations.
-This figure appears in color in the electronic version of this article.
-}
-   \label{webfig:simulations_Nobs}
-\end{figure}
-\clearpage
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-\begin{tikzpicture} [auto,node distance=0cm]
-    \node at (0,0) (A) {
-    \begin{tikzpicture}
-     \node[anchor=south west,inner sep=0] at (0,0) (c) {\includegraphics[width=0.75	\columnwidth]{Figures/competeROC.pdf}};
-         \begin{scope} [x={(c.south east)},y={(c.north west)}]
-                 \node at (0,1) [font=\tiny\sffamily] {A} ;
-	\end{scope}
-     \end{tikzpicture}
-     };
-     \node [below=of A] (B) {
-     \begin{tikzpicture}
-    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=0.75\columnwidth]{Figures/competeROCfdr.pdf}};
-        \begin{scope} [x={(d.south east)},y={(d.north west)}]
-            \node at (0,1) [font=\tiny\sffamily] {B} ;
-\end{scope}
-    \end{tikzpicture}
-    };
-  \end{tikzpicture}
-   \caption{Effect of deviations of the assumptions of 100\% true responders at the post-vaccine time point on A) ROC and B) FDR curves.
-The true response rate is shown in the gray box above each plot.
-}
-   \label{webfig:simulations_Nobs}
 \end{figure}

### 2013-05-29 MIMOSA.tex
Fully revised and shortened. Ready to submit

@@
 We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
-We compare our method against frequentist approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
+We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
 Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
@@
 More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e.
-DNA content, cell--surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
+DNA content, cell-surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
 Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
@@
 The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against cell--surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
@@
 By univariate, we mean that we have only one positive cell subset.
-This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$).
+This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ or B$+$ (\textit{vs.} A$-$/B$-$).
 Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$).
@@
 It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p^{(u)}_{i}$ and $p^{(s)}_{i}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
-Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
+Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using an Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
 As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
@@
 \end{equation}
-Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i, i=1,\dots,I$, are known, we define the complete data log-likelihood:
+Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i,i=1,\dots,I$, are known, we define the complete data log-likelihood:
 \begin{equation}
@@
 \noindent\textbf{MCMC algorithm}\\
-We generated realizations from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}.
+We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
 Details about the algorithms are given in Appendix B of the Supplementary Material. We used the
-method of \cite{Raftery:1992vp} and \cite{Raftery:1996ws} to determine
+method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
 the number of iterations, based on a short pilot run of the
@@
 In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder.
+We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder.
+For MIMOSA, the FDR was computed using the approach of~\citet{Newton2004} whereas for all other methods FDR control was done using the approach of~\citet{Benjamini1995}.
 
@@
 These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (see Supplementary Figure 1).
-We also evaluated,through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.
+We also evaluated, through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.
+Nonetheless, the ranking of the methods is consistent and the conclusions from the ROC analysis still hold.
 
@@
 The effect of information sharing on the sensitivity and specificity of the model was also investigated by varying the number of subjects in simulations (Supplementary Figure 4).
-Our model performed well when as few as 20 subjects were included, although the observed false discovery rate was below the nominal rate with so few subjects.
+Our model performed well when as few as 20 subjects were included, although the nominal false discovery rate overestimated observed false discovery rate.
 
@@
 Although this leads to a sensible clustering by stimulation, the observed patterns do not agree as well with the empirical differences.
-A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust p-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation).
+A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation.
 At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes.
 The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA.
+These results suggest that fitting a stimulation--specific model is the right approach since information is shared across subjects that are expected to exhibit similar behavior.
 
@@
 As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
-In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B).
+In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B).
 The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing IL2$+$ or IFN-$\gamma+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
-The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, black labels).
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
+The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels).
 In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B).
@@
 Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once.
-Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Figure~\ref{fig:mvsimulations} A-C).
-These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Figure~\ref{fig:mvsimulations} B).
+Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Supplementary Figure 6 A,B).
+These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Supplementary Figure 6 B).
 
@@
 This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.
-In some cases, the FDR is not well controlled by any method tested here (Supplementary Figure 1), volcano plots of effect size vs posterior probability of response for different cytokine subsets demonstrate these are subjects at a pre--vaccine time--point with a response at the assay--level (Supplementary Figure 6).
-Such biological variation is occasionally expected.
+In some cases, the FDR is not well controlled by any method tested here and the observed false discovery rate is greater than the nominal false discovery rate (Supplementary Figure 1).
+Volcano plots of effect size vs posterior probability of response for different cytokine subsets show these false positives are pre--vaccine subjects exhibiting response response to stimulation (Supplementary Figure 7).
+Such variability is occasionally expected and may be due to biological variation or technical assay problems.
 
@@
 While this is the standard approach to ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information.
-Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal, though some have addressed both changes in frequency of expression and expression level\cite{McDavid2012}.
+Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal, though some have addressed both changes in frequency of expression and expression level~\citep{McDavid2012}.
 Extensions of the MIMSOA model, incorporating the continuous MFI or gene expression level are warranted in the future.
@@
 In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.}, the gene is not expressed in the cell), or the information is used for pre-filtering~\citep{Flatz:2011jb}.
-The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C).
+The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm}).
 
 Detecting differences in poly-functional cell populations (\textit{i.e.}, identifying changes in cell populations that co-express multiple proteins, cytokines, or genes) is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\citep{Milush:2009bz}.
-In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies (e.g.
-in Leishmania) poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
+In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
 In the ICS data used here, the stimulation is expected to increase only the number of antigen specific cells detected.
@@
 In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential biomarker expression.
-Here, we have shown that MIMOSA has higher sensitivity and specificity than the competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single-cell assay count data obtained from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).
+Here, we have shown that MIMOSA has higher sensitivity and specificity than the competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Supplementary Figure 6 A,B), and the model generally provides a better fit to the single-cell assay count data obtained from studies with these types of experimental designs (Supplementary Figure 6 B).
 Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
@@
 As an example, we would need both, a large number of subjects and a large number of events (cells) collected, to properly estimate the 33 parameters for $K=4$.
-
 A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.
@@
 
-Appendix A, B, and C and Supplementary Figures 1--6 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
+Appendix A, B, and C and Supplementary Figures 1--7 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
 
@@
 Pendleton Charitable Trust for their generous equipment donation.
-
+\vspace{-10pt}
 \bibliographystyle{biorefs}
@@
 \bibliography{MIMOSA}
-\clearpage
 
-\begin{table}
+\begin{table}[h]
 \centering
@@
 
-\begin{figure} %  figure placement: here, top, bottom, or page
+\begin{figure}[h] %  figure placement: here, top, bottom, or page
    \centering
@@
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2.
-E-F) ENV-1-PTEG and expressing IFN-$\gamma$ and/or IL2.
+E-F) ENV-1-PTEG and expressing IFN-$\gamma$ or IL2.
 ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
@@
 
-\begin{figure} %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics{TIKZFig5.eps}
-%\begin{tikzpicture} [auto, node distance=0cm]
-%\node at (0,0) (A) {
-%\begin{tikzpicture}
-%    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
-%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
-%    \node at (0,1) [font=\small\sffamily] {A} ;
-%\end{scope}
-%\end{tikzpicture}
-%};
-%\node [right=of A] (B) {
-%\begin{tikzpicture}
-%   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
-%   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-%       \node at (0,1) [font=\small\sffamily] {B} ;
-%\end{scope}
-%\end{tikzpicture}
-%};
-%\end{tikzpicture}
-   \caption{Multivariate simulations from a two-sided model.
-Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
-Multivariate MIMOSA was compared against Fisher's exact test, and the likelihood ratio test.
-A) Average ROC curves for the competing methods over 10 simulations.
-B) Average observed and nominal false discovery rate for each method over 10 simulations.
-This figure appears in color in the electronic version of this article.}
-   \label{fig:mvsimulations}
-\end{figure}
-
 \end{document}

### 2013-05-29 SupplementaryMaterialMIMOSA.tex
Revised figures based on Raphael's comments
Fully revised and shortened. Ready to submit

@@
  };
+ \node [below=of C] (E) {
+\begin{tikzpicture}
+    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/revised_8.pdf}};
+        \begin{scope}[x={(d.south east)},y={(d.north west)}]
+        \node at (0,1) [font=\tiny\sffamily] {E} ;
+                \end{scope}
+        \end{tikzpicture}
+ };
    % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
@@
     \end{tikzpicture}
-   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.
-This figure appears in color in the electronic version of this article.}
+   \caption{Comparison of MIMOSA on other cytokines and cytokine combinations for ENV-1-PTEG stimulated CD4+ T-cells from the HVTN065 trial.}
 \label{webfig:HVTN065Results}
@@
 B) Average observed and nominal FDR from 10 simulations for 20, 50, and 100 observations.
-This figure appears in color in the electronic version of this article.
 }
@@
    \centering
+   \includegraphics{TIKZFig5.eps}
+%\begin{tikzpicture} [auto, node distance=0cm]
+%\node at (0,0) (A) {
+%\begin{tikzpicture}
+%    \node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_ROC_rev.pdf}};
+%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
+%    \node at (0,1) [font=\small\sffamily] {A} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\node [right=of A] (B) {
+%\begin{tikzpicture}
+%   \node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.47\columnwidth]{Figures/Multivariate_Sims_FDR_rev.pdf}};
+%   \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
+%       \node at (0,1) [font=\small\sffamily] {B} ;
+%\end{scope}
+%\end{tikzpicture}
+%};
+%\end{tikzpicture}
+   \caption{Multivariate simulations from a two-sided model.
+Ten, eight-dimensional data sets were simulated from a two-sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the eight dimensions (N=1,500).
+Multivariate MIMOSA was compared against Fisher's exact test, and the likelihood ratio test.
+A) Average ROC curves for the competing methods over 10 simulations.
+B) Average observed and nominal false discovery rate for each method over 10 simulations.
+This figure appears in color in the electronic version of this article.}
+   \label{fig:mvsimulations}
+\end{figure}
+
+\begin{figure} %  figure placement: here, top, bottom, or page
+   \centering
 \begin{tikzpicture} [auto,node distance=0cm]

### 2013-06-03 MIMOSA.tex
Cut

@@
 
-\title{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
+\title{Mixture Models for Single-Cell Assays with Applications to Vaccine Studies}
 %\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
@@
 We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
-Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods.
+Using several experimental assays measuring proteins or genes at the single-cell level and simulations, we show that our method has higher sensitivity and specificity than alternative methods.
 Additional simulations show that our framework is also robust to model misspecification.
-Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
+Finally, we demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
 Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
@@
 \label{s:intro}
-Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
-This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
-For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
-In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
-Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
-More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e.
+Cell populations, particularly in the immune system, are never truly homogeneous.
+Individual cells may be in different biochemical states that define functional but measurable differences between them.
+This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures (\textit{e.g.,} gene expression of bulk samples).
+For this reason, single-cell assays (\textit{e.g.,} flow, mass cytometry, single-cell gene expression) are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
+These assays typically measure multiple (from tens to 100's) of biomarkers (\textit{e.g.,} DNA content, RNA or protein expression levels, or protein phosphorylation) simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of hematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+\citet{McDavid2012} review the special considerations for analyzing single cell assay data.
+In general, special treatment of single-cell level data is warranted in order to preserve information about cell population heterogeneity.
+%
+%Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
+%This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
+%For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
+%In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
+%Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
+%More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e.
 DNA content, cell-surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
-Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
-A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
-While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
-This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
-Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
-For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
-These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+%Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
+%A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
+%While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
+%This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
+%Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
+%For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
+%These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 
-During analysis, cell level biomarker fluorescence intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
+During data analysis, cell-level biomarker measurements are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
 For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology.
-This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
+This is the case for the Fluidigm qPCR-based single-cell gene expression technology, where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level~\citep{McDavid2012}.
  After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers.
@@
 A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation).
-Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (\textit{e.g.}, animal studies).
 
-A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine.
+%Another motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine.
 Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
- While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response.
- After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}.
-The antigen--specific T-cell subpopulations (i.e.
-the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.1\% for HIV vaccine studies like the one shown here).
-The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine.
+A motivating example from vaccine research is the need to assess the immunogenicity of a vaccine.
+Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
+ While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
+ After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.
+Measuring these antigen--specific T-cell subpopulations is a challenge because they
+ constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.01\%--0.1\% for HIV vaccine studies like the one considered here).
+
+ The intracellular cytokine staining (ICS) assay can measure the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
 Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
-A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations can be detected.
+A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations and small biological effect sizes can be reliably detected.
 Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
@@
 Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
-Vaccine--specific response is determined when a subject has no response at a pre--vaccine time point and a response at a post--vaccine time point, and furthermore, when placebo recipients exhibit no response at either time point.
-The comparisons between stimulated and unstimulated samples are typically done within a time point rather than directly between pre--and post--vaccine time points to avoid confounding vaccine effect with date or time of blood draw, and additionally because baseline (pre--vaccination) samples are not always available.
-In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (on the order of 10s of cells, with effect sizes on the order of $10^{-4}$, relative to the total number of cells), and real biological differences might be difficult to detect.
+Vaccine--specific response is determined when a subject has no response at a pre--vaccine time point and a response at a post--vaccine time point. 
+Specificity is further assessed by measuring response in placebo recipients.
+Importantly, to avoid confounding vaccine effect with time of blood draw and because baseline samples are not always available, comparisons between stimulated and unstimulated samples are typically done within a time point rather than directly between pre--and post--vaccine time points.
+%In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.
 
@@
 
-The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly.
-In our model, cell counts are modeled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood.
+The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues.
+In our model, cell counts are modeled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects through a prior distribution on the unknown proportion(s) of the binomial (or multinomial) likelihood.
 In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet) distributions where the hyper-parameters for each mixture component are shared across subjects.
 This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with the single-cell assays considered here, and increases sensitivity and specificity when detecting responders.
+Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units.
 Because our framework is multivariate in nature, multiple cell subsets can be modeled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.
@@
 Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model.
-In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present some results using a single-cell gene expression data.
+In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present results using single-cell gene expression data.
 Finally, in Section \ref{s:discussion} we discuss our findings and future work.
@@
 Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
-Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine.
+Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured for each subject and time point, each with and without antigen stimulation (here we focus on the HIV Envelope peptide pool).
 For ease of presentation we restricted ourselves to the CD4+ T-cell subsets.
-Samples on day 0 were taken just before vaccination and no response is expected there.
-The corresponding samples can be used as negative controls.
-Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.
+Samples on day 0 were taken before vaccination and no response is expected there.
+Conversely, day 182 (26 weeks) is close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.
 In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively).
-The effect sizes (differences in the proportions of stimulated and unstimulated cells) are on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, with a few samples exhibiting large numbers of positive events.
+The effect sizes (differences in the proportions of stimulated and unstimulated cells) were on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, depending on the cytokine, with a few samples exhibiting larger numbers of positive events.
 
@@
 T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
-The expression from the simulated samples  was compared to paired, unstimulated controls.
+The expression from stimulated samples  was compared to paired, unstimulated controls.
 In this data set, we have approximately 90 single cells per subject per stimulation condition with 96 gene expression measurements per cell.
@@
 
-For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample.
+For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the un-stimulated sample.
 Subjects that show differential expression will be called responders for that biomarker.
-In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described below.
+In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model.
 
@@
 \label{s:DE}
-For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows:
-
-\begin{equation*}
-(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}
-\end{equation*}
-where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively.
-In order to detect responding subjects, we consider two competing models:
-\begin{equation*}
-{\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}
-\end{equation*}
+For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively.
+In order to detect responding subjects, we consider two competing models,
+${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
 Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ).
@@
 
-In our multivariate model, the beta distribution is replaced by a multinomial distribution, as follows:
+In our multivariate model, the beta distribution is replaced by a multinomial distribution.
 \begin{equation*}
@@
 \subsection{Prior}
-As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, such that
-\begin{align*}
-(\mathbf{p}^{(u)}_{i}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\\\nonumber
- (\mathbf{p}^{(s)}_{i}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}^{(s)}) \quad \text{and}\quad (\mathbf{p}^{(u)}_{i}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\end{align*}
-where the indicator variable $z_i$ is defined in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
+As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. The beta priors in  \ref{ss:priors} are replaced by Dirichlet distributions.
+%\begin{align*}
+%(\mathbf{p}^{(u)}_{i}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\\\nonumber
+ %(\mathbf{p}^{(s)}_{i}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}^{(s)}) \quad \text{and}\quad (\mathbf{p}^{(u)}_{i}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\end{align*}
+The indicator variable $z_i$ is defined as in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
 As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation.
@@
 Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Appendix C of the Supplementary Material). 
-For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
-\begin{align*}
-\mathrm{L}_0(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \cdot \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!},
-\end{align*}
-where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$.
+%For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
+%\begin{align*}
+%\mathrm{L}_0(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \cdot \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!},
+%\end{align*}
+%where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$.
 Similarly the marginal likelihood for the alternative model is given by
-\[
-\mathrm{L}_1(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}_{i}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) =\frac{\mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}) \mathrm{B}(\boldsymbol{\alpha}^{(s)}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(s)})\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!}.
-\]
+%\[
+%\mathrm{L}_1(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}_{i}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) =\frac{\mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}) \mathrm{B}(\boldsymbol{\alpha}^{(s)}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(s)})\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!}.
+%\]
 The estimation procedures (both EM and MCMC based) for the Dirichlet--multinomial distribution are the same as for the beta-binomial model except that the number of parameters to estimate is larger.
@@
 \label{s:discussion}
-Experimentalists already have access to a myriad of single-cell assays such as flow cytometry, mass cytometry and multiplexed quantitative-PCR.
-The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays will become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}.
-Current approaches for single-cell assays such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
-Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy,Gottardo:2006}.
+The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}.
+Current data analysis approaches such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
+Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton2004,Smyth:2005iy,Gottardo:2006}.
 In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.
@@
 The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders).
-Information is shared across responders and non-responders through exchangeable beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5).
+Information is shared across responders and non-responders through exchangeable priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5).
 The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
-This has proven to be useful for the ICS data where stimulation induced changes are expected to be one-sided.
-In some cases, the FDR is not well controlled by any method tested here and the observed false discovery rate is greater than the nominal false discovery rate (Supplementary Figure 1).
-Volcano plots of effect size vs posterior probability of response for different cytokine subsets show these false positives are pre--vaccine subjects exhibiting response response to stimulation (Supplementary Figure 7).
-Such variability is occasionally expected and may be due to biological variation or technical assay problems.
+This has proven to be useful for ICS data where stimulation induced changes are expected to be one-sided.
+Although in some cases the FDR is not well controlled by any method tested (Supplementary Figure 1A), volcano plots of effect size vs posterior probability of response show these false positives are pre--vaccine subjects exhibiting response  to stimulation (Supplementary Figure 7).
+Such variability is occasionally expected and may be biological or a technical assay problem.
 
@@
 While this is the standard approach to ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information.
-Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal, though some have addressed both changes in frequency of expression and expression level~\citep{McDavid2012}.
+Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}.
+We have shown elsewhere that changes in both the frequency of expression and expression level are informative~{McDavid2012}.
+The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but importantly, further demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm}).
+Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
 Extensions of the MIMSOA model, incorporating the continuous MFI or gene expression level are warranted in the future.
 
-Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
-In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.}, the gene is not expressed in the cell), or the information is used for pre-filtering~\citep{Flatz:2011jb}.
-The ability of MIMOSA to identify stimulation-specific expression patterns in single-cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm}).
+%In the case of the Fluidigm data, most analysis methods have been focused on identifying differences in the continuous part of the signal ignoring cells that are undetected (\textit{i.e.}, the gene is not expressed in the cell), or the information is used for pre-filtering~\citep{Flatz:2011jb}.
 
-Detecting differences in poly-functional cell populations (\textit{i.e.}, identifying changes in cell populations that co-express multiple proteins, cytokines, or genes) is important in immunology, since it allows the identification of more precisely defined, more homogeneous cell populations~\citep{Milush:2009bz}.
+Detecting differences in poly-functional cell populations (\textit{i.e.}, identifying changes in cell populations that co-express multiple proteins, cytokines, or genes) is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
 In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
-In the ICS data used here, the stimulation is expected to increase only the number of antigen specific cells detected.
+In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected.
 Hence, if a specific cell subset expressing multiple biomarkers is being differentially expressed, differential expression based on the marginal cell counts should also be detected.
@@
 For two--sided changes, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
-In this case, the use of multivariate model, as our Dirichlet-multinomial model, will become important to detect differential biomarker expression.
+In this case, the use of multivariate models, as our Dirichlet-multinomial model, will become important to detect differences.
 Here, we have shown that MIMOSA has higher sensitivity and specificity than the competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Supplementary Figure 6 A,B), and the model generally provides a better fit to the single-cell assay count data obtained from studies with these types of experimental designs (Supplementary Figure 6 B).
 Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
-In the case of more than two biomarkers, the number of parameters to estimate for our Dirichlet--multinomial model is $2^{K+1}+1$, which is large even for moderate values of K.
+In the case of more than two biomarkers, the number of parameters for our model is $2^{K+1}+1$, which is large even for moderate values of K.
 As an example, we would need both, a large number of subjects and a large number of events (cells) collected, to properly estimate the 33 parameters for $K=4$.
@@
 
-All of the results presented here were obtained with a software implementation of the EM and MCMC MIMOSA models in R and C++, and is freely available from GitHub \\
-(\url{http://www.github.org/RGLab/MIMOSA}). An R package will soon be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
+All of the results presented here were obtained with an R and C++ implementation the MIMOSA models, freely available at (\url{http://www.github.org/RGLab/MIMOSA}).
+An R package will be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
 
@@
 
-Appendix A, B, and C and Supplementary Figures 1--7 referenced in Sections \ref{s:data} and \ref{s:DEone},\ref{s:results} and \ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
+Appendix A, B, and C and Supplementary Figures 1--7 referenced in Sections \ref{s:data}--\ref{s:demarkercombos} are available in the Supplementary Material.\vspace*{-8pt}
 
 \section*{Acknowledgments}
-This work was supported by the Intramural Research Program of the  National Institute of Allergy and Infectious Diseases (NIAID) and the National Institutes of Health (NIH) [R01 EB008400, U01 AI068635-01 to RG], grants from the Bill and Melinda Gates Foundation VISC (Vaccine Immunology Statistical Center) [grant numbers OPP38744,  OPP1032317], a grant from the Bill \& Melinda Gates Foundation to the CAVD (Collaboration for AIDS Vaccine Discovery) [grant number OPP1032325].
-Study HVTN065 was conducted by the HIV Vaccine Trials Network (HVTN), and supported by the National Institute of Allergy and Infectious Diseases (NIAID).
-The the HVTN Laboratory Program is supported by the NIH [grant  number UM1AI068618].
-Funding was also provided by the Public Health Service from the NIH and the University of Washington Center for AIDS Research (CfAR), an NIH-funded program [grant numbers UM1 AI068618, P30 AI027757].
-We thank the James B.
-Pendleton Charitable Trust for their generous equipment donation.
+This work was supported by the Intramural Research Program of NIAID and NIH, grants [R01 EB008400, U01 AI068635-01], the Bill and Melinda Gates Foundation through grants to VISC [OPP38744,  OPP1032317] and CAVD [OPP1032325].
+HVTN065 was conducted by the HVTN and supported by NIAID.
+The the HVTN Lab Program is supported by the NIH [UM1AI068618], the Public Health Service, and the University of Washington Center for AIDS Research [UM1 AI068618, P30 AI027757].
+Thanks to the James B.
+Pendleton Charitable Trust for a generous equipment donation.
 \vspace{-10pt}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 
-\begin{figure}[h] %  figure placement: here, top, bottom, or page
+\begin{figure} %  figure placement: here, top, bottom, or page
    \centering
-   \includegraphics{TIKZFig1.eps}
-%\begin{tikzpicture} [auto, node distance=0cm]
-%\node at (0,0) (A){
-%    \begin{tikzpicture}
-%    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=0.95\columnwidth]{Figures/revised_2.pdf}};
-%    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
-%    \node at (0,1) [font=\small\sffamily] {A} ;
-%    \node at (0.5,1) [font=\small\sffamily] {B} ;
-%    \end{scope}
-%\end{tikzpicture}
-% };
-% \node [below=of A] (B) {
-% \begin{tikzpicture}
-%    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=0.95\columnwidth]{Figures/revised_1.pdf}};
-%    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
-%    \node at (0,1) [font=\small\sffamily] {C} ;
-%    \node at (0.5,1) [font=\small\sffamily] {D} ;
-%    \end{scope}
-%    \end{tikzpicture}
-% };
-%  \node [below=of B] (E) {
-% \begin{tikzpicture}
-%    \node[anchor=south west, inner sep=0] at (0,0) (baz){\includegraphics[width=0.95\columnwidth]{Figures/revised_15.pdf}};
-%    \begin{scope} [x={(baz.south east)},y={(baz.north west)}]
-%    \node at (0,1) [font=\small\sffamily] {E} ;
-%    \node at (0.5,1) [font=\small\sffamily] {F} ;
-%    \end{scope}
-%    \end{tikzpicture}
-% };
-%\end{tikzpicture}
+   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
    \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.

### 2013-06-03 MIMOSA.tex
cut cut cut

@@
  PRATIP CHATTOPADHYAY$^{3}$,
- MARIA DOMINQUEZ$^{3}$,
+ MARIA DOMINGUEZ$^{3}$,
  STEVE DE ROSA$^{1,2}$,
@@
 Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays.
-Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions.
+Our model allows the inference to be subject specific, as is typically required when assessing vaccine responses, while borrowing strength across subjects through common prior distributions.
 We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
@@
 Individual cells may be in different biochemical states that define functional but measurable differences between them.
-This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures (\textit{e.g.,} gene expression of bulk samples).
-For this reason, single-cell assays (\textit{e.g.,} flow, mass cytometry, single-cell gene expression) are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
-These assays typically measure multiple (from tens to 100's) of biomarkers (\textit{e.g.,} DNA content, RNA or protein expression levels, or protein phosphorylation) simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of hematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
+This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures (\textit{e.g.,} bulk gene expression).
+For this reason, single-cell assays (\textit{e.g.,} flow, mass cytometry, single-cell gene expression) are an important tool in immunology, providing a functional snapshot of the immune system at a given time.
+These assays typically measure multiple biomarkers (\textit{e.g.,} DNA content, RNA or protein expression levels, or protein phosphorylation) simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of hematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
 \citet{McDavid2012} review the special considerations for analyzing single cell assay data.
@@
 
-During data analysis, cell-level biomarker measurements are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
+Typically, cell-level measurements are thresholded as positive or negative so that subsets with different multivariate $+/-$ biomarkers can be obtained as Boolean combinations.
 For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology.
-This is the case for the Fluidigm qPCR-based single-cell gene expression technology, where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level~\citep{McDavid2012}.
- After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers.
-Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations.
+This is the case for the Fluidigm's qPCR-based single-cell gene expression technology, where genes are recorded as expressed or not, at the single-cell level~\citep{McDavid2012}.
+ After thresholding, we obtain a Boolean matrix of $N$ cells $\times$ $K$ biomarkers and we can form $2^K$ putative cell subsets.
 When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
-A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation).
+A common statistical problem is to identify subjects for whom the proportion of cells expressing a specific combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation).
 
@@
 Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
- While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
- After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.
-Measuring these antigen--specific T-cell subpopulations is a challenge because they
- constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.01\%--0.1\% for HIV vaccine studies like the one considered here).
+T-cells that recognize the antigen become \emph{activated}, produce cytokines that potentiate the immune response, and proliferate and persist in the immune system providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.
+This vaccine memory can be interrogated through the intracellular cytokine staining (ICS) assay, by measuring cytokine production in antigen-specific T-cells in response to antigen stimulation~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+ %While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
+ %After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.
 
- The intracellular cytokine staining (ICS) assay can measure the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
-Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
-A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations and small biological effect sizes can be reliably detected.
-Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.
+The antigen--specific subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%).  
+% The intracellular cytokine staining (ICS) assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring intracellular cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
+%Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
+Thus, a sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that changes in these rare cell populations can be reliably detected.
+Subsequently, each individual cell is classified as either positive or negative for each cytokine based on predetermined thresholds, and the number of cells matching each subpopulation phenotype is counted.
 
-These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences.
-Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}.
-Vaccine--specific response is determined when a subject has no response at a pre--vaccine time point and a response at a post--vaccine time point. 
-Specificity is further assessed by measuring response in placebo recipients.
-Importantly, to avoid confounding vaccine effect with time of blood draw and because baseline samples are not always available, comparisons between stimulated and unstimulated samples are typically done within a time point rather than directly between pre--and post--vaccine time points.
+Cell counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences.
+\emph{Responders} are subjects whose T-cells exhibit a response to stimulation, and the response is vaccine-specific if it exists at the post-vaccine time point but not the pre-vaccine time point.
+%Response is vaccine--specific when a subject responds at a post--vaccine time point but not at a pre--vaccine time point. 
+%Response in placebo recipients is further used to assess vaccine-specificity. 
+These comparisons are typically performed within a time point  in order to avoid confounding vaccine effects with time of blood draw and because baseline samples are not always available. 
 %In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.
@@
 
-In the remainder of the paper, we use the following notation to describe our model.
-We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated.
+We use the following notation to describe our model.
+We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and unstimulated.
 Each cell can either be positive or negative for a biomarker.
 Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations.
-We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples, respectively.
-We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample, respectively.
-For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$.
-Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
+We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and unstimulated samples, respectively.
+We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample.
+For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$, and finally $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
 
-We consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.
+We consider two types of immunological single-cell assays.
 
 \textit{Flow cytometry}:
-The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
-The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
-Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
-Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured for each subject and time point, each with and without antigen stimulation (here we focus on the HIV Envelope peptide pool).
-For ease of presentation we restricted ourselves to the CD4+ T-cell subsets.
-Samples on day 0 were taken before vaccination and no response is expected there.
-Conversely, day 182 (26 weeks) is close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.
-In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively).
-The effect sizes (differences in the proportions of stimulated and unstimulated cells) were on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, depending on the cytokine, with a few samples exhibiting larger numbers of positive events.
+Our primary dataset is ICS data generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen~\citep{Goepfert:2011ci}.
+The study goal was to assess the immunogenicity of different prime-boost regimens.
+Here, we analyze a subset of the data consisting of 98 vaccine and placebo recipients at days 0 (pre-vaccine) and 182 (two weeks post-vaccine), focusing on CD4+ T-cell response to stimulation with HIV Envelope peptide.
+Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured for each subject and time point, each with and without antigen stimulation.
+No response is expected at day 0, while day 182 is close to the immunogenicity peak and many vaccinees are expected to respond for at least some cytokines.
+A median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively).
+The empirical differences in the proportions of stimulated and unstimulated cells were on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, depending on cytokine subset.
 
-\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects.
-T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
+\textit{Fluidigm single-cell gene expression}: Tetramer-sorted CD8$+$ T-cells from sixteen subjects.
+T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation was measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
 The expression from stimulated samples  was compared to paired, unstimulated controls.
@@
 \label{s:DEone}
-Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations.
+Datasets like those presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many combinations.
 By univariate, we mean that we have only one positive cell subset.
@@
 Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$).
-In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.
+In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the unstimulated and stimulated samples (Table \ref{tab:twobytwo}).
 
@@
 
-For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the un-stimulated sample.
+For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the unstimulated sample.
 Subjects that show differential expression will be called responders for that biomarker.
@@
 \label{s:DE}
-For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively.
+For a given subject $i$, the positive cell counts for the stimulated and unstimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and unstimulated paired samples, respectively.
 In order to detect responding subjects, we consider two competing models,
 ${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
-Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ).
+Under the null model, ${\cal M}_0$, there is no difference between the stimulated and unstimulated samples, and the proportions are equal (yet the cell counts can differ).
 Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder.
@@
 \label{ss:priors}
-Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows:
- \begin{align*}
-(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
-(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
-\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
- \end{align*}
+Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions.
+For non-responders, $(p^{(u)}_{i} | z_{i}=0)  \sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})$, and for responders $(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  \sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)})$, %, as follows:
+% \begin{align*}
+%(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
+%(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
+%\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
+ %\end{align*}
  where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}$ are unknown hyper-parameters shared across all subjects.
-Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are explicitly shared across the two models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
+Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are  shared across both models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
 Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
@@
 Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using an Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
-As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
+We also describe a fully Bayesian model where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
 In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.
@@
 In our proposed EM and MCMC algorithms, we greatly simplify our calculations by directly utilizing the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after marginalizing $p^{(s)}_{i}$ and $p^{(u)}_{i}$ from the null and alternative likelihoods.
-Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Appendix B, Supplementary Material), and are given by,
+Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Appendix B, Supplementary Material).%, and are given by,
 
- \begin{equation*}
-  	\mathrm{L}_0(\alpha^{(u)},\beta^{(u)}|\mathbf{y}_i)
-=\binom{N^{(u)}_{i}}{n^{(u)}_{i}}\binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+n^{(u)}_{i}+\alpha^{(u)},N^{(s)}_{i}-n^{(s)}_{i}+N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}
- \end{equation*}
-
-and
-\begin{equation}
-\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
-=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot\frac{\mathrm{B}(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}
-\label{model2:unconstrained}
-\end{equation}
-Above, $\mathrm{B}$ is the Beta function. Assuming that the missing data, $z_i,i=1,\dots,I$, are known, we define the complete data log-likelihood:
+% \begin{equation*}
+%  	\mathrm{L}_0(\alpha^{(u)},\beta^{(u)}|\mathbf{y}_i)
+%=\binom{N^{(u)}_{i}}{n^{(u)}_{i}}\binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+n^{(u)}_{i}+\alpha^{(u)},N^{(s)}_{i}-n^{(s)}_{i}+N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}
+% \end{equation*}
+%
+%and
+%\begin{equation}
+%\mathrm{L}_1(\alpha^{(u)},\beta^{(u)},\alpha^{(s)},\beta^{(s)}|\mathbf{y}_i)
+%=\binom{N^{(u)}_{i}}{n^{(u)}_{i}} \binom{N^{(s)}_{i}}{n^{(s)}_{i}}\cdot\frac{\mathrm{B}(n^{(u)}_{i}+\alpha^{(u)},N^{(u)}_{i}-n^{(u)}_{i}+\beta^{(u)})}{\mathrm{B}(\alpha^{(u)},\beta^{(u)})}\cdot \frac{\mathrm{B}(n^{(s)}_{i}+\alpha^{(s)},N^{(s)}_{i}-n^{(s)}_{i}+\beta^{(s)})}{\mathrm{B}(\alpha^{(s)},\beta^{(s)})}
+%\label{model2:unconstrained}
+%\end{equation}
+%Above, $\mathrm{B}$ is the Beta function. 
+Assuming that the missing data, $z_i,i=1,\dots,I$, are known, we define the complete data log-likelihood:
 \begin{equation}
@@
 We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters~\citep{Ihaka:1996ud}.
-Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence.
-In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components.
-We then use the estimated $z_i$'s to estimate the $p^{(u)}_{i}$'s and $p^{(s)}_{i}$'s and use these to set the hyper-parameters to their method-of-moments estimates.
+We initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative components and then use the estimated $z_i$'s to estimate the $p^{(u)}_{i}$'s and $p^{(s)}_{i}$'s and use these to set the hyper-parameters to their method-of-moments estimates.
 
@@
 We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}.
-All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs samplings.
-Details about the algorithms are given in Appendix B of the Supplementary Material. We used the
-method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
-the number of iterations, based on a short pilot run of the
-sampler. For each dataset presented here, we calculated that no more
-than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
+All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
+Details about the algorithms and implementation for the data presented here are given in Appendix B of the Supplementary Material. 
+%We used the
+%method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
+%the number of iterations, based on a short pilot run of the
+%sampler. For each dataset presented here, we calculated that no more
+%than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
 To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.
@@
 \label{s:results}
-In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data.
-We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder.
+We evaluated and compared the performance of MIMOSA against Fisher's exact test, a likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by compared the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for the ICS data.
+A false discovery is defined as a day 0 sample or a sample from a placebo recipient (expected non-responders) that is identified as a responder by an algorithm.
 For MIMOSA, the FDR was computed using the approach of~\citet{Newton2004} whereas for all other methods FDR control was done using the approach of~\citet{Benjamini1995}.
@@
 \subsection{ICS}
-In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives.
-This assumption was required, since we do not have independent, external measurement of the true response for the day 182 vaccinees.
+\label{s:ics}
+In the ICS data we considered observations from vaccinees at day 0 or from placebos at day 0 or 182 as \emph{true negatives} in that we do not expect to observe a response in those samples.
+Observations from vaccinees at day 182 were considered \emph{true positives} in that a response is expected assuming a 100\% response rate.
+This assumption was required, since we do not have independent, external measurement of the true response.
 Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
@@
 Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
-These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (see Supplementary Figure 1).
-We also evaluated, through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.
-Nonetheless, the ranking of the methods is consistent and the conclusions from the ROC analysis still hold.
+These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (Supplementary Figure 1).
+We also evaluated, through simulation, what would happen to the ROC curves when our assumption of 100\% response rate was false (\textit{i.e.,} when not all day 182 vaccinees are responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to non-responders at day 182.
+Nonetheless, the ranking of the methods remained consistent and the conclusions from the ROC analysis still hold.
 
@@
 \subsection{Single-cell gene expression}
-We applied the MIMOSA model to a Fluidigm single-cell gene expression data set.
+We applied the MIMOSA model to the Fluidigm single-cell gene expression data set.
 We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation.
-In order to detect stimulation specific changes of expression, we fit our model to each gene within each stimulation.
-The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A).
-These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) and closely match the empirical estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B).
-As a comparison, we also fit the model across all stimulations (Figure~\ref{fig:fluidigm} C).
-Although this leads to a sensible clustering by stimulation, the observed patterns do not agree as well with the empirical differences.
-A similar analysis using a two-sided Fisher's exact test and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation.
+In order to detect stimulation specific changes of expression, we fit our model separately to each gene within each stimulation.
+The results show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A).
+These patterns evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) closely match the empirical estimates of the differences in proportions (Figure~\ref{fig:fluidigm} B).
+As a comparison, we also fit the model combining information across all stimulations (Figure~\ref{fig:fluidigm} C).
+Although this leads to a sensible clustering by stimulation, the observed pattern does not agree as well with the empirical differences.
+A similar analysis applying a two-sided Fisher's exact test to each sample and gene, and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation.
 At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes.
@@
 We examined the performance of the constrained ($p^{(s)}>p^{(u)}$) and unconstrained ($p^{(s)} \ne p^{(u)}$) beta-binomial mixture models via simulations.
-For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point.
-We simulated data from this constrained model with 200 subjects, a response rate of 50\%, $N=$ 10,000, 25,000, and 50,000 cells, with ten independent realizations of data for each $N$.
-We fit the one-sided MIMOSA model to this data.
-We evaluated the sensitivity and specificity of the model's ability to correctly identify  subjects from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change.
-We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A, C).
-In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).
-
-For both the constrained and unconstrained simulations, MIMOSA was superior to competing methods, including Fisher's exact test, with respect to sensitivity and specificity even fewer subjects and smaller cell counts (Supplementary Figures 4 A,B and 5 A,B).
-Additionally, the estimated FDR for MIMOSA more closely reflected the nominal FDR compared to Fisher's exact test and competing methods (Figure~\ref{fig:simulations}, panels B, D).
+For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (univariate IL2+) from the primary immunogenicity time point.
+We simulated data from this constrained model with 200 subjects, a response rate of 50\%, $N=$ 10,000, 25,000, and 50,000 cells, and ten independent realizations of data for each $N$.
+We fit the one-sided MIMOSA model to this data and evaluated the sensitivity and specificity to identify responders and non-responders through ROC analysis.
+MIMOSA was compared against the same methods as in section~\ref{s:ics}.
+We repeated this procedure for the two-sided models fit to two-sided data, and compared the ability of each method to properly control the FDR.
+For both the constrained and unconstrained simulations, MIMOSA exhibited superior sensitivity and specificity than competing methods, even with fewer subjects and smaller cell counts (Figure~\ref{fig:simulations} A,C, and Supplementary Figures 4 A and 5 A).
+MIMOSA also provided better control of the FDR than competing approaches (Figure~\ref{fig:simulations}, panels B, D, Supplementary Figures 4 B and 5 B).
 
@@
 
-To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the event (cell) proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
-The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 5 panels C and D).
-Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
+To assess the sensitivity of the model to deviations from model assumptions, we simulated data with the cell proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions.
+The means and variances of the truncated normal distributions were set to the MLE of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (Supplementary Figure 5 C,D).
+Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed competing methods.
 
@@
 
-\subsection{Model}
+%\subsection{Model}
+In our multivariate model, the beta distribution is replaced by a multinomial distribution.
+Thus the observed data are drawn from $(\mathbf{n}^{(u)}_{i}|\mathbf{p}^{(u)}_{i},) \sim \mathcal{M}(N^{(u)}_{i},\mathbf{p}^{(u)}_{i})$ and $(\mathbf{n}^{(s)}_{i}|\mathbf{p}^{(s)}_{i}) \sim \mathcal{M}(N^{(s)}_{i},\mathbf{p}^{(s)}_{i})$, where $N^{\{s,u\}}_{i}=\sum\limits_{k=1}\limits^{2^K} n^{\{s,u\}}_{ik}$ are the number of cells collected and $\mathbf{p}^{(u)}_{i}$ and $\mathbf{p}^{(s)}_{i}$ are the unknown proportions for the unstimulated and stimulated samples, respectively.
 
-In our multivariate model, the beta distribution is replaced by a multinomial distribution.
-\begin{equation*}
- (\mathbf{n}^{(u)}_{i}|\mathbf{p}^{(u)}_{i},) \sim \mathcal{M}(N^{(u)}_{i},\mathbf{p}^{(u)}_{i})\quad\text{and}\quad (\mathbf{n}^{(s)}_{i}|\mathbf{p}^{(s)}_{i}) \sim \mathcal{M}(N^{(s)}_{i},\mathbf{p}^{(s)}_{i})\label{eq:mult_likeliehood}
- \end{equation*}
+%\quad\text{and}\quad (\mathbf{n}^{(s)}_{i}|\mathbf{p}^{(s)}_{i}) \sim \mathcal{M}(N^{(s)}_{i},\mathbf{p}^{(s)}_{i})\label{eq:mult_likeliehood}
 
-where $N^{\{s,u\}}_{i}=\sum\limits_{k=1}\limits^{2^K} n^{\{s,u\}}_{ik}$ are the number of cells collected and $\mathbf{p}^{(u)}_{i}$ and $\mathbf{p}^{(s)}_{i}$ are the unknown proportions for the un-stimulated and stimulated samples, respectively.
+%\begin{equation*}
+ %(\mathbf{n}^{(u)}_{i}|\mathbf{p}^{(u)}_{i},) \sim \mathcal{M}(N^{(u)}_{i},\mathbf{p}^{(u)}_{i})\quad\text{and}\quad (\mathbf{n}^{(s)}_{i}|\mathbf{p}^{(s)}_{i}) \sim \mathcal{M}(N^{(s)}_{i},\mathbf{p}^{(s)}_{i})\label{eq:mult_likeliehood}
+% \end{equation*}
 
-\subsection{Prior}
+%\subsection{Prior}
 As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. The beta priors in  \ref{ss:priors} are replaced by Dirichlet distributions.
@@
 
-\subsection{Parameter estimation}
+%\subsection{Parameter estimation}
 Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Appendix C of the Supplementary Material). 
@@
 Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large.
-More details about our multivariate MCMC algorithm is given in Appendix C of the Supplementary Material.
+More details are given in Appendix C of the Supplementary Material.
 
@@
 \bibliography{MIMOSA}
-
+\clearpage
 \begin{table}[ht]

### 2013-06-03 SupplementaryMaterialMIMOSA.tex
cut cut cut

@@
 In our case each parameter has the same exponential prior with mean $1,000$
+
+\noindent\textbf{Implementation Details for MCMC Algorithm}\\
+We used the method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
+the number of iterations, based on a short pilot run of the
+sampler. For each dataset presented in the manuscript, we calculated that no more
+than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities.
+To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.
 

### 2013-06-03 MIMOSA.tex
cut cut cut 23 pages

@@
 The indicator variable $z_i$ is defined as in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
-As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation.
+As in the beta-binomial case, both EM and MCMC algorithms can be used for parameter estimation.
 When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}^{\{u,s\}}$ and $w$ as for the beta-binomial model.
@@
 %\subsection{Parameter estimation}
-Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Appendix C of the Supplementary Material). 
+We again make use of the marginal likelihoods in closed form (see Appendix C of the Supplementary Material). 
 %For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 %\]
-The estimation procedures (both EM and MCMC based) for the Dirichlet--multinomial distribution are the same as for the beta-binomial model except that the number of parameters to estimate is larger.
+The estimation procedures (both EM and MCMC) for the Dirichlet--multinomial are the same as for the beta-binomial model except that the number of parameters to estimate is larger.
 We initialize the $z_i$ in the EM algorithm with the positivity calls from the multivariate Fisher's exact test.
@@
 \subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
-As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
-In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B).
+As a proof-of-concept, we applied our multivariate MIMOSA model to two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example $K=2$ and we have four possible subsets.
+In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the $log_2$ of the sum of the counts of the 3 positive subsets.
 The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions.
-The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels).
-In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B).
-Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts are bidirectional and their sum is near zero.
+The multivariate MIMOSA model tests for any difference across all polyfunctional cell subsets and detects significant differences in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels), whereas the standard approach detects no differences (Figure~\ref{fig:polyfunctionality} B).
+Testing all combinations simultaneously is also an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.
 
@@
 Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once.
-Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Supplementary Figure 6 A,B).
-These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Supplementary Figure 6 B).
+Therefore, we performed simulations in eight dimensions to assess the performance of the multivariate MIMOSA  model on the resulting 2x8 tables (Supplementary Figure 6 A,B).
+These results show that multivariate MIMOSA outperformed competing methods even when cell counts were small and at small effect sizes.
+MIMOSA also provided better FDR control than competing approaches (Supplementary Figure 6 B).
 
@@
 The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}.
-Current data analysis approaches such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
+Current approaches, such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
 Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton2004,Smyth:2005iy,Gottardo:2006}.
@@
 
-The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders).
-Information is shared across responders and non-responders through exchangeable priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5).
-The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.
-This has proven to be useful for ICS data where stimulation induced changes are expected to be one-sided.
-Although in some cases the FDR is not well controlled by any method tested (Supplementary Figure 1A), volcano plots of effect size vs posterior probability of response show these false positives are pre--vaccine subjects exhibiting response  to stimulation (Supplementary Figure 7).
-Such variability is occasionally expected and may be biological or a technical assay problem.
+The MIMOSA model uses a mixture framework to model counts in experimental subjects across multiple conditions.
+Information is shared across subjects through exchangeable priors, increasing the power to detect true differences compared to competing methods, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5 B).
+Although in some instances the FDR was not well controlled by any method tested (Supplementary Figure 1A), volcano plots of effect size vs posterior probability of response show these false positives are pre--vaccine subjects exhibiting response  to stimulation (Supplementary Figure 7).
+Such variability is occasionally expected due to assay failures or biological variation.
 
-Our model is fit to dichotomized data  (cells are either positive or negative).
-While this is the standard approach to ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information.
-Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}.
-We have shown elsewhere that changes in both the frequency of expression and expression level are informative~{McDavid2012}.
-The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but importantly, further demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm}).
-Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions.
-Extensions of the MIMSOA model, incorporating the continuous MFI or gene expression level are warranted in the future.
+MIMOSA is fit to dichotomized data  (cells are either positive or negative).
+While this is the standard approach for ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information.
+For single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}.
+The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but demonstrates that biologically relevant signal is present in the frequency of gene expression (Figure~\ref{fig:fluidigm}).
+We have shown elsewhere that changes in both the frequency of expression and expression level are informative~{McDavid2012}, and extensions of the model that incorporate the continuous component (MFI or gene expression) are warranted in the future.
+Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative subsets.
 
@@
 
-Detecting differences in poly-functional cell populations (\textit{i.e.}, identifying changes in cell populations that co-express multiple proteins, cytokines, or genes) is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
+Detecting differences in poly-functional cell populations is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
 In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
-In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected.
-Hence, if a specific cell subset expressing multiple biomarkers is being differentially expressed, differential expression based on the marginal cell counts should also be detected.
-As such, identifying poly-functional cytokine profiles from ICS data can be done in an iterative way.
-First, univariate tests on marginal populations are performed, and then specific cell subsets expressing the positive biomarkers detected are tested.
-However, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that need to be tested, and a multivariate approach might be preferable.
-In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\citep{Nason:2006dx}.
+In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells.
+The univariate MIMOSA model allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the unstimulated sample, and is ideal for this situation.
+Hence, if a specific cell subset expressing multiple biomarkers is differentially expressed, then differential expression based on the marginal cell counts should also be detectable.
+As such, identifying poly-functional cytokine profiles from ICS data could be done in an iterative way.
+First, univariate tests on marginal populations are performed, then specific cell subsets expressing the positive biomarkers detected are tested.
+However, this iterative (univariate) approach might not be satisfactory due to the large number of combinations that need to be tested. A multivariate approach might be preferable.
+In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should only take into account the cytokine combinations of interest~\citep{Nason:2006dx}.
 
-For two--sided changes, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
-In this case, the use of multivariate models, as our Dirichlet-multinomial model, will become important to detect differences.
-Here, we have shown that MIMOSA has higher sensitivity and specificity than the competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Supplementary Figure 6 A,B), and the model generally provides a better fit to the single-cell assay count data obtained from studies with these types of experimental designs (Supplementary Figure 6 B).
+When changes are two--sided (as with Fluidigm), differences in poly-functional cell populations are not always detectable  since differences may cancel in the marginal populations (Figure~\ref{fig:polyfunctionality} B).
+In this case, the use of multivariate models, as our Dirichlet-multinomial model, will be important (Figure~\ref{fig:polyfunctionality} A).
 Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
 In the case of more than two biomarkers, the number of parameters for our model is $2^{K+1}+1$, which is large even for moderate values of K.
-As an example, we would need both, a large number of subjects and a large number of events (cells) collected, to properly estimate the 33 parameters for $K=4$.
+Even for  $K=4$, we would need a large number of subjects and cells to properly estimate the 33 parameters.
 A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.

### 2013-06-04 MIMOSA.tex
Cut some text.

@@
 
- \author{GREG FINAK$^{1\ast}$, 
+ \author{GREG FINAK$^{1}$, 
  ANDREW MCDAVID$^{1}$,
@@
  MARIO ROEDERER$^{3}$,\\
- RAPHAEL GOTTARDO$^{1}$
+ RAPHAEL GOTTARDO$^{1\ast}$
  \\[4pt]
@@
 
-The antigen--specific subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%).  
+The antigen-specific subpopulations constitute a small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%), 
 % The intracellular cytokine staining (ICS) assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring intracellular cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
 %Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
-Thus, a sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that changes in these rare cell populations can be reliably detected.
-Subsequently, each individual cell is classified as either positive or negative for each cytokine based on predetermined thresholds, and the number of cells matching each subpopulation phenotype is counted.
+and a large number of cells must be collected (on the order of 50,000 to 100,000 T-cells) to ensure that changes in these cell populations can be reliably detected.
+Then, each individual cell is classified as either positive or negative for each cytokine based on fixed thresholds, and the number of cells matching each phenotype is counted.
 
 Cell counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences.
-\emph{Responders} are subjects whose T-cells exhibit a response to stimulation, and the response is vaccine-specific if it exists at the post-vaccine time point but not the pre-vaccine time point.
+\emph{Responders} are subjects whose T-cells respond to stimulation, and the response is vaccine-specific if it exists at the post-vaccine time point but not the pre-vaccine time point.
 %Response is vaccine--specific when a subject responds at a post--vaccine time point but not at a pre--vaccine time point. 
 %Response in placebo recipients is further used to assess vaccine-specificity. 
-These comparisons are typically performed within a time point  in order to avoid confounding vaccine effects with time of blood draw and because baseline samples are not always available. 
+These comparisons are typically performed within a time point in order to avoid confounding vaccine effects time and because baseline samples are not always available. 
 %In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.
@@
 Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model.
-In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present results using single-cell gene expression data.
+In Section \ref{s:demarkercombos}, we present a multivariate extension that can be used to test multivariate biomarker differential expression and present results using single-cell gene expression data.
 Finally, in Section \ref{s:discussion} we discuss our findings and future work.
-
 \section{Notation and Data}
@@
 Each cell can either be positive or negative for a biomarker.
-Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations.
+Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ Boolean combinations.
 We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and unstimulated samples, respectively.
 We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample.
-For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$, and finally $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
+For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(\cdot)}_{i}=\{n^{(\cdot)}_{ik}: k=1,\dots,2^K\}$, and finally $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
 
@@
 
-\textit{Flow cytometry}:
+\noindent\textit{Flow cytometry}:
 Our primary dataset is ICS data generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen~\citep{Goepfert:2011ci}.
@@
 No response is expected at day 0, while day 182 is close to the immunogenicity peak and many vaccinees are expected to respond for at least some cytokines.
-A median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively).
+A median of 51K and 58K T-cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively).
 The empirical differences in the proportions of stimulated and unstimulated cells were on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, depending on cytokine subset.
 
-\textit{Fluidigm single-cell gene expression}: Tetramer-sorted CD8$+$ T-cells from sixteen subjects.
+\noindent\textit{Fluidigm single-cell gene expression}: Tetramer-sorted CD8$+$ T-cells from sixteen subjects.
 T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation was measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays.
@@
 In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model.
-
 \subsection{Beta-binomial model}
@@
 In order to detect responding subjects, we consider two competing models,
-${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
+${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}$ and ${\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
 Under the null model, ${\cal M}_0$, there is no difference between the stimulated and unstimulated samples, and the proportions are equal (yet the cell counts can differ).
@@
 
-\noindent\textbf{EM algorithm}\\
-Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
-\[
-\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.
-\]
+\noindent\textbf{EM algorithm:} Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
+$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$; see Appendix A in Supplementary Material for details.
+%=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.\]
 The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to
@@
 
-\noindent\textbf{MCMC algorithm}\\
-We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}.
+\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
@@
  %(\mathbf{p}^{(s)}_{i}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}^{(s)}) \quad \text{and}\quad (\mathbf{p}^{(u)}_{i}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\end{align*}
-The indicator variable $z_i$ is defined as in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$.
+The indicator variable $z_i$ is defined as in Section \ref{ss:priors}.
 As in the beta-binomial case, both EM and MCMC algorithms can be used for parameter estimation.
@@
 %\subsection{Parameter estimation}
-We again make use of the marginal likelihoods in closed form (see Appendix C of the Supplementary Material). 
+We again make use of the marginal likelihoods available in closed form (see Appendix C of the Supplementary Material). 
 %For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
@@
 The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but demonstrates that biologically relevant signal is present in the frequency of gene expression (Figure~\ref{fig:fluidigm}).
-We have shown elsewhere that changes in both the frequency of expression and expression level are informative~{McDavid2012}, and extensions of the model that incorporate the continuous component (MFI or gene expression) are warranted in the future.
+We have shown elsewhere that changes in both the frequency of expression and expression level are informative~\citep{McDavid2012}, and extensions of the model that incorporate the continuous component (MFI or gene expression) are warranted in the future.
 Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative subsets.
@@
 For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha^{\{s,u\}}_{k}=\alpha^{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
-As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
+In reality, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
 More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.
@@
 All of the results presented here were obtained with an R and C++ implementation the MIMOSA models, freely available at (\url{http://www.github.org/RGLab/MIMOSA}).
-An R package will be released as part of the Bioconductor project~\citep{Gentleman:2004tt}.
 
@@
 \section*{Acknowledgments}
-This work was supported by the Intramural Research Program of NIAID and NIH, grants [R01 EB008400, U01 AI068635-01], the Bill and Melinda Gates Foundation through grants to VISC [OPP38744,  OPP1032317] and CAVD [OPP1032325].
+This work was supported by the Intramural Research Program of NIAID and NIH, grants [R01 EB008400, U01 AI068635-01], the Bill and Melinda Gates Foundation through grants [OPP38744,  OPP1032317, OPP1032325].
 HVTN065 was conducted by the HVTN and supported by NIAID.
-The the HVTN Lab Program is supported by the NIH [UM1AI068618], the Public Health Service, and the University of Washington Center for AIDS Research [UM1 AI068618, P30 AI027757].
+The the HVTN Lab Program is supported by the NIH [UM1AI068618], the Public Health Service, and the University of Washington CFAR [UM1 AI068618, P30 AI027757].
 Thanks to the James B.
@@
    \centering
-   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
+%   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
    \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.

### 2013-06-04 MIMOSA.tex
Added calculation of \tilde{z}_i to the supplementary material
Removed some references. Almost there!

@@
 Current approaches, such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
-Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton2004,Smyth:2005iy,Gottardo:2006}.
+Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep[e.g.][]{Newton2004,Gottardo:2006}.
 In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.
@@
 Detecting differences in poly-functional cell populations is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
-In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
+In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep[][e.g.]{Betts:2006dw,Precopio:2007ht}.
 In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells.
@@
    \centering
-%   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
+   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
    \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set.

### 2013-06-04 MIMOSA.tex
fix dash and endash

--- 2013-06-04 MIMOSA.tex
@@
 Our model allows the inference to be subject specific, as is typically required when assessing vaccine responses, while borrowing strength across subjects through common prior distributions.
-We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
+We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation--Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm.
 We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes.
@@
 Additional simulations show that our framework is also robust to model misspecification.
-Finally, we demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
+Finally, we demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this approach using single-cell gene expression data and simulations.}{Expectation--Maximization, Markov Chain Monte Carlo,
 Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
@@
 
-Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
+Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non-parametric methods~\citep{Sinclair:2004hs}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
 All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (or non-responders).
@@
 \noindent\textit{Flow cytometry}:
-Our primary dataset is ICS data generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen~\citep{Goepfert:2011ci}.
-The study goal was to assess the immunogenicity of different prime-boost regimens.
+Our primary dataset is ICS data generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime--boost regimen~\citep{Goepfert:2011ci}.
+The study goal was to assess the immunogenicity of different prime--boost regimens.
 Here, we analyze a subset of the data consisting of 98 vaccine and placebo recipients at days 0 (pre-vaccine) and 182 (two weeks post-vaccine), focusing on CD4+ T-cell response to stimulation with HIV Envelope peptide.

### 2013-06-04 MIMOSA.tex
shorten again

--- 2013-06-04 MIMOSA.tex
@@
  STEVE DE ROSA$^{1,2}$,
- MARIO ROEDERER$^{3}$,\\
+ MARIO ROEDERER$^{3}$,
  RAPHAEL GOTTARDO$^{1\ast}$
@@
  \\[2pt]
- {gfinak@fhcrc.org}}
+ {rgottard@fhcrc.org}}
 
@@
 T-cells that recognize the antigen become \emph{activated}, produce cytokines that potentiate the immune response, and proliferate and persist in the immune system providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.
-This vaccine memory can be interrogated through the intracellular cytokine staining (ICS) assay, by measuring cytokine production in antigen-specific T-cells in response to antigen stimulation~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
- %While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
- %After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.
-
+This vaccine memory can be interrogated through the intracellular cytokine staining (ICS) assay, by measuring antigen-specific cytokine production in response to stimulation~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 The antigen-specific subpopulations constitute a small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%),
@@
 
-\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}.
+\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms \citep{Gelfand1990}.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
-Details about the algorithms and implementation for the data presented here are given in Appendix B of the Supplementary Material. 
+Details about the algorithms and implementation are given in Appendix B of the Supplementary Material. 
 %We used the
@@
 
-MIMOSA is fit to dichotomized data  (cells are either positive or negative).
+MIMOSA is fit to dichotomized data (cells are positive or negative).
 While this is the standard approach for ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information.
@@
 The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but demonstrates that biologically relevant signal is present in the frequency of gene expression (Figure~\ref{fig:fluidigm}).
-We have shown elsewhere that changes in both the frequency of expression and expression level are informative~\citep{McDavid2012}, and extensions of the model that incorporate the continuous component (MFI or gene expression) are warranted in the future.
+We have shown elsewhere that changes in both the frequency of expression and expression level are informative~\citep{McDavid2012}, and extensions of the model that incorporate the continuous component are warranted in the future.
 Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative subsets.
@@
 
-When changes are two--sided (as with Fluidigm), differences in poly-functional cell populations are not always detectable  since differences may cancel in the marginal populations (Figure~\ref{fig:polyfunctionality} B).
+When changes are two-sided (as with Fluidigm), differences in poly-functional cell populations are not always detectable  since differences may cancel in the marginal populations (Figure~\ref{fig:polyfunctionality} B).
 In this case, the use of multivariate models, as our Dirichlet-multinomial model, will be important (Figure~\ref{fig:polyfunctionality} A).

### 2013-06-04 MIMOSA.tex
added tilde zi

@@
 \noindent\textbf{EM algorithm:} Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
-$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$; see Appendix A in Supplementary Material for details.
+$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$; see Appendix B in Supplementary Material for details.
 %=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.\]
@@
 
-\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms \citep{Gelfand1990}.
+\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.

### 2013-06-04 MIMOSA.tex
Added some references and citations.

--- 2013-06-04 MIMOSA.tex
@@
 Current approaches, such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small.
-Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep[e.g.][]{Newton2004,Gottardo:2006}.
-In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.
+Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Newton2004,Gottardo:2006,Lo2007}.
+In addition, most of these methods are univariate in nature and inappropriate for high-dimensional, next-generation single-cell assays.
 
@@
 Detecting differences in poly-functional cell populations is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
-In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep[][e.g.]{Betts:2006dw,Precopio:2007ht}.
+In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep[][e.g.]{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
 In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells.
@@
 \section*{Acknowledgments}
-This work was supported by the Intramural Research Program of NIAID and NIH, grants [R01 EB008400, U01 AI068635-01], the Bill and Melinda Gates Foundation through grants [OPP38744,  OPP1032317, OPP1032325].
+This work was supported by the Intramural Research Program of NIAID and NIH, grants [R01 EB008400, U01 AI068635-01], the Bill and Melinda Gates Foundation through grants to VISC [OPP38744,  OPP1032317] and CAVD [OPP1032325].
 HVTN065 was conducted by the HVTN and supported by NIAID.
-The the HVTN Lab Program is supported by the NIH [UM1AI068618], the Public Health Service, and the University of Washington CFAR [UM1 AI068618, P30 AI027757].
+The the HVTN Lab Program is supported by the NIH [UM1AI068618], the Public Health Service, and the University of Washington Center for AIDS Research [UM1 AI068618, P30 AI027757].
 Thanks to the James B.
 Pendleton Charitable Trust for a generous equipment donation.
-\vspace{-10pt}
 \bibliographystyle{biorefs}

### 2013-06-04 MIMOSA.tex
update

@@
 \noindent\textbf{EM algorithm:} Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
-$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$; see Appendix B in Supplementary Material for details.
-%=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.\]
+$\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})$:
+\[
+ \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)},|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)}|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}^{(u)},\tilde{\beta}^{(u)},\tilde{\alpha}^{(s)},\tilde{\beta}^{(s)}|\mathbf{y}_i)}.\]
 The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to
@@
 
-\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms.
+\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms~\citep{Gelfand1990}.
 All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
@@
 \bibliography{MIMOSA}
-\clearpage
+
 \begin{table}[ht]

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.Rmd
add mimosafigures

Load the data and fit the MIMOSA model
===============

```{r }
source("utility.R")
```{r echo=FALSE}
setup()
```

Read and merge the hvtn data. Cache it.

```{r cache=TRUE}
hvtndata<-readHVTNData()
hvtndata<-processHVTNData(hvtndata)
hvtndata<-constructEset(hvtndata)
```

read the fluidigm data

```{r cache=FALSE}
fluidigm<-readFluidigmData()
```

Process it

```{r cache=FALSE}
fluidigm<-processFluidigmData()
```

Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC

```{r }
if(!file.exists("hvtnEMfit.rds")){
  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
}else{
  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
}
if(!file.exists("hvtnMCMCfit.rds")){
  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE)
  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
}else{
  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
}
```

drop the stuff fit via mcmc and summarize, fit LRT, Fisher's and rank by log fold change

```{r cache=TRUE}
hvtn.summary.EM<-summarizeMIMOSAFit(hvtn.result.EM)
hvtn.summary.EM<-subset(hvtn.summary.EM,Method%in%"EM")
hvtn.summary.mcmc<-summarizeMIMOSAFit(hvtn.result.mcmc)
hvtn.summary.mcmc<-FisherFromSummary(hvtn.summary.mcmc)
hvtn.summary.mcmc<-LRT(hvtn.summary.mcmc)
hvtn.summary.mcmc<-FoldChangeRank(hvtn.summary.mcmc)
```

Compute ROC and FDR curves and calculate the AUC.
+ cache=TRUE

```{r }
ROC<-ComputeROCs(hvtn.summary.mcmc,hvtn.summary.EM)
FDR<-ComputeFDRs(hvtn.summary.mcmc,hvtn.summary.EM)
AUC<-ComputeAUCs(ROC)
```

FDR and ROC plots
--------
We see that MIMOSA performs as well or better than competing methods. The MCMC routine performs better than the EM routine for some data.

```{r cache=TRUE,fig.width=10,fig.height=8,fig.cap='The TNF expressing cytokine specific cells have high background as evidenced by the poor performance of all methods fit to TNF+ subsets of cells, with the exception of the triple-positive cell subset. As the number of simultaneously expressed cytokines being examined increases, the background decreases.'}
ROC$Method<-relevel(ROC$Method,"MIMOSA (EM)")
ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=Method),size=3,data=AUC,show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
```{r fig.width=10,fig.height=8,fig.cap='The FDR is well controlled for cytokine subsets with low background. The TNF expressing subsets have high background and consequently positive assay results in the "Visit 2" samples, which are labelled as "true negatives" in this scheme. The result is observed as a large observed false discovery rate for all methods.'}
FDR$Method<-relevel(FDR$Method,"MIMOSA (EM)")
ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
```

Process the fluidigm data to construct an expression set for MIMOSA
---------

```{r warning=FALSE,cache=TRUE}
fl.bystim<-lapply(fluidigm,exprs)
cd.bystim<-lapply(fluidigm,cData)
fl<-lapply(names(fl.bystim),function(n){
  m<-melt(cbind(fl.bystim[[n]],cd.bystim[[n]]),id=c("Stim.Agent","Stim.Condition","Time.of.Stim","Sero.Status","Patient.ID","Chip.Number","Well","Number.of.Cells"))
  ddply(m,.(Stim.Condition,Patient.ID,variable),summarize,N=length(value),n=sum(value>0))
})
fl<-lapply(fl,function(x)setnames(x,"variable","Gene"))
Esets.fluidigm<-lapply(fl,function(x)ConstructMIMOSAExpressionSet(x,reference=Stim.Condition%in%"Unstim",measure.columns=c("N","n"),other.annotations=c("Gene","Patient.ID","Stim.Condition"),default.cast.formula=component~Patient.ID+Gene,.variables=.(Patient.ID,Gene),featureCols=1,ref.append.replace="_REF"))
nms<-names(fl.bystim)
Esets.fluidigm<-lapply(1:length(Esets.fluidigm),function(x){pData(Esets.fluidigm[[x]])<-cbind(pData(Esets.fluidigm[[x]]),Stim=nms[x]);Esets.fluidigm[[x]]})
Eset.combined<-combine(Esets.fluidigm[[1]],Esets.fluidigm[[2]],Esets.fluidigm[[3]],Esets.fluidigm[[4]])
```

Fit MIMOSA by gene, grouping by stimulation and for all samples together.
----

```{r fluidigm.fit, cache=TRUE}
if(!file.exists("fluidigm.fits.by.stim.allgenes.rds")){
  fluidigm.fits.by.stim.allgenes<-lapply(list(Eset.combined),function(x)MIMOSA(N+n~Stim+Patient.ID|Gene,x,ref=RefTreat%in%"Reference",subset=RefTreat%in%"Treatment",method="mcmc",EXPRATE=1e-4,iter=105000,burn=5000,pXi=1,alternative="not equal",run.parallel=TRUE,getP=TRUE))
  fluidigm.fits.by.stim.bygene<-mclapply(Esets.fluidigm,function(x)MIMOSA(N+n~Stim+Patient.ID|Gene,x,ref=RefTreat%in%"Reference",subset=RefTreat%in%"Treatment",method="mcmc",EXPRATE=1e-4,iter=105000,burn=50000,pXi=1,alternative="not equal",getP=TRUE,run.parallel=TRUE))
  saveRDS(fluidigm.fits.by.stim.allgenes,file="fluidigm.fits.by.stim.allgenes.rds")
  saveRDS(fluidigm.fits.by.stim.bygene,file="fluidigm.fits.by.stim.bygene.rds")
}else{
  fluidigm.fits.by.stim.allgenes<-readRDS("fluidigm.fits.by.stim.allgenes.rds")
  fluidigm.fits.by.stim.bygene<-readRDS("fluidigm.fits.by.stim.bygene.rds")
}
```

Heatmap of posterior probabilities of response for fit of all samples by gene.
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.probs.bygene,cache=TRUE}
foo<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)x@z[,2]))
props<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))
S<-sign(props)
S[S==0]<-1
colnames(S)<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])[,c("Patient.ID")]
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=2),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
```

Heatmap of posterior differences of proportions fit by gene for all samples.
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.props.bygene,cache=TRUE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
```

Heatmap of posterior probability of response for fit of samples by stimulation and by gene
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.probs.bystim.bygene,cache=TRUE}
foo<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)x@z[,2]))))
props<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))))
S<-sign(props)
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])[,c("Patient.ID")])))
S[S==0]<-1
stim<-as.numeric(factor(do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Stim)))))
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Patient.ID)))
rownames(S)<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene[[1]],function(x)as.character(unique(pData(x)$Gene))))
heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=1),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
```

Heatmap of posterior differences in proportions for fit of samples by stimulation and by gene. 
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.props.bystim.bygene,cache=TRUE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
```

Fisher's exact test and heatmaps of -log10(q-values) from Fisher's
---
There is little clustering by stimulation evident from testing for differential expression with Fisher's exact test.

```{r cache=TRUE}
fisher<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
  p<-vector('numeric',nrow(x@result@n.stim))
  for(i in 1:nrow(x@result@n.stim)){
    p[i]<-fisher.test(as.matrix(rbind(x@result@n.stim[i,],x@result@n.unstim[i,])),alternative="two.sided")$p.value
  }
  p
}))
S<-sign(lapply(fluidigm.fits.by.stim.allgenes,function(x)do.call(rbind,lapply(x,function(x)(prop.table(as.matrix(x@result@n.stim),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2])))[[1]])
S[S==0]<-1
colnames(fisher)<-gsub("_B3GAT1_Treatment","",colnames(S))
rownames(fisher)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
heatmap.2(fisher*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
```

Multivariate MIMOSA on fluidigm data looking at interactions
---------
There is a difference when comparing stimulations and looking at the multivariate vector of expression for GZMK and CCR7 (i.e [+/+,+/-,-/+,--]), but no difference is detected when we look at expression of GZMK | CCR7, since one cell subset increases while the other decreases upon stimultion.

```{r cache=FALSE}
fl.comb<-combine(fluidigm[[1]],fluidigm[[2]],fluidigm[[3]],fluidigm[[4]])
melted<-melt(fl.comb)
require(glmnet)
require(Matrix)
melted$Et.bin<-as.numeric(as.logical(melted$Et))
#setnames(melted,"__wellKey","wellKey")
melted<-within(melted,group<-Patient.ID:Stim.Agent)
melted<-cast(melt(melted,id=c("group","Stim.Condition","primerid","wellKey"),measure="Et.bin"),group+Stim.Condition+wellKey~primerid)

interact.scan<-combn(colnames(melted)[4:99],2,function(x){
  g1<-x[1];g2<-x[2]
  gg<-paste(g1,g2,sep=":")
  bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
  bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
  setnames(bar,"foo",gg)
  bar[,5]<-factor(bar[,5],levels=c("0:0","0:1","1:0","1:1"))
  C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
  expected<-c("0:0","0:1","1:0","1:1")
  unused<-setdiff(expected,colnames(C))
  newframe<-NULL
  for(i in seq_along(expected)){
    if(expected[i]%in%unused){
      newframe<-cbind(newframe,rep(0,nrow(C)))
    }else{
      newframe<-cbind(newframe,C[,expected[i]])
    }
  }
  colnames(newframe)<-expected
  C<-cbind(C[,1:2],newframe)
  CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])
  foo<-lapply(CC,function(x)prop.table(as.matrix(x),1))
  psum<-mean(rowSums(foo[[1]][,2:4])-rowSums(foo[[2]][,2:4]))
  pij<-colMeans(foo[[1]][,2:4]-foo[[2]][,2:4])
  res<-c(psum,pij)
  res
})

genes<-combn(colnames(melted)[6:101],2)[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
interact.scan[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
ggg<-genes[,6]
g1<-ggg[1];g2<-ggg[2] #CCR7 and GZMK
g1<-"CCR7";g2<-"GZMK"
gg<-paste(g1,g2,sep=":")
bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
setnames(bar,"foo",gg)
C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])

toplot<-data.frame(rbind(data.frame(CC$n.stim),data.frame(CC$n.unstim)),stim=c(rep("Stim",nrow(CC$n.stim)),rep("Unstim",nrow(CC$n.unstim))))
toplot$id<-rep(unlist(lapply(strsplit(as.character(subset(C,Stim.Condition%in%"Unstim")[,1]),":"),function(x)x[1]),use.names=FALSE),2)
colnames(toplot)[1:4]<-gsub("X","",paste(g1,g2,colnames(toplot)[1:4],sep=":"))
toplot<-melt(toplot,id=c("id","stim"))
ggplot(toplot)+geom_raster(aes(x=variable,y=id,fill=value))+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray",breaks=round(seq(0,96,l=10)))

DD<-CC
DD[[1]]<-cbind(CC[[1]][,1],rowSums(CC[[1]][,2:4]))
DD[[2]]<-cbind(CC[[2]][,1],rowSums(CC[[2]][,2:4]))
toplot2<-data.frame(rbind(data.frame(DD$n.stim),data.frame(DD$n.unstim)),stim=c(rep("Stim",nrow(DD$n.stim)),rep("Unstim",nrow(DD$n.unstim))))
toplot2$id<-rep(unlist(lapply(strsplit(as.character(subset(C,Stim.Condition%in%"Unstim")[,1]),":"),function(x)x[1]),use.names=FALSE),2)
colnames(toplot2)[1:2]<-gsub("X","",paste(g1,g2,colnames(toplot2)[1:2],sep=":"))
toplot2<-melt(toplot2,id=c("id","stim"))
ggplot(toplot2)+geom_raster(aes(x=variable,y=id,fill=value))+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray",breaks=round(seq(0,96,l=10)))

```{r fitmulti,echo=FALSE,message=FALSE,warning=FALSE,error=FALSE,results='hide'}
flres<-.fitMCMC(data=CC,inits=MIMOSA:::MDMix(CC,initonly=TRUE),EXPRATE=1e-2,pXi=1,alternative="not equal")
flsum<-.fitMCMC(data=DD,inits=MIMOSA:::MDMix(DD,initonly=TRUE),EXPRATE=1e-2,pXi=1,alternative="not equal")
sapply(list.files(pattern=c("\\.dat")),unlink)

A<-cbind(melt(data.frame(matrix(prop.table(as.matrix(CC$n.stim),1)-prop.table(as.matrix(CC$n.unstim),1),ncol=4))),flres$z[,2],1:length(flres$z[,2]))
B<-cbind(melt(data.frame(matrix(prop.table(as.matrix(DD$n.stim),1)-prop.table(as.matrix(DD$n.unstim),1),ncol=2))),flsum$z[,2],1:length(flres$z[,2]))
colnames(A)<-c("Component","Effect Size","Pr(response)","ID")
colnames(B)<-c("Component","Effect Size","Pr(response)","ID")
A$Component<-factor(A$Component,labels=c("-/-","-/+","+/-","+/+"))
B$Component<-factor(B$Component,labels=c("-","+"))
A<-rbind(A,B)
ggplot(A)+geom_point(aes(x=`Effect Size`,y=`Pr(response)`,shape=factor(ID)))+facet_wrap(~Component)+geom_vline(aes(x=0),lty=3)+scale_shape_manual(values=1:16,name="Subject")
```

Simulations
==========
One sided
------
We'll use the model fit from IFNg+ for ENV-1-PTEG based on the ROC and FDR plots
Note that MIMOSA has greater sensitivity and specificity than Fisher's exact test or the Likelihood ratio test. When there are 5 observations, the methods being to perform equivalently.

```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5}
sims<-OneSidedSims(hvtn.result.mcmc)
ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
```

Two-sided
--------

```{r two.sided.sims, cache=FALSE, echo=FALSE,fig.width=5,fig.height=5}
sims<-TwoSidedSims(hvtn.result.mcmc)
ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
```

Multivariate Simulations
-----
Average ROC for 10 multivariate simulations with effect size of 2.5x10-3 and -2.5x10-3 in two of eight components (N=4743 events).

```{r cache=TRUE}
set.seed(101)
as<-c(1,1,1,1,3.5,1,1)
au<-c(1,1,1,1,1,3.5,1)
as<-c(1000-sum(as),as)*100
au<-c(1000-sum(au),au)*100
if(!file.exists("mvsims.rds")){
mdsim<-sapply(1:10,function(i){
simd<-MIMOSA:::simMD(alpha.s=as,alpha.u=au,N=100,w=0.5,n=3.5,alternative="not equal")
foo<-.fitMCMC(simd,inits=MIMOSA:::MDMix(simd,initonly=TRUE))
foo
})
saveRDS(mdsim,file="mvsims.rds")
}else{
  mdsim<-readRDS("mvsims.rds")
}
fisher.p<-vector('list',ncol(mdsim))
for(i in 1:ncol(mdsim)){    
  fisher.p[[i]]<-rep(NA,nrow(mdsim[,i]$n.stim))
for(j in 1:nrow(mdsim[,i]$n.stim)){
  fisher.p[[i]][j]<-  fisher.test(cbind(mdsim[,i]$n.stim[j,],mdsim[,i]$n.unstim[j,]),alternative="two.sided",simulate.p.value=TRUE)$p
}
}

FDR<-function(x,truth){
  o<-order(x,decreasing=FALSE)
  cbind(tpr=cumsum(truth[o])/sum(truth),fpr=cumsum(!truth[o])/sum(!truth))
}

fisher.p.adj<-lapply(fisher.p,function(x)p.adjust(x,"fdr"))
fisher.roc<-do.call(rbind,by(cbind(do.call(rbind,lapply(fisher.p.adj,function(i)FDR(i,gl(2,50)==2)))),rep(gl(100,1),10),function(x)colMeans(x)))
mimosa.adj<-lapply(1:ncol(mdsim),function(i)MIMOSA:::fdr(mdsim[,i]$z))
MIMOSA.roc<-do.call(rbind,by(do.call(rbind,lapply(mimosa.adj,function(i)FDR(i,gl(2,50)==2))),rep(gl(100,1),10),function(x)colMeans(x)))
foo<-data.frame(as.data.frame(rbind(fisher.roc,MIMOSA.roc)),Method=gl(2,100,labels=c("Fisher","MIMOSA")))
sz<-element_text(size=18)
ggplot(foo)+geom_line(aes(x=fpr,y=tpr,col=Method))+scale_y_continuous("TPR")+scale_x_continuous("FPR")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)
```

Competing ROC curves to evaluate the effect of inaccuracies in the true positives and false positives.
-----
Simulate from models with 20,35,50,65,80 percent response rates but treat the ROC evaluation as 100% response.

```{r results='hide',cache=TRUE}
comproc<-CompareROC(hvtn.result.mcmc)
comproc[[1]]$true.resp.rate<-as.numeric(as.character(comproc[[1]]$q))*2
comproc[[2]]$true.resp.rate<-as.numeric(as.character(comproc[[2]]$q))*2

ggplot(comproc[[1]])+geom_line(aes(x=fdr.hat,y=true.fdr.hat,color=method))+facet_wrap(~true.resp.rate)+theme_bw()+geom_abline(1,lty=3)+labs(title="Observed vs True FDR when \ntrue response rate varies but is assumed to be 100%")
ggplot(comproc[[2]])+geom_line(aes(x=FPR.hat,y=TPR.hat,color=method))+facet_wrap(~true.resp.rate)+theme_bw()+labs(title="ROC curves when \ntrue response rate varies but is assumed to be 100%")+scale_x_continuous("False Positive Rate")+scale_y_continuous("True Positive Rate")
```

Supplementary Figure 2
------
Histogram of the empirical proportions of unstimulated cells and overlaid posterior densities of the beta distribution with $\alpha_s$ and $\alpha_u$ estimated from the data for ENV-1-PTEG stimulated, IFN-$\gamma$+ CD4+ T-cells, demonstrating that the assumption of a common distribution for piu across subjects is reasonable.

```{r supp.fig.2,message=FALSE,echo=FALSE}
res<-hvtn.result.mcmc[[7]]
ggplot(data.frame(p.hat=prop.table(as.matrix(res@result@n.unstim),1)[,2]))+geom_histogram(aes(x=p.hat,y=..density..),col="black",fill="gray")+theme_bw()+scale_x_continuous(name=expression(hat(p)[u]))+geom_line(data=data.frame(x=seq(0,3e-4,l=10000),y=dbeta(seq(0,3e-4,l=10000),res@result@params[2,4],res@result@params[2,3]),stim="ENV-1-PTEG"),aes(x=x,y=y),col="blue",lwd=2,lty=3)+facet_wrap(~stim)+theme(strip.text.x=element_text(size=24),axis.text.x=element_text(size=18),axis.text.y=element_text(size=18),axis.title.x=element_text(size=24),axis.title.y=element_text(size=24))
```

Unconstrained MIMOSA fit to data violating model assumptions
-----

```{r violated.assumptions, cache=FALSE,echo=FALSE,fig.width=8,fig.height=5}
sims<-TwoSidedSimsTruncated(hvtn.result.mcmc)
ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
```

How many CD4 and CD8 T-cells?
-----

```{r echo=FALSE,message=FALSE,warning=FALSE,result='hide',fig.width=5,fig.height=2.5}
suppressPackageStartupMessages(require(scales))
df<-data.frame(tcellsub="CD4",treatment=gl(2,224/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[1]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[1]]@result@n.unstim)[,1]))
df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,224/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[30]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[30]]@result@n.unstim)[,1])))
ggplot(df)+geom_density(aes(x=value,fill=tcellsub),alpha=0.75,position="identity")+theme_bw()+scale_x_continuous(name="Cell Count",labels=comma)
```

Table of median cell counts for different subsets
-----

```{r results='asis',message=FALSE,echo=FALSE}
suppressPackageStartupMessages(require(plyr))
suppressPackageStartupMessages(require(reshape))
suppressPackageStartupMessages(require(xtable))
d<-cast(melt(ddply(df,.(tcellsub),summarize,median=round(quantile(value,c(0.5)))),measure="median"),~tcellsub)
print(xtable(d[,-1L],display=c("fg","fg","fg")),type="html",include.rownames=FALSE)
```

Magnitude of the proportions for the HVTN data
----

```{r echo=FALSE, message=FALSE, warning=FALSE, result='hide',fig.width=5,fig.height=2.5}
df<-data.frame(tcellsub="CD4",treatment=gl(2,224/2,labels=c("stim","unstim")),melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.unstim),1)[,2])[,2]))
df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,224/2,labels=c("stim","unstim")),proportion=melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.unstim),1)[,2])[,2])))
ggplot(df)+geom_density(aes(x=value,fill=tcellsub),alpha=0.75,position="identity")+theme_bw()+scale_x_continuous(name="Cell Count",labels=scientific)+facet_wrap(~tcellsub,scales="free")
```

Table of median cell proportions for HVTN data
----

```{r results='asis',message=FALSE,echo=FALSE}
d<-cast(melt(df,measure='value'),~tcellsub,fun.aggregate=function(x)mean(x))[,-1L]
print(xtable(d,display=c("fg","fg","fg")),type="html",include.rownames=FALSE)
```

### 2013-06-25 MIMOSA/mimosafigures/compile.R

@@ -1,0 +1,8 @@
+library(knitr)
+knitr::spin("manuscriptFigures.R",FALSE)
+knit2html("manuscriptFigures.Rmd")
+#system("pandoc --jsmath manuscriptFigures.md>manuscriptFigures.html")
+system("(echo '<head><script src=\"https://raw.github.com/Polychart/polychart2/develop/polychart2.standalone.js\"></script><script src=\"\"></script></head>'; cat manuscriptFigures.html)>manuscriptFigures2.html")
+system("sed 's/&quot;/\"/g' < manuscriptFigures2.html>manuscriptFigures.html")
+system("cp manuscriptFigures.html /Users/gfinak/Dropbox/Public/MIMOSA/manuscriptFigures.html")
+system("cp -r figure /Users/gfinak/Dropbox/Public/MIMOSA/")

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.R

@@ -1,0 +1,547 @@
+#'Load the data and fit the MIMOSA model
+#'===============
+source("utility.R")
+#+ echo=FALSE
+setup()
+
+#'Read and merge the hvtn data. Cache it.
+#+ cache=TRUE
+hvtndata<-readHVTNData()
+hvtndata<-processHVTNData(hvtndata)
+hvtndata<-constructEset(hvtndata)
+
+#'read the fluidigm data
+#+ cache=FALSE
+fluidigm<-readFluidigmData()
+
+#'Process it
+#+ cache=FALSE
+fluidigm<-processFluidigmData()
+
+
+#'Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC
+if(!file.exists("hvtnEMfit.rds")){
+  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
+  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
+}else{
+  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
+}
+if(!file.exists("hvtnMCMCfit.rds")){
+  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
+  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
+}else{
+  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
+}
+
+#'drop the stuff fit via mcmc and summarize, fit LRT, Fisher's and rank by log fold change
+#+ cache=TRUE
[…]

### 2013-06-25 MIMOSA/mimosafigures/utility.R

@@ -1,0 +1,667 @@
+#'setup the environment
+#'
+#'load MIMOSA, parallel, filehash,gdata,pracma,reshape and ggplot2
+setup<-function(){
+  suppressPackageStartupMessages({require(MIMOSA)
+                                  require(multicore)
+                                  require(parallel)
+                                  require(filehash)
+                                  require(gdata)
+                                  require(pracma)
+                                  require(ggplot2)
+                                  require(reshape)
+                                  require(data.table)
+                                  require(SingleCellAssay)
+                                  require(gplots)
+                                  
+  })
+}
+
+#'construct an expected vs observed fdr curve
+#'
+#'Expected vs observed FDr taking the expected fdr and the truth.
+fdrCurves<-function(fdr,truth){
+  truth<-truth[order(fdr)]
+  true.fdr<-cumsum(1-truth)/1:length(truth)
+  fdr<-sort(fdr)
+  cbind(fdr,true.fdr)
+}
+
+#'Construct and ROC curve
+#'
+#'Given the p-value and the truth, construct an ROC curve.
+rocCurves<-function(p,truth){
+  #s<-seq(0,1,l=1000)
+  #table<-t(sapply(s,function(th){
+  #  prop.table(table(test=factor(p<=th,levels=c("FALSE","TRUE")),truth=truth),margin=2)["TRUE",]
+  #}))
+  #colnames(table)<-c("FPR","TPR")
+  #table
+  o<-order(p,decreasing=FALSE)
+  intervals<-lapply(sapply(sort(unique(p)),function(x)which(as.numeric(p[o])%in%x)),max)
[…]

### 2013-06-25 MIMOSA.tex
update paper with 2nd round of revisions

@@
 %Response in placebo recipients is further used to assess vaccine-specificity. 
-These comparisons are typically performed within a time point in order to avoid confounding vaccine effects time and because baseline samples are not always available. 
+These comparisons are typically performed within a time point in order to avoid confounding vaccine effects with time of visit and because baseline samples are not always available. 
 %In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.
@@
 \label{s:data}
-
-We use the following notation to describe our model.
-We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and unstimulated.
-Each cell can either be positive or negative for a biomarker.
-Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ Boolean combinations.
-We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and unstimulated samples, respectively.
-We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample.
-For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(\cdot)}_{i}=\{n^{(\cdot)}_{ik}: k=1,\dots,2^K\}$, and finally $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
 
@@
 \label{s:DEone}
+
 Datasets like those presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many combinations.
 By univariate, we mean that we have only one positive cell subset.
-This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ or B$+$ (\textit{vs.} A$-$/B$-$).
+This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ OR B$+$ (\textit{vs.} A$-$/B$-$).
 Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$).
@@
 \label{s:DE}
-For a given subject $i$, the positive cell counts for the stimulated and unstimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and unstimulated paired samples, respectively.
+
+We use the following notation to describe our model.
+We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and unstimulated.
+Each cell can either be positive or negative for a biomarker.
+Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ Boolean combinations.
+We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and unstimulated samples, respectively.
+We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample.
+For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(\cdot)}_{i}=\{n^{(\cdot)}_{ik}: k=1,\dots,2^K\}$, and finally $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.
+
+For a given subject $i$, the positive cell counts for the stimulated and unstimulated samples are jointly modeled as $(n^{(s)}_{i}|p^{(s)}_{i}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and unstimulated paired samples, respectively.
 In order to detect responding subjects, we consider two competing models,
@@
 For MIMOSA, the FDR was computed using the approach of~\citet{Newton2004} whereas for all other methods FDR control was done using the approach of~\citet{Benjamini1995}.
+The posterior mean of the $z_i$ was used for ranking the results in the FDR calculations.
 
@@
 \label{s:ics}
-In the ICS data we considered observations from vaccinees at day 0 or from placebos at day 0 or 182 as \emph{true negatives} in that we do not expect to observe a response in those samples.
+In applying the MIMOSA model to the ICS data set, we fit a separate model for each cytokine subset (and each stimulation if we had analyzed more than one).
+We do not compare pre-vaccination and post-vaccination time points against each other directly.
+Rather, we have paired samples for each time points with the corresponding negative controls.
+Observations within each time point are contrasted against the corresponding negative controls, and we fit the model to observations from all time points simultaneously.
+Consequently all time points that are fit simultaneously will contribute to the priors for $p^{(u)}$.
+A subsequent evaluation of pre- and post-vaccination time points would involve comparing the positivity calls for paired samples at the two time points.
+However, data are not typically analyzed this way, as this could lead to confounding of date with vaccine effect (i.e.
+batch effects).
+Furthermore, vaccine trials do not always collect baseline samples.
+
+Within this framework, we considered observations from vaccinees at day 0 or from placebos at day 0 or 182 as \emph{true negatives} in that we do not expect to observe a response in those samples.
 Observations from vaccinees at day 182 were considered \emph{true positives} in that a response is expected assuming a 100\% response rate.
@@
 In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F).
-Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
+Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 OR IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}.
 These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (Supplementary Figure 1).
@@
 A similar analysis applying a two-sided Fisher's exact test to each sample and gene, and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation.
+We note that the differences in the clustering are not surprising given that the MIMOSA model is sharing information across subjects within a stimulation.
 At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes.
@@
 The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases.
-The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts are bidirectional and their sum is near zero.
+The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ OR GZMK$+$), would not be appropriate in this case, since changes in the counts are bidirectional and their sum is near zero.
 The multivariate MIMOSA model tests for any difference across all polyfunctional cell subsets and detects significant differences in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels), whereas the standard approach detects no differences (Figure~\ref{fig:polyfunctionality} B).
@@
 All of the results presented here were obtained with an R and C++ implementation the MIMOSA models, freely available at (\url{http://www.github.org/RGLab/MIMOSA}).
+The code to replicate the analyisis is available at (\url{http://www.github.org/RGLab/MIMOSA_manuscript/MIMOSA/mimosafigures}).
 
@@
 Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2.
-E-F) ENV-1-PTEG and expressing IFN-$\gamma$ or IL2.
+E-F) ENV-1-PTEG and expressing IFN-$\gamma$ OR IL2.
 ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
@@
 D) Clustering of the signed q-values from Fisher's exact test.
+The differences in clustering are not surprising given that the MIMOSA model shares information within stimulation.
 All genes  were significant at, or below, the 10\% FDR level.

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.Rmd
Update code to use bootstrap etc..

@@
 if(!file.exists("hvtnMCMCfit.rds")){
-  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE)
+  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
   saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
@@
 ROC$Method<-relevel(ROC$Method,"MIMOSA (EM)")
-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=Method),size=3,data=AUC,show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+levs <-
+  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
+    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
+    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
+    "IL2+IFNg+", "IL2+TNF+", "TNF+")
+ROC$CYTOKINE<-factor(ROC$CYTOKINE,labels=levs)
+AUC$CYTOKINE<-factor(AUC$CYTOKINE,labels=levs)
+rocplots<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=Method),size=3,data=AUC,show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
 ```{r fig.width=10,fig.height=8,fig.cap='The FDR is well controlled for cytokine subsets with low background.
@@
 FDR$Method<-relevel(FDR$Method,"MIMOSA (EM)")
-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+levs<-levels(FDR$CYTOKINE)
+levs <-
+  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
+    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
+    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
+    "IL2+IFNg+", "IL2+TNF+", "TNF+")
+FDR$CYTOKINE<-factor(FDR$CYTOKINE,labels=levs)
+fdrplots<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+```
+
+Figure 1 and S1. Write to Figures folder.
+Don't evaluate here.
+
+```{r echo=FALSE,eval=FALSE}
+suppressPackageStartupMessages(library(gridExtra))
+f1levs<-levs[c(8,11,13,17,15,16,10,7)]
+sflevs<-setdiff(levs,f1levs)
+AUC$y[AUC$Method%in%"Fisher"]<-0.4
+AUC$y[AUC$Method%in%"MIMOSA (EM)"]<-0.5
+AUC$y[AUC$Method%in%"MIMOSA (MCMC)"]<-0.3
+
+fnames<-c("revised_2.pdf","revised_1.pdf","revised_15.pdf","revised_3.pdf","revised_5.pdf","revised_6.pdf","revised_7.pdf","revised_8.pdf")
+for(i in seq_along(f1levs)){
+  fi.1<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"&CYTOKINE%in%f1levs[i]))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=factor(Method)),data=subset(AUC,CYTOKINE%in%f1levs[i]),show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+  fi.2<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")&CYTOKINE%in%f1levs[i]))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+  pdf(file=sprintf("/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/%s",fnames[i]),width=10,height=3)
+  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
+  dev.off()
+  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
+ }
 ```
@@
 rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=2),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
+heatmap.2((S*(foo))[apply(foo,1,function(x)sum(x>0.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
+```
+
+Figure 2 C, (fit of all samples simultaneously)
+
+```{r eval=FALSE,echo=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmCombinedFit.pdf")
+foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+dev.off()
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 ```
@@
 rownames(props)<-rownames(S)
-heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
+heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
 ```
@@
 rownames(S)<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene[[1]],function(x)as.character(unique(pData(x)$Gene))))
-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=1),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
+h<-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
+hmaporder<-h$colInd
+geneinds<-apply(foo,1,function(x)sum(x>.6)>=1)
+```
+
+Figure 2 A
+
+```{r eval=FALSE,echo=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Fluidigm_plot1.pdf")
+foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+dev.off()
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 ```
@@
 rownames(props)<-rownames(S)
-heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
+heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
 ```
@@
 }))
+adj.fisher<-matrix(p.adjust(fisher,"fdr"),nrow=96)
 S<-sign(lapply(fluidigm.fits.by.stim.allgenes,function(x)do.call(rbind,lapply(x,function(x)(prop.table(as.matrix(x@result@n.stim),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2])))[[1]])
@@
 stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
-heatmap.2(fisher*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
+heatmap.2(fisher*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
+```
+
+Figure 2 D
+
+```{r eval=FALSE,echo=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmFisher.pdf")
+colnames(adj.fisher)<-colnames(fisher)
+rownames(adj.fisher)<-rownames(fisher)
+heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+dev.off()
+heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+```
+
+Figure 2C empirical proportions
+
+```{r eval=FALSE,echo=FALSE}
+emprop<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
+  p<-vector('numeric',nrow(x@result@n.stim))
+  p<-(prop.table(as.matrix((x@result@n.stim)),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2]
+}))
+rownames(emprop)<-rownames(fisher)
+colnames(emprop)<-colnames(fisher)
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmEmpiricalProps.pdf")
+heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
+dev.off()
+heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
 ```
@@
 toplot<-melt(toplot,id=c("id","stim"))
-ggplot(toplot)+geom_raster(aes(x=variable,y=id,fill=value))+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray",breaks=round(seq(0,96,l=10)))
+p1<-ggplot(toplot)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray")
 
@@
 toplot2<-melt(toplot2,id=c("id","stim"))
-ggplot(toplot2)+geom_raster(aes(x=variable,y=id,fill=value))+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray",breaks=round(seq(0,96,l=10)))
+p2<-ggplot(toplot2)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray")
 
@@
 sapply(list.files(pattern=c("\\.dat")),unlink)
+```
+
+Figure 4
+
+```{r eval=FALSE,echo=FALSE}
+library(gridExtra)
+missed<-!MIMOSA:::fdr(flres$z)<0.05
+toplot$variable<-factor(toplot$variable,labels=c("CCR7-GZMK-","CCR7-GZMK+","CCR7+GZMK-","CCR7+GZMK+"))
+toplot2$variable<-factor(toplot2$variable,labels=c("CCR7-GZMK-","CCR7+ or GZMK+"))
+p1<-ggplot(toplot)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous("log2 count",na.value="gray",breaks=round(seq(0,6,l=10)))+theme(axis.text.y=element_text(color=c("black","red")[missed+1]))
+p2<-ggplot(toplot2)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90,hjust=1))+facet_wrap(~stim)+scale_fill_continuous("log2 count",na.value="gray",breaks=round(seq(0,6,l=10)))+theme(axis.text.y=element_text(color=c("black","red")[2]))
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/revised_Fluidigm_Multivariate_CCR7_GZMK.pdf",width=6,height=4)
+grid.arrange(p1,ncol=1)
+dev.off()
+grid.arrange(p1,ncol=1)
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/revised_Fluidigm_Marginal_CCR7_GZMK.pdf",width=6,height=4)
+grid.arrange(p2,ncol=1)
+dev.off()
+grid.arrange(p2,ncol=1)
 
@@
 ```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5}
-sims<-OneSidedSims(hvtn.result.mcmc)
-ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+if(!file.exists("cache/sims.50K.RData")){
+  sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
+  sims.25K<-OneSidedSims(hvtn.result.mcmc,N=25000)
+  sims.10K<-OneSidedSims(hvtn.result.mcmc,N=10000)
+  cache('sims.50K')
+  cache('sims.25K')
+  cache('sims.10K')
+}else{
+  load("cache/sims.10K.RData")
+  load("cache/sims.25K.RData")
+  load("cache/sims.50K.RData")
+}
+ggplot(sims.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims.10K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims.10K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
 ```
@@
 ```{r two.sided.sims, cache=FALSE, echo=FALSE,fig.width=5,fig.height=5}
-sims<-TwoSidedSims(hvtn.result.mcmc)
-ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+if(!file.exists("cache/sims2.10K.RData")){
+sims2.50K<-TwoSidedSims(hvtn.result.mcmc,N=50000)
+sims2.25K<-TwoSidedSims(hvtn.result.mcmc,N=25000)
+sims2.10K<-TwoSidedSims(hvtn.result.mcmc,N=10000)
+cache('sims2.50K')
+cache('sims2.25K')
+cache('sims2.10K')
+}else{
+  load("cache/sims2.10K.RData")
+  load("cache/sims2.25K.RData")
+  load("cache/sims2.50K.RData")
+}
+ggplot(sims2.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims2.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims2.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims2.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims2.10K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims2.10K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+```
+
+Figure 3
+
+```{r eval=FALSE,echo=FALSE}
+levs<-levels(sims.50K$sim.ROC$method)
+levs[2]<-c("MIMOSA (one sided)")
+sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
+p1.1<-ggplot(subset(sims.50K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+levs<-levels(sims2.50K$sim.ROC$method)
+levs[2]<-c("MIMOSA (two sided)")
+sims2.50K$sim.ROC$method<-factor(sims2.50K$sim.ROC$method,labels=levs)
+p2.1<-ggplot(subset(sims2.50K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (two sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+
+levs<-levels(sims.50K$sim.FDR$method)
+levs[1]<-"MIMOSA (one sided)"
+sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
+p1.2<-ggplot(subset(sims.50K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+levs<-levels(sims2.50K$sim.FDR$method)
+levs[1]<-c("MIMOSA (two sided)")
+sims2.50K$sim.FDR$method<-factor(sims2.50K$sim.FDR$method,labels=levs)
+p2.2<-ggplot(subset(sims2.50K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (two sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_ROC_50K.pdf",width=6,height=4)
+p1.1
+dev.off()
+p1.1
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_TwoSided_ROC_50K.pdf",width=6,height=4)
+p2.1
+dev.off()
+p2.1
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_TwoSided_FDR_50K.pdf",width=6,height=4)
+p2.2
+dev.off()
+p2.2
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_FDR_50K.pdf",width=6,height=4)
+p1.2
+dev.off()
+p1.2
 ```
@@
 
-FDR<-function(x,truth){
+ROCfun<-function(x,truth){
   o<-order(x,decreasing=FALSE)
@@
 fisher.p.adj<-lapply(fisher.p,function(x)p.adjust(x,"fdr"))
-fisher.roc<-do.call(rbind,by(cbind(do.call(rbind,lapply(fisher.p.adj,function(i)FDR(i,gl(2,50)==2)))),rep(gl(100,1),10),function(x)colMeans(x)))
+fisher.roc<-do.call(rbind,by(cbind(do.call(rbind,lapply(fisher.p.adj,function(i)ROCfun(i,gl(2,50)==2)))),rep(gl(100,1),10),function(x)colMeans(x)))
 mimosa.adj<-lapply(1:ncol(mdsim),function(i)MIMOSA:::fdr(mdsim[,i]$z))
-MIMOSA.roc<-do.call(rbind,by(do.call(rbind,lapply(mimosa.adj,function(i)FDR(i,gl(2,50)==2))),rep(gl(100,1),10),function(x)colMeans(x)))
+MIMOSA.roc<-do.call(rbind,by(do.call(rbind,lapply(mimosa.adj,function(i)ROCfun(i,gl(2,50)==2))),rep(gl(100,1),10),function(x)colMeans(x)))
 foo<-data.frame(as.data.frame(rbind(fisher.roc,MIMOSA.roc)),Method=gl(2,100,labels=c("Fisher","MIMOSA")))
 sz<-element_text(size=18)
-ggplot(foo)+geom_line(aes(x=fpr,y=tpr,col=Method))+scale_y_continuous("TPR")+scale_x_continuous("FPR")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)
+p1<-ggplot(foo)+geom_line(aes(x=fpr,y=tpr,col=Method))+scale_y_continuous("TPR")+scale_x_continuous("FPR")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)
+#FDR
+mim.fdr<-as.data.frame(cbind(do.call(rbind,lapply(mimosa.adj,function(x)fdrCurves(x,gl(2,50)==2))),rep=gl(100,1)))
+ddply(mim.fdr,.(rep),summarize,true.fdr.hat=mean(true.fdr),fdr.hat=mean(fdr))
+fisher.fdr<-as.data.frame(cbind(do.call(rbind,lapply(fisher.p.adj,function(x)fdrCurves(x,gl(2,50)==2))),rep=gl(100,1)))
+plot(ddply(mim.fdr,.(rep),summarize,true.fdr.hat=mean(true.fdr),fdr.hat=mean(fdr)))
+```
+
+Figure 5
+
+```{r eval=FALSE,echo=FALSE}
 ```
@@
 
-ggplot(comproc[[1]])+geom_line(aes(x=fdr.hat,y=true.fdr.hat,color=method))+facet_wrap(~true.resp.rate)+theme_bw()+geom_abline(1,lty=3)+labs(title="Observed vs True FDR when \ntrue response rate varies but is assumed to be 100%")
-ggplot(comproc[[2]])+geom_line(aes(x=FPR.hat,y=TPR.hat,color=method))+facet_wrap(~true.resp.rate)+theme_bw()+labs(title="ROC curves when \ntrue response rate varies but is assumed to be 100%")+scale_x_continuous("False Positive Rate")+scale_y_continuous("True Positive Rate")
+p8<-ggplot(comproc[[1]])+geom_line(aes(x=fdr.hat,y=true.fdr.hat,color=method))+facet_wrap(~true.resp.rate,ncol=5)+theme_bw()+geom_abline(1,lty=3)+labs(title="Observed vs True FDR when \ntrue response rate varies but is assumed to be 100%")+theme(axis.text.x=element_text(angle=90))
+p9<-ggplot(comproc[[2]])+geom_line(aes(x=FPR.hat,y=TPR.hat,color=method))+facet_wrap(~true.resp.rate,ncol=5)+theme_bw()+labs(title="ROC curves when \ntrue response rate varies but is assumed to be 100%")+scale_x_continuous("False Positive Rate")+scale_y_continuous("True Positive Rate")+theme(axis.text.x=element_text(angle=90))
+
+library(gridExtra)
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/competeROC.pdf",width=10,height=3)
+p9
+dev.off()
+p9
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/competeROCfdr.pdf",width=10,height=3)
+p8
+dev.off()
+p8
 ```
@@
 ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+```
+
+Figure S3
+
+```{r eval=FALSE,echo=FALSE}
+p1<-ggplot(subset(sims$sim.ROC,Num==50000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+p2<-ggplot(subset(sims$sim.ROC,Num==10000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+p3<-ggplot(subset(sims$sim.FDR,Num==50000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+p4<-ggplot(subset(sims$sim.FDR,Num==10000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_ROC_50K.pdf",width=5,height=3)
+p1
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_ROC_10K.pdf",width=5,height=3)
+p2
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_FDR_50K.pdf",width=5,height=3)
+p3
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_FDR_10K.pdf",width=5,height=3)
+p4
+dev.off()
+p1
+p2
+p3
+p4
+
+#rerun one sided sims at 50K, more than 10 reps
+sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
+levs<-levels(sims.50K$sim.FDR$method)
+levs[1]<-"MIMOSA (one sided)"
+sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
+levs<-levels(sims.50K$sim.ROC$method)
+levs[2]<-"MIMOSA (one sided)"
+sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
+p5<-ggplot(subset(sims.50K$sim.FDR,Nobs!=200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+p6<-ggplot(subset(sims.50K$sim.ROC,Nobs!=200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_ROC_varyNobs.pdf",width=9,height=3)
+p6
+dev.off()
+p6
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_FDR_varyNobs.pdf",width=8,height=3)
+p5
+dev.off()
+p5
+```
+
+Figure of effect size vs probability of response
+
+```{r }
+pd<-do.call(rbind,lapply(hvtn.result.mcmc,function(x)pData(x)))
+pd<-cbind(pd,do.call(rbind,lapply(hvtn.result.mcmc,function(x)data.frame(response=MIMOSA:::fdr(x@z)<0.01,Prob.resp=x@z[,2],effect=do.call(c,lapply(x@result@p,function(x)diff(rev(x[,2]))))))))
+pd$VACCINE<-factor(pd$RX_CODE%in%c("T1","T2","T3","T4"))
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/volcanoplots.pdf",width=15,height=10)
+ggplot(subset(pd,TCELLSUB%in%"cd3+/cd4+"))+geom_point(aes(x=effect,y=Prob.resp,color=response,shape=VISITNO:VACCINE))+facet_wrap(TCELLSUB~CYTOKINE,scale="free_x")+theme(axis.text.x=element_text(angle=90))+scale_color_discrete("Response",labels=c("No","Yes"))+theme_bw()
+dev.off()
+ggplot(subset(pd,TCELLSUB%in%"cd3+/cd4+"))+geom_point(aes(x=effect,y=Prob.resp,color=response,shape=VISITNO:VACCINE))+facet_wrap(TCELLSUB~CYTOKINE,scale="free_x")+theme(axis.text.x=element_text(angle=90))+scale_color_discrete("Response",labels=c("No","Yes"))+theme_bw()
 ```
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, result='hide',fig.width=5,fig.height=2.5}
-df<-data.frame(tcellsub="CD4",treatment=gl(2,224/2,labels=c("stim","unstim")),melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.unstim),1)[,2])[,2]))
-df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,224/2,labels=c("stim","unstim")),proportion=melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.unstim),1)[,2])[,2])))
+require(scales)
+df<-data.frame(tcellsub="CD4",treatment=gl(2,nrow(hvtn.result.mcmc[[1]]@result@n.stim)/2,labels=c("stim","unstim")),melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.unstim),1)[,2])[,2]))
+df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,nrow(hvtn.result.mcmc[[30]]@result@n.unstim)/2,labels=c("stim","unstim")),proportion=melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.unstim),1)[,2])[,2])))
 ggplot(df)+geom_density(aes(x=value,fill=tcellsub),alpha=0.75,position="identity")+theme_bw()+scale_x_continuous(name="Cell Count",labels=scientific)+facet_wrap(~tcellsub,scales="free")

### 2013-06-25 MIMOSA/mimosafigures/mimosafigures-test/manuscriptFigures.Rmd
Update code to use bootstrap etc..

Load the data and fit the MIMOSA model
===============

```{r }
source("utility.R")
```{r echo=FALSE}
setup()
```

### Read and merge the hvtn data.

```{r cache=TRUE}
hvtndata<-readHVTNData()
hvtndata<-processHVTNData(hvtndata)
hvtndata<-constructEset(hvtndata)
```

### Read the fluidigm data.

```{r cache=FALSE}
fluidigm<-readFluidigmData()
```

Process it.

```{r cache=FALSE}
fluidigm<-processFluidigmData()
```

### Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC implementations.

```{r }
if(!file.exists("hvtnEMfit.rds")){
  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
}else{
  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
}
if(!file.exists("hvtnMCMCfit.rds")){
  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
}else{
  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
}
```

### Summarize the results. Do the likelihood ratio test, Fisher's test, and rank by log fold change for comparison.

```{r cache=TRUE,message=FALSE,warning=FALSE}
hvtn.summary.EM<-summarizeMIMOSAFit(hvtn.result.EM)
hvtn.summary.EM<-subset(hvtn.summary.EM,Method%in%"EM")
hvtn.summary.mcmc<-summarizeMIMOSAFit(hvtn.result.mcmc)
hvtn.summary.mcmc<-FisherFromSummary(hvtn.summary.mcmc)
hvtn.summary.mcmc<-LRT(hvtn.summary.mcmc)
hvtn.summary.mcmc<-FoldChangeRank(hvtn.summary.mcmc)
```

### Compute ROC and FDR curves and calculate the AUC for all the results.

```{r cache=TRUE}
ROC<-ComputeROCs(hvtn.summary.mcmc,hvtn.summary.EM)
FDR<-ComputeFDRs(hvtn.summary.mcmc,hvtn.summary.EM)
AUC<-ComputeAUCs(ROC)
```

### FDR and ROC plots of the ICS data.
--------
We see that MIMOSA performs as well or better than competing methods. The MCMC routine performs better than the EM routine for some data.

```{r cache=TRUE,fig.width=10,fig.height=8,fig.cap='The TNF expressing cytokine specific cells have high background as evidenced by the poor performance of all methods fit to TNF+ subsets of cells, with the exception of the triple-positive cell subset. As the number of simultaneously expressed cytokines being examined increases, the background decreases.'}
ROC$Method<-relevel(ROC$Method,"MIMOSA (EM)")
levs <-
  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
    "IL2+IFNg+", "IL2+TNF+", "TNF+")
ROC$CYTOKINE<-factor(ROC$CYTOKINE,labels=levs)
AUC$CYTOKINE<-factor(AUC$CYTOKINE,labels=levs)
rocplots<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=Method),size=3,data=AUC,show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
```{r fig.width=10,fig.height=8,fig.cap='The FDR is well controlled for cytokine subsets with low background. The TNF expressing subsets have high background and consequently positive assay results in the "Visit 2" samples, which are labelled as "true negatives" in this scheme. The result is observed as a large observed false discovery rate for all methods.'}
FDR$Method<-relevel(FDR$Method,"MIMOSA (EM)")
levs<-levels(FDR$CYTOKINE)
levs <-
  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
    "IL2+IFNg+", "IL2+TNF+", "TNF+")
FDR$CYTOKINE<-factor(FDR$CYTOKINE,labels=levs)
fdrplots<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
rocplots
fdrplots
```

### Figure 1 and S1.

```{r echo=FALSE,eval=FALSE}
suppressPackageStartupMessages(library(gridExtra))
f1levs<-levs[c(8,11,13,17,15,16,10,7)]
sflevs<-setdiff(levs,f1levs)
AUC$y[AUC$Method%in%"Fisher"]<-0.4
AUC$y[AUC$Method%in%"MIMOSA (EM)"]<-0.5
AUC$y[AUC$Method%in%"MIMOSA (MCMC)"]<-0.3

fnames<-c("revised_2.pdf","revised_1.pdf","revised_15.pdf","revised_3.pdf","revised_5.pdf","revised_6.pdf","revised_7.pdf","revised_8.pdf")
for(i in seq_along(f1levs)){
  fi.1<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"&CYTOKINE%in%f1levs[i]))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=factor(Method)),data=subset(AUC,CYTOKINE%in%f1levs[i]),show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1)) 
  fi.2<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")&CYTOKINE%in%f1levs[i]))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
  pdf(file=sprintf("/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/%s",fnames[i]),width=10,height=3)
  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
  dev.off()
  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
}
```

### Process the fluidigm data to construct an expression set for MIMOSA.
---------

```{r warning=FALSE,cache=TRUE}
fl.bystim<-lapply(fluidigm,exprs)
cd.bystim<-lapply(fluidigm,cData)
fl<-lapply(names(fl.bystim),function(n){
  m<-melt(cbind(fl.bystim[[n]],cd.bystim[[n]]),id=c("Stim.Agent","Stim.Condition","Time.of.Stim","Sero.Status","Patient.ID","Chip.Number","Well","Number.of.Cells"))
  ddply(m,.(Stim.Condition,Patient.ID,variable),summarize,N=length(value),n=sum(value>0))
})
fl<-lapply(fl,function(x)setnames(x,"variable","Gene"))
Esets.fluidigm<-lapply(fl,function(x)ConstructMIMOSAExpressionSet(x,reference=Stim.Condition%in%"Unstim",measure.columns=c("N","n"),other.annotations=c("Gene","Patient.ID","Stim.Condition"),default.cast.formula=component~Patient.ID+Gene,.variables=.(Patient.ID,Gene),featureCols=1,ref.append.replace="_REF"))
nms<-names(fl.bystim)
Esets.fluidigm<-lapply(1:length(Esets.fluidigm),function(x){pData(Esets.fluidigm[[x]])<-cbind(pData(Esets.fluidigm[[x]]),Stim=nms[x]);Esets.fluidigm[[x]]})
Eset.combined<-combine(Esets.fluidigm[[1]],Esets.fluidigm[[2]],Esets.fluidigm[[3]],Esets.fluidigm[[4]])
```

### Fit MIMOSA by gene, grouping by stimulation and for all samples together.
----

```{r fluidigm.fit, cache=TRUE}
if(!file.exists("fluidigm.fits.by.stim.allgenes.rds")){
  fluidigm.fits.by.stim.allgenes<-lapply(list(Eset.combined),function(x)MIMOSA(N+n~Stim+Patient.ID|Gene,x,ref=RefTreat%in%"Reference",subset=RefTreat%in%"Treatment",method="mcmc",EXPRATE=1e-4,iter=105000,burn=5000,pXi=1,alternative="not equal",run.parallel=TRUE,getP=TRUE))
  fluidigm.fits.by.stim.bygene<-mclapply(Esets.fluidigm,function(x)MIMOSA(N+n~Stim+Patient.ID|Gene,x,ref=RefTreat%in%"Reference",subset=RefTreat%in%"Treatment",method="mcmc",EXPRATE=1e-4,iter=105000,burn=50000,pXi=1,alternative="not equal",getP=TRUE,run.parallel=TRUE))
  saveRDS(fluidigm.fits.by.stim.allgenes,file="fluidigm.fits.by.stim.allgenes.rds")
  saveRDS(fluidigm.fits.by.stim.bygene,file="fluidigm.fits.by.stim.bygene.rds")
}else{
  fluidigm.fits.by.stim.allgenes<-readRDS("fluidigm.fits.by.stim.allgenes.rds")
  fluidigm.fits.by.stim.bygene<-readRDS("fluidigm.fits.by.stim.bygene.rds")
}
```

### Heatmaps for Fluidigm data fit to all stimulations simultaneously.
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.probs.bygene,cache=TRUE}
foo<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)x@z[,2]))
props<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))
S<-sign(props)
S[S==0]<-1
colnames(S)<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])[,c("Patient.ID")]
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
#heatmap.2((S*(foo))[apply(foo,1,function(x)sum(x>0.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
```

### Figure 2 B, (fit of all samples simultaneously)

```{r eval=FALSE,echo=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmCombinedFit.pdf")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
dev.off()
```

Heatmap of posterior differences in proportions from simultaneous fit (not shown).
----
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).
-When fitting the data to all samples simultaneously, we still observe clustering of samples by stimulation (i.e. CMV stimulated samples cluster together, and HIV stimulated samples cluster together).

```{r heatmap.props.bygene,cache=TRUE,include=FALSE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
```

### Heatmaps for Fluidigm data fit per stimulation.
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.probs.bystim.bygene,cache=TRUE}
foo<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)x@z[,2]))))
props<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene,function(x)do.call(rbind,lapply(x,function(x)do.call(c,lapply(x@result@p,function(x)-diff(x[,2])))))))
S<-sign(props)
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])[,c("Patient.ID")])))
S[S==0]<-1
stim<-as.numeric(factor(do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Stim)))))
colnames(S)<-do.call(c,lapply(fluidigm.fits.by.stim.bygene,function(x)as.character(pData(x[[1]])$Patient.ID)))
rownames(S)<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene[[1]],function(x)as.character(unique(pData(x)$Gene))))
#h<-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
h<-heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
hmaporder<-h$colInd
geneinds<-apply(foo,1,function(x)sum(x>.6)>=1)
```

### Figure 2 A (fit per stimulation)
Fit MIMOSA per stimulation.

```{r eval=FALSE,echo=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Fluidigm_plot1.pdf")
foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
dev.off()
```

Heatmap of posterior differences in proportions for fit of samples by stimulation and by gene (not shown).
---
-Red and orange are CMV (pp65). Yellow and green are HIV (gag and nef).

```{r heatmap.props.bystim.bygene,cache=TRUE,include=FALSE}
colnames(props)<-colnames(S)
rownames(props)<-rownames(S)
heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
```

### Fisher's exact test and heatmaps of -log10(q-values) from Fisher's test.
---
There is little clustering by stimulation evident from testing for differential expression with Fisher's exact test.

```{r cache=TRUE}
fisher<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
  p<-vector('numeric',nrow(x@result@n.stim))
  for(i in 1:nrow(x@result@n.stim)){
    p[i]<-fisher.test(as.matrix(rbind(x@result@n.stim[i,],x@result@n.unstim[i,])),alternative="two.sided")$p.value
  }
  p
}))
adj.fisher<-matrix(p.adjust(fisher,"fdr"),nrow=96)
S<-sign(lapply(fluidigm.fits.by.stim.allgenes,function(x)do.call(rbind,lapply(x,function(x)(prop.table(as.matrix(x@result@n.stim),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2])))[[1]])
S[S==0]<-1
colnames(fisher)<-gsub("_B3GAT1_Treatment","",colnames(S))
rownames(fisher)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
#heatmap.2(fisher*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
colnames(adj.fisher)<-colnames(fisher)
rownames(adj.fisher)<-rownames(fisher)
heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
```

### Figure 2 D (q-values from Fisher's test)

```{r eval=FALSE,echo=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmFisher.pdf")
colnames(adj.fisher)<-colnames(fisher)
rownames(adj.fisher)<-rownames(fisher)
heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
dev.off()

```{r eval=TRUE,echo=FALSE}
emprop<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
  p<-vector('numeric',nrow(x@result@n.stim))
  p<-(prop.table(as.matrix((x@result@n.stim)),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2]
}))
rownames(emprop)<-rownames(fisher)
colnames(emprop)<-colnames(fisher)

```{r echo=FALSE,warning=FALSE}
heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
```

### Figure 2C empirical proportions.

```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmEmpiricalProps.pdf")
heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
dev.off()
```

### Multivariate MIMOSA fit to fluidigm data, looking at interactions of genes.
---------
There is a difference when comparing stimulations and looking at the multivariate vector of expression for GZMK and CCR7 (i.e [+/+,+/-,-/+,--]), but no difference is detected when we look at expression of GZMK | CCR7, since one cell subset increases while the other decreases upon stimultion.

```{r cache=TRUE}
fl.comb<-combine(fluidigm[[1]],fluidigm[[2]],fluidigm[[3]],fluidigm[[4]])
melted<-melt(fl.comb)
require(glmnet)
require(Matrix)
melted$Et.bin<-as.numeric(as.logical(melted$Et))
#setnames(melted,"__wellKey","wellKey")
melted<-within(melted,group<-Patient.ID:Stim.Agent)
melted<-cast(melt(melted,id=c("group","Stim.Condition","primerid","wellKey"),measure="Et.bin"),group+Stim.Condition+wellKey~primerid)

# interact.scan<-combn(colnames(melted)[4:99],2,function(x){
#   g1<-x[1];g2<-x[2]
#   gg<-paste(g1,g2,sep=":")
#   bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
#   bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
#   setnames(bar,"foo",gg)
#   bar[,5]<-factor(bar[,5],levels=c("0:0","0:1","1:0","1:1"))
#   C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
#   expected<-c("0:0","0:1","1:0","1:1")
#   unused<-setdiff(expected,colnames(C))
#   newframe<-NULL
#   for(i in seq_along(expected)){
#     if(expected[i]%in%unused){
#       newframe<-cbind(newframe,rep(0,nrow(C)))
#     }else{
#       newframe<-cbind(newframe,C[,expected[i]])
#     }
#   }
#   colnames(newframe)<-expected
#   C<-cbind(C[,1:2],newframe)
#   CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])
#   foo<-lapply(CC,function(x)prop.table(as.matrix(x),1))
#   psum<-mean(rowSums(foo[[1]][,2:4])-rowSums(foo[[2]][,2:4]))
#   pij<-colMeans(foo[[1]][,2:4]-foo[[2]][,2:4])
#   res<-c(psum,pij)
#   res
# })
#cache("interact.scan")

#genes<-combn(colnames(melted)[6:101],2)[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
#interact.scan[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
#ggg<-genes[,6]
#g1<-ggg[1];g2<-ggg[2] #CCR7 and GZMK
g1<-"CCR7";g2<-"GZMK"
gg<-paste(g1,g2,sep=":")
bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
setnames(bar,"foo",gg)
C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])

toplot<-data.frame(rbind(data.frame(CC$n.stim),data.frame(CC$n.unstim)),stim=c(rep("Stim",nrow(CC$n.stim)),rep("Unstim",nrow(CC$n.unstim))))
toplot$id<-rep(unlist(lapply(strsplit(as.character(subset(C,Stim.Condition%in%"Unstim")[,1]),":"),function(x)x[1]),use.names=FALSE),2)
colnames(toplot)[1:4]<-gsub("X","",paste(g1,g2,colnames(toplot)[1:4],sep=":"))
toplot<-melt(toplot,id=c("id","stim"))
p1<-ggplot(toplot)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray")

DD<-CC
DD[[1]]<-cbind(CC[[1]][,1],rowSums(CC[[1]][,2:4]))
DD[[2]]<-cbind(CC[[2]][,1],rowSums(CC[[2]][,2:4]))
toplot2<-data.frame(rbind(data.frame(DD$n.stim),data.frame(DD$n.unstim)),stim=c(rep("Stim",nrow(DD$n.stim)),rep("Unstim",nrow(DD$n.unstim))))
toplot2$id<-rep(unlist(lapply(strsplit(as.character(subset(C,Stim.Condition%in%"Unstim")[,1]),":"),function(x)x[1]),use.names=FALSE),2)
colnames(toplot2)[1:2]<-gsub("X","",paste(g1,g2,colnames(toplot2)[1:2],sep=":"))
toplot2<-melt(toplot2,id=c("id","stim"))
p2<-ggplot(toplot2)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous(na.value="gray")

```{r fitmulti,echo=FALSE,message=FALSE,warning=FALSE,error=FALSE,results='hide'}
flres<-.fitMCMC(data=CC,inits=MIMOSA:::MDMix(CC,initonly=TRUE),EXPRATE=1e-2,pXi=1,alternative="not equal")
flsum<-.fitMCMC(data=DD,inits=MIMOSA:::MDMix(DD,initonly=TRUE),EXPRATE=1e-2,pXi=1,alternative="not equal")
sapply(list.files(pattern=c("\\.dat")),unlink)
```

### Figure 4. Heatmaps of counts for all combinations of CCR7 and GZMK vs marginal counts.
Samples in red show no difference by MIMOSA.

```{r eval=TRUE,echo=FALSE}
library(gridExtra)
missed<-!MIMOSA:::fdr(flres$z)<0.05
toplot$variable<-factor(toplot$variable,labels=c("CCR7-GZMK-","CCR7-GZMK+","CCR7+GZMK-","CCR7+GZMK+"))
toplot2$variable<-factor(toplot2$variable,labels=c("CCR7-GZMK-","CCR7+ or GZMK+"))
p1<-ggplot(toplot)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous("log2 count",na.value="gray",breaks=round(seq(0,6,l=10)))+theme(axis.text.y=element_text(color=c("black","red")[missed+1]))
p2<-ggplot(toplot2)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90,hjust=1))+facet_wrap(~stim)+scale_fill_continuous("log2 count",na.value="gray",breaks=round(seq(0,6,l=10)))+theme(axis.text.y=element_text(color=c("black","red")[2]))
grid.arrange(p1,ncol=1)
grid.arrange(p2,ncol=1)

```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/revised_Fluidigm_Multivariate_CCR7_GZMK.pdf",width=6,height=4)
grid.arrange(p1,ncol=1)
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/revised_Fluidigm_Marginal_CCR7_GZMK.pdf",width=6,height=4)
grid.arrange(p2,ncol=1)
dev.off()

A<-cbind(melt(data.frame(matrix(prop.table(as.matrix(CC$n.stim),1)-prop.table(as.matrix(CC$n.unstim),1),ncol=4))),flres$z[,2],1:length(flres$z[,2]))
B<-cbind(melt(data.frame(matrix(prop.table(as.matrix(DD$n.stim),1)-prop.table(as.matrix(DD$n.unstim),1),ncol=2))),flsum$z[,2],1:length(flres$z[,2]))
colnames(A)<-c("Component","Effect Size","Pr(response)","ID")
colnames(B)<-c("Component","Effect Size","Pr(response)","ID")
A$Component<-factor(A$Component,labels=c("-/-","-/+","+/-","+/+"))
B$Component<-factor(B$Component,labels=c("-","+"))
A<-rbind(A,B)
ggplot(A)+geom_point(aes(x=`Effect Size`,y=`Pr(response)`,shape=factor(ID)))+facet_wrap(~Component)+geom_vline(aes(x=0),lty=3)+scale_shape_manual(values=1:16,name="Subject")
```

Simulations
==========
One sided
------
We'll use the model fit from IFNg+ for ENV-1-PTEG based on the ROC and FDR plots
Note that MIMOSA has greater sensitivity and specificity than Fisher's exact test or the Likelihood ratio test. When there are 5 observations, the methods being to perform equivalently.

```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5,include=FALSE}
if(!file.exists("cache/sims.10K.RData")){
sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
sims.25K<-OneSidedSims(hvtn.result.mcmc,N=25000)
sims.10K<-OneSidedSims(hvtn.result.mcmc,N=10000)
cache('sims.50K')
cache('sims.25K')
cache('sims.10K')
}else{
load("cache/sims.10K.RData")
load("cache/sims.25K.RData")
load("cache/sims.50K.RData")
}
ggplot(sims.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
ggplot(sims.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
ggplot(sims.10K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims.10K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
```

Two-sided
--------

```{r two.sided.sims, cache=FALSE, echo=FALSE,fig.width=5,fig.height=5,include=FALSE}
if(!file.exists("cache/sims2.10K.RData")){
sims2.50K<-TwoSidedSims(hvtn.result.mcmc,N=50000)
sims2.25K<-TwoSidedSims(hvtn.result.mcmc,N=25000)
sims2.10K<-TwoSidedSims(hvtn.result.mcmc,N=10000)
cache('sims2.50K')
cache('sims2.25K')
cache('sims2.10K')
}else{
load("cache/sims2.10K.RData")
load("cache/sims2.25K.RData")
load("cache/sims2.50K.RData")
}
ggplot(sims2.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims2.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
ggplot(sims2.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
ggplot(sims2.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
smlcnts.roc<-ggplot(subset(sims2.10K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations (small counts)")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
smlcnts.fdr<-ggplot(subset(sims2.10K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations (small counts)")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
```

### Figure 3: One-sided and two-sided simulations.

```{r eval=TRUE,echo=FALSE}
levs<-levels(sims.50K$sim.ROC$method)
levs[2]<-c("MIMOSA (one sided)")
sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
p1.1<-ggplot(subset(sims.50K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
levs<-levels(sims2.50K$sim.ROC$method)
levs[2]<-c("MIMOSA (two sided)")
sims2.50K$sim.ROC$method<-factor(sims2.50K$sim.ROC$method,labels=levs)
p2.1<-ggplot(subset(sims2.50K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (two sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))

levs<-levels(sims.50K$sim.FDR$method)
levs[1]<-"MIMOSA (one sided)"
sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
p1.2<-ggplot(subset(sims.50K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
levs<-levels(sims2.50K$sim.FDR$method)
levs[1]<-c("MIMOSA (two sided)")
sims2.50K$sim.FDR$method<-factor(sims2.50K$sim.FDR$method,labels=levs)
p2.2<-ggplot(subset(sims2.50K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (two sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)

```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_ROC_50K.pdf",width=6,height=4)
p1.1
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_TwoSided_ROC_50K.pdf",width=6,height=4)
p2.1
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_TwoSided_FDR_50K.pdf",width=6,height=4)
p2.2
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_FDR_50K.pdf",width=6,height=4)
p1.2
dev.off()

```{r echo=FALSE}
p1.1
p2.1
p2.2
p1.2
```

### Multivariate Simulations
-----
Average ROC for 10 multivariate simulations with effect size of 2.5x10-3 and -2.5x10-3 in two of eight components (N=4743 events).

```{r cache=TRUE,results='hide',message=FALSE,warning=FALSE}
set.seed(101)
as<-c(1,1,1,1,3.5,1,1)
au<-c(1,1,1,1,1,3.5,1)
as<-c(1000-sum(as),as)*100
au<-c(1000-sum(au),au)*100
if(!file.exists("mvsims.rds")){
mdsim<-sapply(1:10,function(i){
simd<-MIMOSA:::simMD(alpha.s=as,alpha.u=au,N=100,w=0.5,n=3.5,alternative="not equal")
foo<-.fitMCMC(simd,inits=MIMOSA:::MDMix(simd,initonly=TRUE))
foo
})
saveRDS(mdsim,file="mvsims.rds")
}else{
  mdsim<-readRDS("mvsims.rds")
}
fisher.p<-vector('list',ncol(mdsim))
for(i in 1:ncol(mdsim)){    
  fisher.p[[i]]<-rep(NA,nrow(mdsim[,i]$n.stim))
for(j in 1:nrow(mdsim[,i]$n.stim)){
  fisher.p[[i]][j]<-  fisher.test(cbind(mdsim[,i]$n.stim[j,],mdsim[,i]$n.unstim[j,]),alternative="two.sided",simulate.p.value=TRUE)$p
}
}

ROCfun<-function(x,truth){
  o<-order(x,decreasing=FALSE)
  cbind(tpr=cumsum(truth[o])/sum(truth),fpr=cumsum(!truth[o])/sum(!truth))
}

fisher.p.adj<-lapply(fisher.p,function(x)p.adjust(x,"fdr"))
fisher.roc<-do.call(rbind,by(cbind(do.call(rbind,lapply(fisher.p.adj,function(i)ROCfun(i,gl(2,50)==2)))),rep(gl(100,1),10),function(x)colMeans(x)))
mimosa.adj<-lapply(1:ncol(mdsim),function(i)MIMOSA:::fdr(mdsim[,i]$z))
MIMOSA.roc<-do.call(rbind,by(do.call(rbind,lapply(mimosa.adj,function(i)ROCfun(i,gl(2,50)==2))),rep(gl(100,1),10),function(x)colMeans(x)))
foo<-data.frame(as.data.frame(rbind(fisher.roc,MIMOSA.roc)),Method=gl(2,100,labels=c("Fisher","MIMOSA")))
sz<-element_text(size=18)
p1<-ggplot(foo)+geom_line(aes(x=fpr,y=tpr,col=Method))+scale_y_continuous("TPR")+scale_x_continuous("FPR")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)
#FDR
mim.fdr<-as.data.frame(cbind(do.call(rbind,lapply(mimosa.adj,function(x)fdrCurves(x,gl(2,50)==2))),rep=gl(100,1)))
mim.fdr<-ddply(mim.fdr,.(rep),summarize,true.fdr.hat=mean(true.fdr),fdr.hat=mean(fdr))
fisher.fdr<-as.data.frame(cbind(do.call(rbind,lapply(fisher.p.adj,function(x)fdrCurves(x,gl(2,50)==2))),rep=gl(100,1)))
fisher.fdr<-ddply(fisher.fdr,.(rep),summarize,true.fdr.hat=mean(true.fdr),fdr.hat=mean(fdr))
foo<-data.frame(rbind(fisher.fdr,mim.fdr),Method=gl(2,100,labels=c("Fisher","MIMOSA")))
p2<-ggplot(foo)+geom_line(aes(y=true.fdr.hat,x=fdr.hat,col=Method))+scale_y_continuous("observed fdr")+scale_x_continuous("nominal fdr")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)+geom_abline(1,lty=3)
grid.arrange(p1,p2)
```

Competing ROC curves to evaluate the effect of inaccuracies in the true positives and false positives.
-----
Simulate from models with 20,35,50,65,80 percent response rates but treat the ROC evaluation as 100% response.

```{r results='hide',cache=TRUE}
comproc<-CompareROC(hvtn.result.mcmc)
comproc[[1]]$true.resp.rate<-as.numeric(as.character(comproc[[1]]$q))*2
comproc[[2]]$true.resp.rate<-as.numeric(as.character(comproc[[2]]$q))*2

p8<-ggplot(comproc[[1]])+geom_line(aes(x=fdr.hat,y=true.fdr.hat,color=method))+facet_wrap(~true.resp.rate,ncol=5)+theme_bw()+geom_abline(1,lty=3)+labs(title="Observed vs True FDR when \ntrue response rate varies but is assumed to be 100%")+theme(axis.text.x=element_text(angle=90))
p9<-ggplot(comproc[[2]])+geom_line(aes(x=FPR.hat,y=TPR.hat,color=method))+facet_wrap(~true.resp.rate,ncol=5)+theme_bw()+labs(title="ROC curves when \ntrue response rate varies but is assumed to be 100%")+scale_x_continuous("False Positive Rate")+scale_y_continuous("True Positive Rate")+theme(axis.text.x=element_text(angle=90))

library(gridExtra)
```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/competeROC.pdf",width=10,height=3)
p8
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/competeROCfdr.pdf",width=10,height=3)
p9
dev.off()

```{r echo=FALSE,fig.width=12,fig.height=6}
p8
p9
```

Supplementary Figure 2
------
Histogram of the empirical proportions of unstimulated cells and overlaid posterior densities of the beta distribution with $\alpha_s$ and $\alpha_u$ estimated from the data for ENV-1-PTEG stimulated, IFN-$\gamma$+ CD4+ T-cells, demonstrating that the assumption of a common distribution for piu across subjects is reasonable.

```{r supp.fig.2,message=FALSE,echo=FALSE}
res<-hvtn.result.mcmc[[7]]
ggplot(data.frame(p.hat=prop.table(as.matrix(res@result@n.unstim),1)[,2]))+geom_histogram(aes(x=p.hat,y=..density..),col="black",fill="gray")+theme_bw()+scale_x_continuous(name=expression(hat(p)[u]))+geom_line(data=data.frame(x=seq(0,3e-4,l=10000),y=dbeta(seq(0,3e-4,l=10000),res@result@params[2,4],res@result@params[2,3]),stim="ENV-1-PTEG"),aes(x=x,y=y),col="blue",lwd=2,lty=3)+facet_wrap(~stim)+theme(strip.text.x=element_text(size=24),axis.text.x=element_text(size=18),axis.text.y=element_text(size=18),axis.title.x=element_text(size=24),axis.title.y=element_text(size=24))
```

Unconstrained MIMOSA fit to data violating model assumptions
-----

```{r violated.assumptions, cache=FALSE,echo=FALSE,fig.width=20,fig.height=10}
sims<-TwoSidedSimsTruncated(hvtn.result.mcmc)
violated.roc<-ggplot(subset(sims$sim.ROC,Num==50000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations Violating Model Assumptions")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
violated.fdr<-ggplot(subset(sims$sim.FDR,Num==50000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations Violating Model Assumptions")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
grid.arrange(violated.roc,violated.fdr,smlcnts.roc,smlcnts.fdr)
```

### Figure S3

```{r eval=TRUE,echo=FALSE}
p1<-ggplot(subset(sims$sim.ROC,Num==50000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
p2<-ggplot(subset(sims$sim.ROC,Num==10000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
p3<-ggplot(subset(sims$sim.FDR,Num==50000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
p4<-ggplot(subset(sims$sim.FDR,Num==10000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)

```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_ROC_50K.pdf",width=5,height=3)
p1
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_ROC_10K.pdf",width=5,height=3)
p2
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_FDR_50K.pdf",width=5,height=3)
p3
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_FDR_10K.pdf",width=5,height=3)
p4
dev.off()

#rerun one sided sims at 50K, more than 10 reps
sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
levs<-levels(sims.50K$sim.FDR$method)
levs[1]<-"MIMOSA (one sided)"
sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
levs<-levels(sims.50K$sim.ROC$method)
levs[2]<-"MIMOSA (one sided)"
sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
p5<-ggplot(subset(sims.50K$sim.FDR,Nobs!=200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
p6<-ggplot(subset(sims.50K$sim.ROC,Nobs!=200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_ROC_varyNobs.pdf",width=9,height=3)
p6
dev.off()
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_FDR_varyNobs.pdf",width=8,height=3)
p5
dev.off()
```

### One-sided simulations, 50K events. N=20,50,100

```{r echo=FALSE}
p6
p5
```

### Figure of effect size vs probability of response

```{r }
pd<-do.call(rbind,lapply(hvtn.result.mcmc,function(x)pData(x)))
pd<-cbind(pd,do.call(rbind,lapply(hvtn.result.mcmc,function(x)data.frame(response=MIMOSA:::fdr(x@z)<0.01,Prob.resp=x@z[,2],effect=do.call(c,lapply(x@result@p,function(x)diff(rev(x[,2]))))))))
pd$VACCINE<-factor(pd$RX_CODE%in%c("T1","T2","T3","T4"))
```{r eval=FALSE}
pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/volcanoplots.pdf",width=15,height=10)
ggplot(subset(pd,TCELLSUB%in%"cd3+/cd4+"))+geom_point(aes(x=effect,y=Prob.resp,color=response,shape=VISITNO:VACCINE))+facet_wrap(TCELLSUB~CYTOKINE,scale="free_x")+theme(axis.text.x=element_text(angle=90))+scale_color_discrete("Response",labels=c("No","Yes"))+theme_bw()
dev.off()
```{r echo=FALSE,fig.width=15}
ggplot(subset(pd,TCELLSUB%in%"cd3+/cd4+"))+geom_point(aes(x=effect,y=Prob.resp,color=response,shape=VISITNO:VACCINE))+facet_wrap(TCELLSUB~CYTOKINE,scale="free_x")+theme(axis.text.x=element_text(angle=90))+scale_color_discrete("Response",labels=c("No","Yes"))+theme_bw()
```

Distribution of total CD4 and CD8 T-cell counts?
-----

```{r echo=FALSE,message=FALSE,warning=FALSE,result='hide',fig.width=5,fig.height=2.5}
suppressPackageStartupMessages(require(scales))
df<-data.frame(tcellsub="CD4",treatment=gl(2,205/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[1]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[1]]@result@n.unstim)[,1]))
df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,224/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[30]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[30]]@result@n.unstim)[,1])))
ggplot(df)+geom_density(aes(x=value,fill=tcellsub),alpha=0.75,position="identity")+theme_bw()+scale_x_continuous(name="Cell Count",labels=comma)
```

Median cell counts for T-cell subsets.
-----

```{r results='asis',message=FALSE,echo=FALSE}
suppressPackageStartupMessages(require(plyr))
suppressPackageStartupMessages(require(reshape))
suppressPackageStartupMessages(require(xtable))
d<-cast(melt(ddply(df,.(tcellsub),summarize,median=round(quantile(value,c(0.5)))),measure="median"),~tcellsub)
print(xtable(d[,-1L],display=c("fg","fg","fg")),type="html",include.rownames=FALSE)
```

Distribution of the proportions for the HVTN data
----

```{r echo=FALSE, message=FALSE, warning=FALSE, result='hide',fig.width=5,fig.height=2.5}
require(scales)
df<-data.frame(tcellsub="CD4",treatment=gl(2,nrow(hvtn.result.mcmc[[1]]@result@n.stim)/2,labels=c("stim","unstim")),melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[1]]@result@n.unstim),1)[,2])[,2]))
df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,nrow(hvtn.result.mcmc[[30]]@result@n.unstim)/2,labels=c("stim","unstim")),proportion=melt(data.frame(stim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.stim),1)[,2],unstim=prop.table(as.matrix(hvtn.result.mcmc[[30]]@result@n.unstim),1)[,2])[,2])))
ggplot(df)+geom_density(aes(x=value,fill=tcellsub),alpha=0.75,position="identity")+theme_bw()+scale_x_continuous(name="Cell Count",labels=scientific)+facet_wrap(~tcellsub,scales="free")
```

Median cell proportions for ICS data.
----

```{r results='asis',message=FALSE,echo=FALSE}
d<-cast(melt(df,measure='value'),~tcellsub,fun.aggregate=function(x)mean(x))[,-1L]
print(xtable(d,display=c("fg","fg","fg")),type="html",include.rownames=FALSE)
```

### 2013-06-25 MIMOSA/mimosafigures/compile.R

@@ -1,8 +1,5 @@
 library(knitr)
+library(knitrBootstrap)
 knitr::spin("manuscriptFigures.R",FALSE)
-knit2html("manuscriptFigures.Rmd")
-#system("pandoc --jsmath manuscriptFigures.md>manuscriptFigures.html")
-system("(echo '<head><script src=\"https://raw.github.com/Polychart/polychart2/develop/polychart2.standalone.js\"></script><script src=\"\"></script></head>'; cat manuscriptFigures.html)>manuscriptFigures2.html")
-system("sed 's/&quot;/\"/g' < manuscriptFigures2.html>manuscriptFigures.html")
-system("cp manuscriptFigures.html /Users/gfinak/Dropbox/Public/MIMOSA/manuscriptFigures.html")
-system("cp -r figure /Users/gfinak/Dropbox/Public/MIMOSA/")
+knit_bootstrap("manuscriptFigures.Rmd")
+system("cp manuscriptFigures.html /home/gfinak/Dropbox/Public/MIMOSA/manuscriptFigures.html")

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.R

@@ -89,6 +89,7 @@ for(i in seq_along(f1levs)){
   pdf(file=sprintf("/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/%s",fnames[i]),width=10,height=3)
   grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
   dev.off()
+  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
  }
 
 #'Process the fluidigm data to construct an expression set for MIMOSA
@@ -139,6 +140,7 @@ pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmCombi
 foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
 heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 dev.off()
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 
 #'Heatmap of posterior differences of proportions fit by gene for all samples.
 #'----
@@ -172,6 +174,7 @@ pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Fluidigm_plot
 foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
[…]

### 2013-06-25 MIMOSA/mimosafigures/mimosafigures-test/manuscriptFigures.R

@@ -1,0 +1,602 @@
+#'Load the data and fit the MIMOSA model
+#'===============
+source("utility.R")
+#+ echo=FALSE
+setup()
+
+#'### Read and merge the hvtn data.
+#+ cache=TRUE
+hvtndata<-readHVTNData()
+hvtndata<-processHVTNData(hvtndata)
+hvtndata<-constructEset(hvtndata)
+
+#'### Read the fluidigm data.
+#+ cache=FALSE
+fluidigm<-readFluidigmData()
+
+#'Process it.
+#+ cache=FALSE
+fluidigm<-processFluidigmData()
+
+ 
+#'### Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC implementations.
+if(!file.exists("hvtnEMfit.rds")){
+  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
+  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
+}else{
+  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
+}
+if(!file.exists("hvtnMCMCfit.rds")){
+  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
+  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
+}else{
+  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
+}
+
+#'### Summarize the results. Do the likelihood ratio test, Fisher's test, and rank by log fold change for comparison.
+#+ cache=TRUE,message=FALSE,warning=FALSE
[…]

### 2013-06-25 MIMOSA/mimosafigures/mimosafigures-test/utility.R

@@ -1,0 +1,667 @@
+#'setup the environment
+#'
+#'load MIMOSA, parallel, filehash,gdata,pracma,reshape and ggplot2
+setup<-function(){
+  suppressPackageStartupMessages({require(MIMOSA)
+                                  require(multicore)
+                                  require(parallel)
+                                  require(filehash)
+                                  require(gdata)
+                                  require(pracma)
+                                  require(ggplot2)
+                                  require(reshape)
+                                  require(data.table)
+                                  require(SingleCellAssay)
+                                  require(gplots)
+                                  
+  })
+}
+
+#'construct an expected vs observed fdr curve
+#'
+#'Expected vs observed FDr taking the expected fdr and the truth.
+fdrCurves<-function(fdr,truth){
+  truth<-truth[order(fdr)]
+  true.fdr<-cumsum(1-truth)/1:length(truth)
+  fdr<-sort(fdr)
+  cbind(fdr,true.fdr)
+}
+
+#'Construct and ROC curve
+#'
+#'Given the p-value and the truth, construct an ROC curve.
+rocCurves<-function(p,truth){
+  #s<-seq(0,1,l=1000)
+  #table<-t(sapply(s,function(th){
+  #  prop.table(table(test=factor(p<=th,levels=c("FALSE","TRUE")),truth=truth),margin=2)["TRUE",]
+  #}))
+  #colnames(table)<-c("FPR","TPR")
+  #table
+  o<-order(p,decreasing=FALSE)
+  intervals<-lapply(sapply(sort(unique(p)),function(x)which(as.numeric(p[o])%in%x)),max)
[…]

### 2013-06-25 MIMOSA/mimosafigures/compile.R
Mucked up the directory structure. Try again

@@ -1,5 +1,0 @@
-library(knitr)
-library(knitrBootstrap)
-knitr::spin("manuscriptFigures.R",FALSE)
-knit_bootstrap("manuscriptFigures.Rmd")
-system("cp manuscriptFigures.html /home/gfinak/Dropbox/Public/MIMOSA/manuscriptFigures.html")

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.R

@@ -1,579 +1,0 @@
-#'Load the data and fit the MIMOSA model
-#'===============
-source("utility.R")
-#+ echo=FALSE
-setup()
-
-#'Read and merge the hvtn data. Cache it.
-#+ cache=TRUE
-hvtndata<-readHVTNData()
-hvtndata<-processHVTNData(hvtndata)
-hvtndata<-constructEset(hvtndata)
-
-#'read the fluidigm data
-#+ cache=FALSE
-fluidigm<-readFluidigmData()
-
-#'Process it
-#+ cache=FALSE
-fluidigm<-processFluidigmData()
-
-
-#'Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC
-if(!file.exists("hvtnEMfit.rds")){
-  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
-  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
-}else{
-  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
-}
-if(!file.exists("hvtnMCMCfit.rds")){
-  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
-  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
-}else{
-  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
-}
-
-#'drop the stuff fit via mcmc and summarize, fit LRT, Fisher's and rank by log fold change
-#+ cache=TRUE
[…]

### 2013-06-25 MIMOSA/mimosafigures/mimosafigures-test/compile.R

@@ -1,8 +1,0 @@
-library(knitr)
-knitr::spin("manuscriptFigures.R",FALSE)
-knit2html("manuscriptFigures.Rmd")
-#system("pandoc --jsmath manuscriptFigures.md>manuscriptFigures.html")
-system("(echo '<head><script src=\"https://raw.github.com/Polychart/polychart2/develop/polychart2.standalone.js\"></script><script src=\"\"></script></head>'; cat manuscriptFigures.html)>manuscriptFigures2.html")
-system("sed 's/&quot;/\"/g' < manuscriptFigures2.html>manuscriptFigures.html")
-system("cp manuscriptFigures.html /Users/gfinak/Dropbox/Public/MIMOSA/manuscriptFigures.html")
-system("cp -r figure /Users/gfinak/Dropbox/Public/MIMOSA/")

### 2013-06-25 MIMOSA/mimosafigures/mimosafigures-test/manuscriptFigures.R

@@ -1,602 +1,0 @@
-#'Load the data and fit the MIMOSA model
-#'===============
-source("utility.R")
-#+ echo=FALSE
-setup()
-
-#'### Read and merge the hvtn data.
-#+ cache=TRUE
-hvtndata<-readHVTNData()
-hvtndata<-processHVTNData(hvtndata)
-hvtndata<-constructEset(hvtndata)
-
-#'### Read the fluidigm data.
-#+ cache=FALSE
-fluidigm<-readFluidigmData()
-
-#'Process it.
-#+ cache=FALSE
-fluidigm<-processFluidigmData()
-
- 
-#'### Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC implementations.
-if(!file.exists("hvtnEMfit.rds")){
-  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
-  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
-}else{
-  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
-}
-if(!file.exists("hvtnMCMCfit.rds")){
-  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
-  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
-}else{
-  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
-}
-
-#'### Summarize the results. Do the likelihood ratio test, Fisher's test, and rank by log fold change for comparison.
-#+ cache=TRUE,message=FALSE,warning=FALSE
[…]

### 2013-06-25 MIMOSA/mimosafigures/mimosafigures-test/utility.R

@@ -1,667 +1,0 @@
-#'setup the environment
-#'
-#'load MIMOSA, parallel, filehash,gdata,pracma,reshape and ggplot2
-setup<-function(){
-  suppressPackageStartupMessages({require(MIMOSA)
-                                  require(multicore)
-                                  require(parallel)
-                                  require(filehash)
-                                  require(gdata)
-                                  require(pracma)
-                                  require(ggplot2)
-                                  require(reshape)
-                                  require(data.table)
-                                  require(SingleCellAssay)
-                                  require(gplots)
-                                  
-  })
-}
-
-#'construct an expected vs observed fdr curve
-#'
-#'Expected vs observed FDr taking the expected fdr and the truth.
-fdrCurves<-function(fdr,truth){
-  truth<-truth[order(fdr)]
-  true.fdr<-cumsum(1-truth)/1:length(truth)
-  fdr<-sort(fdr)
-  cbind(fdr,true.fdr)
-}
-
-#'Construct and ROC curve
-#'
-#'Given the p-value and the truth, construct an ROC curve.
-rocCurves<-function(p,truth){
-  #s<-seq(0,1,l=1000)
-  #table<-t(sapply(s,function(th){
-  #  prop.table(table(test=factor(p<=th,levels=c("FALSE","TRUE")),truth=truth),margin=2)["TRUE",]
-  #}))
-  #colnames(table)<-c("FPR","TPR")
-  #table
-  o<-order(p,decreasing=FALSE)
-  intervals<-lapply(sapply(sort(unique(p)),function(x)which(as.numeric(p[o])%in%x)),max)
[…]

### 2013-06-25 MIMOSA/mimosafigures/utility.R

@@ -1,667 +1,0 @@
-#'setup the environment
-#'
-#'load MIMOSA, parallel, filehash,gdata,pracma,reshape and ggplot2
-setup<-function(){
-  suppressPackageStartupMessages({require(MIMOSA)
-                                  require(multicore)
-                                  require(parallel)
-                                  require(filehash)
-                                  require(gdata)
-                                  require(pracma)
-                                  require(ggplot2)
-                                  require(reshape)
-                                  require(data.table)
-                                  require(SingleCellAssay)
-                                  require(gplots)
-                                  
-  })
-}
-
-#'construct an expected vs observed fdr curve
-#'
-#'Expected vs observed FDr taking the expected fdr and the truth.
-fdrCurves<-function(fdr,truth){
-  truth<-truth[order(fdr)]
-  true.fdr<-cumsum(1-truth)/1:length(truth)
-  fdr<-sort(fdr)
-  cbind(fdr,true.fdr)
-}
-
-#'Construct and ROC curve
-#'
-#'Given the p-value and the truth, construct an ROC curve.
-rocCurves<-function(p,truth){
-  #s<-seq(0,1,l=1000)
-  #table<-t(sapply(s,function(th){
-  #  prop.table(table(test=factor(p<=th,levels=c("FALSE","TRUE")),truth=truth),margin=2)["TRUE",]
-  #}))
-  #colnames(table)<-c("FPR","TPR")
-  #table
-  o<-order(p,decreasing=FALSE)
-  intervals<-lapply(sapply(sort(unique(p)),function(x)which(as.numeric(p[o])%in%x)),max)
[…]

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.Rmd
Add code for generating manuscript figures

--- 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.Rmd
@@
 
-Read and merge the hvtn data. Cache it.
+### Read and merge the hvtn data.
 
@@
 
-read the fluidigm data
+### Read the fluidigm data.
 
@@
 
-Process it
+Process it.
 
@@
 
-Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC
+### Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC implementations.
 
@@
 
-drop the stuff fit via mcmc and summarize, fit LRT, Fisher's and rank by log fold change
+### Summarize the results. Do the likelihood ratio test, Fisher's test, and rank by log fold change for comparison.
 
-```{r cache=TRUE}
+```{r cache=TRUE,message=FALSE,warning=FALSE}
 hvtn.summary.EM<-summarizeMIMOSAFit(hvtn.result.EM)
@@
 
-Compute ROC and FDR curves and calculate the AUC.
-+ cache=TRUE
+### Compute ROC and FDR curves and calculate the AUC for all the results.
 
-```{r }
+```{r cache=TRUE}
 ROC<-ComputeROCs(hvtn.summary.mcmc,hvtn.summary.EM)
@@
 
-FDR and ROC plots
+### FDR and ROC plots of the ICS data.
 --------
@@
 FDR$Method<-relevel(FDR$Method,"MIMOSA (EM)")
-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+levs<-levels(FDR$CYTOKINE)
+levs <-
+  c("IFN-IL2-TNF+", "IFN-IL2+TNF-", "IFN-IL2+TNF+", "IFN+IL2-TNF-", 
+    "IFN+IL2-TNF+", "IFN+IL2+TNF-", "IFN+IL2+TNF+", "IFNg+", "IFNg+IL2+", 
+    "IFNg+TNF+", "IL2+", "IL2-IFNg+", "IL2+ OR IFNg+", "IL2+IFNg-", 
+    "IL2+IFNg+", "IL2+TNF+", "TNF+")
+FDR$CYTOKINE<-factor(FDR$CYTOKINE,labels=levs)
+fdrplots<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+rocplots
+fdrplots
 ```
 
-Process the fluidigm data to construct an expression set for MIMOSA
+### Figure 1 and S1.
+
+```{r echo=FALSE,eval=FALSE}
+suppressPackageStartupMessages(library(gridExtra))
+f1levs<-levs[c(8,11,13,17,15,16,10,7)]
+sflevs<-setdiff(levs,f1levs)
+AUC$y[AUC$Method%in%"Fisher"]<-0.4
+AUC$y[AUC$Method%in%"MIMOSA (EM)"]<-0.5
+AUC$y[AUC$Method%in%"MIMOSA (MCMC)"]<-0.3
+
+fnames<-c("revised_2.pdf","revised_1.pdf","revised_15.pdf","revised_3.pdf","revised_5.pdf","revised_6.pdf","revised_7.pdf","revised_8.pdf")
+for(i in seq_along(f1levs)){
+  fi.1<-ggplot(subset(ROC,TCELLSUB%in%"cd3+/cd4+"&CYTOKINE%in%f1levs[i]))+geom_line(aes(x=FPR,y=TPR,color=Method))+theme_bw()+facet_wrap(~CYTOKINE)+geom_text(aes(x=x,y=y,label=sprintf("AUC=%s",signif(AUC,2)),col=factor(Method)),data=subset(AUC,CYTOKINE%in%f1levs[i]),show_guide=FALSE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+  fi.2<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")&CYTOKINE%in%f1levs[i]))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
+  pdf(file=sprintf("/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/%s",fnames[i]),width=10,height=3)
+  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
+  dev.off()
+  grid.arrange(fi.1,fi.2,ncol=2,nrow=1)
+}
+```
+
+### Process the fluidigm data to construct an expression set for MIMOSA.
 ---------
@@
 
-Fit MIMOSA by gene, grouping by stimulation and for all samples together.
+### Fit MIMOSA by gene, grouping by stimulation and for all samples together.
 ----
@@
 
-Heatmap of posterior probabilities of response for fit of all samples by gene.
+### Heatmaps for Fluidigm data fit to all stimulations simultaneously.
 ----
@@
 rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x)as.character(unique(pData(x)$Gene))))
-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=2),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
+#heatmap.2((S*(foo))[apply(foo,1,function(x)sum(x>0.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="signed Posterior Probabilities\n(fit by gene, all stimulations)")
+foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 ```
 
-Heatmap of posterior differences of proportions fit by gene for all samples.
+### Figure 2 B, (fit of all samples simultaneously)
+
+```{r eval=FALSE,echo=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmCombinedFit.pdf")
+foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+dev.off()
+```
+
+Heatmap of posterior differences in proportions from simultaneous fit (not shown).
 ----
@@
 
-```{r heatmap.props.bygene,cache=TRUE}
+```{r heatmap.props.bygene,cache=TRUE,include=FALSE}
 colnames(props)<-colnames(S)
 rownames(props)<-rownames(S)
-heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
+heatmap.2(S*sqrt(abs(props)),col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior Differences in Proportions\n(fit by gene all stimulations)")
 ```
 
-Heatmap of posterior probability of response for fit of samples by stimulation and by gene
+### Heatmaps for Fluidigm data fit per stimulation.
 ---
@@
 rownames(S)<-do.call(cbind,lapply(fluidigm.fits.by.stim.bygene[[1]],function(x)as.character(unique(pData(x)$Gene))))
-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.1)>=1),],col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
+#h<-heatmap.2((S*foo)[apply(foo,1,function(x)sum(x>.6)>=1),],col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],main="Signed Posterior Probabilities\n(fit by gene and stimulation).")
+foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
+h<-heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+hmaporder<-h$colInd
+geneinds<-apply(foo,1,function(x)sum(x>.6)>=1)
 ```
 
-Heatmap of posterior differences in proportions for fit of samples by stimulation and by gene. 
+### Figure 2 A (fit per stimulation)
+Fit MIMOSA per stimulation.
+
+```{r eval=FALSE,echo=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Fluidigm_plot1.pdf")
+foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
+heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+dev.off()
+```
+
+Heatmap of posterior differences in proportions for fit of samples by stimulation and by gene (not shown).
 ---
@@
 
-```{r heatmap.props.bystim.bygene,cache=TRUE}
+```{r heatmap.props.bystim.bygene,cache=TRUE,include=FALSE}
 colnames(props)<-colnames(S)
 rownames(props)<-rownames(S)
-heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
+heatmap.2(sqrt(abs(props))*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="Posterior differences in proportions\n(fit by gene and stimulations)")
 ```
 
-Fisher's exact test and heatmaps of -log10(q-values) from Fisher's
+### Fisher's exact test and heatmaps of -log10(q-values) from Fisher's test.
 ---
@@
 stim<-pData(fluidigm.fits.by.stim.allgenes[[1]][[1]])$Stim
-heatmap.2(fisher*S,col=colorpanel(n=10,high="green",low="red",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
+#heatmap.2(fisher*S,col=colorpanel(n=10,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.5,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,main="-log10(p-values)\nfrom Fisher's Exact Test")
+colnames(adj.fisher)<-colnames(fisher)
+rownames(adj.fisher)<-rownames(fisher)
+heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 ```
 
-Multivariate MIMOSA on fluidigm data looking at interactions
+### Figure 2 D (q-values from Fisher's test)
+
+```{r eval=FALSE,echo=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmFisher.pdf")
+colnames(adj.fisher)<-colnames(fisher)
+rownames(adj.fisher)<-rownames(fisher)
+heatmap.2((((adj.fisher))*S)[apply(adj.fisher,1,function(x)sum(x<0.1)>1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
+dev.off()
+
+```{r eval=TRUE,echo=FALSE}
+emprop<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x){
+  p<-vector('numeric',nrow(x@result@n.stim))
+  p<-(prop.table(as.matrix((x@result@n.stim)),1)-prop.table(as.matrix(x@result@n.unstim),1))[,2]
+}))
+rownames(emprop)<-rownames(fisher)
+colnames(emprop)<-colnames(fisher)
+
+```{r echo=FALSE,warning=FALSE}
+heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
+```
+
+### Figure 2C empirical proportions.
+
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmEmpiricalProps.pdf")
+heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
+dev.off()
+```
+
+### Multivariate MIMOSA fit to fluidigm data, looking at interactions of genes.
 ---------
@@
 
-```{r cache=FALSE}
+```{r cache=TRUE}
 fl.comb<-combine(fluidigm[[1]],fluidigm[[2]],fluidigm[[3]],fluidigm[[4]])
@@
 
-interact.scan<-combn(colnames(melted)[4:99],2,function(x){
-  g1<-x[1];g2<-x[2]
-  gg<-paste(g1,g2,sep=":")
-  bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
-  bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
-  setnames(bar,"foo",gg)
-  bar[,5]<-factor(bar[,5],levels=c("0:0","0:1","1:0","1:1"))
-  C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
-  expected<-c("0:0","0:1","1:0","1:1")
-  unused<-setdiff(expected,colnames(C))
-  newframe<-NULL
-  for(i in seq_along(expected)){
-    if(expected[i]%in%unused){
-      newframe<-cbind(newframe,rep(0,nrow(C)))
-    }else{
-      newframe<-cbind(newframe,C[,expected[i]])
-    }
-  }
-  colnames(newframe)<-expected
-  C<-cbind(C[,1:2],newframe)
-  CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])
-  foo<-lapply(CC,function(x)prop.table(as.matrix(x),1))
-  psum<-mean(rowSums(foo[[1]][,2:4])-rowSums(foo[[2]][,2:4]))
-  pij<-colMeans(foo[[1]][,2:4]-foo[[2]][,2:4])
-  res<-c(psum,pij)
-  res
-})
+# interact.scan<-combn(colnames(melted)[4:99],2,function(x){
+#   g1<-x[1];g2<-x[2]
+#   gg<-paste(g1,g2,sep=":")
+#   bar<-data.frame(melted[,c("group","Stim.Condition",g1,g2)])
+#   bar$foo<-factor(bar[,c(g1)]):factor(bar[,c(g2)])
+#   setnames(bar,"foo",gg)
+#   bar[,5]<-factor(bar[,5],levels=c("0:0","0:1","1:0","1:1"))
+#   C<-cast(melt(bar,id=c("group","Stim.Condition"),measure=c(gg)),group+Stim.Condition~value,fun.aggregate=function(x)length(x))
+#   expected<-c("0:0","0:1","1:0","1:1")
+#   unused<-setdiff(expected,colnames(C))
+#   newframe<-NULL
+#   for(i in seq_along(expected)){
+#     if(expected[i]%in%unused){
+#       newframe<-cbind(newframe,rep(0,nrow(C)))
+#     }else{
+#       newframe<-cbind(newframe,C[,expected[i]])
+#     }
+#   }
+#   colnames(newframe)<-expected
+#   C<-cbind(C[,1:2],newframe)
+#   CC<-list(n.unstim=subset(C,Stim.Condition%in%"Unstim")[,-c(1:2)],n.stim=subset(C,Stim.Condition%in%"Stim")[,-c(1:2)])
+#   foo<-lapply(CC,function(x)prop.table(as.matrix(x),1))
+#   psum<-mean(rowSums(foo[[1]][,2:4])-rowSums(foo[[2]][,2:4]))
+#   pij<-colMeans(foo[[1]][,2:4]-foo[[2]][,2:4])
+#   res<-c(psum,pij)
+#   res
+# })
+#cache("interact.scan")
 
-genes<-combn(colnames(melted)[6:101],2)[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
-interact.scan[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
-ggg<-genes[,6]
-g1<-ggg[1];g2<-ggg[2] #CCR7 and GZMK
+#genes<-combn(colnames(melted)[6:101],2)[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
+#interact.scan[,which(abs(interact.scan[1,])<0.01&abs(interact.scan[4,])>0.3)]
+#ggg<-genes[,6]
+#g1<-ggg[1];g2<-ggg[2] #CCR7 and GZMK
 g1<-"CCR7";g2<-"GZMK"
@@
 sapply(list.files(pattern=c("\\.dat")),unlink)
+```
+
+### Figure 4. Heatmaps of counts for all combinations of CCR7 and GZMK vs marginal counts.
+Samples in red show no difference by MIMOSA.
+
+```{r eval=TRUE,echo=FALSE}
+library(gridExtra)
+missed<-!MIMOSA:::fdr(flres$z)<0.05
+toplot$variable<-factor(toplot$variable,labels=c("CCR7-GZMK-","CCR7-GZMK+","CCR7+GZMK-","CCR7+GZMK+"))
+toplot2$variable<-factor(toplot2$variable,labels=c("CCR7-GZMK-","CCR7+ or GZMK+"))
+p1<-ggplot(toplot)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90))+facet_wrap(~stim)+scale_fill_continuous("log2 count",na.value="gray",breaks=round(seq(0,6,l=10)))+theme(axis.text.y=element_text(color=c("black","red")[missed+1]))
+p2<-ggplot(toplot2)+geom_tile(aes(x=variable,y=id,fill=log2(value)),interpolate=FALSE)+theme_bw()+theme(axis.text.x=element_text(angle=90,hjust=1))+facet_wrap(~stim)+scale_fill_continuous("log2 count",na.value="gray",breaks=round(seq(0,6,l=10)))+theme(axis.text.y=element_text(color=c("black","red")[2]))
+grid.arrange(p1,ncol=1)
+grid.arrange(p2,ncol=1)
+
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/revised_Fluidigm_Multivariate_CCR7_GZMK.pdf",width=6,height=4)
+grid.arrange(p1,ncol=1)
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/revised_Fluidigm_Marginal_CCR7_GZMK.pdf",width=6,height=4)
+grid.arrange(p2,ncol=1)
+dev.off()
 
@@
 ==========
-One sided
-------
+Simulations for the one-sided and two-sided MIMOSA model.
 We'll use the model fit from IFNg+ for ENV-1-PTEG based on the ROC and FDR plots
-Note that MIMOSA has greater sensitivity and specificity than Fisher's exact test or the Likelihood ratio test. When there are 5 observations, the methods being to perform equivalently.
 
-```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5}
-sims<-OneSidedSims(hvtn.result.mcmc)
-ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5,include=FALSE}
+if(!file.exists("cache/sims.10K.RData")){
+sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
+sims.25K<-OneSidedSims(hvtn.result.mcmc,N=25000)
+sims.10K<-OneSidedSims(hvtn.result.mcmc,N=10000)
+cache('sims.50K')
+cache('sims.25K')
+cache('sims.10K')
+}else{
+load("cache/sims.10K.RData")
+load("cache/sims.25K.RData")
+load("cache/sims.50K.RData")
+}
+ggplot(sims.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims.10K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims.10K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+
+```{r two.sided.sims, cache=FALSE, echo=FALSE,fig.width=5,fig.height=5,include=FALSE}
+if(!file.exists("cache/sims2.10K.RData")){
+sims2.50K<-TwoSidedSims(hvtn.result.mcmc,N=50000)
+sims2.25K<-TwoSidedSims(hvtn.result.mcmc,N=25000)
+sims2.10K<-TwoSidedSims(hvtn.result.mcmc,N=10000)
+cache('sims2.50K')
+cache('sims2.25K')
+cache('sims2.10K')
+}else{
+load("cache/sims2.10K.RData")
+load("cache/sims2.25K.RData")
+load("cache/sims2.50K.RData")
+}
+ggplot(sims2.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims2.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+ggplot(sims2.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+ggplot(sims2.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+smlcnts.roc<-ggplot(subset(sims2.10K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations (small counts)")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+smlcnts.fdr<-ggplot(subset(sims2.10K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations (small counts)")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
 ```
 
-Two-sided
---------
+### Figure 3: One-sided and two-sided simulations.
 
-```{r two.sided.sims, cache=FALSE, echo=FALSE,fig.width=5,fig.height=5}
-sims<-TwoSidedSims(hvtn.result.mcmc)
-ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+```{r eval=TRUE,echo=FALSE}
+levs<-levels(sims.50K$sim.ROC$method)
+levs[2]<-c("MIMOSA (one sided)")
+sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
+p1.1<-ggplot(subset(sims.50K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+levs<-levels(sims2.50K$sim.ROC$method)
+levs[2]<-c("MIMOSA (two sided)")
+sims2.50K$sim.ROC$method<-factor(sims2.50K$sim.ROC$method,labels=levs)
+p2.1<-ggplot(subset(sims2.50K$sim.ROC,Nobs==200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (two sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+
+levs<-levels(sims.50K$sim.FDR$method)
+levs[1]<-"MIMOSA (one sided)"
+sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
+p1.2<-ggplot(subset(sims.50K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+levs<-levels(sims2.50K$sim.FDR$method)
+levs[1]<-c("MIMOSA (two sided)")
+sims2.50K$sim.FDR$method<-factor(sims2.50K$sim.FDR$method,labels=levs)
+p2.2<-ggplot(subset(sims2.50K$sim.FDR,Nobs==200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (two sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_ROC_50K.pdf",width=6,height=4)
+p1.1
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_TwoSided_ROC_50K.pdf",width=6,height=4)
+p2.1
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_TwoSided_FDR_50K.pdf",width=6,height=4)
+p2.2
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_FDR_50K.pdf",width=6,height=4)
+p1.2
+dev.off()
+
+```{r echo=FALSE}
+p1.1
+p2.1
+p2.2
+p1.2
 ```
 
-Multivariate Simulations
+### Multivariate Simulations
 -----
@@
 
-```{r cache=TRUE}
+```{r cache=TRUE,results='hide',message=FALSE,warning=FALSE}
 set.seed(101)
@@
 fisher.p.adj<-lapply(fisher.p,function(x)p.adjust(x,"fdr"))
-fisher.roc<-do.call(rbind,by(cbind(do.call(rbind,lapply(fisher.p.adj,function(i)FDR(i,gl(2,50)==2)))),rep(gl(100,1),10),function(x)colMeans(x)))
+fisher.roc<-do.call(rbind,by(cbind(do.call(rbind,lapply(fisher.p.adj,function(i)ROCfun(i,gl(2,50)==2)))),rep(gl(100,1),10),function(x)colMeans(x)))
 mimosa.adj<-lapply(1:ncol(mdsim),function(i)MIMOSA:::fdr(mdsim[,i]$z))
-MIMOSA.roc<-do.call(rbind,by(do.call(rbind,lapply(mimosa.adj,function(i)FDR(i,gl(2,50)==2))),rep(gl(100,1),10),function(x)colMeans(x)))
+MIMOSA.roc<-do.call(rbind,by(do.call(rbind,lapply(mimosa.adj,function(i)ROCfun(i,gl(2,50)==2))),rep(gl(100,1),10),function(x)colMeans(x)))
 foo<-data.frame(as.data.frame(rbind(fisher.roc,MIMOSA.roc)),Method=gl(2,100,labels=c("Fisher","MIMOSA")))
 sz<-element_text(size=18)
-ggplot(foo)+geom_line(aes(x=fpr,y=tpr,col=Method))+scale_y_continuous("TPR")+scale_x_continuous("FPR")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)
+p1<-ggplot(foo)+geom_line(aes(x=fpr,y=tpr,col=Method))+scale_y_continuous("TPR")+scale_x_continuous("FPR")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)
+#FDR
+mim.fdr<-as.data.frame(cbind(do.call(rbind,lapply(mimosa.adj,function(x)fdrCurves(x,gl(2,50)==2))),rep=gl(100,1)))
+mim.fdr<-ddply(mim.fdr,.(rep),summarize,true.fdr.hat=mean(true.fdr),fdr.hat=mean(fdr))
+fisher.fdr<-as.data.frame(cbind(do.call(rbind,lapply(fisher.p.adj,function(x)fdrCurves(x,gl(2,50)==2))),rep=gl(100,1)))
+fisher.fdr<-ddply(fisher.fdr,.(rep),summarize,true.fdr.hat=mean(true.fdr),fdr.hat=mean(fdr))
+foo<-data.frame(rbind(fisher.fdr,mim.fdr),Method=gl(2,100,labels=c("Fisher","MIMOSA")))
+p2<-ggplot(foo)+geom_line(aes(y=true.fdr.hat,x=fdr.hat,col=Method))+scale_y_continuous("observed fdr")+scale_x_continuous("nominal fdr")+theme_bw()+theme(axis.title.x=sz,axis.title.y=sz,axis.text.x=sz,axis.text.y=sz)+geom_abline(1,lty=3)
+grid.arrange(p1,p2)
 ```
@@
 
-ggplot(comproc[[1]])+geom_line(aes(x=fdr.hat,y=true.fdr.hat,color=method))+facet_wrap(~true.resp.rate)+theme_bw()+geom_abline(1,lty=3)+labs(title="Observed vs True FDR when \ntrue response rate varies but is assumed to be 100%")
-ggplot(comproc[[2]])+geom_line(aes(x=FPR.hat,y=TPR.hat,color=method))+facet_wrap(~true.resp.rate)+theme_bw()+labs(title="ROC curves when \ntrue response rate varies but is assumed to be 100%")+scale_x_continuous("False Positive Rate")+scale_y_continuous("True Positive Rate")
+p8<-ggplot(comproc[[1]])+geom_line(aes(x=fdr.hat,y=true.fdr.hat,color=method))+facet_wrap(~true.resp.rate,ncol=5)+theme_bw()+geom_abline(1,lty=3)+labs(title="Observed vs True FDR when \ntrue response rate varies but is assumed to be 100%")+theme(axis.text.x=element_text(angle=90))
+p9<-ggplot(comproc[[2]])+geom_line(aes(x=FPR.hat,y=TPR.hat,color=method))+facet_wrap(~true.resp.rate,ncol=5)+theme_bw()+labs(title="ROC curves when \ntrue response rate varies but is assumed to be 100%")+scale_x_continuous("False Positive Rate")+scale_y_continuous("True Positive Rate")+theme(axis.text.x=element_text(angle=90))
+
+library(gridExtra)
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/competeROC.pdf",width=10,height=3)
+p8
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/competeROCfdr.pdf",width=10,height=3)
+p9
+dev.off()
+
+```{r echo=FALSE,fig.width=12,fig.height=6}
+p8
+p9
 ```
@@
 
-```{r violated.assumptions, cache=FALSE,echo=FALSE,fig.width=8,fig.height=5}
+```{r violated.assumptions, cache=FALSE,echo=FALSE,fig.width=20,fig.height=10}
 sims<-TwoSidedSimsTruncated(hvtn.result.mcmc)
-ggplot(sims$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+violated.roc<-ggplot(subset(sims$sim.ROC,Num==50000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations Violating Model Assumptions")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+violated.fdr<-ggplot(subset(sims$sim.FDR,Num==50000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from Two-Sided Simulations Violating Model Assumptions")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+grid.arrange(violated.roc,violated.fdr,smlcnts.roc,smlcnts.fdr)
 ```
 
-How many CD4 and CD8 T-cells?
+### Figure S3
+
+```{r eval=TRUE,echo=FALSE}
+p1<-ggplot(subset(sims$sim.ROC,Num==50000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+p2<-ggplot(subset(sims$sim.ROC,Num==10000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+p3<-ggplot(subset(sims$sim.FDR,Num==50000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+p4<-ggplot(subset(sims$sim.FDR,Num==10000))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_ROC_50K.pdf",width=5,height=3)
+p1
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_ROC_10K.pdf",width=5,height=3)
+p2
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_FDR_50K.pdf",width=5,height=3)
+p3
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_Truncated_FDR_10K.pdf",width=5,height=3)
+p4
+dev.off()
+
+#rerun one sided sims at 50K, more than 10 reps
+sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
+levs<-levels(sims.50K$sim.FDR$method)
+levs[1]<-"MIMOSA (one sided)"
+sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
+levs<-levels(sims.50K$sim.ROC$method)
+levs[2]<-"MIMOSA (one sided)"
+sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
+p5<-ggplot(subset(sims.50K$sim.FDR,Nobs!=200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+p6<-ggplot(subset(sims.50K$sim.ROC,Nobs!=200))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_ROC_varyNobs.pdf",width=9,height=3)
+p6
+dev.off()
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/Sim_OneSided_FDR_varyNobs.pdf",width=8,height=3)
+p5
+dev.off()
+```
+
+### One-sided simulations, 50K events. N=20,50,100
+
+```{r echo=FALSE}
+p6
+p5
+```
+
+### Figure of effect size vs probability of response
+
+```{r }
+pd<-do.call(rbind,lapply(hvtn.result.mcmc,function(x)pData(x)))
+pd<-cbind(pd,do.call(rbind,lapply(hvtn.result.mcmc,function(x)data.frame(response=MIMOSA:::fdr(x@z)<0.01,Prob.resp=x@z[,2],effect=do.call(c,lapply(x@result@p,function(x)diff(rev(x[,2]))))))))
+pd$VACCINE<-factor(pd$RX_CODE%in%c("T1","T2","T3","T4"))
+```{r eval=FALSE}
+pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/volcanoplots.pdf",width=15,height=10)
+ggplot(subset(pd,TCELLSUB%in%"cd3+/cd4+"))+geom_point(aes(x=effect,y=Prob.resp,color=response,shape=VISITNO:VACCINE))+facet_wrap(TCELLSUB~CYTOKINE,scale="free_x")+theme(axis.text.x=element_text(angle=90))+scale_color_discrete("Response",labels=c("No","Yes"))+theme_bw()
+dev.off()
+```{r echo=FALSE,fig.width=15}
+ggplot(subset(pd,TCELLSUB%in%"cd3+/cd4+"))+geom_point(aes(x=effect,y=Prob.resp,color=response,shape=VISITNO:VACCINE))+facet_wrap(TCELLSUB~CYTOKINE,scale="free_x")+theme(axis.text.x=element_text(angle=90))+scale_color_discrete("Response",labels=c("No","Yes"))+theme_bw()
+```
+
+Distribution of total CD4 and CD8 T-cell counts?
 -----
@@
 suppressPackageStartupMessages(require(scales))
-df<-data.frame(tcellsub="CD4",treatment=gl(2,224/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[1]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[1]]@result@n.unstim)[,1]))
+df<-data.frame(tcellsub="CD4",treatment=gl(2,205/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[1]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[1]]@result@n.unstim)[,1]))
 df<-rbind(df,data.frame(tcellsub="CD8",treatment=gl(2,224/2,labels=c("stim","unstim")),count=melt(data.frame(stim=hvtn.result.mcmc[[30]]@result@n.stim[,1],unstim=hvtn.result.mcmc[[30]]@result@n.unstim)[,1])))
@@
 
-Table of median cell counts for different subsets
+Median cell counts for T-cell subsets.
 -----
@@
 
-Magnitude of the proportions for the HVTN data
+Distribution of the proportions for the HVTN data
 ----
@@
 
-Table of median cell proportions for HVTN data
+Median cell proportions for ICS data.
 ----

### 2013-06-25 MIMOSA/mimosafigures/compile.R

@@ -1,0 +1,5 @@
+library(knitr)
+library(knitrBootstrap)
+knitr::spin("manuscriptFigures.R",FALSE)
+knit_bootstrap("manuscriptFigures.Rmd")
+system("cp manuscriptFigures.html /home/gfinak/Dropbox/Public/MIMOSA/manuscriptFigures.html")

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.R

@@ -1,0 +1,599 @@
+#'Load the data and fit the MIMOSA model
+#'===============
+source("utility.R")
+#+ echo=FALSE
+setup()
+
+#'### Read and merge the hvtn data.
+#+ cache=TRUE
+hvtndata<-readHVTNData()
+hvtndata<-processHVTNData(hvtndata)
+hvtndata<-constructEset(hvtndata)
+
+#'### Read the fluidigm data.
+#+ cache=FALSE
+fluidigm<-readFluidigmData()
+
+#'Process it.
+#+ cache=FALSE
+fluidigm<-processFluidigmData()
+
+ 
+#'### Fit the MIMOSA model to the HVTN data all cytokines and ENV-1-PTEG using EM and MCMC implementations.
+if(!file.exists("hvtnEMfit.rds")){
+  hvtn.result.EM<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="EM",run.parallel=TRUE)
+  saveRDS(hvtn.result.EM,file="hvtnEMfit.rds")
+}else{
+  hvtn.result.EM<-readRDS("hvtnEMfit.rds")
+}
+if(!file.exists("hvtnMCMCfit.rds")){
+  hvtn.result.mcmc<-MIMOSA(NSUB+CYTNUM~ASSAYID+VISITNO+PTID+RX_CODE+RefTreat|TCELLSUB+ANTIGEN+CYTOKINE,hvtndata,ref=RefTreat%in%"Reference"&ANTIGEN%in%"ENV-1-PTEG",subset=RefTreat%in%"Treatment"&ANTIGEN%in%"ENV-1-PTEG",method="mcmc",run.parallel=TRUE,pXi=0.1,getP=TRUE)
+  saveRDS(hvtn.result.mcmc,file="hvtnMCMCfit.rds")
+}else{
+  hvtn.result.mcmc<-readRDS("hvtnMCMCfit.rds")
+}
+
+#'### Summarize the results. Do the likelihood ratio test, Fisher's test, and rank by log fold change for comparison.
+#+ cache=TRUE,message=FALSE,warning=FALSE
[…]

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.Rmd
update html

@@
 -----
+Top row shows results for MIMOSA fit to data simulated from a model violating model assumptions.
+Bottorm row shows results of two-sided MIMOSA fit to data with low counts (10K events).
 
@@
 sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
+
+```{r eval=TRUE}
 p5<-ggplot(subset(sims.50K$sim.FDR,Nobs!=200))+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA (one sided)")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
@@
 dev.off()
-```
-
-### One-sided simulations, 50K events. N=20,50,100
-
-```{r echo=FALSE}
-p6
-p5
 ```

### 2013-06-25 MIMOSA/mimosafigures/manuscriptFigures.R

@@ -507,6 +507,8 @@ ggplot(data.frame(p.hat=prop.table(as.matrix(res@result@n.unstim),1)[,2]))+geom_
 
 #'Unconstrained MIMOSA fit to data violating model assumptions
 #'-----
+#'Top row shows results for MIMOSA fit to data simulated from a model violating model assumptions.
+#'Bottorm row shows results of two-sided MIMOSA fit to data with low counts (10K events).
 #+ violated.assumptions, cache=FALSE,echo=FALSE,fig.width=20,fig.height=10
 sims<-TwoSidedSimsTruncated(hvtn.result.mcmc)
 violated.roc<-ggplot(subset(sims$sim.ROC,Num==50000))+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from Two-Sided Simulations Violating Model Assumptions")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Num)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
@@ -542,6 +544,8 @@ sims.50K$sim.FDR$method<-factor(sims.50K$sim.FDR$method,labels=levs)
 levs<-levels(sims.50K$sim.ROC$method)
 levs[2]<-"MIMOSA (one sided)"
 sims.50K$sim.ROC$method<-factor(sims.50K$sim.ROC$method,labels=levs)
+
+#+eval=TRUE
[…]

### 2013-06-26 MIMOSA/mimosafigures/manuscriptFigures.R
manuscriptFigures.html

@@ -135,7 +135,7 @@ rownames(S)<-do.call(rbind,lapply(fluidigm.fits.by.stim.allgenes[[1]],function(x
 foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
 heatmap.2((S*(foo))[apply(foo2,1,function(x)sum(x<0.1)>=1),],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),cexRow=0.75,hclustfun=function(x)hclust(x,method="complete"),trace="none",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)],Colv=TRUE,denscol="black")
 
-#'### Figure 2 B, (fit of all samples simultaneously)
+#'### Figure 2 C, (fit of all samples simultaneously)
 #+eval=FALSE,echo=FALSE
 pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmCombinedFit.pdf")
 foo2<-apply(foo,2,function(x)MIMOSA:::fdr(cbind(1-x,x)))
@@ -228,7 +228,7 @@ colnames(emprop)<-colnames(fisher)
 #+echo=FALSE,warning=FALSE
 heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])
 
-#'### Figure 2C empirical proportions.
+#'### Figure 2 B empirical proportions.
 #+eval=FALSE
 pdf(file="/Users/gfinak/Documents/manuscripts/MIMOSA_Paper/Figures/FluidigmEmpiricalProps.pdf")
 heatmap.2(emprop[geneinds,hmaporder],col=colorpanel(n=100,high="yellow",low="blue",mid="white"),margins=c(6,5),Colv=NULL,denscol="black",tracecol="black",symbreaks=TRUE,ColSideColors=c("red","orange","yellow","green")[as.numeric(stim)[hmaporder]])

### 2013-06-26 MIMOSA/mimosafigures/manuscriptFigures.Rmd
manuscriptFigures.html
manuscriptFigures

@@
 
-### Figure 1 and S1.
+### Figure 1 and Supplementary Figure 1.
 
@@
 
-### Figure 2 B, (fit of all samples simultaneously)
+### Figure 2 C, (fit of all samples simultaneously)
 
@@
 
-### Figure 2C empirical proportions.
+### Figure 2 B empirical proportions.
 
@@
 
-```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5,include=FALSE}
+```{r simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5,include=TRUE}
 if(!file.exists("cache/sims.10K.RData")){
@@
 ggplot(sims.50K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
-ggplot(sims.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
-ggplot(sims.10K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
-ggplot(sims.10K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+#ggplot(sims.25K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+#ggplot(sims.25K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+#ggplot(sims.10K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
+#ggplot(sims.10K$sim.FDR)+geom_line(aes(x=fdr.hat,y=true.fdr.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Observed vs Nominal FDR for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("Nominal FDR")+scale_y_continuous("Observed FDR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))+geom_abline(1,lty=3)
+```
+
+### Supplementary Figure 4
+One-sided simulations with varying values of I (number of subjects).
 
@@
 
+### Supplementary Figure 6
 Competing ROC curves to evaluate the effect of inaccuracies in the true positives and false positives.
@@
 
-Supplementary Figure 2
-------
-Histogram of the empirical proportions of unstimulated cells and overlaid posterior densities of the beta distribution with $\alpha_s$ and $\alpha_u$ estimated from the data for ENV-1-PTEG stimulated, IFN-$\gamma$+ CD4+ T-cells, demonstrating that the assumption of a common distribution for piu across subjects is reasonable.
+### Supplementary Figure 2
 
@@
 
+Supplementary Figure 3 
+------
+Histogram of the empirical proportions of unstimulated cells and overlaid posterior densities of the beta distribution with $\alpha_s$ and $\alpha_u$ estimated from the data for ENV-1-PTEG stimulated, IFN-$\gamma$+ CD4+ T-cells, demonstrating that the assumption of a common distribution for piu across subjects is reasonable.
 Unconstrained MIMOSA fit to data violating model assumptions
@@
 
-### Figure S3
+### Supplementary Figure 5
 
@@
 
-Distribution of total CD4 and CD8 T-cell counts?
+### Supplementary Figure 7
+Distribution of total CD4 and CD8 T-cell counts.
 -----

### 2013-06-26 MIMOSA/mimosafigures/manuscriptFigures.R

@@ -73,7 +73,7 @@ FDR$CYTOKINE<-factor(FDR$CYTOKINE,labels=levs)
 fdrplots<-ggplot(subset(FDR,TCELLSUB%in%c("cd3+/cd4+")))+geom_line(aes(y=true.fdr,x=fdr,color=Method))+geom_abline(1,lty=3)+theme_bw()+scale_x_continuous(name="Nominal FDR")+scale_y_continuous(name="Observed FDR")+facet_wrap(~CYTOKINE)+theme(axis.text.x=element_text(angle=90,hjust=1))
 rocplots
 fdrplots
-#'### Figure 1 and S1. 
+#'### Figure 1 and Supplementary Figure 1. 
 #+echo=FALSE,eval=FALSE
 suppressPackageStartupMessages(library(gridExtra))
 f1levs<-levs[c(8,11,13,17,15,16,10,7)]
@@ -346,7 +346,7 @@ ggplot(A)+geom_point(aes(x=`Effect Size`,y=`Pr(response)`,shape=factor(ID)))+fac
 #'==========
 #'Simulations for the one-sided and two-sided MIMOSA model.
 #'We'll use the model fit from IFNg+ for ENV-1-PTEG based on the ROC and FDR plots
-#+ simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5,include=FALSE
+#+ simulations,cache=FALSE,echo=FALSE,fig.width=5,fig.height=5,include=TRUE
 if(!file.exists("cache/sims.10K.RData")){
 sims.50K<-OneSidedSims(hvtn.result.mcmc,N=50000)
 sims.25K<-OneSidedSims(hvtn.result.mcmc,N=25000)
@@ -361,12 +361,13 @@ load("cache/sims.50K.RData")
 }
 ggplot(sims.50K$sim.ROC)+geom_line(aes(x=FPR.hat,y=TPR.hat,col=relevel(method,"MIMOSA")),direction="vh",lwd=1.5,alpha=0.8)+theme_bw()+labs(title="Average ROC for 10 Data Sets from One-Sided Simulations")+scale_x_continuous("FPR")+scale_y_continuous("TPR")+coord_cartesian(xlim=c(-0.01,1.01),ylim=c(-0.01,1.01))+scale_color_discrete("Method")+facet_wrap(~Nobs)+theme(strip.text.x=element_text(size=18),axis.title.x=element_text(size=18),axis.title.y=element_text(size=18),axis.text.x=element_text(angle=90))
[…]
